### 2023-08-23 paper.tex
setup repo

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}

\title{bit-smc-paper}

\begin{document}

\linenumbers
\title{backwards-in-time SMC}

% First authors
\author[1, $\dagger$]{Gertjan Bisschop}

% Middle Authors
\author[2]{Alison Etheridge}

\author[3]{Peter Ralph}
\author[1]{Jerome Kelleher}
\affil[$\dagger$]{Denotes corresponding author}

\maketitle

\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
\affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
\affil[3]{University of Oregon, USA}

\section{Abstract}

\section{Introduction}

\section{Methods}

\section{Results}

\section{Discussion}

\end{document}

### 2023-11-06 paper.tex
smc-description

@@
 \usepackage[round]{natbib}
-\usepackage{amsmath,amssymb,amsfonts}%
+\usepackage{amsmath,amssymb,amsfonts, bbm}%
 \usepackage{geometry}%
@@
 \usepackage{tikz}
+\usetikzlibrary{math,calc,positioning}
 \usetikzlibrary{calc,positioning}
+\usepackage{subcaption}
 
-\title{bit-smc-paper}
+\newcommand{\noderef}[1]{\textsf{#1}}
+\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
+\newcommand{\kwarg}[0]{\texttt{KwARG}}
+\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
+\newcommand{\argweaverD}[0]{\texttt{ARGweaver-D}}
+\newcommand{\relate}[0]{\texttt{Relate}}
+\newcommand{\espalier}[0]{\texttt{Espalier}}
+\newcommand{\arbores}[0]{\texttt{Arbores}}
 
@@
 \linenumbers
-\title{backwards-in-time SMC}
+\title{Computing likelihoods under the SMC for a general class of ARGs.}
 
@@
 \author[2]{Alison Etheridge}
+\author[3]{Peter Ralph}
 
-\author[3]{Peter Ralph}
+% last author
 \author[1]{Jerome Kelleher}
@@
 \section{Abstract}
+%% WHAT IS THE MAIN FOCUS OF THE PAPER?
 
 \section{Introduction}
+% What are ARGs, their relevance and importance
 
-\section{Methods}
+% associating likelihoods with ARGs
 
-\section{Results}
+% previous likelihood calculations on ARGsL Kuhner and Felsenstein (2000), ARGweaver
+
+% history of approximating the coalescent: 
+SMC (McVean and Cardin, 2005) SMC' (Marjoram and Wall, 2006).
+Also heuristic approximations: Li and Stephens (2003) extending Stephens and Donnelly 
+(2000)
+These have enabled (scalable) ARG inference methods: ARGweaver (relying on SMC, limited 
+to smaller samples, but ability to sample from posterior, and full ARG), and Relate, 
+tsinfer/tsdate (copying algorithm), ArgNeedle (using SMC) 
+Clear distinction into two groups: ARGWeaver infers actual ARG.
+Whereas the other approaches are heuristic in nature. For example, where ARGweaver has 
+the built in assumption of neighbouring trees being separated by single Subtree Prune 
+Regraft (SPR) (Song citation?), this is not the case for any of the heuristic data 
+structures. (other peculiarities: ...).
+
+% SMC and relatives are useful: lot of focus however on integrating over the distribution of trees to get Ne(t). \cite{PSMC, ...}
+% Here the idea is to use the SMC in a different way that resonates more with the true direction of flow
+% of genetic/haplotype information.
+
+% results in a concrete way to compute likelihoods for *any* inferred ARG (integrating out the recombination time)
+% meaning: e.g. no requirements on adjacent trees only differing by a single SPR, (other?)
+
+% In this paper, we attempt to address these problems by
+% providing a precise definition of an ARG that is general enough to encompass
+% the output of popular simulation and inference methods.
+
+% ADD: remark on fact that we are only dealing with topology and branch length information
+% adding likelihood of observing particular mutation configuration given an ARG is straightforward.
+
+\section{Methods and Results}
+\subsection{SMC backwards-in-time}\label{par:description}
+
+Formulated backwards-in-time the SMC \citep{mcvean_approximating_2005} only requires a simple modification to the 
+coalescent with recombination. By only considering those pairs of 
+lineages that share 
+genomic intervals with ancestral material, as eligible for coalescence, we obtain the 
+sequential Markovian structure along the chromosome the SMC is best known for.\\
+
+More formally, and following the notation of \cite{mcvean_approximating_2005}, at any 
+point in time, the process can be described by $L(t)$, the set of labeled lineages 
+extant at time $t$ each represented by a union of ordered non-overlapping half open 
+intervals detailing the ancestral material 
+carried by that lineage $X_i = \{[x_{i0}, y_{i0}), \dotsc, [x_{im}, y_{im})\}$.
+$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$, each represented by 
+single interval spanning the entire genome.
+Backwards in time the process evolves by a succession of coalescence and recombination 
+events until all segments of ancestral material are each present in only one lineage. 
+The waiting time to next event is determined by these two competing processes with exponentially 
+distributed waiting times as described below.\\
+
+Recombination is described by a Poisson process of rate $r$ per unit of length and time. 
+A recombination to the 
+left of $x$ with $x>x_{i0}$ and $x<y_{im}$ splits a lineage $i$ into two new, uniquely 
+labelled lineages.\\
+
+Any two \emph{overlapping} lineages coalesce at rate $\lambda = 1/(2N_e)$ in the case
+of a diploid population of constant size. The newly formed lineage acquires the 
+union of both ordered intervals.
+Although coalescence is reciprocal, here we'll define a strict total order 
+(see \ref{par:liks}) on 
+all lineages in $X(t)$ based on their leftmost starting points. 
+For any such strict total order on $X_l(t)$ and at any point in the process 
+the instantaneous rate of coalescence then equals $\lambda \sum_{i \neq j} I_{ij}$.
+
+\begin{equation} \label{def:coal}
+I_{ij} = \begin{cases}
+1 & X_i > X_j \wedge X_i \cap X_j \neq \emptyset \\
+0 & \text{otherwise}
+\end{cases}
+\end{equation}
+
+We can now define $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$ such that $|C_i(t)| = 
+I_{i}(t) = \sum_{j} I_{ij}(t)$.
+Because $I_{ij}$ is not reciprocal, each extant lineage has their own unique set of 
+lineages $C_i(t)$ it can coalesce with. This will simplify the likelihood computations 
+(see next paragraph). Note that although recombination affects $C_i(t)$ it does not 
+affect $I_{i}$.\\
+
+% note on the disjoint set of intervals for each lineage under the SMC
+As noted by \cite{mcvean_approximating_2005}, the backwards-in-time description of the 
+SMC is slightly different from the left-to-right approach in that recombination 
+in non-ancestral material is possible. Although this does not affect the distribution 
+of marginal genealogies, it does mean that not all lineages in $L$ can be represented 
+by a single interval. More precisely, all lineages will be associated with 
+a single interval up to the point where one section of the 
+genome has reached its most recent common ancestor but the neighbouring parts have not.
+
+\subsection{Recording an ARG} \label{par:recording}
+
+The process described in the previous paragraph details the transitions of the 
+set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
+$\mathcal{G} = (N, E)$ is obtained by registering a tuple $(c, p, X_{cp})$ whenever a 
+lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
+replaced by one (coalescence) or two lineage(s) (recombination), $p$ (and $p^{\prime}$).
+The last element of the tuple is the extent of the edge. It is itself a  
+disjoint set of intervals such that $X_{cp} \cup X_{cp^{\prime}} = X_c$. 
+The set of edges $E$, containing all such tuples $(c, p, X_{cp})$, 
+and $N$, the set of nodes representing all genomes in the ARG, together
+define this data structure.\\
+
+Recording the full process as described above will result in unary nodes.
+These are nodes that will have only one child in one or more local trees 
+(see fig. \ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
+for the nodes registered for those lineages hit by a recombination event. 
+In case of a coalescence event, a node is registered along the full 
+extent of the original lineage $X_c$. 
+Coalescence between two lineages without perfectly 
+overlapping intervals will therefore result in locally unary nodes.\\
+
+The resulting data structure can be simplified to various degrees \citep{Wong-2023}. 
+This equates to removing certain types of unary nodes, 
+depending on the degree of information 
+we want to retain (or can reconstruct) from the original process. 
+Unary nodes therefore are a vital link between 
+ARGs and local trees. In a fully simplified ARG, without unary nodes, 
+the local trees only store information on local coalescence events.
+
+Here we will compute the likelihood for an ARG
+under the SMC for which we have no information on the time to any 
+of the recombination events. We will assume however, that the identity 
+of the lineage involved in each recombination event can be correctly 
+inferred by relying on the unary nodes (see section X.X on locally unary nodes). 
+This enables us to extend likelihood computations to the 
+a more general class of ARGs.
+
+\begin{figure}[h!] \label{fig:smc-unary}
+\centering
+\includegraphics[width=\textwidth]{smc-custom-haplotypes-mini.png}
+\caption{SMC with unary nodes}
+\end{figure}
+
+\subsection{Unary nodes} \label{par:unary}
+
+The SMC with unary nodes
+Unary nodes connect marginal trees to the ARG.
+These are inferred by \tsinfer, \kwarg, \argweaver. %add citation
+
+% what happens in the partial or complete absence of unary nodes
+% even in the absence of unary nodes we can compute a likelihood, however, this has an impact.
+
+% The SMC and unary nodes
+% ADD: figure here that tries to deal with all possible causes of confusion
+% unary nodes mainly
+
+% highlight the amount of information that needs to be tracked to compute a likelihood
+% for the exact Hudson CwR
+
+\subsection{Calculating likelihoods} \label{par:liks}
+
+Retracing the ARG backwards in time, each coalescence event is associated with two child 
+lineages, $c$ and $d$, and a single parent, $p$. By defining $C_i(t)$ asymmetrically, 
+the history of each these two child lineages can be considered independently without 
+risking to double count any of the coalescence events. The likelihood of 
+observing the edge $(c, p, X_{cp})$ can thus be considered independently from 
+the likelihood of all other edges, and in particular from the likelihood 
+of $(d, p, X_{dp})$.\\
+
+% extends from here means min(X_{cp})=x, max(X_{cp})=y
+% improve this description here
+Assuming $(c, p, X_{cp})$ extends from $x$ to $y$ along the genome and $t_i$ 
+indicates the age of 
+node $i$ in the ARG, then the probability of not observing a recombination across the 
+entire area (span x depth) of this edge is
+
+% full extent of segment
+\begin{equation}\label{eq:span}
+A_{(c, p)}(\theta) = e^{-r (y-x)(t_p - t_{c})}
+\end{equation}
+% also x and y not necessarily equal to x_{c_{1}0} and y_{c_{1}m}.
+
+What's left to compute the likelihood, is to take into account all events 
+associated with $x$, 
+the leftmost coordinate of $(c, p, X_{cp})$. This is either only the 
+coalescence event involving 
+$c$ and $d$. In this case $x=x_{c0}$. Or, $x$ can a new recombination break point that 
+occurred on the lineage represented by $c$ at some time $s$ between $t_{c}$ 
+and $\min(t_p, t_{p^{\prime}})$, where $p^{\prime}$ is the parent of the 
+edge $(c, p^{\prime}, X_{cp^{\prime}})$ ending at $x$.
+In the latter case the new lineage starting at $x$ can be given a temporary label 
+$c^{\prime}$ such that with the same logic from the previous paragraph we 
+can define $F(c^{\prime}, s, t) = \int_{s}^{t} I_{c^{\prime}}(u)du$.
+
+\begin{equation}\label{eq:depth}
+B_{(c, p)}(\theta) = \begin{cases}
+e^{-\lambda F(c, t_c, t_p)} \lambda^{I_{cd}(t_p)} & x=x_{c0} \\
+\int_{t_c}^{t_{p} \wedge t_{p^{\prime}}} r e^{-rs} e^{-\lambda F(c^{\prime}, s, t_{p})} ds \lambda^{I_{c^{\prime}d}(t_p)} & x=x_{c^{\prime}0}>x_{c0} \\
+\end{cases}
+\end{equation}
+
+Here, the (second) exponential term gives the probability of not observing a 
+coalescence event before $t_p$. In case of a recombination, $re^{-rs}$ further 
+gives the probability of observing a recombination 
+at time $s$ given a Poisson process along position $x$ with rate $r$. The 
+time of the recombination is integrated out.
+The last term gives the point probability density of the coalescence event 
+that terminates the edge. The indicator function is used to resolve 
+the non-reciprocal nature of coalescence events as defined here.\\
+
+The likelihood of ARG $\mathcal{G}$ defined by the set of edges $E$ and
+given parameters $\theta$ then is
+
+\begin{equation}\label{eq:full-lik}
+\mathcal{L}(\mathcal{G}|\theta) = \prod_{(c, p) \in E} A_{(c, p)}(\theta) * B_{(c, p)}(\theta)
+\end{equation}
+
+% Does this hold for the case beyond the root of the genealogy?
+
+Because the likelihood is computed based on edges,
+and edges are no longer recorded once the local most recent common ancestor 
+has been reached, we will underestimate the number of positions where 
+a recombination could have occurred in case a recombination hits a 
+lineage with trapped non-ancestral material (see end \ref{par:description}). 
+If we assume that only a single recombination 
+occurred, then it suffices however to provide a correction factor $g$ in equation
+\ref{eq:depth}. Here $g$ represents the 
+length of non-ancestral material along which a recombination would have resulted  
+in the same observed local sequence of tree changes.
+
+\begin{figure}[h!] \label{fig:algo}
+\centering
+\includegraphics[width=0.9\textwidth]{smc_algo.png}
+\caption{Description of algorithm}
+\end{figure}
+
+\begin{figure}[h!] \label{fig:algo-flat}
+\centering
+\includegraphics[width=\textwidth]{smc_algo_flat.png}
+\caption{Description of algorithm}
+\end{figure}
+
+\subsection{Tree sequence format} \label{par:algo}
+
+Although the argument outlined in the previous paragraph is valid for any ARG, here we 
+will describe how to (efficiently) extract the information required to associate a
+tree sequence encoding of an ARG with a likelihood under the SMC.\\
+
+Formally, we represent a tree sequence using a set of tables \citep{kelleher_efficient_2018}. 
+This simple tabular representation of any ARG $\mathcal{G} = (N, E)$, can be 
+augmented with additional tables with information on sequence variation
+using a site and mutation table. The node table stores information on all 
+haplotypes $N$ in the tree sequence. Information about how nodes relate to 
+each other along the genome is defined in a tabular encoding of $E$.
+Each row in this edge table constitutes a tuple \texttt{(left, right, parent, child)}.
+Note that we defined the extent of any edge $\epsilon \in \mathcal{G}$ as the union of 
+disjoint intervals. In the tree sequence encoding however, $\epsilon$ will be represented 
+by a single entry for each contiguous interval. As remarked before, under the SMC, this 
+distinction only matters beyond the root of the genealogy. The distinction between both 
+types of edges will therefore not be made in the remainder of the paper.\\
+
+The likelihood computation as described above can be done with a single pass 
+of the edge table.
+The central problem however is to efficiently maintain the state $I_e(t)$, which counts 
+the number of lineages (the lineage associated with) the focal edge $\epsilon$ can coalesce 
+with at time $t$.
+
+The definition of $I_{kl}$ (\ref{def:coal}) is based on the assumption 
+that we can impose a strict total order on all edges based on their starting point.
+
+Because of the SMC:
+an edge starting at $x$ can only coalesce 
+with a subset of the other edges that make up the local tree at position $x$.
+
+Because local tree changes only occur whenever edges are removed and/or inserted,
+we will rely on the ordering of these insertions and removals as implemented in
+\texttt{tskit}, to ensure this state can be correctly and efficiently maintained.\\
+
+% edge diff algorithm
+The order in which edges are inserted and removed is determined by the index 
+vectors $\textbf{i}$ and $\textbf{o}$ \citep{kelleher_efficient_2016}. The edge insertion vector 
+$\textbf{i}$ gives the ordering of edges sorted by left endpoint, and among edges 
+with the same left endpoint, sorted so that edges closer to the root appear later. 
+The edge removal vector $\textbf{o}$ is similar, except gives the ordering of edges 
+by right endpoint and with edges closer to the root appearing sooner. This order imposes the 
+strong ordering on all edges as required by the definition of $I_{kl}$. The variables $\texttt{i}$ and 
+$\texttt{o}$ variables are used to keep track of our position in the edge insertion and edge removal 
+indexes.\\
+
+% the counting algorithm
+To transition from one tree to the next, we first iterate across the indices of all edges 
+in $\textbf{o}$ from its current index up to $\texttt{o}$. For each edge in we decrement $I$ 
+from index \texttt{edge.child} to \texttt{edge.parent}. We subsequently iterate across the 
+edges in $\textbf{i}$ up to index $\texttt{i}$. For each new edge, we first compute the likelihood 
+associated with that edge and then 
+increment $I$ from \texttt{edge.child} to \texttt{edge.parent}.\\
+
+% improve this section, then transition to limitations of unary nodes
+We further rely on the assumption that the (inferrable, see next paragraph) unary nodes 
+provide us with the required information on both the extent of all extant lineages through 
+time as well as on the time boundaries for any recombination event and the identity of the 
+lineage hit by that recombination.
+
+% section on how deviations from the ideal assumptions affect the algorithm
+% extent of lineages
+% version of the algorithm with polytomies
+
+% version of the algorithm in the absence of unary nodes
+
+\subsection{Flexibility}
+
+By explicitly associating a single likelihood with each edge, we can compute a likelihood 
+under the SMC for any valid tree sequence.
+Strength of ARGs is the ability to gain temporal resolution. This approach allows us to fully 
+benefit from this. %elaborate
+We can readily incorporate changes in $N_e$ and $r$, in both dimensions (time and along the genome).
+Similarly, any subset of the tree sequence that is itself a valid tree sequence can again 
+be used to compute a likelihood. Such operations include: slicing genome both in time, and space.
+
+\subsubsection{Slicing the ARG in time}
+% TO DO: give example why this might be useful.
+
+\subsubsection{Slicing the ARG in (genome) space}
+%TO DO: give example why this might be useful.
+
+\subsubsection{Variable Ne}
+
+\subsubsection{Variable r}
+
+%\subsubsection{Node labellings}
+% implement idea from Guo et al. 2022.
 
 \section{Discussion}
+% summary: what did we do here
+
+% future directions: 
+incorporating node labels (see \citep{guo_recombination-aware_2022}).
+
+First step towards a genealogy sampler (for any ARG, and at scale): quantifying uncertainty 
+(with inferred ARG as starting point)
+
+resolve polytomies in a model-based way
+
+% explore the information in unary nodes
+
+\bibliographystyle{plainnat}
+\bibliography{paper, temp}
+
+\section{Supplementary material}
 

### 2023-11-06 paper.tex
add figures

@@
 SMC is slightly different from the left-to-right approach in that recombination 
-in non-ancestral material is possible. Although this does not affect the distribution 
+in non-ancestral material is possible (see fig.\ref{fig:smc-unary}). 
+Although this does not affect the distribution 
 of marginal genealogies, it does mean that not all lineages in $L$ can be represented 
@@
 These are nodes that will have only one child in one or more local trees 
-(see fig. \ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
+(see fig.\ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
 for the nodes registered for those lineages hit by a recombination event. 
@@
 under the SMC for which we have no information on the time to any 
-of the recombination events. We will assume however, that the identity 
+of the recombination events. 
+We will assume however, that the identity 
 of the lineage involved in each recombination event can be correctly 
 inferred by relying on the unary nodes (see section X.X on locally unary nodes).
-This enables us to extend likelihood computations to the 
-a more general class of ARGs.
+As highlighted in fig.\ref{fig:smc-unary}, unary nodes can be inferred from the 
+available haplotype information. 
+Currently, the following methods return this information by default 
+as part of their ARG reconstruction algorithm: 
+\argweaver \citep{rasmussen_genome_wide_2014} 
+\tsinfer \citep{kelleher_inferring_2019}, 
+\kwarg \citep{ignatieva_kwarg_2021}.
+The presence of these unary nodes enables us to extend likelihood computations 
+to the a more general class of ARGs.
 
-\begin{figure}[h!] \label{fig:smc-unary}
+\begin{figure}[!ht]
 \centering
-\includegraphics[width=\textwidth]{smc-custom-haplotypes-mini.png}
-\caption{SMC with unary nodes}
+\includegraphics[width=\textwidth]{figures/smc-custom-haplotypes-mini.png}
+\caption{SMC with unary nodes: Tree sequence encoding of an ARG (top left) 
+generated under the SMC. The haplotypes that make up the lineages coalescing 
+at the root of each marginal tree are highlighted at the top. This ancestral 
+material is coloured based on the samples that inherit from it.
+These reveal the details on the presence of the unary nodes. 
+Note how the presence of the unary nodes 
+modifies the left-to-right logic of the prune regrafts 
+under the SMC (blue dotted arrows).
+The unary nodes allow us to keep track of the full extent of all lineages 
+through time (under the SMC) and thus infer the lineage along which a 
+recombination event happened.}
+\label{fig:smc-unary}
 \end{figure}
@@
 
-The SMC with unary nodes
-Unary nodes connect marginal trees to the ARG.
-These are inferred by \tsinfer, \kwarg, \argweaver. %add citation
+Recording/inferring unary nodes allow us to completely reconstruct 
+$X(t)$ through time under the SMC up to the exact 
+time of the recombination events. \\
 
-% what happens in the partial or complete absence of unary nodes
-% even in the absence of unary nodes we can compute a likelihood, however, this has an impact.
+The SMC and unary nodes. \\
 
-% The SMC and unary nodes
-% ADD: figure here that tries to deal with all possible causes of confusion
-% unary nodes mainly
-
-% highlight the amount of information that needs to be tracked to compute a likelihood
-% for the exact Hudson CwR
+inferring recombination = change of parent. \\
 
@@
 \centering
-\includegraphics[width=0.9\textwidth]{smc_algo.png}
+\includegraphics[width=0.9\textwidth]{figures/smc_algo.png}
 \caption{Description of algorithm}
@@
 \centering
-\includegraphics[width=\textwidth]{smc_algo_flat.png}
+\includegraphics[width=\textwidth]{figures/smc_algo_flat.png}
 \caption{Description of algorithm}
@@
 will describe how to (efficiently) extract the information required to associate a
-tree sequence encoding of an ARG with a likelihood under the SMC.\\
+tree sequence encoding of an ARG with a likelihood under the SMC (see fig. \ref{fig:smc-unary}).\\
 
@@
 distinction only matters beyond the root of the genealogy. The distinction between both 
-types of edges will therefore not be made in the remainder of the paper.\\
+types of edges will therefore not be made in the remainder of the paper.
+% Although it does requires us to keep track of some additional information to avoid
+% wrongfully assuming a recombination event happened.
+% Similarly, the distinction between the lineage associated with an edge, and the edge 
+% itself can render things very verbose.
+\\
 
@@
 with at time $t$. 
+The definition of $I_{kl}$ (\ref{def:coal}) is based on the assumption 
+that we can impose a strict total order on all (overlapping) lineages based on their 
+starting point.
 
-The definition of $I_{kl}$ (\ref{def:coal}) is based on the assumption 
-that we can impose a strict total order on all edges based on their starting point.
-
-Because of the SMC:
+%%%% improve this %%%%%%5
+The lineages associated with a particular edge can only coalesce with other lineages 
 an edge starting at $x$ can only coalesce 
 with a subset of the other edges that make up the local tree at position $x$.
-
+%%%% 
 Because local tree changes only occur whenever edges are removed and/or inserted,
@@
 under the SMC for any valid tree sequence.
-Strength of ARGs is the ability to gain temporal resolution. This approach allows us to fully 
-benefit from this. %elaborate
-We can readily incorporate changes in $N_e$ and $r$, in both dimensions (time and along the genome).
+
+The \texttt{tskit} library provides a whole range  
+of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
+and simplification.
 Similarly, any subset of the tree sequence that is itself a valid tree sequence can again 
@@
 
+Strength of ARGs is the ability to gain temporal resolution. This approach allows us to fully 
+benefit from this.
+
+We can readily incorporate changes in $N_e$ and $r$, in both dimensions (time and along the genome).
+
+% HOW DO WE HIGHLIGHT THESE FEATURES: SUPPLEMENTS?
 \subsubsection{Slicing the ARG in time}
@@
 % summary: what did we do here
+Here, we have introduced a scalable algorithm to compute likelihoods 
+under the SMC for a general class of ARGs. 
+Important step moving forwards towards bridging the gap between the tremendous 
+recent advances in ARG inference and current state of the art statistical inference of 
+past demography and selection (but see). % some/many of these approaches 
+% ignore recombination
 
@@
 
-First step towards a genealogy sampler (for any ARG, and at scale): quantifying uncertainty 
-(with inferred ARG as starting point)
+First step towards a genealogy sampler (for any ARG, and at scale): 
+quantifying uncertainty 
+(with inferred ARG as starting point). Compared to existing MCMC approaches, 
+we further restrict the size of ARG space by 
+integrating out the exact time to the recombination event.
 
-resolve polytomies in a model-based way
+% discuss accurracy:
+% there will be scenarios where the assumptions made here are not met: e.g. polytomies
+% yet we can still always compute a likelihood.
+% we can alwyas associate a value, what remains to be verified what the information is in that
+% case.
 

### 2023-11-07 workflow/cwr-v-smc.py
add workflow

@@ -1,0 +1,78 @@
+import msprime
+import sys
+import tskit
+
+NE="10000" #diploid
+MU="1.25e-8"
+RE="1.25e-8"
+SEED="1024"
+ID="chr1"
+SEQLEN=1000000
+SAMPLES=100
+
+OUTPUT_DIR = '../true-trees'
+
+def sim_ts(seed):
+	ts = msprime.sim_ancestry(
+		samples=SAMPLES,
+		population_size=NE,
+		random_seed=seed,
+		model='hudson',
+		record_full_arg=True,
+		sequence_length=SEQLEN
+	)
+	return ts
+
+def simplify_ts(ts):
+	node_type = msprime.NODE_IS_CA_EVENT | msprime.NODE_IS_RE_EVENT
+	retain_nodes = np.bitwise_and(ts.tables.nodes.flags, node_type) == 0
+	nodes = np.arange(ts.num_nodes)[retain_nodes]
+	tss = ts.simplify(
+        samples=nodes,
+    )
+	num_samples = ts.num_samples
+	tables = tss.dump_tables()
+	# modify tables to correctly mark sample nodes again
+	for j in range(num_samples, tss.num_nodes):
+	    node_obj = tables.nodes[j]
+	    node_obj = node_obj.replace(flags=0)
+	    tables.nodes[j] = node_obj
+	ts2 = tables.tree_sequence()
+	return ts2.simplify(keep_unary=True)
+
+def compute_likelihood(ts):
+	pass
+
+
+def main():
+	if len(sys.argv) == 4:
+		unique_id = sys.argv[1]
+		num_replicates = sys.argv[2]
+		start_seed = sys.argv[3]
+	else:
+		unique_id = "test_sim"
+		start_seed = 42
+		num_replicates = 10
+	
+	# Set a fixed random seed for reproducibility
+	rng = np.random.default_rng(start_seed)
+	random_seeds = rng.integers(1, 2**31, size=num_replicates)
+    results = np.zeros((2, random_seeds.size), dtype=np.float64)
+
+    for i in random_seeds.size:
+    	ts = sim_ts(random_seeds[i])
+    	results[0, i] = msprime.log_arg_likelihood(
+    		ts, 
+    		Ne=NE,
+    		recombination_rate=RE,
+    	)
+    	tss = simplify_ts(ts)
+    	results[1,i] = compute_likelihood(tss)
+
+    output = f'results_{unique_id}_{num_replcates}_{start_seed}.tsv'
[…]

### 2023-11-07 workflow/infer_relate_trees.sh

@@ -1,0 +1,50 @@
+#binaries from website
+RELATE_DIR=`pwd`/../../../relate
+#git clone https://github.com/leospeidel/relate_lib.git
+#cd relate_lib && mkdir -p build && cd build
+#cmake .. && make && cd ../..
+RELATE_LIB=`pwd`/../../../relate_lib 
+
+#hapmap
+printf '
+cl <- %s
+cat("Position(bp) Rate(cM/Mb) Map(cM)\n")
+cat("0 1.0 0.0\n")
+cat(as.integer(cl), "0.0", cl/1e6, "\n")
+' $SEQLEN | R --slave >$ID.hapmap
+
+NE="20000" #haploid
+MU="1.25e-8"
+SEED="1024"
+ID="chr1"
+SEQLEN=1000000
+SAMPLES=100
+
+WORK_DIR=`pwd`/sim_$SEED
+IN_DIR=$WORK_DIR/relate_inputs
+OUT_DIR=$WORK_DIR/relate_outputs
+mkdir -p $WORK_DIR
+mkdir -p $IN_DIR
+mkdir -p $OUT_DIR
+
+#dump into haplotypes
+$RELATE_DIR/bin/RelateFileFormats \
+  --mode ConvertFromVcf --haps $ID.haps \
+  --sample samples -i $ID
+
+#ancestral states
+awk 'NR > 1 { printf(">%s\n%s\n", $3, $4); }' ID.vcf >$ID.anc.fa
+
+#infer trees
+$RELATE_DIR/scripts/PrepareInputFiles/PrepareInputFiles.sh \
+  --haps $ID.haps \
+  --sample samples \
+  --ancestor $ID.anc.fa \
+  -o $ID
+$RELATE_DIR/bin/Relate --mode All -m $MU -N $NE \
+  --haps $ID.haps.gz --sample sample.gz --annot $ID.annot \
+  --dist $ID.dist.gz --map $ID.hapmap -o $ID
+
+# convert to tree sequence with edges linked across trees
+$RELATE_LIB/bin/Convert --mode ConvertToTreeSequence \
+  --anc $OUT_DIR/$ID.anc.gz --mut $OUT_DIR/$ID.mut.gz --iterations 1000 --compress -o $OUT_DIR/$ID

### 2023-11-07 workflow/ts_to_argneedle.py

@@ -1,0 +1,55 @@
+# vcf to .sites
+import os
+import tskit
+import numpy as np
+import subprocess
+
+from tqdm import tqdm
+
+INPUT_DIR = '/scratch/gbisshop/output/n_100/seq_1g/temp'
+INFO_DIR = '/scratch/gbisshop/output/n_100/seq_1g/argneedle-info'
+OUTPUT_DIR = '/scratch/gbisshop/output/n_100/seq_1g/argneedle-arg'
+ARGNEEDLE_TS_DIR = '/scratch/gbisshop/output/n_100/seq_1g/argneedle-trees'
+TS_DIR = '/ceph/users/gbisshop/true-trees/n_100/seq_1g'
+
+NE = 10000
+MU = 1.25e-8
+REC = 1.25e-8
+NUM_PROCESSES = 20
+
+def run_argneedle(seed):
+    subprocess.run([
+        "arg_needle",
+        "--hap_gz", INPUT_DIR + f'/{seed}.haps',
+        "--map", INPUT_DIR + f'/{seed}.map',
+        "--normalize", "1",
+        "--normalize_demography", INFO_DIR + '/demo.demo',
+        "--out", OUTPUT_DIR + f'/{seed}'
+    ])
+
+
+def generate_demo():
+    with open(INFO_DIR + '/demo.demo', 'w') as file:
+        print(f'0.0\t{NE * 2}')
+
+def generate_map(seed):
+    ts = tskit.load(TS_DIR + f'/{seed}.trees')
+    with open(INFO_DIR + f'/{seed}.map', 'w') as file:
+        for variant in ts.variants():
+            if variant.num_alleles <= 2:
+                # chr1, snp_id, cM, bp
+                rec_pos = variant.position * (REC * 100)
+                print(f'1\tsnp{variant.site.id}\t{rec_pos}\t{variant.position+1}', file=file)
+
+def convert_argweaver(infile):
+    pass
+
+
+def main():
+    seeds = [file.rstrip('.trees') for file in os.listdir(TS_DIR)]
+    for seed in tqdm(seeds, total=len(seeds)):
+        run_argneedle(seed)
+
+
+if __name__ == '__main__':
+    main()

### 2023-11-07 workflow/ts_to_argweaver.py

@@ -1,0 +1,226 @@
+# vcf to .sites
+import os, sys
+import tskit
+import numpy as np
+import subprocess
+import networkx as nx
+import pandas as pd
+import collections
+import msprime
+
+from multiprocessing import Pool
+
+from tqdm import tqdm
+
+TS_DIR = '/ceph/users/gbisshop/true-trees/n_100/seq_1g'
+OUTPUT_DIR = '/scratch/gbisshop/output/n_100/seq_1g/argweaver-sites'
+ARGWEAVER_SMC_DIR = '/scratch/gbisshop/output/n_100/seq_1g/argweaver-smc'
+ARGWEAVER_ARG_DIR = '/scratch/gbisshop/output/n_100/seq_1g/argweaver-arg'
+ARGWEAVER_DIR = '/ceph/users/gbisshop/software/argweaver/bin/arg-sample'
+ARGWEAVER_CONV_DIR = '/ceph/users/gbisshop/software/argweaver/bin/smc2arg'
+ARGWEAVER_PATH = '/ceph/users/gbisshop/software/argweaver/'
+NE = 10000
+MU = 1.25e-8
+REC = 1.25e-8
+NUM_PROCESSES = 20
+
+def run_argweaver(sites):
+    seed = sites.rstrip('.sites')
+    file = OUTPUT_DIR + f'/{sites}'
+    subprocess.run([
+        ARGWEAVER_DIR,
+        "--sites", file,
+        "--output", ARGWEAVER_SMC_DIR + f'/{seed}',
+        "--overwrite",
+        "--popsize", str(NE),
+        "--mutrate", str(MU),
+        "--recombrate", str(REC),
+        "--iters", "1000",
+        "--sample-step", "10",
+        "--no-compress-output",
+    ])
+
+
+def convert_argweaver(infile):
+    """
+    Convert an ARGweaver .arg file to a tree sequence. `infile` should be a filehandle,
+    e.g. as returned by the `open` command. An example .arg file is at
+
+    https://github.com/CshlSiepelLab/argweaver/blob/master/test/data/test_trans/0.arg
+    """
+    start, end = next(infile).strip().split()
+    assert start.startswith("start=")
+    start = int(start[len("start=") :])
+    assert end.startswith("end=")
+    end = int(end[len("end=") :])
+    # the "name" field can be a string. Force it to be so, in case it is just numbers
+    df = pd.read_csv(infile, header=0, sep="\t", dtype={"name": str, "parents": str})
[…]

### 2023-11-07 workflow/ts_to_argweaver_mini_short.py

@@ -1,0 +1,271 @@
+# vcf to .sites
+import os, sys
+import tskit
+import numpy as np
+import subprocess
+import networkx as nx
+import pandas as pd
+import collections
+import msprime
+import tsdate
+from multiprocessing import Pool
+
+from tqdm import tqdm
+from algorithm import *
+
+OUTPUT_DIR = '/scratch/gbisshop/output/argweaver-mini-short'
+ARGWEAVER_DIR = '/ceph/users/gbisshop/software/argweaver/bin/arg-sample'
+ARGWEAVER_PATH = '/ceph/users/gbisshop/software/argweaver/'
+ARGWEAVER_CONV_DIR = '/ceph/users/gbisshop/software/argweaver/bin/smc2arg'
+
+NE = 10000
+MU = 2.5e-8
+REC = 1.25e-8
+NUM_PROCESSES = 1
+
+ANC_SEED = 124511
+MUT_SEED = 452121
+
+def run_sim():
+    ts = msprime.sim_ancestry(
+        samples=10,
+        population_size=NE,
+        recombination_rate=REC,
+        sequence_length=1e4,
+        random_seed=ANC_SEED,
+        record_full_arg=True
+    )
+    ts = msprime.sim_mutations(ts, rate=MU, random_seed=MUT_SEED)
+    ts.dump(OUTPUT_DIR + f"/true-arg-{ANC_SEED}-{MUT_SEED}.trees")
+    return ts
+
+
+def run_argweaver(sites, max_time):
+    ntimes = 100
+    seed = sites.rstrip('.sites')
+    file = OUTPUT_DIR + f'/{sites}'
+    subprocess.run([
+        ARGWEAVER_DIR,
+        "--sites", file,
+        "--output", OUTPUT_DIR + f'/{seed}',
+        "--overwrite",
+        "--popsize", str(NE),
+        "--mutrate", str(MU),
+        "--recombrate", str(REC),
+        "--iters", "1000",
+        "--sample-step", "10",
+        "--no-compress-output",
+        "--maxtime", f"{max_time}",
+        "--ntimes", f"{ntimes}"
+    ])
+
+
+def convert_argweaver(infile):
+    """
+    Convert an ARGweaver .arg file to a tree sequence. `infile` should be a filehandle,
+    e.g. as returned by the `open` command. An example .arg file is at
+
+    https://github.com/CshlSiepelLab/argweaver/blob/master/test/data/test_trans/0.arg
+    """
+    start, end = next(infile).strip().split()
[…]

### 2023-11-07 paper.tex
update algorithm description

@@
 \usepackage{tikz}
-\usetikzlibrary{math,calc,positioning}
-\usetikzlibrary{calc,positioning}
-\usepackage{subcaption}
+
+%\usepackage{subcaption}
 
@@
 
-More formally, and following the notation of \cite{mcvean_approximating_2005}, at any 
+More formally, and following the notation of \citet{mcvean_approximating_2005}, at any 
 point in time, the process can be described by $L(t)$, the set of labeled lineages 
@@
 lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
-replaced by one (coalescence) or two lineage(s) (recombination), $p$ (and $p^{\prime}$).
+replaced by one or two lineage(s), $p$ (and $p^{\prime}$), 
+in case of a coalescence or recombination respectively.
 The last element of the tuple is the extent of the edge. It is itself a  
@@
 Recording the full process as described above will result in unary nodes.
-These are nodes that will have only one child in one or more local trees 
+These are nodes that have only one child in one or more local trees 
 (see fig.\ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
 for the nodes registered for those lineages hit by a recombination event. 
-In case of a coalescence event, a node is registered along the full 
-extent of the original lineage $X_c$. 
+In case of a coalescence event, an edge is registered along the full 
+extent of both involved lineages. 
 Coalescence between two lineages without perfectly 
 overlapping intervals will therefore result in locally unary nodes.\\
+
+These unary nodes ensure the one to one correspondence between an ARG and its 
+representation as a sequence of local trees (see fig.\ref{fig:smc-unary}). 
+Indeed, the presence of these unary nodes allow us to completely reconstruct 
+the state of all extant lineages through time from the local trees.
+
+%under the SMC up to the exact time of the recombination events.
 
@@
 we want to retain (or can reconstruct) from the original process. 
-Unary nodes therefore are a vital link between 
-ARGs and local trees. In a fully simplified ARG, without unary nodes, 
+In a fully simplified ARG, without unary nodes, 
 the local trees only store information on local coalescence events.
+Under the SMC, if we only omit recombination nodes, % is there a better word?
+we are left with the coalescent nodes, those nodes that are coalescent 
+along some fraction of the genome. We can however still identify \emph{all} 
+lineages that were hit by a recombination event and constrain the time to 
+this event to within a window defined by the age of the node $c$ associated with that 
+lineage and the minimum age of the parents of the edges connecting to $c$.
 
+As highlighted in fig.\ref{fig:smc-unary}, (coalescing) unary nodes can be inferred 
+from the available haplotype information. Currently, the following methods 
+return this information by default as part of their ARG reconstruction algorithm: 
+\argweaver \citep{rasmussen_genome-wide_2014} 
+\tsinfer \citep{kelleher_inferring_2019}, 
+\kwarg \citep{ignatieva_kwarg_2021}.
+
+The presence of these (coalescing) unary nodes enables us to extend likelihood 
+computations to a more general class of ARGs.
 Here we will compute the likelihood for an ARG
@@
 of the lineage involved in each recombination event can be correctly 
-inferred by relying on the unary nodes (see section X.X on locally unary nodes).
-As highlighted in fig.\ref{fig:smc-unary}, unary nodes can be inferred from the 
-available haplotype information. 
-Currently, the following methods return this information by default 
-as part of their ARG reconstruction algorithm: 
-\argweaver \citep{rasmussen_genome_wide_2014} 
-\tsinfer \citep{kelleher_inferring_2019}, 
-\kwarg \citep{ignatieva_kwarg_2021}.
-The presence of these unary nodes enables us to extend likelihood computations 
-to the a more general class of ARGs.
+inferred by relying on the unary nodes.
 
@@
 \includegraphics[width=\textwidth]{figures/smc-custom-haplotypes-mini.png}
-\caption{SMC with unary nodes: Tree sequence encoding of an ARG (top left) 
-generated under the SMC. The haplotypes that make up the lineages coalescing 
-at the root of each marginal tree are highlighted at the top. This ancestral 
+\caption{ARG (top left) generated under the SMC and its representation as a 
+series of local trees. The presence of the unary nodes ensure these two 
+representations contain the same information.
+The haplotypes that make up the lineages coalescing 
+at the root of each marginal tree are highlighted at the top. Ancestral 
 material is coloured based on the samples that inherit from it.
-These reveal the details on the presence of the unary nodes. 
 Note how the presence of the unary nodes 
-modifies the left-to-right logic of the prune regrafts 
-under the SMC (blue dotted arrows).
-The unary nodes allow us to keep track of the full extent of all lineages 
-through time (under the SMC) and thus infer the lineage along which a 
-recombination event happened.}
+modifies the left-to-right logic of single-prune-regrafts induced by 
+recombination under the SMC (blue dotted arrows).
+Even in the absence of the blue recombination nodes, 
+the unary coalescing nodes allow us to reconstruct the state of all lineages 
+up to the exact time of the recombination event.
+}
 \label{fig:smc-unary}
@@
 \subsection{Unary nodes} \label{par:unary}
-
-Recording/inferring unary nodes allow us to completely reconstruct 
-$X(t)$ through time under the SMC up to the exact 
-time of the recombination events. \\
-
-The SMC and unary nodes. \\
-
-inferring recombination = change of parent. \\
+% Do we need a seperate section on the unary nodes?
+% is there more to be said about the SMC inm the presence of unary nodes?
+Recording unary nodes has interesting consequences on the left-to-right 
+description of the SMC. Looking at a single marginal tree, the presence of 
+unary nodes tells us that either \\
 
@@
 \centering
-\includegraphics[width=0.9\textwidth]{figures/smc_algo.png}
+\includegraphics[width=0.8\textwidth]{figures/smc_algo.png}
 \caption{Description of algorithm}
@@
 
-%%%% improve this %%%%%%5
-The lineages associated with a particular edge can only coalesce with other lineages 
-an edge starting at $x$ can only coalesce 
-with a subset of the other edges that make up the local tree at position $x$.
-%%%% 
-Because local tree changes only occur whenever edges are removed and/or inserted,
-we will rely on the ordering of these insertions and removals as implemented in
+The tree sequence encoding provides us with a natural order on all lineages.  
+For each local tree with left coordinate $x$, lineages associated with an edge 
+starting at $x$ can only coalesce with the subset of lineages associated with an 
+edge that is represented in that same local tree. Because local tree changes 
+only occur whenever edges are removed and/or inserted,
+we can rely on the ordering of these insertions and removals as implemented in
 \texttt{tskit}, to ensure this state can be correctly and efficiently maintained.\\
@@
 % improve this section, then transition to limitations of unary nodes
-We further rely on the assumption that the (inferrable, see next paragraph) unary nodes 
+We further rely on the assumption that the unary (coalescent) nodes 
 provide us with the required information on both the extent of all extant lineages through 
@@
 % section on how deviations from the ideal assumptions affect the algorithm
+However,
+
+% keep track of last parent to identify recombination events
+% recombination events beyond root: we keep track of
+% parent of prior tree
+
 % extent of lineages
+
 % version of the algorithm with polytomies
@@
 \bibliographystyle{plainnat}
-\bibliography{paper, temp}
+\bibliography{paper.bib, temp.bib}
 

### 2023-11-07 paper.tex
algo details

@@
 Clear distinction into two groups: ARGWeaver infers actual ARG.
+% what does kwarg do?
 Whereas the other approaches are heuristic in nature. For example, where ARGweaver has 
@@
 structures. (other peculiarities: ...).
+
+% computing likelihoods on ARGs; from Kuhner and Felsenstein to ARGWeaver
 
@@
 \includegraphics[width=0.8\textwidth]{figures/smc_algo.png}
-\caption{Description of algorithm}
+\caption{Description of algorithm: the algorithm progresses from left to right and 
+keeps track of the . Let's assume we have visited the first 5 edges present in the 
+first tree (tabel 1). Moving onto the next tree, need to remove single edge 
+$(2, 5, [0,42))$, update the counts $I$ by decrementing all intervals between 5 and 
+2. The state of the vector is now $I_6$, the number of extant lineages the lineage 
+associated with edge $(1, 6, [42, 100)$ can coalesce with. Now compute the likelihood 
+associated with this edge. Then insert this edge and increment the count vector between 
+1 and 6. Two intermediate steps to move to the second tree. Similarly again two steps 
+to move to the last tree and insert edge $(5, 7, [70,100))$
+}
+
 \end{figure}
@@
 that we can impose a strict total order on all (overlapping) lineages based on their 
-starting point.
-
-The tree sequence encoding provides us with a natural order on all lineages.  
+starting point. The tree sequence encoding provides us with a natural order on all lineages.  
 For each local tree with left coordinate $x$, lineages associated with an edge 
-starting at $x$ can only coalesce with the subset of lineages associated with an 
-edge that is represented in that same local tree. Because local tree changes 
-only occur whenever edges are removed and/or inserted,
-we can rely on the ordering of these insertions and removals as implemented in
-\texttt{tskit}, to ensure this state can be correctly and efficiently maintained.\\
+starting at $x$ can only coalesce with the subset of lineages associated with one of the
+edges that make up that local tree. Each of these edges will either already have 
+been present in the previous local tree, or will have the same starting point $x$.
+In \texttt{tskit}, the order in which edges are inserted and removed 
+while moving from one tree to the next, is determined by the index 
+vectors $\textbf{i}$ and $\textbf{o}$ \citep{kelleher_efficient_2016}.
+The edge insertion vector $\textbf{i}$ gives the ordering of edges sorted by left 
+endpoint, and among edges with the same left endpoint, sorted so that edges closer 
+to the root appear later. The edge removal vector $\textbf{o}$ is similar, 
+except gives the ordering of edges by right endpoint and with edges closer to the 
+root appearing sooner. The variables $\texttt{i}$ and 
+$\texttt{o}$ are used to keep track of our position in the edge insertion and edge removal 
+indexes as we move along the tree sequence. This order imposes the 
+strong ordering on all edges as required by the definition of $I_{kl}$.\\
 
 % edge diff algorithm
-The order in which edges are inserted and removed is determined by the index 
-vectors $\textbf{i}$ and $\textbf{o}$ \citep{kelleher_efficient_2016}. The edge insertion vector 
-$\textbf{i}$ gives the ordering of edges sorted by left endpoint, and among edges 
-with the same left endpoint, sorted so that edges closer to the root appear later. 
-The edge removal vector $\textbf{o}$ is similar, except gives the ordering of edges 
-by right endpoint and with edges closer to the root appearing sooner. This order imposes the 
-strong ordering on all edges as required by the definition of $I_{kl}$. The variables $\texttt{i}$ and 
-$\texttt{o}$ variables are used to keep track of our position in the edge insertion and edge removal 
-indexes.\\
+The size of the subset of lineages in each internode interval each lineage 
+can coalesce with can thus be maintained by keeping track of a single vector $I$ 
+that is decremented or incremented as defined by
+$\textbf{o}[\texttt{o}]$ and $\textbf{i}[\texttt{i}]$, 
+Further keep track of a last-parent array to detect recombination events.
+We can thus compute a likelihood for the tree sequence, while visiting each edge only once.
 
-% the counting algorithm
+More concretely, prior to inserting an edge, $I$ represents the number of lineages 
+the lineage associated with 
+that edge can coalesce with. Then insert the edge and update the counts.
+
 To transition from one tree to the next, we first iterate across the indices of all edges 
@@
 
-% improve this section, then transition to limitations of unary nodes
+% improve this section, transition to modifications in the algorithm in case of 
+% incomplete unary nodes, polytomies, ...
 We further rely on the assumption that the unary (coalescent) nodes 
@@
 % section on how deviations from the ideal assumptions affect the algorithm
-However,
 
@@
 
-\subsection{Flexibility}
+\subsection{Flexibility / general class of ARGs}
 
+% highlight in what ways we can now compute likelihoods for a GENERAL class of ARGs
+
+% slicing of the ARG
 By explicitly associating a single likelihood with each edge, we can compute a likelihood 
-under the SMC for any valid tree sequence.
+under the SMC for any valid tree sequence. The \texttt{tskit} library provides a whole range  
+of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
+as well as reduce the number of tracked samples. 
+Any such subset of the tree sequence is itself a valid tree sequence 
+that can again be used to compute a likelihood with.
 
-The \texttt{tskit} library provides a whole range  
-of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
-and simplification.
-Similarly, any subset of the tree sequence that is itself a valid tree sequence can again 
-be used to compute a likelihood. Such operations include: slicing genome both in time, and space.
+Our approach therefore strenghtens the inherent ability of ARGs to provide us with temporal 
+resolution on the inferred genealogical history of a sample.
 
-Strength of ARGs is the ability to gain temporal resolution. This approach allows us to fully 
-benefit from this.
-
+% something on the process backwards in time
 We can readily incorporate changes in $N_e$ and $r$, in both dimensions (time and along the genome).
@@
 
-%\subsubsection{Node labellings}
-% implement idea from Guo et al. 2022.
-
 \section{Discussion}
@@
 recent advances in ARG inference and current state of the art statistical inference of 
-past demography and selection (but see). % some/many of these approaches 
+past demography and selection (but see ). % some/many of these approaches
+
 % ignore recombination

### 2023-11-08 paper.tex
update figures

@@
 \newcommand{\arbores}[0]{\texttt{Arbores}}
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% deal with supplementary items
+\newcommand{\supplementarysection}{%
+  \setcounter{figure}{0}% Reset figure counter
+  \let\oldthefigure\thefigure% Capture figure numbering scheme
+  \renewcommand{\thefigure}{S\oldthefigure}% Prefix figure number with S
+  \section{Supplementary section}% Set supplementary section
+}
 
@@
 \section{Introduction}
+
+% Main motivation:
+Major progress in ARG inference.
+
+Result is whole range of inference methods, all inferring an ARG with their own specific 
+characteristics.
+Many of these scalable heuristic methods do not explicitly infer recombination events.
+
+%ARGs that are inferred here lack information to enable us to compute 
+% a likelihood under the Hudson coalescent.
+% Missing: exact time to recombination + non-inferrable recombinations
+
+Currently no way to connect the output of the diverse range of ARG inference methods back 
+to a model that would allow us to perform a probabilistic exploration of the 
+ARG space around the point estimate, (or likelihood-based inference, ...)
+
 % What are ARGs, their relevance and importance
@@
 
-% previous likelihood calculations on ARGsL Kuhner and Felsenstein (2000), ARGweaver
+% previous likelihood calculations on ARGs Kuhner and Felsenstein (2000), ARGweaver
 
 % history of approximating the coalescent: 
-SMC (McVean and Cardin, 2005) SMC' (Marjoram and Wall, 2006).
-Also heuristic approximations: Li and Stephens (2003) extending Stephens and Donnelly 
-(2000)
-These have enabled (scalable) ARG inference methods: ARGweaver (relying on SMC, limited 
-to smaller samples, but ability to sample from posterior, and full ARG), and Relate, 
-tsinfer/tsdate (copying algorithm), ArgNeedle (using SMC) 
-Clear distinction into two groups: ARGWeaver infers actual ARG.
+%SMC (McVean and Cardin, 2005) SMC' (Marjoram and Wall, 2006).
+%Also heuristic approximations: Li and Stephens (2003) extending Stephens and Donnelly (2000)
+%These have enabled (scalable) ARG inference methods: ARGweaver (relying on SMC, limited 
+%to smaller samples, but ability to sample from posterior, and full ARG), and Relate, 
+%tsinfer/tsdate (copying algorithm), ArgNeedle (using SMC) 
+%Clear distinction into two groups: ARGWeaver infers actual ARG.
 % what does kwarg do?
-Whereas the other approaches are heuristic in nature. For example, where ARGweaver has 
-the built in assumption of neighbouring trees being separated by single Subtree Prune 
-Regraft (SPR) (Song citation?), this is not the case for any of the heuristic data 
-structures. (other peculiarities: ...).
+%Whereas the other approaches are heuristic in nature. For example, where ARGweaver has 
+%the built in assumption of neighbouring trees being separated by single Subtree Prune 
+%Regraft (SPR) (Song citation?), this is not the case for any of the heuristic data 
+%structures. (other peculiarities: ...).
 
@@
 description of the SMC. Looking at a single marginal tree, the presence of 
-unary nodes tells us that either \\
+unary nodes are telltale signs of future or past coalescence events (in a 
+left-to-right sense).\\
 
@@
 \centering
-\includegraphics[width=\textwidth]{figures/smc_algo_flat.png}
-\caption{Description of algorithm}
+\includegraphics[width=\textwidth]{figures/ts_algo_2rows.png}
+\caption{Description of algorithm: Moving along the genome we transition from the first 
+local tree (1) to the next (4). For each edge that is removed from the first tree we 
+decrement the count in the corresponding internode intervals and update the identity of 
+the last parent. Subsequently, for each edge that will be inserted we compute the likelihood 
+associated with that edge prior to incrementing the counts.}
 \end{figure}
@@
 
-% improve this section, transition to modifications in the algorithm in case of 
-% incomplete unary nodes, polytomies, ...
-We further rely on the assumption that the unary (coalescent) nodes 
-provide us with the required information on both the extent of all extant lineages through 
-time as well as on the time boundaries for any recombination event and the identity of the 
-lineage hit by that recombination.
+% ADD IN RESULTS
+Figure \ref{sup:fig:vs-hudson} shows the correlation between the likelihood under the hudson 
+for an ARG with detailed recombination information simulated and computed using 
+\texttt{msprime} \citep{baumdicker_efficient_2021}. Show impact of unary nodes on $r^2$.
+Simulated mutation data under the SMC and passed this to ARGweaver.
+Scatterplot of coalescent prior as returned by ARGweaver \citep{rasmussen_genome-wide_2014} 
+for a 100 samples from the posterior against the likelihood as described here fore these 
+sampled ARGs \ref{sup:fig:vs-argweaver}
 
-% section on how deviations from the ideal assumptions affect the algorithm
-
-% keep track of last parent to identify recombination events
-% recombination events beyond root: we keep track of
-% parent of prior tree
-
-% extent of lineages
-
-% version of the algorithm with polytomies
-
-% version of the algorithm in the absence of unary nodes
+Implementation of the algorithm is available on https://github.com/GertjanBisschop/runsmc.
 
 \subsection{Flexibility / general class of ARGs}
-
-% highlight in what ways we can now compute likelihoods for a GENERAL class of ARGs
 
@@
 resolution on the inferred genealogical history of a sample. 
+We can further readily incorporate changes in $N_e$ and $r$, 
+both in time as well as along the genome.
 
-% something on the process backwards in time
-We can readily incorporate changes in $N_e$ and $r$, in both dimensions (time and along the genome).
-
-% HOW DO WE HIGHLIGHT THESE FEATURES: SUPPLEMENTS?
-\subsubsection{Slicing the ARG in time}
-% TO DO: give example why this might be useful.
-
-\subsubsection{Slicing the ARG in (genome) space}
-%TO DO: give example why this might be useful.
-
-\subsubsection{Variable Ne}
-
-\subsubsection{Variable r}
+% Algorithm is suitable for a general class of ARGs each with their own specific features.
+% What happens in the case of ARGs that do not meet our assumptions: absent unary nodes, polytomies, ...
+The above algorithm treats each edge independently which implies we can deal with polytomies.
+Furthermore, although we rely heavily on unary nodes to provide us with the necessary 
+information on recombination events, our algorithm can also accomodate for 
+those inferred ARGs that lack unary nodes (e.g.\ \relate, and see figure in supplements).
+However, in case of (sub)tree height changes along the ARG, the algorithm would infer 
+two recombination events along the lineages associated 
+with both edges switching parent nodes. This can be mitigated by limiting the number of 
+'inferrable' recombination events per coalescence event to 1 (see supplementary figure).
 
@@
 under the SMC for a general class of ARGs.
-Important step moving forwards towards bridging the gap between the tremendous 
-recent advances in ARG inference and current state of the art statistical inference of 
-past demography and selection (but see ). % some/many of these approaches
 
-% ignore recombination
+Impact both on ways of doing ARG inference itself and quantifying the uncertainty associated 
+with that as well as performing statistical inference of past demography.
 
-% future directions: 
-incorporating node labels (see \citep{guo_recombination-aware_2022}).
+* statistical inference *
+This is an important step towards bridging the gap between the tremendous 
+recent advances in ARG inference and the current state of the art statistical inference of 
+past demography (but see ). 
+% some/many of these approaches only use the information present in marginal trees.
+The provided theory and associated algorithm will function as a building block of a 
+demography-aware genealogy sampler. Current approach couple of major advantages:
+* any type of inferred ARG as a starting point. Potentially implying a massive reduction in 
+MCMC runtime by reducing the necessary burnin period.
+* by integrating out the exact time to the recombination event we further restrict the size 
+of ARG space that needs to be explored.
+* exploit the efficiencies of the tree sequence data structure (see \citep{Mahmoudi_bayesian_2022})
+* we can readily incorporate $N_e$-changes as well as multiple populations and migration.
 
-First step towards a genealogy sampler (for any ARG, and at scale): 
-quantifying uncertainty 
-(with inferred ARG as starting point). Compared to existing MCMC approaches, 
-we further restrict the size of ARG space by 
-integrating out the exact time to the recombination event.
+* ARG inference *
+Combining ARG inference methods: tsinfer for first few generations (large sample sizes), where 
+the initial phase is biased relative to the Wright-Fisher model when the sample size is large 
+\citep{bhaskar_distortion_2014}, or when the considered events are very recent \citep{wakeley_gene_2012}
+% tsinfer not model-based, minimal assumptions, heuristic, scalable
 
-% discuss accurracy:
-% there will be scenarios where the assumptions made here are not met: e.g. polytomies
-% yet we can still always compute a likelihood.
-% we can alwyas associate a value, what remains to be verified what the information is in that
-% case.
-
-% explore the information in unary nodes
+quantify uncertainty
+* polytomies represent uncertainty:
+*  systematicallty break them by sampling from the coalescent rather then randomly
+* associate uncertainty as metadata to parent node representing uncertainty in the
+* interval spanned by the polytomy
+* can something similar be done for stacked recombinations?
 
@@
 
-\section{Supplementary material}
+\pagebreak
+
+\supplementarysection
+\section*{Supplementary Information}
+
+\end{figure}
+
+\begin{figure}[h!] \label{sup:fig:vs-argweaver}
+\centering
+\includegraphics[width=\textwidth]{example-image-a}
+\caption{Compare likelihoods inferred ARGs ARGweaver}
+\end{figure}
+
+\end{figure}
+
+\begin{figure}[h!] \label{sup:fig:vs-hudson}
+\centering
+\includegraphics[width=\textwidth]{example-image-a}
+\caption{Compare likelihoods Hudson and SMC}
+\end{figure}
 

### 2023-11-08 paper.tex
compile error correction

@@
 \centering
-\includegraphics[width=\textwidth]{figures/smc-custom-haplotypes-mini.png}
+\includegraphics[width=\textwidth]{figures/smc_custom_2rows.png}
 \caption{ARG (top left) generated under the SMC and its representation as a 
@@
 
-\begin{figure}[h!] \label{fig:algo}
-\centering
-\includegraphics[width=0.8\textwidth]{figures/smc_algo.png}
-\caption{Description of algorithm: the algorithm progresses from left to right and 
-keeps track of the . Let's assume we have visited the first 5 edges present in the 
-first tree (tabel 1). Moving onto the next tree, need to remove single edge 
-$(2, 5, [0,42))$, update the counts $I$ by decrementing all intervals between 5 and 
-2. The state of the vector is now $I_6$, the number of extant lineages the lineage 
-associated with edge $(1, 6, [42, 100)$ can coalesce with. Now compute the likelihood 
-associated with this edge. Then insert this edge and increment the count vector between 
-1 and 6. Two intermediate steps to move to the second tree. Similarly again two steps 
-to move to the last tree and insert edge $(5, 7, [70,100))$
-}
-
-\end{figure}
-
-\begin{figure}[h!] \label{fig:algo-flat}
+\begin{figure}[!ht] \label{fig:algo}
 \centering
@@
 
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=0.5\textwidth]{example-image-a}
+\caption{Compare likelihoods inferred ARGs ARGweaver}
+ \label{sup:fig:vs-argweaver}
 \end{figure}
 
-\begin{figure}[h!] \label{sup:fig:vs-argweaver}
+\begin{figure}[!ht]
 \centering
-\includegraphics[width=\textwidth]{example-image-a}
-\caption{Compare likelihoods inferred ARGs ARGweaver}
-\end{figure}
-
-\end{figure}
-
-\begin{figure}[h!] \label{sup:fig:vs-hudson}
-\centering
-\includegraphics[width=\textwidth]{example-image-a}
+\includegraphics[width=0.5\textwidth]{example-image-a}
 \caption{Compare likelihoods Hudson and SMC}
+\label{sup:fig:vs-hudson}
 \end{figure}

### 2023-11-10 paper.tex
update ARG recording section

@@
 
-Formulated backwards-in-time the SMC \citep{mcvean_approximating_2005} only requires a simple modification to the 
+Formulated backwards-in-time, the SMC \citep{mcvean_approximating_2005} only requires a 
+simple modification to the 
 coalescent with recombination. By only considering those pairs of 
@@
 genomic intervals with ancestral material, as eligible for coalescence, we obtain the 
-sequential Markovian structure along the chromosome the SMC is best known for.\\
+sequential Markovian structure along the chromosome.\\
 
@@
 carried by that lineage $X_i = \{[x_{i0}, y_{i0}), \dotsc, [x_{im}, y_{im})\}$.
-$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$, each represented by 
+$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$, each represented by a 
 single interval spanning the entire genome.
@@
 The waiting time to next event is determined by these two competing processes with exponentially 
-distributed waiting times as described below.\\
+distributed waiting times as outlined below.\\
 
@@
 all lineages in $X(t)$ based on their leftmost starting points. 
-For any such strict total order on $X_l(t)$ and at any point in the process 
+For any such strict total order on $X(t)$ and at any point in the process 
 the instantaneous rate of coalescence then equals $\lambda \sum_{i \neq j} I_{ij}$.
@@
 lineages $C_i(t)$ it can coalesce with. This will simplify the likelihood computations 
-(see next paragraph). Note that although recombination affects $C_i(t)$ it does not 
+(see \ref{par:liks). Note that although recombination affects $C_i(t)$ it does not 
 affect $I_{i}$.\\
@@
 set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
-$\mathcal{G} = (N, E)$ is obtained by registering a tuple $(c, p, X_{cp})$ whenever a 
+$\mathcal{G} = (N, E)$ capturing the various inheritance paths along the genome 
+is obtained by registering a tuple $(c, p, X_{cp})$ whenever a 
 lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
@@
 (see fig.\ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
-for the nodes registered for those lineages hit by a recombination event. 
-In case of a coalescence event, an edge is registered along the full 
-extent of both involved lineages. 
+for the nodes registered for those lineages affected by a recombination event. 
 Coalescence between two lineages without perfectly 
-overlapping intervals will therefore result in locally unary nodes.\\
-
-These unary nodes ensure the one to one correspondence between an ARG and its 
-representation as a sequence of local trees (see fig.\ref{fig:smc-unary}). 
+overlapping intervals will similarly result in locally 
+unary nodes (see \ref{fig:smc-unary}).These unary nodes ensure the one 
+to one correspondence between an ARG and its 
+representation as a sequence of local trees. 
 Indeed, the presence of these unary nodes allow us to completely reconstruct 
-the state of all extant lineages through time from the local trees.
+the state of all extant lineages through time from the local trees.\\
 
@@
 from the available haplotype information. Currently, the following methods 
-return this information by default as part of their ARG reconstruction algorithm: 
+return the full this information by default as part of their ARG reconstruction algorithm: 
 \argweaver \citep{rasmussen_genome-wide_2014} 
@@
 \caption{ARG (top left) generated under the SMC and its representation as a 
-series of local trees. The presence of the unary nodes ensure these two 
-representations contain the same information.
-The haplotypes that make up the lineages coalescing 
-at the root of each marginal tree are highlighted at the top. Ancestral 
-material is coloured based on the samples that inherit from it.
-Note how the presence of the unary nodes 
-modifies the left-to-right logic of single-prune-regrafts induced by 
-recombination under the SMC (blue dotted arrows).
-Even in the absence of the blue recombination nodes, 
-the unary coalescing nodes allow us to reconstruct the state of all lineages 
-up to the exact time of the recombination event.
+series of local trees. The presence of the unary nodes in the local trees 
+ensure these two representations contain the same information. 
+The ARG is augmented with haplotype information, coloured according to the 
+samples that inherit from it. 
+Note how the presence of a unary node on a local tree is 
+indicative of either a past (to the left of the local tree) 
+or future (to the right) coalescence event. It therefore modifies the 
+left-to-right logic of the SMC whereby the floating lineage coalesces randomly
+with a the remaining portion of the local tree. 
 }
@@
 
-\subsection{Unary nodes} \label{par:unary}
+%\subsection{Unary nodes} \label{par:unary}
 % Do we need a seperate section on the unary nodes?
 % is there more to be said about the SMC inm the presence of unary nodes?
-Recording unary nodes has interesting consequences on the left-to-right 
-description of the SMC. Looking at a single marginal tree, the presence of 
-unary nodes are telltale signs of future or past coalescence events (in a 
-left-to-right sense).\\
+%Recording unary nodes has interesting consequences on the left-to-right 
+%description of the SMC. Looking at a single marginal tree, the presence of 
+%unary nodes are telltale signs of future or past coalescence events (in a 
+%left-to-right sense).\\
 

### 2023-11-13 paper.tex
start introduction
update intro and discussion

@@
 \section{Abstract}
-%% WHAT IS THE MAIN FOCUS OF THE PAPER?
+This means there currently is no way to connect the output of the diverse 
+range of ARG inference methods back 
+to a model that would allow us to perform a probabilistic exploration of the 
+ARG space around the point estimate, or likelihood-based inference
 
@@
 
-% Main motivation:
-Major progress in ARG inference.
+The underlying genealogical history that relates a sample of individuals back in time
+contains all possible information this sample can provide on the history of demography, 
+selection and recombination for the population of interest. 
+Reconstructing these genealogies with certainty is impossible however. 
+As such, in the past the data structure capturing these inheritance path 
+or Ancestral Recombination Graph (ARG), has usually been considered a nuisance 
+variable to be integrated out during statistical inference 
+\citep{kuhner_2000, griffiths_ancestral_1996}.
 
-Result is whole range of inference methods, all inferring an ARG with their own specific 
-characteristics.
-Many of these scalable heuristic methods do not explicitly infer recombination events.
+% explicitly outputs an ARG
+ARGweaver, generates these ARGs explicitly again using MCMC sampling, by discretizing time 
+they can apply HMM-based algorithms to efficiently generate new samples from the distribution of 
+possible ARGs.
 
-%ARGs that are inferred here lack information to enable us to compute 
-% a likelihood under the Hudson coalescent.
-% Missing: exact time to recombination + non-inferrable recombinations
+% model-based approaches vs
 
-Currently no way to connect the output of the diverse range of ARG inference methods back 
+% recent advancements: heuristics
+Recently, major advancements 
+relying on heuristic rather than model-based approaches
+made ARG inference scalable.
+\citep{rasmussen_2014, kelleher_2019, speidel_2019, schaefer_2021, wohns_2022, 
+zhang_biobank-scale_2023, zhan_towards_2023}
+have raised the realistic prospect of ARG-based analysis becoming a standard 
+part of the population and statistical genetics toolkit
+This has also meant however that
+we are now using the term ARG for a varying range of approxiate structures
+as generated by these inference methods \citep{wong_a-general_2023}.
+
+Firstly, in contrast to model-based inference methods like \argweaver, 
+these models only deal with recombination in an implicit sense.
+
+For example, where ARGweaver has 
+the built in assumption of neighbouring trees being separated by single Subtree Prune 
+Regraft (SPR) [REF Song et al.], this is not the case for any of the heuristic data 
+structures.
+Lack recombination nodes in the ARG, crucial in ability to compute likelihoods 
+under Hudson. \citep{Wong_a-general_2023}
+
+Secondly, the lack of an underlying model also means that
+we have no way to deal with uncertainty, an issue that is 
+resolved differently by each inference tool.
+
+Besides not having a unified output format, The degree of completeness 
+in which genetic inheritance from ancestors to 
+descendants is documented by each of these methods varies extensively.
+
+This means there currently is no way to connect the output of the diverse 
+range of ARG inference methods back 
 to a model that would allow us to perform a probabilistic exploration of the 
-ARG space around the point estimate, (or likelihood-based inference, ...)
+ARG space around the point estimate, or likelihood-based inference 
+(that does not only consider individual marginal trees).
 
-% What are ARGs, their relevance and importance
+[TO ADD] mention the SMC \citep{mcvean_approximating_2004}.
+SMC and relatives are useful: lot of focus however on integrating over 
+the distribution of trees to get Ne(t) [ADD REF: PSMC, MSMC]
+Here the idea is to use the SMC in a different way that resonates more 
+with the true direction of flow of genetic/haplotype information.
+Results in a concrete way to compute likelihoods for *any* inferred ARG 
+by integrating out the time to recombination.
+(integrating out the recombination time)
+and formulate an algorithm 
+that is general in the sense that it can deal with the various levels of 
+completeness that characterise the fields currents attempts at reconstucting genome-wide 
+genealogies.
 
-% associating likelihoods with ARGs
-
-% previous likelihood calculations on ARGs Kuhner and Felsenstein (2000), ARGweaver
-
-% history of approximating the coalescent: 
-%SMC (McVean and Cardin, 2005) SMC' (Marjoram and Wall, 2006).
-%Also heuristic approximations: Li and Stephens (2003) extending Stephens and Donnelly (2000)
-%These have enabled (scalable) ARG inference methods: ARGweaver (relying on SMC, limited 
-%to smaller samples, but ability to sample from posterior, and full ARG), and Relate, 
-%tsinfer/tsdate (copying algorithm), ArgNeedle (using SMC) 
-%Clear distinction into two groups: ARGWeaver infers actual ARG.
-% what does kwarg do?
-%Whereas the other approaches are heuristic in nature. For example, where ARGweaver has 
-%the built in assumption of neighbouring trees being separated by single Subtree Prune 
-%Regraft (SPR) (Song citation?), this is not the case for any of the heuristic data 
-%structures. (other peculiarities: ...).
-
-% computing likelihoods on ARGs; from Kuhner and Felsenstein to ARGWeaver
-
-% SMC and relatives are useful: lot of focus however on integrating over the distribution of trees to get Ne(t). \cite{PSMC, ...}
-% Here the idea is to use the SMC in a different way that resonates more with the true direction of flow
-% of genetic/haplotype information.
-
-% results in a concrete way to compute likelihoods for *any* inferred ARG (integrating out the recombination time)
-% meaning: e.g. no requirements on adjacent trees only differing by a single SPR, (other?)
-
-% In this paper, we attempt to address these problems by
-% providing a precise definition of an ARG that is general enough to encompass
-% the output of popular simulation and inference methods.
-
-% ADD: remark on fact that we are only dealing with topology and branch length information
-% adding likelihood of observing particular mutation configuration given an ARG is straightforward.
+[TO ADD] remark on fact that we are only dealing with topology and branch length information
+adding likelihood of observing particular mutation configuration given an ARG is straightforward.)
 
@@
 lineages $C_i(t)$ it can coalesce with. This will simplify the likelihood computations 
-(see \ref{par:liks). Note that although recombination affects $C_i(t)$ it does not 
+(see \ref{par:liks}). Note that although recombination affects $C_i(t)$ it does not 
 affect $I_{i}$.\\
@@
 Here, we have introduced a scalable algorithm to compute likelihoods 
-under the SMC for a general class of ARGs.
+under the SMC for a general class of ARGs. This work will contribute significantly 
+to ARG-based analysis becoming a standard part of the population 
+and statistical genetics toolkit. 
+In particular, we see three main applications for the work presented here. 
+Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
+model and a set of parameters will help bridge the gap between the tremendous recent 
+advances in ARG inference and the current state of the art statistical inference of 
+past demography. Secondly, hybrid ARG-inference methods combining the best of both 
+heuristic and model based inference approaches. Thirdly, quantify uncertainty.
 
-Impact both on ways of doing ARG inference itself and quantifying the uncertainty associated 
-with that as well as performing statistical inference of past demography.
-
-* statistical inference *
+1/ statistical inference
 This is an important step towards bridging the gap between the tremendous 
 recent advances in ARG inference and the current state of the art statistical inference of 
-past demography (but see ). 
-% some/many of these approaches only use the information present in marginal trees.
+past demography (but see ): currently most/all methods perform ARG-based analysis 
+by extracting mariginal trees from the ARG \citep{hejase_2022} % likelihood-based demographic inference on local trees (ref?) 
 The provided theory and associated algorithm will function as a building block of a 
@@
 * exploit the efficiencies of the tree sequence data structure (see \citep{Mahmoudi_bayesian_2022})
-* we can readily incorporate $N_e$-changes as well as multiple populations and migration.
+* we can readily incorporate $N_e$-changes as well as multiple populations and migration.\\
 
-* ARG inference *
+Biobank scale datasets, first few generations violate coalescent assumptions. Here we use heuristic 
+method. Once number of lineages has reduced sufficiently, model-based approach can be used. One 
+of the advantages of formulating this backwards in time. 
 Combining ARG inference methods: tsinfer for first few generations (large sample sizes), where 
 the initial phase is biased relative to the Wright-Fisher model when the sample size is large 
-\citep{bhaskar_distortion_2014}, or when the considered events are very recent \citep{wakeley_gene_2012}
-% tsinfer not model-based, minimal assumptions, heuristic, scalable
+\citep{bhaskar_distortion_2014}, or when the considered events are very recent \citep{wakeley_gene_2012}\\
 
-quantify uncertainty
-* polytomies represent uncertainty:
-*  systematicallty break them by sampling from the coalescent rather then randomly
-* associate uncertainty as metadata to parent node representing uncertainty in the
-* interval spanned by the polytomy
-* can something similar be done for stacked recombinations?
+Both of these future directions tie in with the ability to quantify uncertainty by sampling 
+from an underlying distribution. This will in particular enable us to deal with those 
+parts in both time and along the genome for which we have insufficient information to confidently 
+reconstruct the ARG. For example the presence of polytomies in \tsinfer represents 
+uncertainty. Systematicallty break them by sampling from the coalescent rather then randomly.
+Quantifying uncertainty would not necessarily always have to be associated with dealing with 
+the entire subset of all compatible genealogies. Uncertainty could also be quantified and passed 
+on as metadata to a parent node representing the uncertainty in the interval spanned by the 
+polytomy.
+
+% can something similar be done for stacked recombinations?
+
+[TO DO] Comment on accuracy. The described algorithm has been implemented in such a way that 
+any valid tree sequence will result in a likelihood value. This also holds for any tree sequence 
+that does not meet the assumptions. We see it as future work to determine to what extent 
+particular deviations from these assumptions will affect the shape of the likelihood surface given 
+a set of parameters.
+
+% final conclusion
 

### 2023-11-14 paper.tex
update discussion

@@
 Lack recombination nodes in the ARG, crucial in ability to compute likelihoods 
-under Hudson. \citep{Wong_a-general_2023}
+under Hudson. \citep{wong_a-general_2023}
 
@@
 lineage with trapped non-ancestral material (see end \ref{par:description}). 
-If we assume that only a single recombination 
+If we assume that only a single recombination could have  
 occurred, then it suffices however to provide a correction factor $g$ in equation
@@
 \includegraphics[width=\textwidth]{figures/ts_algo_2rows.png}
-\caption{Description of algorithm: Moving along the genome we transition from the first 
-local tree (1) to the next (4). For each edge that is removed from the first tree we 
-decrement the count in the corresponding internode intervals and update the identity of 
-the last parent. Subsequently, for each edge that will be inserted we compute the likelihood 
-associated with that edge prior to incrementing the counts.}
+\caption{Description of algorithm: Moving along the genome we transition from  
+local tree $T_1$ (1) to the next, $T_2$ (4). Edges that do not persist in $T_2$ are 
+removed (2). Edges starting at the left coordinate of $T_2$ are inserted (3). 
+After deletion or insertion of an edge the count vector $I$ is updated along the 
+corresponding internode intervals. Prior to updating $I$ following an insertion, 
+the likelihood for that edge is computed as $I$ then represents the total number of 
+lineages the lineage associated with this new edge can 
+coalesce with. By moving along the genome in this way, each edge gets visited exactly once.}
 \end{figure}
@@
 % ADD IN RESULTS
-Figure \ref{sup:fig:vs-hudson} shows the correlation between the likelihood under the hudson 
-for an ARG with detailed recombination information simulated and computed using 
-\texttt{msprime} \citep{baumdicker_efficient_2021}. Show impact of unary nodes on $r^2$.
-Simulated mutation data under the SMC and passed this to ARGweaver.
-Scatterplot of coalescent prior as returned by ARGweaver \citep{rasmussen_genome-wide_2014} 
-for a 100 samples from the posterior against the likelihood as described here fore these 
-sampled ARGs \ref{sup:fig:vs-argweaver}
-
-Implementation of the algorithm is available on https://github.com/GertjanBisschop/runsmc.
+Figure \ref{sup:fig:vs-argweaver} shows the expected near-perfect correlation 
+between the coalescent prior as returned by \argweaver and the likelihood as computed here. 
+We simulated 1Mb of data for 10 diploid individuals 
+under the SMC using human-like parameters. 100 samples were taking from the posterior every for 
+a total of 1000 MCMC iterations.
 
@@
 Any such subset of the tree sequence is itself a valid tree sequence 
-that can again be used to compute a likelihood with.
-
-Our approach therefore strenghtens the inherent ability of ARGs to provide us with temporal 
+that can again be used to compute a likelihood with. Our approach therefore strenghtens 
+the inherent ability of ARGs to provide us with temporal 
 resolution on the inferred genealogical history of a sample. 
 We can further readily incorporate changes in $N_e$ and $r$, 
-both in time as well as along the genome.
+both in time as well as along the genome by respectively adding intervals to $I$ 
+or by splitting edges when/wherever the parameter set changes.
+[NOTE: these options have already been implemented in the current implementation]
 
-% Algorithm is suitable for a general class of ARGs each with their own specific features.
-% What happens in the case of ARGs that do not meet our assumptions: absent unary nodes, polytomies, ...
-The above algorithm treats each edge independently which implies we can deal with polytomies.
-Furthermore, although we rely heavily on unary nodes to provide us with the necessary 
-information on recombination events, our algorithm can also accomodate for 
-those inferred ARGs that lack unary nodes (e.g.\ \relate, and see figure in supplements).
-However, in case of (sub)tree height changes along the ARG, the algorithm would infer 
-two recombination events along the lineages associated 
-with both edges switching parent nodes. This can be mitigated by limiting the number of 
-'inferrable' recombination events per coalescence event to 1 (see supplementary figure).
+Although we rely heavily on unary nodes to provide us with the necessary 
+information on recombination events, our algorithm only implicitly assumes there presence.
+The algorithm can therefore also accomodate for 
+those inferred ARGs that lack unary nodes (e.g.\ \relate). 
+Note that without unary nodes, our algorithm will detect a recombination event along 
+both coalescing lineages in case of a (sub)tree height changes along the ARG as both 
+edges would switch parent nodes. This can be mitigated by limiting the number of 
+'inferrable' recombination events per coalescence event to 1.
+% simulated results
+We simulated 1000 ARGs under the Hudson coalescent retaining all recombination information 
+and computed the corresponding likelihood using \texttt{msprime} \citep{baumdicker_efficient_2021}. 
+After removing the recombination nodes the likelihoods as defined here were computed both with and 
+without the unary nodes. Figure \ref{sup:fig:vs-hudson} shows that the proportion of variation 
+explained is higher when the unary nodes are retained.
+
+% polytomies
+[TO DO: polytomies] The above algorithm treats each edge independently which implies 
+we can deal with polytomies.
 
 \section{Discussion}
-% summary: what did we do here
+
 Here, we have introduced a scalable algorithm to compute likelihoods 
@@
 Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
-model and a set of parameters will help bridge the gap between the tremendous recent 
+model and an associated set of parameters will help bridge the gap between the tremendous recent 
 advances in ARG inference and the current state of the art statistical inference of 
-past demography. Secondly, hybrid ARG-inference methods combining the best of both 
-heuristic and model based inference approaches. Thirdly, quantify uncertainty.
+past demography. Secondly, hybrid ARG-inference methods that combine the best of both 
+heuristic and model-based inference approaches are now possible.
+Thirdly, using a model we can quantify the uncertainty inherently associated 
+with ARG-reconstruction.
 
-1/ statistical inference
-This is an important step towards bridging the gap between the tremendous 
-recent advances in ARG inference and the current state of the art statistical inference of 
-past demography (but see ): currently most/all methods perform ARG-based analysis 
-by extracting mariginal trees from the ARG \citep{hejase_2022} % likelihood-based demographic inference on local trees (ref?) 
-The provided theory and associated algorithm will function as a building block of a 
-demography-aware genealogy sampler. Current approach couple of major advantages:
-* any type of inferred ARG as a starting point. Potentially implying a massive reduction in 
-MCMC runtime by reducing the necessary burnin period.
-* by integrating out the exact time to the recombination event we further restrict the size 
-of ARG space that needs to be explored.
-* exploit the efficiencies of the tree sequence data structure (see \citep{Mahmoudi_bayesian_2022})
-* we can readily incorporate $N_e$-changes as well as multiple populations and migration.\\
+% 1/ statistical inference
+Currently most/all ARG-based inference methods base their analysis on marginal 
+trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
+provided here can function as a building block of a 
+demography-aware genealogy sampler that would allow us to incorporate the information 
+provided by recombination events.
+More concretely, given our algorithm, any inferred ARG can be used a good starting point 
+for a future MCMC-sampler. This potentially implies a massive runtime reduction by 
+reducing the necessary burn-in period. Evenmore, by integrating out the exact time 
+to the recombination event we further restrict the size of the ARG space that needs to 
+be explored. 
+This algorithm will also allow us to exploit the efficiencies of the tree sequence 
+data structure (see \citep{mahmoudi_bayesian_2022}) to evaluate any new proposal.
+We can readily incorporate $N_e$-changes as well as multiple populations and migration.\\
 
+% 2/ hybrid ARG-inference
 Biobank scale datasets, first few generations violate coalescent assumptions. Here we use heuristic 
@@
 
+% 3/ quantifying uncertainty
 Both of these future directions tie in with the ability to quantify uncertainty by sampling 
@@
 % final conclusion
+\section{Data availability}
+
+All scripts and data used for this manuscript are available on https://github.com/gertjanbisschop/smc-bit-paper
+An implementation of the algorithm is available on https://github.com/gertjanbisschop/runsmc.
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{example-image-a}
-\caption{Compare likelihoods inferred ARGs ARGweaver}
+\includegraphics[width=0.75\textwidth]{figures/supplementary-figs/argweaver_vs_runsmc.png}
+\caption{Compare likelihoods inferred ARGs ARGweaver: placeholder for final figure with similar content}
  \label{sup:fig:vs-argweaver}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{example-image-a}
-\caption{Compare likelihoods Hudson and SMC}
+\includegraphics[width=0.75\textwidth]{figures/supplementary-figs/v_hudson_unary_simpl.png}
+\caption{Compare likelihoods Hudson and SMC with and without unary nodes: placeholder for final figure with similar content}
 \label{sup:fig:vs-hudson}

### 2023-11-14 paper.tex
Updates from Overleaf
update intro and discussion

@@
 Reconstructing these genealogies with certainty is impossible however. 
-As such, in the past the data structure capturing these inheritance path 
+As such, in the past the data structure capturing these inheritance paths 
 or Ancestral Recombination Graph (ARG), has usually been considered a nuisance 
@@
 
+% first major step forwards in ARG inference
 % explicitly outputs an ARG
@@
 possible ARGs.
-
+% [TO ADD] also argweaver is the first to compute likelihoods for an ARG under the SMC, musn't forget to include this.
 % model-based approaches vs
@@
 % recent advancements: heuristics
-Recently, major advancements 
-relying on heuristic rather than model-based approaches
-made ARG inference scalable.
+Recently, major advancements relying on heuristic rather than model-based approaches
+have made ARG inference scalable.
 \citep{rasmussen_2014, kelleher_2019, speidel_2019, schaefer_2021, wohns_2022, 
-zhang_biobank-scale_2023, zhan_towards_2023}
+zhang_biobank-scale_2023, zhan_towards_2023}. These methods
 have raised the realistic prospect of ARG-based analysis becoming a standard 
-part of the population and statistical genetics toolkit
+part of the population and statistical genetics toolkit.
 This has also meant however that
-we are now using the term ARG for a varying range of approxiate structures
-as generated by these inference methods \citep{wong_a-general_2023}.
+we are now using the term ARG for a varying range of approximate structures
+as generated by these different inference methods \citep{wong_a-general_2023}. 
+Besides not having a unified output format, The degree of completeness 
+in which genetic inheritance from ancestors to 
+descendants is documented by each of these methods varies extensively. 
+[TO DO:] discuss: each of these tools will deal with or resolve uncertainty 
+in a different way.
+This has complicated the ability to connect the output of this diverse range of 
+ARG inference methods to a model-based deescription that would allow us to perform  
+probabilistic exploration of the ARG space around the point estimate, 
+or likelihood-based inference (that does not just consider individual marginal trees).\\
 
-Firstly, in contrast to model-based inference methods like \argweaver, 
-these models only deal with recombination in an implicit sense.
+Recombination in particular:
+In contrast to model-based inference methods like \argweaver, 
+these models only deal with recombination in an implicit sense.\\
 
-For example, where ARGweaver has 
+[TO DO: finish this section] For example, where ARGweaver has 
 the built in assumption of neighbouring trees being separated by single Subtree Prune 
@@
 
-Secondly, the lack of an underlying model also means that
-we have no way to deal with uncertainty, an issue that is 
-resolved differently by each inference tool.
-
-Besides not having a unified output format, The degree of completeness 
-in which genetic inheritance from ancestors to 
-descendants is documented by each of these methods varies extensively.
-
-This means there currently is no way to connect the output of the diverse 
-range of ARG inference methods back 
-to a model that would allow us to perform a probabilistic exploration of the 
-ARG space around the point estimate, or likelihood-based inference 
-(that does not only consider individual marginal trees).
-
-[TO ADD] mention the SMC \citep{mcvean_approximating_2004}.
+% then onto the SMC
+[TO ADD: discuss SMC] mention the SMC \citep{mcvean_approximating_2004}.
 SMC and relatives are useful: lot of focus however on integrating over 
-the distribution of trees to get Ne(t) [ADD REF: PSMC, MSMC]
-Here the idea is to use the SMC in a different way that resonates more 
-with the true direction of flow of genetic/haplotype information.
+the distribution of trees to get $N_e(t)$ [ADD REF: PSMC, MSMC]
+Here, the idea is to use the SMC in a different way that resonates more 
+with the true direction of flow of genetic information.
 Results in a concrete way to compute likelihoods for *any* inferred ARG 
 by integrating out the time to recombination.
-(integrating out the recombination time)
 and formulate an algorithm 
 that is general in the sense that it can deal with the various levels of 
-completeness that characterise the fields currents attempts at reconstucting genome-wide 
-genealogies.
-
-[TO ADD] remark on fact that we are only dealing with topology and branch length information
-adding likelihood of observing particular mutation configuration given an ARG is straightforward.)
+completeness that characterise the field's currents attempts [TO DO: make this sound more positive] 
+at reconstucting genome-wide genealogies.
 
@@
 
+Note that we are only considering the topology and branch length information of any ARG. 
+Extending this expression by adding the likelihood of observing a set of mutations 
+given an ARG is straightforward.
+
 \subsection{Tree sequence format} \label{par:algo}
@@
 Here, we have introduced a scalable algorithm to compute likelihoods 
-under the SMC for a general class of ARGs. This work will contribute significantly 
+under the SMC for a general class of ARGs. By formulating the SMC backwards in time 
+we were able to associate a single likelihood value with each edge in the ARG. 
+The compatibility of the algorithm with the \tskit library allows us to 
+exploit its associated efficiencies both for the current work and its future 
+applications. This work will contribute significantly 
 to ARG-based analysis becoming a standard part of the population 
@@
 for a future MCMC-sampler. This potentially implies a massive runtime reduction by 
-reducing the necessary burn-in period. Evenmore, by integrating out the exact time 
+reducing the necessary burn-in period. Furthermore, by integrating out the exact time 
 to the recombination event we further restrict the size of the ARG space that needs to 
-be explored. 
-This algorithm will also allow us to exploit the efficiencies of the tree sequence 
-data structure (see \citep{mahmoudi_bayesian_2022}) to evaluate any new proposal.
-We can readily incorporate $N_e$-changes as well as multiple populations and migration.\\
+be explored. This MCMC-sampler would again rely on the tree sequence data structure to 
+efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}). 
+$N_e$-changes as well as multiple populations and migration can be readily incorporated 
+in the current algorithm.\\
 
 % 2/ hybrid ARG-inference
-Biobank scale datasets, first few generations violate coalescent assumptions. Here we use heuristic 
-method. Once number of lineages has reduced sufficiently, model-based approach can be used. One 
-of the advantages of formulating this backwards in time. 
-Combining ARG inference methods: tsinfer for first few generations (large sample sizes), where 
-the initial phase is biased relative to the Wright-Fisher model when the sample size is large 
-\citep{bhaskar_distortion_2014}, or when the considered events are very recent \citep{wakeley_gene_2012}\\
+Hybrid backwards-in-time simulations that combine Wright-Fisher dynamics in the 
+very recent past and coalescent simulations for the remaining time are a tried and tested 
+approach \citep{bhaskar_distortion_2014, nelson_accounting_2020} to enable large scale 
+simulations while avoiding the documented biases of the coalescent relative to the 
+Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. The same approach 
+could be used for ARG-inference where one can rely on a heuristic inference method for 
+the recent past, while using a model-based approach beyond that cutoff point.\\
 
-% 3/ quantifying uncertainty
-Both of these future directions tie in with the ability to quantify uncertainty by sampling 
-from an underlying distribution. This will in particular enable us to deal with those 
+% 3/ quantifying uncertainty [unfinished]
+Both of these future directions tie in with the desired ability of any ARG-related method to 
+quantify the uncertainty associated with
+
+to quantify uncertainty by sampling 
+from an underlying distribution.
+
+This will in particular enable us to deal with those 
 parts in both time and along the genome for which we have insufficient information to confidently 
-reconstruct the ARG. For example the presence of polytomies in \tsinfer represents 
-uncertainty. Systematicallty break them by sampling from the coalescent rather then randomly.
-Quantifying uncertainty would not necessarily always have to be associated with dealing with 
-the entire subset of all compatible genealogies. Uncertainty could also be quantified and passed 
-on as metadata to a parent node representing the uncertainty in the interval spanned by the 
-polytomy.
+reconstruct the ARG. The presence of polytomies in ARGs inferred by \tsinfer, for example, 
+represents uncertainty. Systematicallty breaking them by sampling from the coalescent rather 
+then randomly would already be a major step forward.
 
+Quantifying uncertainty would not necessarily always have to be associated 
+with sampling 
+with dealing with the entire subset of all compatible genealogies. Uncertainty could also 
+be quantified and passed on as metadata to a parent node representing the uncertainty 
+in the interval spanned by the polytomy.
 % can something similar be done for stacked recombinations?
@@
 % final conclusion
+
 \section{Data availability}

### 2023-11-16 paper.tex
update recording ARGs

@@
 
-%\usepackage{subcaption}
-
-\newcommand{\noderef}[1]{\textsf{#1}}
 \newcommand{\tsinfer}[0]{\texttt{tsinfer}}
@@
 \newcommand{\arbores}[0]{\texttt{Arbores}}
+\newcommand{\tskit}[0]{\texttt{tskit}}
 
@@
 
+[great point from Wong et al] Note that it is important to distinguish 
+here between structures and models: whether an inference method is based on 
+heuristics or a rigorous mathematical model is orthogonal to the level 
+of detail provided in its estimate. One could heuristically estimate a 
+fully precise ARG, or statistically sample a partial, approximate ARG under 
+a model such as the coalescent.
+What we want to achieve with this paper is to bring the two concepts closer 
+to each other. Compute a likelihood for an approximate (define) ARG under a 
+approximate model of the coalescent. 
+Approximate in two senses: SMC + leave out precise recombination time info
+
 The underlying genealogical history that relates a sample of individuals back in time
@@
 \citep{kuhner_2000, griffiths_ancestral_1996}.
+
+The breakthroughs in scale achieved by recent methods 
+are all based, in different ways, on inferring approximate 
+structures instead of a complete and fully detailed history
+% in particular with respect to recombination nodes
 
@@
 
+[TO DO:] what does *general* mean? and when in the paper do we discuss it?
+
 [TO DO: finish this section] For example, where ARGweaver has 
@@
 with the true direction of flow of genetic information.
+
 Results in a concrete way to compute likelihoods for *any* inferred ARG 
@@
 Backwards in time the process evolves by a succession of coalescence and recombination 
-events until all segments of ancestral material are each present in only one lineage. 
-The waiting time to next event is determined by these two competing processes with exponentially 
-distributed waiting times as outlined below.\\
+events until each segment of ancestral material is only present in one lineage. 
+The waiting time to next event is determined by these two competing processes 
+with exponentially distributed waiting times as outlined below.\\
 
@@
 $\mathcal{G} = (N, E)$ capturing the various inheritance paths along the genome 
-is obtained by registering a tuple $(c, p, X_{cp})$ whenever a 
-lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
+is obtained by registering an edge represented by a tuple $(c, p, X_{cp})$ 
+whenever a lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
 replaced by one or two lineage(s), $p$ (and $p^{\prime}$), 
@@
 
-Recording the full process as described above will result in unary nodes.
-These are nodes that have only one child in one or more local trees 
-(see fig.\ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
-for the nodes registered for those lineages affected by a recombination event. 
-Coalescence between two lineages without perfectly 
-overlapping intervals will similarly result in locally 
-unary nodes (see \ref{fig:smc-unary}).These unary nodes ensure the one 
-to one correspondence between an ARG and its 
-representation as a sequence of local trees. 
-Indeed, the presence of these unary nodes allow us to completely reconstruct 
-the state of all extant lineages through time from the local trees.\\
+The ARG as described here records the impact of every possible event along each 
+inheritance path in full detail. We could however simplify this graph by 
+throwing away the precise time information of all recombination events. This 
+corresponds to removing the recombinant internal nodes within the ARG such that some of 
+the internal nodes representing coalescence events now have 2 outgoing (backwards in time) 
+edges. On the ARG in fig.\ref{fig:smc-unary}, the output of a simulation under the SMC, 
+this corresponds to removing the blue internal nodes. Crucially however, 
+we can still identify all lineages that were hit by a recombination event. Yet we 
+can only constrain the time to this event to within a window defined by the age of the 
+child node $c$ associated with that lineage and the minimum age of the parents of the 
+edges connecting to $c$.\\
 
-%under the SMC up to the exact time of the recombination events.
+ARG simplification therefore touches upon the question of what is knowable.  
+[TO DO: Add link with inference methods]\\
 
-The resulting data structure can be simplified to various degrees \citep{Wong-2023}. 
-This equates to removing certain types of unary nodes, 
-depending on the degree of information 
-we want to retain (or can reconstruct) from the original process. 
-In a fully simplified ARG, without unary nodes, 
-the local trees only store information on local coalescence events.
-Under the SMC, if we only omit recombination nodes, % is there a better word?
-we are left with the coalescent nodes, those nodes that are coalescent 
-along some fraction of the genome. We can however still identify \emph{all} 
-lineages that were hit by a recombination event and constrain the time to 
-this event to within a window defined by the age of the node $c$ associated with that 
-lineage and the minimum age of the parents of the edges connecting to $c$.
+Many ARG reconstruction methods return a data structure that 
+is in essence a series of local tries [REFs]. This does not need to be a contradiction. 
+All ARGs can be represented as a series of local trees \citep{wong-2023}.
+As long as we register each observed node when building the local tree that 
+records the path through the ARG, then we can guarantee a one-to-one correspondence 
+with the original ARG. Crucially, this will result in nodes that 
+are locally unary. These are nodes that only have a single child along some of the 
+intervals within the genome.\\
 
+Recombination nodes are one such type of unary nodes. Figure \ref{fig:smc-unary} (bottom)
+reveals however that all internal nodes but node 4 are unary along some part of 
+the genome. These nodes all represent coalescence events.
+
+Note that the presence of these unary nodes allows us to reconstruct the full extent 
+of each lineage going back through time. Similarly, all lineages involved in a recombination 
+event can be uniquely identified by a change in parent going from one tree to the next 
+for the child node associated with that lineage. Both the ARG and the sequence of 
+local trees contain the same information, as long as they are simplified to the 
+same degree.\\
+
+[TO DO: do I need to expand on this?] Simplification is thus tied in with the concept of locally unary nodes.\\
+
+Which ARG inference methods return locally unary nodes and how can they be inferred.
+[TO DO] give a good explanation (using the figure) why these methods can infer the extent 
+of coalescing lineages.
 As highlighted in fig.\ref{fig:smc-unary}, (coalescing) unary nodes can be inferred 
@@
 
-The presence of these (coalescing) unary nodes enables us to extend likelihood 
-computations to a more general class of ARGs.
-Here we will compute the likelihood for an ARG
-under the SMC for which we have no information on the time to any 
-of the recombination events. 
-We will assume however, that the identity 
-of the lineage involved in each recombination event can be correctly 
-inferred by relying on the unary nodes.
+[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.
+
+In the next paragraph we will derive an expression to compute the 
+likelihood for an ARG under the SMC for which we have no information on the 
+time to any of the recombination events. We will then detail an algorithm 
+that allows us to compute this value given an ARG 
+represented as a series of local trees. To do this we will rely on the unary nodes 
+to identify each lineage involved in a recombination event 
+as described in this section.
 
@@
 \end{figure}
-
-%\subsection{Unary nodes} \label{par:unary}
-% Do we need a seperate section on the unary nodes?
-% is there more to be said about the SMC inm the presence of unary nodes?
-%Recording unary nodes has interesting consequences on the left-to-right 
-%description of the SMC. Looking at a single marginal tree, the presence of 
-%unary nodes are telltale signs of future or past coalescence events (in a 
-%left-to-right sense).\\
 
@@
 
+[TO DO: improve this remark: when recording an ARG we do record edges in this situation,
+however, msprime simulations will stop recording edges for local trees that have reached 
+their local mrca.] 
 Because the likelihood is computed based on edges,
@@
 length of non-ancestral material along which a recombination would have resulted  
-in the same observed local sequence of tree changes.
+in the same observed local sequence of tree changes.\\
 
@@
 we can deal with polytomies.
+[TO DO: discuss, what if the identity of internal nodes is not preserved across trees]
 
@@
 we were able to associate a single likelihood value with each edge in the ARG. 
-The compatibility of the algorithm with the \tskit library allows us to 
+The compatibility of the algorithm with the \texttt{tskit} library allows us to 
 exploit its associated efficiencies both for the current work and its future 

### 2023-11-20 paper.tex
update intro

--- 2023-11-16 paper.tex
@@
 
-[great point from Wong et al] Note that it is important to distinguish 
-here between structures and models: whether an inference method is based on 
+[TL:DR] Structure the introduction around this great point from Wong et al.:
+Note that it is important to distinguish here between structures and models: 
+whether an inference method is based on 
 heuristics or a rigorous mathematical model is orthogonal to the level 
@@
 a model such as the coalescent.
-What we want to achieve with this paper is to bring the two concepts closer 
-to each other. Compute a likelihood for an approximate (define) ARG under a 
-approximate model of the coalescent. 
-Approximate in two senses: SMC + leave out precise recombination time info
+What we want to achieve with this paper is to compute a likelihood for 
+an approximate (define) ARG under a approximate model of the coalescent. 
+Approximate in two senses: SMC + leave out precise recombination time info.
+Goal of the introduction: bring across the need to do this.
 
@@
 
-The breakthroughs in scale achieved by recent methods 
-are all based, in different ways, on inferring approximate 
-structures instead of a complete and fully detailed history
-% in particular with respect to recombination nodes
+%Although the coalescent with recombination best (?) probabilistic description of 
+%the events generating an ARG for a given sample, ...
 
-% first major step forwards in ARG inference
-% explicitly outputs an ARG
-ARGweaver, generates these ARGs explicitly again using MCMC sampling, by discretizing time 
-they can apply HMM-based algorithms to efficiently generate new samples from the distribution of 
-possible ARGs.
-% [TO ADD] also argweaver is the first to compute likelihoods for an ARG under the SMC, musn't forget to include this.
-% model-based approaches vs
+All these: fully precise ARG under the coalescent with recombination.
+Computationally prohibitive to explore ARG space:
+there are infinitely many ARGs compatible with any genomic dataset, 
+and many of these ARGs will have a comparable likelihood \citep{mcvean_approximating_2004}.
+By dropping the fully precise requirement that (scalable) ARGs became computationally 
+feasible. First ARGwaver.
+% ARGweaver, model-based yet approximate (discretised)
+% [TO ADD] also argweaver is the first to compute likelihoods for an ARG under 
+% the SMC, musn't forget to include this.
+Model-based (an approximation of the coalescent, SMC) + approximate in the sense of discretized.
+and later really scalable methods:\\
 
-% recent advancements: heuristics
+[TL;DR] next step: heuristic methods
 Recently, major advancements relying on heuristic rather than model-based approaches
@@
 part of the population and statistical genetics toolkit.
+% however, scalability has impacted our ability to connect these to any model
 This has also meant however that
@@
 as generated by these different inference methods \citep{wong_a-general_2023}. 
-Besides not having a unified output format, The degree of completeness 
+Besides not having a unified output format, the degree of completeness 
 in which genetic inheritance from ancestors to 
 descendants is documented by each of these methods varies extensively.
-[TO DO:] discuss: each of these tools will deal with or resolve uncertainty 
-in a different way.
 This has complicated the ability to connect the output of this diverse range of 
@@
 
-Recombination in particular:
-In contrast to model-based inference methods like \argweaver, 
-these models only deal with recombination in an implicit sense.\\
+[TL;DR] Specify  different structural approximations:
+1/ Recombination is implicit. Recombinations are not modelled or inferred explicitly, 
+they can be observed however as tree changes. However a tree change might be the result of one or 
+more recombination events.
+2/ each of these tools will deal with or resolve uncertainty 
+in a different way.\\
 
-[TO DO:] what does *general* mean? and when in the paper do we discuss it?
+[TL;DR] Specify different model-based approximations:
+Coalescent with Recombination as the dominant model to describe ancestry.
+Through time different approximations. Wiuf and Hein demonstrating a left-to-right 
+algorithm. No longer Markovian. This idea used by McVean and Cardin: SMC. 
+And followups (SMC', SMC(k)) \citep{mcvean_approximating_2004}.
+SMC and relatives are useful: lot of focus however on integrating over 
+the distribution of trees to get $N_e(t)$ [ADD REF: PSMC, MSMC]\\
 
-[TO DO: finish this section] For example, where ARGweaver has 
-the built in assumption of neighbouring trees being separated by single Subtree Prune 
-Regraft (SPR) [REF Song et al.], this is not the case for any of the heuristic data 
-structures.
-Lack recombination nodes in the ARG, crucial in ability to compute likelihoods 
-under Hudson. \citep{wong_a-general_2023}
-
-% then onto the SMC
-[TO ADD: discuss SMC] mention the SMC \citep{mcvean_approximating_2004}.
-SMC and relatives are useful: lot of focus however on integrating over 
-the distribution of trees to get $N_e(t)$ [ADD REF: PSMC, MSMC]
+[TL;DR] Wrap up by making clear that we will combine the two types of approximations:
 Here, the idea is to use the SMC in a different way that resonates more 
 with the true direction of flow of genetic information.
-
 Results in a concrete way to compute likelihoods for *any* inferred ARG 
@@
 and formulate an algorithm 
-that is general in the sense that it can deal with the various levels of 
+that is \textit{general} in the sense that it can deal with the various levels of 
 completeness that characterise the field's currents attempts [TO DO: make this sound more positive] 
@@
 reveals however that all internal nodes but node 4 are unary along some part of 
-the genome. These nodes all represent coalescence events.
-
-Note that the presence of these unary nodes allows us to reconstruct the full extent 
-of each lineage going back through time. Similarly, all lineages involved in a recombination 
-event can be uniquely identified by a change in parent going from one tree to the next 
-for the child node associated with that lineage. Both the ARG and the sequence of 
+the genome. These nodes all represent coalescence events. Note that the presence of these 
+unary nodes allows us to reconstruct the full extent 
+of each lineage going back through time as well as uniquely identify all lineages involved 
+in a recombination event. Both the ARG and the sequence of 
 local trees contain the same information, as long as they are simplified to the 
@@
 \centering
-\includegraphics[width=\textwidth]{figures/smc_custom_2rows.png}
+\includegraphics[width=\textwidth]{figures/smc_custom_2rows_area.png}
 \caption{ARG (top left) generated under the SMC and its representation as a 
@@
 [TO DO: discuss, what if the identity of internal nodes is not preserved across trees]
+[TO DO: recombination]: We do not *explicitly* assume a single recombination breakpoint to have 
+occurred at each breakpoint.
+[TO DO:] On a similar note and more general: we can compute the likelihood for any 
+valid tree sequence, however, future work to verify to what extent/in which scenarios 
+this is actually helpful.
 
@@
 advances in ARG inference and the current state of the art statistical inference of 
-past demography. Secondly, hybrid ARG-inference methods that combine the best of both 
+past demography. 
+Secondly, inferred ARGs can be used as a good starting point for a future MCMC sampler.
+Thirdly, hybrid ARG-inference methods that combine the best of both 
 heuristic and model-based inference approaches are now possible.
-Thirdly, using a model we can quantify the uncertainty inherently associated 
+Finally, using a model we can quantify the uncertainty inherently associated 
 with ARG-reconstruction.
@@
 trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
-provided here can function as a building block of a 
-demography-aware genealogy sampler that would allow us to incorporate the information 
-provided by recombination events.
-More concretely, given our algorithm, any inferred ARG can be used a good starting point 
+provided here can function as a building block of
+
+$N_e$-changes as well as multiple populations and migration can be readily incorporated 
+in the current algorithm.\\
+
+% 2/ great starting point for MCMC
+Any inferred ARG can be used a good starting point 
 for a future MCMC-sampler. This potentially implies a massive runtime reduction by 
@@
 efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
-$N_e$-changes as well as multiple populations and migration can be readily incorporated 
-in the current algorithm.\\
+[TO DO:]
 
-% 2/ hybrid ARG-inference
+% 3/ hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
 Hybrid backwards-in-time simulations that combine Wright-Fisher dynamics in the 
@@
 
-% 3/ quantifying uncertainty [unfinished]
+% 4/ quantifying uncertainty [unfinished]:
 Both of these future directions tie in with the desired ability of any ARG-related method to 
 quantify the uncertainty associated with
-
-to quantify uncertainty by sampling 
-from an underlying distribution.
 

### 2023-11-28 paper.tex
update intro

@@
 a model such as the coalescent.
-What we want to achieve with this paper is to compute a likelihood for 
-an approximate (define) ARG under a approximate model of the coalescent. 
-Approximate in two senses: SMC + leave out precise recombination time info.
-Goal of the introduction: bring across the need to do this.
 
-The underlying genealogical history that relates a sample of individuals back in time
-contains all possible information this sample can provide on the history of demography, 
-selection and recombination for the population of interest. 
-Reconstructing these genealogies with certainty is impossible however. 
+1/ ARGs are awesome, or general intro about approximations
+
+model is a mathematical simplification of reality that aims to capture essential aspects 
+about what we observe.
+For example, the coalescent, ...
+Formal way of recording what we observe, results in data structure 
+with a certain level of precision.
+For example, ARG, ...
+The structure capturing the details of these intricately 
+interwoven paths of inheritance is referred to as an ancestral recombination graph (ARG)
+
+For a very long time people have assumed that what is meant by an ARG is the 
+fully specified genealogical history  
+where all events are uniquely positioned both in time
+and along the genome. [However see Wong et al. 2023].
+
+Reconstructing these genealogies with certainty about the assumed level of detail 
+is impossible.
 As such, in the past the data structure capturing these inheritance paths 
@@
 variable to be integrated out during statistical inference 
-\citep{kuhner_2000, griffiths_ancestral_1996}.
+\citep{kuhner_2000, griffiths_ancestral_1996}.\\
 
-%Although the coalescent with recombination best (?) probabilistic description of 
-%the events generating an ARG for a given sample, ...
+% 2/ approximations have driven the rate of progress in inference
+Classical approach Hudson, Markovian backwards in time.
+McVean and Cardin, SMC, when described left-to-right also Markovian 
+in that sense. By sacrificing precision on the our ability to capture long 
+range linkage information gain tractability.
 
-All these: fully precise ARG under the coalescent with recombination.
-Computationally prohibitive to explore ARG space:
-there are infinitely many ARGs compatible with any genomic dataset, 
-and many of these ARGs will have a comparable likelihood \citep{mcvean_approximating_2004}.
-By dropping the fully precise requirement that (scalable) ARGs became computationally 
-feasible. First ARGwaver.
-% ARGweaver, model-based yet approximate (discretised)
-% [TO ADD] also argweaver is the first to compute likelihoods for an ARG under 
-% the SMC, musn't forget to include this.
-Model-based (an approximation of the coalescent, SMC) + approximate in the sense of discretized.
-and later really scalable methods:\\
+SMC functioned as the engine of many powerful inference methods 
+if the continuous state space for the Markov chain 
+(consisting of all possible genealogies) is approximated by a moderately 
+sized finite set—typically by enumerating tree topologies and/or discretizing 
+time—then inference can be performed efficiently using well-known algorithms 
+for hidden Markov models (HMMs).
+ARGweaver uses these same insights to develop the first real milestone in ARG inference.\\
 
-[TL;DR] next step: heuristic methods
-Recently, major advancements relying on heuristic rather than model-based approaches
-have made ARG inference scalable.
-\citep{rasmussen_2014, kelleher_2019, speidel_2019, schaefer_2021, wohns_2022, 
-zhang_biobank-scale_2023, zhan_towards_2023}. These methods
+% 3/ Recent progress in ARGs through approximate structures
+
+% bring nuance in the distinction between detail and approximation
+
+% heuristic methods guided by coalescent principles
+
+% what is an approximate structure: implicit recombination, polytomies or not, ...
+
+These methods
 have raised the realistic prospect of ARG-based analysis becoming a standard 
 part of the population and statistical genetics toolkit.
-% however, scalability has impacted our ability to connect these to any model
-This has also meant however that
+
+% 4/ Big problem: how do you connect such approximate structures each 
+% with different levels of detail to models?
+These recent developments have also meant however that
 we are now using the term ARG for a varying range of approximate structures 
@@
 probabilistic exploration of the ARG space around the point estimate, 
-or likelihood-based inference (that does not just consider individual marginal trees).\\
+or likelihood-based inference (that does not just consider individual marginal trees).
 
-[TL;DR] Specify  different structural approximations:
-1/ Recombination is implicit. Recombinations are not modelled or inferred explicitly, 
-they can be observed however as tree changes. However a tree change might be the result of one or 
-more recombination events.
-2/ each of these tools will deal with or resolve uncertainty 
-in a different way.\\
+% 5/ what we will do
 
-[TL;DR] Specify different model-based approximations:
-Coalescent with Recombination as the dominant model to describe ancestry.
-Through time different approximations. Wiuf and Hein demonstrating a left-to-right 
-algorithm. No longer Markovian. This idea used by McVean and Cardin: SMC. 
-And followups (SMC', SMC(k)) \citep{mcvean_approximating_2004}.
-SMC and relatives are useful: lot of focus however on integrating over 
-the distribution of trees to get $N_e(t)$ [ADD REF: PSMC, MSMC]\\
-
-[TL;DR] Wrap up by making clear that we will combine the two types of approximations:
-Here, the idea is to use the SMC in a different way that resonates more 
-with the true direction of flow of genetic information.
-Results in a concrete way to compute likelihoods for *any* inferred ARG 
+Here, the idea is to use the backwards-in-time formulation of the SMC to 
+compute likelihoods for for \textit{any} inferred ARG 
 by integrating out the time to recombination.
-and formulate an algorithm 
+The application of the SMC in a way that resonates more 
+with the true direction of flow of genetic information 
+results in a concrete way to formulate an algorithm 
 that is \textit{general} in the sense that it can deal with the various levels of 
-completeness that characterise the field's currents attempts [TO DO: make this sound more positive] 
-at reconstucting genome-wide genealogies.
+completeness that characterise the current state of the field of ARG reconstruction.
 

### 2023-11-28 paper.tex
update mathematical description

@@
 left of $x$ with $x>x_{i0}$ and $x<y_{im}$ splits a lineage $i$ into two new, uniquely 
-labelled lineages.\\
+labelled lineages. \\ % old lineage is removed from L, two new lineages are added
 
@@
 of a diploid population of constant size. The newly formed lineage acquires the 
-union of both ordered intervals.
+union of both ordered intervals. % L is updated accordingly
 Although coalescence is reciprocal, here we'll define a strict total order 
 (see \ref{par:liks}) on 
-all lineages in $X(t)$ based on their leftmost starting points. 
+all lineages in $L(t)$ based on their leftmost starting points. 
 For any such strict total order on $X(t)$ and at any point in the process 
@@
 I_{i}(t) = \sum_{j} I_{ij}(t)$.
-Because $I_{ij}$ is not reciprocal, each extant lineage has their own unique set of 
+Because $I_{ij}$ is not reciprocal, each extant lineage has its own unique set of 
 lineages $C_i(t)$ it can coalesce with. This will simplify the likelihood computations 
@@
 The ARG as described here records the impact of every possible event along each 
-inheritance path in full detail. We could however simplify this graph by 
-throwing away the precise time information of all recombination events. This 
-corresponds to removing the recombinant internal nodes within the ARG such that some of 
-the internal nodes representing coalescence events now have 2 outgoing (backwards in time) 
-edges. On the ARG in fig.\ref{fig:smc-unary}, the output of a simulation under the SMC, 
-this corresponds to removing the blue internal nodes. Crucially however, 
-we can still identify all lineages that were hit by a recombination event. Yet we 
-can only constrain the time to this event to within a window defined by the age of the 
-child node $c$ associated with that lineage and the minimum age of the parents of the 
-edges connecting to $c$.\\
+inheritance path in full detail. Fig.\ref{fig:smc-unary} shows one such possible 
+ARG generated represented 
+as both a graph (top) as well as a series of local trees (bottom). 
+Note that both structures represent the same information. 
+Crucial for the equivalence between both representations is the presence of
+nodes that (locally) only have a single child along one or more local 
+trees. Not all of these (locally) unary nodes are 
+(blue, fig.\ref{fig:smc-unary}) recombinant nodes. By recording edges along 
+the full extent of each lineage, coalescing nodes can be locally unary 
+along particular genome intervals. Furthermore, the unary nodes enable 
+us to uniquely identify all lineages that were hit by a recombination 
+event, even when we simplify the ARG by removing all 
+(blue, fig.\ref{fig:smc-unary}) recombinant nodes. 
+In the absence of recombination nodes, a recombination can be observed 
+easily enough on the local trees as a change in parent going from 
+one tree to the next. This simplification step 
+does mean however that we lose precise time information on the 
+recombination events. Yet we can constrain the uncertainty to 
+within a window defined by the age of the child node $c$ associated 
+with that lineage and the minimum age of the parents of the 
+edges connecting to $c$. This in turn is easier to visualise using the 
+graph representation (blue shaded area, fig.\ref{fig:smc-unary}).\\
 
-ARG simplification therefore touches upon the question of what is knowable.  
-[TO DO: Add link with inference methods]\\
+ARG simplification is thus linked to the presence of unary nodes in 
+the local trees and simultaneously touches upon the question of what is 
+knowable. Many ARG reconstruction methods return a data structure that 
+is in essence a series of local tries [REFs]. 
+As highlighted, this does not need to be a contradiction. 
+All ARGs can be represented as a series of local trees \citep{wong-2023}.
+The important question is therefore is however about the extent by which any 
+ARG reconstruction methods incorporates the available information 
+on how nodes persist across these local trees.\\
 
-Many ARG reconstruction methods return a data structure that 
-is in essence a series of local tries [REFs]. This does not need to be a contradiction. 
-All ARGs can be represented as a series of local trees \citep{wong-2023}.
-As long as we register each observed node when building the local tree that 
-records the path through the ARG, then we can guarantee a one-to-one correspondence 
-with the original ARG. Crucially, this will result in nodes that 
-are locally unary. These are nodes that only have a single child along some of the 
-intervals within the genome.\\
-
-Recombination nodes are one such type of unary nodes. Figure \ref{fig:smc-unary} (bottom)
-reveals however that all internal nodes but node 4 are unary along some part of 
-the genome. These nodes all represent coalescence events. Note that the presence of these 
-unary nodes allows us to reconstruct the full extent 
-of each lineage going back through time as well as uniquely identify all lineages involved 
-in a recombination event. Both the ARG and the sequence of 
-local trees contain the same information, as long as they are simplified to the 
-same degree.\\
-
-[TO DO: do I need to expand on this?] Simplification is thus tied in with the concept of locally unary nodes.\\
-
-Which ARG inference methods return locally unary nodes and how can they be inferred.
-[TO DO] give a good explanation (using the figure) why these methods can infer the extent 
+[TO DO] give a good explanation (using the represented haplotypes 
+in fig. \ref{fig:smc-unary}) on why these methods can infer the extent 
 of coalescing lineages.
-As highlighted in fig.\ref{fig:smc-unary}, (coalescing) unary nodes can be inferred 
-from the available haplotype information. Currently, the following methods 
+Currently, the following methods 
 return the full this information by default as part of their ARG reconstruction algorithm: 
@@
 \tsinfer \citep{kelleher_inferring_2019}, 
-\kwarg \citep{ignatieva_kwarg_2021}.
+\kwarg \citep{ignatieva_kwarg_2021}.\\
 
-[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.
+[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.\\
 
@@
 \includegraphics[width=\textwidth]{figures/smc_custom_2rows_area.png}
-\caption{ARG (top left) generated under the SMC and its representation as a 
-series of local trees. The presence of the unary nodes in the local trees 
-ensure these two representations contain the same information. 
-The ARG is augmented with haplotype information, coloured according to the 
-samples that inherit from it. 
-Note how the presence of a unary node on a local tree is 
+\caption{ARG generated under the SMC and its representation as a 
+series of local trees (bottom). To ensure the one-to-one correspondance 
+between both representations a node is recorded for each local tree whenever 
+a node is encoundered along its ancestry path in the graph representation. 
+This results in nodes that (locally) have only a single child, and are therefore 
+(locally) unary. Blue (unary) nodes represent recombinant nodes. 
+The blue arrows shows the familiar left-to-right logic of the SMC whereby 
+the floating lineage coalesces randomly
+with a the remaining portion of the local tree following recombination.
+Note how the presence of the unary node on a local tree is 
 indicative of either a past (to the left of the local tree) 
-or future (to the right) coalescence event. It therefore modifies the 
-left-to-right logic of the SMC whereby the floating lineage coalesces randomly
-with a the remaining portion of the local tree. 
+or future (to the right) coalescence event. 
+Both ARG representations can be simplified by removing the recombinant nodes. 
+In that case the information we have on the time to each of these events is 
+limited to the windows (blue shaded array) delimited by by the age of 
+the child node $c$ associated with that lineage and the minimum age of 
+the parents of the edges connecting to $c$. 
+ [TO DO: wanted to use the 
+added in haplotype information to explain why we can actually \textit{infer} the 
+extent of each lineage, and thus create unary nodes in the first place.]
 }
@@
 
-What's left to compute the likelihood, is to take into account all events 
-associated with $x$, 
-the leftmost coordinate of $(c, p, X_{cp})$. This is either only the 
+What's left to compute the likelihood of the focal edge, is to take into account 
+all events associated with $x$, 
+the leftmost coordinate of $(c, p, X_{cp})$. 
+This is either only the 
 coalescence event involving 
-$c$ and $d$. In this case $x=x_{c0}$. Or, $x$ can a new recombination break point that 
+$c$ and $d$. In that case $x=x_{c0}$ and $c$ did not coalesce until $t_p$. 
+Alternatively, $x$ can be a new recombination break point that 
 occurred on the lineage represented by $c$ at some time $s$ between $t_{c}$ 
 and $\min(t_p, t_{p^{\prime}})$, where $p^{\prime}$ is the parent of the 
-edge $(c, p^{\prime}, X_{cp^{\prime}})$ ending at $x$.
-In the latter case the new lineage starting at $x$ can be given a temporary label 
-$c^{\prime}$ such that with the same logic from the previous paragraph we 
-can define $F(c^{\prime}, s, t) = \int_{s}^{t} I_{c^{\prime}}(u)du$.
+edge $(c, p^{\prime}, X_{cp^{\prime}})$ ending at $x$. The new lineage formed 
+by the recombination event and starting at $x$ can be given a temporary label 
+$c_s$. In both cases, the cumulative hazard of a lineage $a$ coalescing 
+between time $s$ and $t$ is given by $F(c, s, t) = \int_{s}^{t} I_{c}(u)du$ 
+such that:
 
@@
 e^{-\lambda F(c, t_c, t_p)} \lambda^{I_{cd}(t_p)} & x=x_{c0} \\
-\int_{t_c}^{t_{p} \wedge t_{p^{\prime}}} r e^{-rs} e^{-\lambda F(c^{\prime}, s, t_{p})} ds \lambda^{I_{c^{\prime}d}(t_p)} & x=x_{c^{\prime}0}>x_{c0} \\
+\int_{t_c}^{t_{p} \wedge t_{p^{\prime}}} r e^{-rs} e^{-\lambda F(c_s, s, t_{p})} ds \lambda^{I_{c_{s}d}(t_p)} & x=x_{c_{s}0}>x_{c0} \\
 \end{cases}
@@
 that terminates the edge. The indicator function is used to resolve 
-the non-reciprocal nature of coalescence events as defined here.\\
+the non-reciprocal nature of coalescence events as defined in \ref{def:coal}.\\
 
@@
 \caption{Description of algorithm: Moving along the genome we transition from  
-local tree $T_1$ (1) to the next, $T_2$ (4). Edges that do not persist in $T_2$ are 
-removed (2). Edges starting at the left coordinate of $T_2$ are inserted (3). 
+local tree $T_1$ to the next, $T_2$ in three steps. 
+Edges that do not persist in $T_2$ are 
+removed first. Edges starting at the left coordinate of $T_2$ are inserted. 
 After deletion or insertion of an edge the count vector $I$ is updated along the 
-corresponding internode intervals. Prior to updating $I$ following an insertion, 
+corresponding internode intervals. Following an insertion, and prior to updating $I$, 
 the likelihood for that edge is computed as $I$ then represents the total number of 
@@
 By explicitly associating a single likelihood with each edge, we can compute a likelihood 
-under the SMC for any valid tree sequence. The \texttt{tskit} library provides a whole range  
+under the SMC for any valid tree sequence. The \tskit library provides a whole range  
 of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
@@
 Although we rely heavily on unary nodes to provide us with the necessary 
-information on recombination events, our algorithm only implicitly assumes there presence.
+information on recombination events, our algorithm only implicitly assumes their presence.
 The algorithm can therefore also accomodate for 
 those inferred ARGs that lack unary nodes (e.g.\ \relate). 
-Note that without unary nodes, our algorithm will detect a recombination event along 
+Without unary nodes though, our algorithm will detect a recombination event along 
 both coalescing lineages in case of a (sub)tree height changes along the ARG as both 
@@
 'inferrable' recombination events per coalescence event to 1.
-% simulated results
+
+[TO DO]: improve results sections with some good viz,
 We simulated 1000 ARGs under the Hudson coalescent retaining all recombination information 

### 2023-12-05 paper.tex
add figures
supplementary figures

@@
 \usepackage{tikz}
+\usepackage{placeins}
 
@@
  [TO DO: wanted to use the 
-added in haplotype information to explain why we can actually \textit{infer} the 
+added in haplotype information to explain why we can actually infer the 
 extent of each lineage, and thus create unary nodes in the first place.]
@@
 An implementation of the algorithm is available on https://github.com/gertjanbisschop/runsmc.
-
+\FloatBarrier
 \bibliographystyle{plainnat}
@@
 \includegraphics[width=0.75\textwidth]{figures/supplementary-figs/argweaver_vs_runsmc.png}
-\caption{Compare likelihoods inferred ARGs ARGweaver: placeholder for final figure with similar content}
+\caption{Simulated 10kb sequence under human-like parameters, 10 diploid samples.
+Inferred ARG using ARGweaver.
+Scatterplot is comparison of prior (discretized SMC) as reported by ARGweaver (x-axis) and likelihood as defined here (y-axis).}
  \label{sup:fig:vs-argweaver}
@@
 \includegraphics[width=0.75\textwidth]{figures/supplementary-figs/v_hudson_unary_simpl.png}
-\caption{Compare likelihoods Hudson and SMC with and without unary nodes: placeholder for final figure with similar content}
+\caption{Compare likelihoods Hudson (full ARG) and SMC with and without unary nodes (without recombination nodes): placeholder for final figure with similar content.}
 \label{sup:fig:vs-hudson}
@@
 
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=0.75\textwidth]{figures/supplementary-figs/without_unary.png}
+\caption{Simulated 1Mb of sequence under human-like parameters.
+Computed likelihood as defined here with unary nodes (y-axis), and without, blue dots (x-axis).
+By limiting the number of recombinations (to 1 event) leading up to a single coalescent node we obtain the values in orange.}
+ \label{sup:fig:rec-correction}
+\end{figure}
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=0.75\textwidth]{figures/supplementary-figs/hudson_liks_vs_tsinfer_smc.png}
+\caption{Simulated full ARG for 1Mb sequence under human-like parameters (100 diploid samples).
+Computed likelihood (x-axis) under Hudson.
+Then inferred ARG with \tsinfer following a random sprinkling of mutations and computed likelihood under SMC (y-axis)}
+ \label{sup:fig:hudson-smc}
+\end{figure}
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=0.75\textwidth]{figures/supplementary-figs/single_rep_human_like_1gb.png}
+\caption{Single replicate from the dataset simulated above with a 100 random sprinklings of mutations.
+Likelihood under the SMC, mutation rate $\mu=1.25e-8$ (x-axis), and with mutation rate $4*\mu$ (y-axis).
+Likelihood for the full ARG under Hudson is indicated with the cross.}
+ \label{sup:fig:single-tsinfer}
+\end{figure}
+
 \end{document}

### 2023-12-11 paper.tex
update introduction

@@
 
-1/ ARGs are awesome, or general intro about approximations
+\subsection{ARGs are awesome, or general intro about approximations}
+% approximation: level of detail
 
-model is a mathematical simplification of reality that aims to capture essential aspects 
-about what we observe.
-For example, the coalescent, ...
-Formal way of recording what we observe, results in data structure 
-with a certain level of precision.
-For example, ARG, ...
-The structure capturing the details of these intricately 
-interwoven paths of inheritance is referred to as an ancestral recombination graph (ARG)
-
+In population genetics, the questions we try to answer are generally 
+about the unobserved processes that gave rise to the genetic variation 
+we can observe given a set of samples. 
+The coalescent with recombination 
+Kingman, 1982a,b; Tajima, 1983; Hudson, 1983) is generally used as 
+the generative model that allows us to understand the genealogy 
+underlying a particular sampleset. 
+Datastructure that captures it: ARG.
 For a very long time people have assumed that what is meant by an ARG is the 
@@
 and along the genome. [However see Wong et al. 2023]. 
+% say something about the difficulty of precise ARG inference (see more scalable ARG inference later on).
 
-Reconstructing these genealogies with certainty about the assumed level of detail 
-is impossible.
-As such, in the past the data structure capturing these inheritance paths 
-or Ancestral Recombination Graph (ARG), has usually been considered a nuisance 
-variable to be integrated out during statistical inference 
-\citep{kuhner_2000, griffiths_ancestral_1996}.\\
+Although ARG, ...,
+something about the level of detail.
 
-% 2/ approximations have driven the rate of progress in inference
-Classical approach Hudson, Markovian backwards in time.
-McVean and Cardin, SMC, when described left-to-right also Markovian 
-in that sense. By sacrificing precision on the our ability to capture long 
-range linkage information gain tractability.
+\subsection{approximations have driven the rate of progress in inference}
 
-SMC functioned as the engine of many powerful inference methods 
-if the continuous state space for the Markov chain 
-(consisting of all possible genealogies) is approximated by a moderately 
-sized finite set—typically by enumerating tree topologies and/or discretizing 
-time—then inference can be performed efficiently using well-known algorithms 
-for hidden Markov models (HMMs).
-ARGweaver uses these same insights to develop the first real milestone in ARG inference.\\
+Instead, approximations of the coalescence with recombination (CwR) have  
+driven our ability to perform statistical genomic inference. In particular, 
+the ability to formulate the inference problem as a hidden Markov model (HMM) 
+has proven to be instrumental [references]. 
+Two crucial approximations: 2 characteristics: left to right, and discetization 
+of state space by emumerating tree topologies and/or discretising time,
+then inference can be performed efficiently using well-known HMM algorithms.
+To improve the ability to simulate the CwR, McVean and Cardin 
+provided an approximation of the CwR that is both Markovian backwards in time 
+as well as left-to-right along the genome \citep{mcvean_2005_approximating}. 
+By sacrificing precision on the our ability to capture long range linkage 
+information, the sequentially Markovian coalescent (SMC) 
+became the engine of many powerful inference methods [references]. 
+Later improvements, SMC', closed the gap between the SMC and the CwR even 
+further.
+A second key approximation is not a generative model, instead the  
+copying model \citep{li_2003_modeling} was specifically designed to 
+generate CwR-like patterns. This heuristic approach has been successfully 
+applied for demographic inference [and other things: ADD] 
+\citep{sheehan_2013_estimating, steinrucken_2018}.
 
-% 3/ Recent progress in ARGs through approximate structures
+Both approaches have further formed the basis of the recent breakthroughs in 
+scalable ARG inference methods [references].
+
+ARGweaver uses 
+sample an ARG from the posterior distribution given the observed variational 
+data. Although require limiting the set of possible node times to be compatible 
+with HMM, ARGs they produce are fully detailed in the sense that recombinations 
+are explitly inferred. 
+All other methods whether partially relying on
+using SMC [ARGneedle] or Li and Stephens [tsinfer, relate] or not [Espallier, ADD] 
+are heuristic in nature and % keep models/method separate from level of detail!!
+
+\subsection{Recent progress in ARGs through approximate structures}
 
@@
 
-% heuristic methods guided by coalescent principles
-
-% what is an approximate structure: implicit recombination, polytomies or not, ...
+% heuristic methods guided by coalescent principles (Li and Stephens)
 
@@
 
-% 4/ Big problem: how do you connect such approximate structures each 
-% with different levels of detail to models?
+Although all different, all these methods share the fact that the ARGs they infer 
+are approximate structures.
+% what is an approximate structure: implicit recombination, polytomies or not, ...
+They lack the full detail: implicit recombination, ...
+In particular, recombination is dealt with implicitly. Recombination events are 
+not directly inferred and thus not dated. Tree topology changes from one local 
+tree to the next without explicit inference of the one or more recombination 
+events required to explain the inferred tree transition.
+
+Note that approximate structure need not be an issue, the fully precise and detailed 
+ARG can never be known. Instead, the main question is how to connect these 
+approximate structures to the existing generative models.
+
+\subsection{The Big problem: how do you connect such approximate structures to a generative model}
+
 These recent developments have also meant however that
@@
 
-% 5/ what we will do
+\subsection{what we will do}
 

### 2023-12-11 paper.tex
update introduction

@@
 % approximation: level of detail
+* [Importance of ARGs: ARGs as object of theoretical reasoning vs what can be known]\\
+* [difference between precision/detail and accuracy. High precision can 
+give a false sense of certainty in the provided estimate].
 
@@
 something about the level of detail.
+% easy by which the CwR can be used to simulate is not matched by the usability of the model for inference
 
@@
 Instead, approximations of the coalescence with recombination (CwR) have  
-driven our ability to perform statistical genomic inference. In particular, 
-the ability to formulate the inference problem as a hidden Markov model (HMM) 
-has proven to be instrumental [references]. 
-Two crucial approximations: 2 characteristics: left to right, and discetization 
-of state space by emumerating tree topologies and/or discretising time,
-then inference can be performed efficiently using well-known HMM algorithms.
+driven our ability to perform statistical genomic inference. Two approximations 
+in particular have proven to be instrumental [...]. Tightly linked to their 
+ability to reformulate these inference problems as hidden markov Models (HMM) 
+and use the well-known efficient HMM-algorithms. 
+Both: 1/ Left to right, and 2/ discetization 
+of state space by emumerating tree topologies and/or discretising time.\\
+
 To improve the ability to simulate the CwR, McVean and Cardin 
@@
 as well as left-to-right along the genome \citep{mcvean_2005_approximating}. 
-By sacrificing precision on the our ability to capture long range linkage 
+By sacrificing precision on our ability to capture long range linkage 
 information, the sequentially Markovian coalescent (SMC) 
-became the engine of many powerful inference methods [references]. 
-Later improvements, SMC', closed the gap between the SMC and the CwR even 
-further.
+became the engine of many powerful inference methods [PSMC]. 
+A later improvement, the SMC', closed the gap between the SMC and the 
+CwR even further.
+
 A second key approximation is not a generative model, instead the 
 copying model \citep{li_2003_modeling} was specifically designed to 
-generate CwR-like patterns. This heuristic approach has been successfully 
+generate CwR-like patterns. 
+Rather than providing a mathematical description of the genealogies 
+underlying the observed data it pragmatically describes the relationship 
+of a single genome to a larger set of samples as the result of an 
+incomplete copying process. This heuristic approach has been successfully 
 applied for demographic inference [and other things: ADD] 
@@
 
-Both approaches have further formed the basis of the recent breakthroughs in 
+Both approaches have further formed the basis of the recent advancements in 
 scalable ARG inference methods [references]. 
-
-ARGweaver uses 
-sample an ARG from the posterior distribution given the observed variational 
-data. Although require limiting the set of possible node times to be compatible 
-with HMM, ARGs they produce are fully detailed in the sense that recombinations 
-are explitly inferred. 
-All other methods whether partially relying on
-using SMC [ARGneedle] or Li and Stephens [tsinfer, relate] or not [Espallier, ADD] 
-are heuristic in nature and % keep models/method separate from level of detail!!
+First real breakthrough, ARGweaver, samples an ARG from the posterior distribution 
+given the observed variational data. Although require limiting the set of 
+possible node times to be compatible with HMM algorithms, the ARGs returned 
+by ARGweaver are fully precise and detailed in the sense that all 
+recombination events are explictly inferred. 
+This is where the real big difference lies with all other scalable methods. 
+Although all different, all these methods share the fact that the ARGs they infer 
+are approximate structures. The ARGs they infer have varying degrees of detail
 
@@
 
-Although all different, all these methods share the fact that the ARGs they infer 
-are approximate structures.
 % what is an approximate structure: implicit recombination, polytomies or not, ...
@@
 
+% Existing formalisms assume that an ARG is known in full detail
 These recent developments have also meant however that

### 2023-12-11 paper.tex
time slice figure

@@
 % approximation: level of detail
-* [Importance of ARGs: ARGs as object of theoretical reasoning vs what can be known]\\
-* [difference between precision/detail and accuracy. High precision can 
-give a false sense of certainty in the provided estimate].
+%Importance of ARGs: ARGs as object of theoretical reasoning vs what can be known\\
+%difference between precision/detail and accuracy. High precision can 
+%give a false sense of certainty in the provided estimate].
 
@@
 
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=\textwidth]{figures/3-slices.png}
+\caption{Simulated ARG under Hudson coalescent (left).
+$N_e$ varied was constant within each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 2500, (5000:\infty):10 000]$.
+Remaining parameters: L=10kb, 20 diploid samples, r=2.5e-8, 148 edges in ARG).
+A single $N_e$ value was estimated for each slice of the ARG as defined by the true population size change time points.
+The inferred posterior distribution (right) using the SMC-likelihood was obtained using a uniform prior.}
+ \label{fig:3-arg-slices}
+\end{figure}
+
 \section{Discussion}

### 2023-12-12 paper.tex
update intro

@@
 \section{Abstract}
-This means there currently is no way to connect the output of the diverse 
+There currently is no way to connect the output of the diverse 
 range of ARG inference methods back 
 to a model that would allow us to perform a probabilistic exploration of the 
-ARG space around the point estimate, or likelihood-based inference
+ARG space around the point estimate, or likelihood-based inference.
 
@@
 \subsection{ARGs are awesome, or general intro about approximations}
-% approximation: level of detail
-%Importance of ARGs: ARGs as object of theoretical reasoning vs what can be known\\
-%difference between precision/detail and accuracy. High precision can 
-%give a false sense of certainty in the provided estimate].
+
+[TL;DR] Importance of ARGs.
+There is a discrepancy between ARGs as object of theoretical reasoning 
+vs what can be known. Explain the difference between precision/detail and accuracy. 
+High precision can give a false sense of certainty in the provided estimate.
 
@@
 Datastructure that captures it: ARG.
+
 For a very long time people have assumed that what is meant by an ARG is the 
-fully specified genealogical history  
-where all events are uniquely positioned both in time
-and along the genome. [However see Wong et al. 2023]. 
-% say something about the difficulty of precise ARG inference (see more scalable ARG inference later on).
-
-Although ARG, ...,
-something about the level of detail.
-% easy by which the CwR can be used to simulate is not matched by the usability of the model for inference
+fully specified genealogical history where all events are uniquely positioned 
+both in time and along the genome. Yet the ease by which the CwR can be simulated,
+and therefore the ease by which an ARG can be generated, is not not matched by 
+the usability of the model for inference.
 
@@
 driven our ability to perform statistical genomic inference. Two approximations 
-in particular have proven to be instrumental [...]. Tightly linked to their 
-ability to reformulate these inference problems as hidden markov Models (HMM) 
-and use the well-known efficient HMM-algorithms. 
-Both: 1/ Left to right, and 2/ discetization 
-of state space by emumerating tree topologies and/or discretising time.\\
+in particular have proven to be instrumental: the sequentially Markovian coalescent, 
+and its extensions (SMC) on the one hand, and the Li and Stephens model on the other. 
+The impact of both models is tightly linked to the fact that they have enabled 
+the reformulation of inference problems as hidden Markov models (HMM) and thus 
+the use of the well-known efficient HMM-algorithms.
+%Both: 1/ Left to right, and 2/ discetization 
+%of state space by emumerating tree topologies and/or discretising time.\\
 
@@
 By sacrificing precision on our ability to capture long range linkage 
-information, the sequentially Markovian coalescent (SMC) 
-became the engine of many powerful inference methods [PSMC]. 
+information, the SMC became the engine of many powerful inference methods [PSMC]. 
 A later improvement, the SMC', closed the gap between the SMC and the 
@@
 
-A second key approximation is not a generative model, instead the 
-copying model \citep{li_2003_modeling} was specifically designed to 
-generate CwR-like patterns. 
-Rather than providing a mathematical description of the genealogies 
-underlying the observed data it pragmatically describes the relationship 
-of a single genome to a larger set of samples as the result of an 
-incomplete copying process. This heuristic approach has been successfully 
-applied for demographic inference [and other things: ADD] 
+The second key approximation is not a generative model. Rather than providing 
+a mathematical description of the genealogies underlying the observed data 
+the Li and Stephens or copying model pragmatically details the relationship 
+of a single genome to a larger set of genomes as the result of an 
+incomplete copying process \citep{li_2003_modeling}. This heuristic approach 
+has been successfully applied for demographic inference [and other things: ADD] 
 \citep{sheehan_2013_estimating, steinrucken_2018}.
 
-Both approaches have further formed the basis of the recent advancements in 
-scalable ARG inference methods [references]. 
-First real breakthrough, ARGweaver, samples an ARG from the posterior distribution 
-given the observed variational data. Although require limiting the set of 
-possible node times to be compatible with HMM algorithms, the ARGs returned 
-by ARGweaver are fully precise and detailed in the sense that all 
-recombination events are explictly inferred. 
-This is where the real big difference lies with all other scalable methods. 
-Although all different, all these methods share the fact that the ARGs they infer 
-are approximate structures. The ARGs they infer have varying degrees of detail
+Both approaches have further formed the basis (of many) of the recent advancements in 
+scalable ARG inference methods [references]. A real distinction can be made however 
+between \argweaver, the first real breakthrough in scalable ARG inference, and the 
+more recent methods. The ARGs sampled by \argweaver from the posterior distribution 
+given the observed variational data, are all fully precise and detailed in the sense 
+that all events are explicitly inferred. Note however that the set of possible node times 
+has to be limited to enable the usage of an HMM. This fully detailed inference holds 
+for recombination, but also for migration events \cite{hubisz_2020_mapping}. 
+All other methods, on the other hand, derive their scalability at least to a 
+certain degree from the fact that 
+the ARGs they infer are approximate structures. In particular, recombination is an 
+emergent property of the inferred ARGs explaining tree topology changes going 
+from one local tree to the next. The recombination events themselves are not 
+explicitly inferred and are thus not part of the resulting data structure.
 
 \subsection{Recent progress in ARGs through approximate structures}
+% onto other features
 
-% bring nuance in the distinction between detail and approximation
+The computational gains from not having to pinpoint recombination events tie in 
+with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022].
 
-% heuristic methods guided by coalescent principles (Li and Stephens)
+Always events that cannot be known. Different ways of dealing with this.
+Some approaches have embraced the approximate nature of the resulting ARG to for 
+example highlight the inability to resolve certain splits (polytomies).
+
+others rigorous treatment of uncertainty (Arbores, ARGinfer): fully precise about recombination.
+
+% onto ARG cannot be fully known
+Note that approximate structure need not be an issue, the fully precise and detailed 
+ARG can never be known. Instead, the main question is how to connect these 
+approximate structures to existing models.
+While this is not a problem when describing the results of simulations 
+(where all details are perfectly known), it is a major issue when we wish to formally 
+describe the output of inference methods, particularly as datasets approach the population 
+scale. 
+As discussed in the introduction, the precise details of all 
+events in these vast ARGs can never be known, and a data structure that enforces 
+complete precision is therefore an impediment to progress.
+
+\subsection{The Big problem: how do you connect such approximate structures to a generative model}
 
@@
 
-% what is an approximate structure: implicit recombination, polytomies or not, ...
-They lack the full detail: implicit recombination, ...
-In particular, recombination is dealt with implicitly. Recombination events are 
-not directly inferred and thus not dated. Tree topology changes from one local 
-tree to the next without explicit inference of the one or more recombination 
-events required to explain the inferred tree transition.
-
-Note that approximate structure need not be an issue, the fully precise and detailed 
-ARG can never be known. Instead, the main question is how to connect these 
-approximate structures to the existing generative models.
-
-\subsection{The Big problem: how do you connect such approximate structures to a generative model}
-
-% Existing formalisms assume that an ARG is known in full detail
 These recent developments have also meant however that
@@
 as generated by these different inference methods \citep{wong_a-general_2023}.
+
 Besides not having a unified output format, the degree of completeness 
@@
 or likelihood-based inference (that does not just consider individual marginal trees).
+% Existing formalisms assume that an ARG is known in full detail
 
@@
 trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
-provided here can function as a building block of
-
+provided here can function as a building block of [...] 
 $N_e$-changes as well as multiple populations and migration can be readily incorporated 
@@
 Both of these future directions tie in with the desired ability of any ARG-related method to 
-quantify the uncertainty associated with
+quantify the uncertainty associated with ARG inference
 

### 2023-12-13 paper.tex
update intro

@@
 \newcommand{\kwarg}[0]{\texttt{KwARG}}
+\newcommand{\argneelde}[0]{\texttt{ARG-Needle}}
 \newcommand{\argweaver}[0]{\texttt{ARGweaver}}
@@
 
-[TL;DR] Importance of ARGs.
-There is a discrepancy between ARGs as object of theoretical reasoning 
-vs what can be known. Explain the difference between precision/detail and accuracy. 
-High precision can give a false sense of certainty in the provided estimate.
+Knowledge of the exact genealogy of a set of samples would make certain problems 
+in population genetics (e.g.\ inferring recombination rates) trivial. 
+Multitude of methods have started to 
+demonstrate the value of attempts to capture the details of these intricately 
+interwoven paths of inheritance referred to as an ancestral recombination graph (ARG).
+Recent advancements: the inference of ARGs at population scale.
+Yet even though potential of ARGs in population and statistical genetics has been 
+demonstrated it remains unclear how to connect these reconstructed ARGs to 
+the stochastic/generative models that have lead us to study these datastructures 
+in the first place.
 
-In population genetics, the questions we try to answer are generally 
-about the unobserved processes that gave rise to the genetic variation 
-we can observe given a set of samples. 
-The coalescent with recombination 
-Kingman, 1982a,b; Tajima, 1983; Hudson, 1983) is generally used as 
-the generative model that allows us to understand the genealogy 
-underlying a particular sampleset. 
-Datastructure that captures it: ARG.
+[Message needs to be here: way to think about genealogies is tied to the CwR. However,
+CwR does not enable us to simply compute a likelihood in the absence of a fully 
+detailed history. The way forward instead has always been driven by approximations.]
+
+% The coalescent with recombination 
+% Kingman, 1982a,b; Tajima, 1983; Hudson, 1983) is generally used as 
+% the generative model that allows us to understand the genealogy 
+% underlying a particular sampleset. 
+% Datastructure that captures it: ARG.
 
@@
 fully specified genealogical history where all events are uniquely positioned 
-both in time and along the genome. Yet the ease by which the CwR can be simulated,
-and therefore the ease by which an ARG can be generated, is not not matched by 
-the usability of the model for inference.
+both in time and along the genome [Wong et al. 2023].
+Yet, the ease by which such a fully detailed history can be simulated under the 
+CwR, is not matched by the usability of the model for inference. 
+In particular, given an ARG, there will always be an infinite amount of other 
+ARGs that could have generated the observed genomic dataset 
+[potentially with very different likelihoods under the CwR]
+\citep{mcvean_2005_approximating}.
+
+% There is a discrepancy between ARGs as object of theoretical reasoning 
+% vs what can be known. Explain the difference between precision/detail and accuracy. 
+% High precision can give a false sense of certainty in the provided estimate.
+
+% In population genetics, the questions we try to answer are generally 
+% about the unobserved processes that gave rise to the genetic variation 
+% we can observe given a set of samples.
 
@@
 Both approaches have further formed the basis (of many) of the recent advancements in 
-scalable ARG inference methods [references]. A real distinction can be made however 
-between \argweaver, the first real breakthrough in scalable ARG inference, and the 
-more recent methods. The ARGs sampled by \argweaver from the posterior distribution 
+ARG inference methods [Rasmussen et al. 2014, Heine et al. 2018, 
+Kelleher et al. 2019, Speidel et al. 2019,
+Rasmussen et al. 2022, Zhang et al.2023]. 
+A further classification of these methods could be made based on the level of detail 
+that is provided on recombination events. The first real breakthrough in large-scale 
+ARG inference, \argweaver, heads the first group. 
+The ARGs sampled by \argweaver from the posterior distribution 
 given the observed variational data, are all fully precise and detailed in the sense 
 that all events are explicitly inferred. Note however that the set of possible node times 
-has to be limited to enable the usage of an HMM. This fully detailed inference holds 
-for recombination, but also for migration events \cite{hubisz_2020_mapping}. 
-All other methods, on the other hand, derive their scalability at least to a 
-certain degree from the fact that 
-the ARGs they infer are approximate structures. In particular, recombination is an 
-emergent property of the inferred ARGs explaining tree topology changes going 
-from one local tree to the next. The recombination events themselves are not 
-explicitly inferred and are thus not part of the resulting data structure.
+has to be limited to enable the usage of an HMM. The other more recent methods, although 
+still SMC-based, have added heuristic pre- and/or post-processing steps 
+that have allowed them to extend their approach to larger sample sizes. We should also 
+mention \kwarg and \arginfer in this category of methods producing fully detailed ARGs 
+although these are not based on the SMC or the copying algorithm. In fact they represent 
+two extremes along the approximation continuum with \kwarg being entirely 
+parsimony-based while \arginfer is a CwR-based MCMC sampler.
+
+The second group of methods (\tsinfer, \relate, \argneedle) has marked a significant  
+leap in scalability. These methods derive their scalability at least to a certain 
+degree from the fact that the ARGs they infer are approximate structures. In particular, 
+recombination is an emergent property of the inferred ARGs explaining tree topology 
+changes going from one local tree to the next. The recombination events themselves 
+are not explicitly inferred and are thus not part of the resulting data structure.
 
 \subsection{Recent progress in ARGs through approximate structures}
-% onto other features
 
 The computational gains from not having to pinpoint recombination events tie in 
-with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022].
+with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022, ADD REFS].
 
-Always events that cannot be known. Different ways of dealing with this.
-Some approaches have embraced the approximate nature of the resulting ARG to for 
-example highlight the inability to resolve certain splits (polytomies).
+Downsides:
+Point estimate:
+Some approaches have embraced the approximate nature of the resulting ARG by, for 
+example, highlighting the inability to resolve certain splits and instead representing 
+them as polytomies \citep{kelleher_2019_inferring}. 
++
+This means that the degree of 
+completeness in which genetic inheritance from ancestors to descendants is documented 
+by each of these methods varies extensively.
 
-others rigorous treatment of uncertainty (Arbores, ARGinfer): fully precise about recombination.
+But: Advantage of being able to infer ARGs at scale.
 
-% onto ARG cannot be fully known
-Note that approximate structure need not be an issue, the fully precise and detailed 
-ARG can never be known. Instead, the main question is how to connect these 
-approximate structures to existing models.
+Connect this to likelihood approach.
+
 While this is not a problem when describing the results of simulations 
@@
 scale. 
-As discussed in the introduction, the precise details of all 
-events in these vast ARGs can never be known, and a data structure that enforces 
-complete precision is therefore an impediment to progress.
+Only successful if we can connect these to statistical models.
+
+Progress driven by approximation: here similarly
+
+Note that approximate structure need not be an issue, the fully precise and detailed 
+ARG can never be known. Instead, the main question is how to connect these 
+approximate structures to existing models.
 
@@
 
-These methods
-have raised the realistic prospect of ARG-based analysis becoming a standard 
-part of the population and statistical genetics toolkit.
-
-These recent developments have also meant however that
-we are now using the term ARG for a varying range of approximate structures 
-as generated by these different inference methods \citep{wong_a-general_2023}.
-
-Besides not having a unified output format, the degree of completeness 
-in which genetic inheritance from ancestors to 
-descendants is documented by each of these methods varies extensively.
 This has complicated the ability to connect the output of this diverse range of 
@@
 probabilistic exploration of the ARG space around the point estimate, 
-or likelihood-based inference (that does not just consider individual marginal trees).
-% Existing formalisms assume that an ARG is known in full detail
+or likelihood-based inference.
+The problem is that there is no direct way of calculating the coalescent 
+likelihood from a set of marginal genealogies (i.e. integrating over the 
+possible ARGs that could have given the set of marginal genealogies).
 
@@
 \caption{Simulated ARG under Hudson coalescent (left).
-$N_e$ varied was constant within each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 2500, (5000:\infty):10 000]$.
+$N_e$ varied was constant within each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000, (5000:\infty):10 000]$.
 Remaining parameters: L=10kb, 20 diploid samples, r=2.5e-8, 148 edges in ARG).
 A single $N_e$ value was estimated for each slice of the ARG as defined by the true population size change time points.
-The inferred posterior distribution (right) using the SMC-likelihood was obtained using a uniform prior.}
+The inferred posterior distribution (right) using the SMC-likelihood was obtained using a uniform prior.
+Number of edges per time slice: [68, 57, 23].
+Locally unary nodes were omitted for clarity.}
 \label{fig:3-arg-slices}
@@
 Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
-model and an associated set of parameters will help bridge the gap between the tremendous recent 
+model and an associated set of parameters, will help bridge the gap between the tremendous recent 
 advances in ARG inference and the current state of the art statistical inference of 
 past demography. 
-Secondly, inferred ARGs can be used as a good starting point for a future MCMC sampler.
-Thirdly, hybrid ARG-inference methods that combine the best of both 
+Secondly, hybrid ARG-inference methods that combine the best of both 
+Thirdly, inferred ARGs can be used as a good starting point for a future MCMC sampler.
 heuristic and model-based inference approaches are now possible.
@@
 
-% 2/ great starting point for MCMC
-Any inferred ARG can be used a good starting point 
-for a future MCMC-sampler. This potentially implies a massive runtime reduction by 
-reducing the necessary burn-in period. Furthermore, by integrating out the exact time 
-to the recombination event we further restrict the size of the ARG space that needs to 
-be explored. This MCMC-sampler would again rely on the tree sequence data structure to 
-efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
-[TO DO:]
-
-% 3/ hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
+% 2/ hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
 Hybrid backwards-in-time simulations that combine Wright-Fisher dynamics in the 
@@
 
+% 3/ great starting point for MCMC
+Any inferred ARG can be used a good starting point 
+for a future MCMC-sampler. This potentially implies a massive runtime reduction by 
+reducing the necessary burn-in period. Furthermore, by integrating out the exact time 
+to the recombination event we further restrict the size of the ARG space that needs to 
+be explored. This MCMC-sampler would again rely on the tree sequence data structure to 
+efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
+
 % 4/ quantifying uncertainty [unfinished]:
-Both of these future directions tie in with the desired ability of any ARG-related method to 
-quantify the uncertainty associated with ARG inference
-
-This will in particular enable us to deal with those 
-parts in both time and along the genome for which we have insufficient information to confidently 
-reconstruct the ARG. The presence of polytomies in ARGs inferred by \tsinfer, for example, 
-represents uncertainty. Systematicallty breaking them by sampling from the coalescent rather 
-then randomly would already be a major step forward.
-
-Quantifying uncertainty would not necessarily always have to be associated 
-with sampling 
-with dealing with the entire subset of all compatible genealogies. Uncertainty could also 
-be quantified and passed on as metadata to a parent node representing the uncertainty 
-in the interval spanned by the polytomy.
+Acknowledging the uncertainty associated with ARG inference does not 
+necessarily require a full-blown MCMC sampler. Instead, recognising those parts of 
+the ARG for which we lack information for confident reconstruction and resolving 
+them stochastically by sampling from the coalescent rather than completely arbitrarily 
+would already be a major step forward.
+Furthermore, quantifying uncertainty would not necessarily always have to be associated 
+with sampling from subset of all compatible genealogies. Uncertainty could also 
+be quantified and passed on as metadata to the nodes in the data format storing the ARG.
 % can something similar be done for stacked recombinations?
 
-[TO DO] Comment on accuracy. The described algorithm has been implemented in such a way that 
-any valid tree sequence will result in a likelihood value. This also holds for any tree sequence 
-that does not meet the assumptions. We see it as future work to determine to what extent 
-particular deviations from these assumptions will affect the shape of the likelihood surface given 
-a set of parameters.
+As a final remark, the described algorithm relies heavily on the presence of locally unary 
+nodes in the tree sequence to correctly identify the full extent of all lineages going back 
+through time. Although their presence is crucial to our ability embed any local tree in the 
+ARG, the concept and their importance are still relatively new (but see 
+\citet{Wong-2023}). Although we have briefly touched upon the observation that in 
+the presence of unary nodes we seem to augment the left-to-right logic of the SMC 
+with some additional information. 
+We believe that the topic of locally unary nodes in general requires further exploration. 
+And more specifically, we see it as future work to fully quantify to what extent the 
+absence of these locally unary nodes will affect the shape of the likelihood surface given 
+any set of parameters.\\
 
-% final conclusion
+[ TO DO final conclusion]
 

### 2023-12-13 paper.tex
update intro

@@
 
-[TL:DR] Structure the introduction around this great point from Wong et al.:
+[inspiration] great point from Wong et al.:
 Note that it is important to distinguish here between structures and models: 
@@
 
-\subsection{ARGs are awesome, or general intro about approximations}
+[TL;DR]: way to think about genealogies is tied to the CwR. However,
+CwR does not enable us to simply compute a likelihood in the absence of a fully 
+detailed history. In the past, this has given rise to approximations to the 
+coalescent that had similar properties yet were significantly easier to use 
+as a basis for inference. 
+A similar process has taken place in ARG-based inference, 
+where scalable ARG inference has been made possible due to both these approximations 
+to the coalescent as well as the idea of ARGs without explicit, fully detailed 
+recombination histories. What is left however is to connect the two.]
 
+%% ARGs are awesome, or general intro about approximations
+
+[FIX ME: start is a mess. Need to find the right way in.]
 Knowledge of the exact genealogy of a set of samples would make certain problems 
@@
 Yet even though potential of ARGs in population and statistical genetics has been 
-demonstrated it remains unclear how to connect these reconstructed ARGs to 
+demonstrated it remains unclear how to connect these reconstructed ARGs 
+(as being approximate structures) to 
 the stochastic/generative models that have lead us to study these datastructures 
 in the first place.
-
-[Message needs to be here: way to think about genealogies is tied to the CwR. However,
-CwR does not enable us to simply compute a likelihood in the absence of a fully 
-detailed history. The way forward instead has always been driven by approximations.]
-
-% The coalescent with recombination 
-% Kingman, 1982a,b; Tajima, 1983; Hudson, 1983) is generally used as 
-% the generative model that allows us to understand the genealogy 
-% underlying a particular sampleset. 
-% Datastructure that captures it: ARG.
 
@@
 
-% There is a discrepancy between ARGs as object of theoretical reasoning 
-% vs what can be known. Explain the difference between precision/detail and accuracy. 
-% High precision can give a false sense of certainty in the provided estimate.
-
-% In population genetics, the questions we try to answer are generally 
-% about the unobserved processes that gave rise to the genetic variation 
-% we can observe given a set of samples.
-
-\subsection{approximations have driven the rate of progress in inference}
-
+%% approximations have driven the rate of progress in inference
 Instead, approximations of the coalescence with recombination (CwR) have  
@@
 the use of the well-known efficient HMM-algorithms.
-%Both: 1/ Left to right, and 2/ discetization 
-%of state space by emumerating tree topologies and/or discretising time.\\
 
@@
 
+%% Recent progress in ARGs through approximate structures}
+
 The second group of methods (\tsinfer, \relate, \argneedle) has marked a significant  
@@
 degree from the fact that the ARGs they infer are approximate structures. In particular, 
-recombination is an emergent property of the inferred ARGs explaining tree topology 
+recombination is an emergent property of the inferred ARGs, explaining tree topology 
 changes going from one local tree to the next. The recombination events themselves 
@@
 
-\subsection{Recent progress in ARGs through approximate structures}
-
 The computational gains from not having to pinpoint recombination events tie in 
 with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022, ADD REFS]. 
-
-Downsides:
-Point estimate:
 Some approaches have embraced the approximate nature of the resulting ARG by, for 
 example, highlighting the inability to resolve certain splits and instead representing 
-them as polytomies \citep{kelleher_2019_inferring}. 
-+
-This means that the degree of 
-completeness in which genetic inheritance from ancestors to descendants is documented 
+them as polytomies \citep{kelleher_2019_inferring}. Taken together this means that the 
+degree of completeness in which genetic inheritance from ancestors to descendants is documented 
 by each of these methods varies extensively.
+[TO ADD: these methods only provide point estimates]
 
-But: Advantage of being able to infer ARGs at scale.
-
-Connect this to likelihood approach.
-
-While this is not a problem when describing the results of simulations 
-(where all details are perfectly known), it is a major issue when we wish to formally 
-describe the output of inference methods, particularly as datasets approach the population 
-scale. 
-Only successful if we can connect these to statistical models.
-
-Progress driven by approximation: here similarly
-
-Note that approximate structure need not be an issue, the fully precise and detailed 
-ARG can never be known. Instead, the main question is how to connect these 
-approximate structures to existing models.
-
-\subsection{The Big problem: how do you connect such approximate structures to a generative model}
+%% The Big problem: how do you connect such approximate structures to a generative model
 
 This has complicated the ability to connect the output of this diverse range of 
-ARG inference methods to a model-based deescription that would allow us to perform  
+ARG inference methods to a model-based description that would allow us to perform  
 probabilistic exploration of the ARG space around the point estimate, 
 or likelihood-based inference.
-The problem is that there is no direct way of calculating the coalescent 
-likelihood from a set of marginal genealogies (i.e. integrating over the 
-possible ARGs that could have given the set of marginal genealogies).
 
-\subsection{what we will do}
-
-Here, the idea is to use the backwards-in-time formulation of the SMC to 
-compute likelihoods for for \textit{any} inferred ARG 
+%% what we will do
+Here, the idea is to bridge this gap and use the backwards-in-time formulation of the SMC 
+to compute likelihoods for for \textit{any} inferred ARG 
 by integrating out the time to recombination.
@@
 
-[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.\\
+[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.
+Is there more to say them the simple remark I made in \ref{fig:smc-unary}?\\
 
@@
 the inherent ability of ARGs to provide us with temporal 
-resolution on the inferred genealogical history of a sample. 
+resolution on the inferred genealogical history of a sample (see \ref{fig:3-arg-slices}). 
 We can further readily incorporate changes in $N_e$ and $r$, 
@@
 
-[TO DO]: improve results sections with some good viz,
+[TO DO]: improve the (scatter) plots that are referred to in this section such that not 
+all questions will be about why likelihood values do not exactly correspond. How to 
+compare these likelihoods in a more meaningful way?
 We simulated 1000 ARGs under the Hudson coalescent retaining all recombination information 
@@
 
-[ TO DO final conclusion]
+[ TO DO final conclusion?]
 

### 2023-12-13 paper.tex
fix references

@@
 \newcommand{\kwarg}[0]{\texttt{KwARG}}
-\newcommand{\argneelde}[0]{\texttt{ARG-Needle}}
+\newcommand{\argneedle}[0]{\texttt{ARG-Needle}}
 \newcommand{\argweaver}[0]{\texttt{ARGweaver}}
@@
 [potentially with very different likelihoods under the CwR]
-\citep{mcvean_2005_approximating}.
+\citep{mcvean_approximating_2005}.
 
@@
 provided an approximation of the CwR that is both Markovian backwards in time 
-as well as left-to-right along the genome \citep{mcvean_2005_approximating}. 
+as well as left-to-right along the genome \citep{mcvean_approximating_2005}. 
 By sacrificing precision on our ability to capture long range linkage 
@@
 of a single genome to a larger set of genomes as the result of an 
-incomplete copying process \citep{li_2003_modeling}. This heuristic approach 
+incomplete copying process \citep{li_modeling_2003}. This heuristic approach 
 has been successfully applied for demographic inference [and other things: ADD] 
-\citep{sheehan_2013_estimating, steinrucken_2018}.
+\citep{sheehan_estimating_2013, steinrucken_inference_2019}.
 
 Both approaches have further formed the basis (of many) of the recent advancements in 
-ARG inference methods [Rasmussen et al. 2014, Heine et al. 2018, 
-Kelleher et al. 2019, Speidel et al. 2019,
-Rasmussen et al. 2022, Zhang et al.2023]. 
+ARG inference methods \citep{rasmussen_genome-wide_2014, heine_bridging_2018, kelleher_inferring_2019, speidel_method_2019, rasmussen_espalier_2022, zhang_biobank-scale_2023}. 
 A further classification of these methods could be made based on the level of detail 
@@
 that have allowed them to extend their approach to larger sample sizes. We should also 
-mention \kwarg and \arginfer in this category of methods producing fully detailed ARGs 
+mention \kwarg and \texttt{ARGinfer} in this category of methods producing fully detailed ARGs 
 although these are not based on the SMC or the copying algorithm. In fact they represent 
 two extremes along the approximation continuum with \kwarg being entirely 
-parsimony-based while \arginfer is a CwR-based MCMC sampler.
+parsimony-based while \texttt{ARGinfer} is a CwR-based MCMC sampler.
 
@@
 example, highlighting the inability to resolve certain splits and instead representing 
-them as polytomies \citep{kelleher_2019_inferring}. Taken together this means that the 
+them as polytomies \citep{kelleher_inferring_2019}. Taken together this means that the 
 degree of completeness in which genetic inheritance from ancestors to descendants is documented 
@@
 As highlighted, this does not need to be a contradiction. 
-All ARGs can be represented as a series of local trees \citep{wong-2023}.
+All ARGs can be represented as a series of local trees \citep{wong_general_2023}.
 The important question is therefore is however about the extent by which any 
@@
 ARG, the concept and their importance are still relatively new (but see 
-\citet{Wong-2023}). Although we have briefly touched upon the observation that in 
+\citet{wong_general_2023}). Although we have briefly touched upon the observation that in 
 the presence of unary nodes we seem to augment the left-to-right logic of the SMC 

### 2023-12-18 paper.tex
update time slice fig

@@
 
-[FIX ME: start is a mess. Need to find the right way in.]
+[FIX ME: introduction is very much work in progress.
+start is a mess. Need to find the right way in.]
 Knowledge of the exact genealogy of a set of samples would make certain problems 
 in population genetics (e.g.\ inferring recombination rates) trivial.
-Multitude of methods have started to 
-demonstrate the value of attempts to capture the details of these intricately 
-interwoven paths of inheritance referred to as an ancestral recombination graph (ARG).
-Recent advancements: the inference of ARGs at population scale.
-Yet even though potential of ARGs in population and statistical genetics has been 
-demonstrated it remains unclear how to connect these reconstructed ARGs 
-(as being approximate structures) to 
-the stochastic/generative models that have lead us to study these datastructures 
-in the first place.
+Recent advancements in large-scale inference methods of the intricately 
+interwoven paths of inheritance referred to as ancestral recombination graphs (ARGs)
+\citep{rasmussen_genome-wide_2014, heine_bridging_2018, 
+kelleher_inferring_2019, speidel_method_2019, rasmussen_espalier_2022, zhang_biobank-scale_2023}, 
+are revolutionising the way we look at large genomic datasets.
+
+Quite a number of studies have started to demonstrate the value of the recent ability to 
+infer ARGs at scale by emphasizing the ability to rephrase many fundamental problems 
+in genomics as questions on the transmission of genetic material from generation 
+to generation [REFS].
+Yet even though the potential of ARGs in population and statistical genetics has 
+been demonstrated, the fact that these inferred ARGs lack explicit detail on 
+recombination events means that it remains unclear how to connect these 
+reconstructed ARGs to the stochastic/generative models that have lead us 
+to study these datastructures in the first place.
 
@@
 Both approaches have further formed the basis (of many) of the recent advancements in 
-ARG inference methods \citep{rasmussen_genome-wide_2014, heine_bridging_2018, kelleher_inferring_2019, speidel_method_2019, rasmussen_espalier_2022, zhang_biobank-scale_2023}. 
+ARG inference methods \citep{rasmussen_genome-wide_2014, heine_bridging_2018, 
+kelleher_inferring_2019, speidel_method_2019, rasmussen_espalier_2022, zhang_biobank-scale_2023}. 
 A further classification of these methods could be made based on the level of detail 
@@
 \centering
-\includegraphics[width=\textwidth]{figures/3-slices.png}
+\includegraphics[width=\textwidth]{figures/3-slices-whitespace.png}
 \caption{Simulated ARG under Hudson coalescent (left).

### 2024-03-14 paper.tex
Updates from Overleaf

@@
 \usepackage{placeins}
+\usepackage[hidelinks]{hyperref}
 
@@
 \newcommand{\tskit}[0]{\texttt{tskit}}
+\newcommand{\comment}[1]{{\it \color{orange} (#1)}}
 
@@
 }
+
+% adjust inter-paragraph spacing - this one's for you, Gertjan
+\setlength{\parskip}{0.5\baselineskip}
 
@@
 \section{Introduction}
+\begin{itemize}
+    \item ARGs are cool: There are now lots of ways to infer ARGs. Large scale either using heuristics or approximations.
+    \item Models are good/essential: CwR useful but terrible for inference -> SMC
+    \item Problem: we can't use models on inferred ARGs (Relate, tsinfer, ARGneedle). These methods produce point estimates. What we can't do:
+        \begin{itemize}
+            \item quantify uncertainty
+            \item explore ARG space systematically around the point estimates. We don't have a metric on what an improvement might be.
+            \item compare models of demography
+            \item can't evaluate goodness of fit of model estimates
+        \end{itemize}
+    \item what we do here
+\end{itemize}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Methods and Results}
+
+First, we describe in more detail what sort of information we have in estimated ARGs.
+Then, we define the SMC process in a ``backwards-in-time'' way,
+rather than the usual ``along-the-genome'' way.
+This makes it relatively straightforward to compute the likelihood,
+as we explain in the next section.
+Finally, we close with a demonstration that the likelihood computation can be used
+to do Bayesian inference on ancestral population sizes.
+
+% % % % % % % % % % % % % % % % % % 
 \subsection{SMC backwards-in-time}\label{par:description}
 
-Formulated backwards-in-time, the SMC \citep{mcvean_approximating_2005} only requires a 
-simple modification to the 
-coalescent with recombination. By only considering those pairs of 
-lineages that share 
-genomic intervals with ancestral material, as eligible for coalescence, we obtain the 
-sequential Markovian structure along the chromosome.\\
+[TO DO: stress fact that this description does not add anything to McVean and Cardin although the SMC is mostly reasoned about in a left to right sense. We use the notation of McVean and Cardin here.]
+In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a 
+simple modification to the coalescent with recombination.
+Instead of allowing any pair of lineages to coalesce,
+the SMC is the process in which only lineage with \emph{overlapping} genomic intervals may coalesce.
+Surprisingly, although this is defined in terms of a backwards-in-time Markov process,
+the resulting tree-valued process along the genome is also Markov.
 
-More formally, and following the notation of \citet{mcvean_approximating_2005}, at any 
-point in time, the process can be described by $L(t)$, the set of labeled lineages 
-extant at time $t$ each represented by a union of ordered non-overlapping half open 
-intervals detailing the ancestral material 
-carried by that lineage $X_i = \{[x_{i0}, y_{i0}), \dotsc, [x_{im}, y_{im})\}$.
-$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$, each represented by a 
-single interval spanning the entire genome.
-Backwards in time the process evolves by a succession of coalescence and recombination 
+We now define the process more formally,
+and following the notation of \citet{mcvean_approximating_2005}.
+At any point in time, the state of the process is $L(t)$,
+the set of labeled lineages extant at time $t$,
+where each ``lineage'' is represented by a union of non-overlapping half-open intervals
+which represent the ancestral material carried by that lineage,
+and labels are integers.
+So, we can write $L(t) = \{X_i(t)\}_{i=1}^{n(t)}$,
+where the $i^\text{th}$ lineage is $X_i(t)$
+and is of the form
+$X_i(t) = ([x_{i0}, y_{i0}), \dotsc, [x_{im_i}, y_{im_i}))$
+for some interval endpoints $x_{i0} < y_{i0} < x_{i1} < \cdot < y_{im_i}$.
+The \emph{span} of this lineage is the interval $s(X_i(t)) = [x_{i0}, y_{im_i}$,
+and its \emph{size} is $|X_i(t)| = y_{im_i} - x_{i0}$.
+Since we look backwards in time, $t=0$ is today, and so
+$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
+each represented by a single interval spanning the entire genome.
+Then, the process evolves by a succession of coalescence and recombination 
 events until each segment of ancestral material is only present in one lineage. 
 The waiting time to next event is determined by these two competing processes 
-with exponentially distributed waiting times as outlined below.\\
+with exponentially distributed waiting times as outlined below.
 
-Recombination is described by a Poisson process of rate $r$ per unit of length and time. 
-A recombination to the 
-left of $x$ with $x>x_{i0}$ and $x<y_{im}$ splits a lineage $i$ into two new, uniquely 
-labelled lineages. \\ % old lineage is removed from L, two new lineages are added
+Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
+so, if $T(t) = \sum_i |X_i(t)|$ is the sum of the total sizes of all lineages,
+then recombination occurs at instaneous rate $rT(t)$.
+A recombination to the left of $x$ 
+(i.e., between $x$ and $x-1$)
+with $x>x_{i0}$ and $x<y_{im}$ splits lineage $i$ into two new lineages,
+which each are given new, unique labels
+(and lineage $i$ is removed).
+This operation keeps the total amount of ancestral material unchanged.
 
-Any two \emph{overlapping} lineages coalesce at rate $\lambda = 1/(2N_e)$ in the case
-of a diploid population of constant size. The newly formed lineage acquires the 
-union of both ordered intervals. % L is updated accordingly
-Although coalescence is reciprocal, here we'll define a strict total order 
-(see \ref{par:liks}) on 
-all lineages in $L(t)$ based on their leftmost starting points. 
-For any such strict total order on $X(t)$ and at any point in the process 
-the instantaneous rate of coalescence then equals $\lambda \sum_{i \neq j} I_{ij}$.
-
+Coalescence occurs between
+two lineages with overlapping spans at rate $\lambda = 1/(2N_e)$
+(if we assume a randomly mating, diploid population of constant size $N_e$).
+The newly formed lineage acquires the union of both ordered intervals, % L is updated accordingly
+and the original lineages are removed.
+Although coalescence is reciprocal, here we'll define a strict total order on lineages
+(see \ref{par:liks}) on all lineages in $L(t)$ based on their leftmost starting points.
+For any such strict total order,
+the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
+i.e., $\lambda \sum_{i \neq j} I_{ij}$,
+where
 \begin{equation} \label{def:coal}
 I_{ij} = \begin{cases}
-1 & X_i > X_j \wedge X_i \cap X_j \neq \emptyset \\
-0 & \text{otherwise}
+    1 & X_i > X_j \text{ and } s(X_i) \cap s(X_j) \neq \emptyset \\
+    0 & \text{otherwise.}
 \end{cases} 
 \end{equation}
+We say that lineage $j$ can coalesce \emph{into} lineaege $i$ if $I_{ij} = 1$
+(and notice that if $I_{ij} = 1$ then $I_{ji} = 0$).
 
-We can now define $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$ such that $|C_i(t)| = 
-I_{i}(t) = \sum_{j} I_{ij}(t)$.
-Because $I_{ij}$ is not reciprocal, each extant lineage has its own unique set of 
-lineages $C_i(t)$ it can coalesce with. This will simplify the likelihood computations 
-(see \ref{par:liks}). Note that although recombination affects $C_i(t)$ it does not 
-affect $I_{i}$.\\
+For future use, define $C_i(t)$ to be the set of lineages that can coalesce into lineage $i$,
+i.e., $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$.
+(Therefore, $|C_i(t)| = I_{i}(t) = \sum_{j} I_{ij}(t)$.)
+Because $I_{ij}$ is defined in terms of a total order, the sets $C_i(t)$ are disjoint,
+which will simplify the likelihood computations.
+In particular,
+although recombination affects \emph{which} lineages can coalesce into $i$,
+it does not change the \emph{number} of such lineages:
+in other words, a recombination event changes $C_i(t)$ but not $I_{i}$.
+Another useful quantity to know is the number of lineages carrying material
+at each point on the genome:
+for a location $z$ and time $t$, this is
+\begin{equation}
+    N(t,z) = \#\{i : x \le z < y \text{ for some } [x,y) \in X_i(t) \} .
+\end{equation}
+Note that here we are not counting the ``gaps'' in a lineage,
+so that $N(t, z)$ might be less than the number of $i$ with $z \in s(X_i(t))$.
+
+The recombination and coalescence operations described above
+either split a lineage or merge two overlapping lineages,
+and so as described so far, each lineage will consist of only a single interval,
+rather than a collection of intervals.
+However, there is one more operation:
+when a sub-segment of a lineage is ancestral to all samples,
+that sub-segment is removed.
+Formally, define $Z(t)$ to be the segments on which the lineages have totally coalesced:
+i.e., $Z(t) = \{z : N(t,z) = 1$.
+Then, each coalescence event replaces the resulting lineage $X_i(t)$
+by $X_i(t) \setminus Z(t)$,
+the set of intervals that results from removing $Z(t)$ from all the intervals in $X_i(t)$.
 
 % note on the disjoint set of intervals for each lineage under the SMC
-As noted by \cite{mcvean_approximating_2005}, the backwards-in-time description of the 
-SMC is slightly different from the left-to-right approach in that recombination 
-in non-ancestral material is possible (see fig.\ref{fig:smc-unary}). 
-Although this does not affect the distribution 
-of marginal genealogies, it does mean that not all lineages in $L$ can be represented 
-by a single interval. More precisely, all lineages will be associated with 
+Because of this last step, a single lineage can consist of more than one disjoint interval.
+As noted by \cite{mcvean_approximating_2005}, 
+this implies that the backwards-in-time description of the 
+SMC is slightly different from the left-to-right description,
+in which single ancestors only ever span a single segment of genome.
+This is depicted in fig.\ref{fig:smc-unary}:
+for instance, node 8 is ancestral to all samples on the middle two trees.
+However, this does not affect the distribution 
+of marginal genealogies up to the roots.
+More precisely, all lineages will be associated with 
 a single interval up to the point where one section of the 
@@
 
-\subsection{Recording an ARG} \label{par:recording}
+% % % % % % % % % % % % % % % % % % 
+\subsection{Estimated ARGs} \label{par:recording}
 
-The process described in the previous paragraph details the transitions of the 
-set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
-$\mathcal{G} = (N, E)$ capturing the various inheritance paths along the genome 
-is obtained by registering an edge represented by a tuple $(c, p, X_{cp})$ 
-whenever a lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
-replaced by one or two lineage(s), $p$ (and $p^{\prime}$), 
-in case of a coalescence or recombination respectively.
-The last element of the tuple is the extent of the edge. It is itself a  
-disjoint set of intervals such that $X_{cp} \cup X_{cp^{\prime}} = X_c$. 
-The set of edges $E$, containing all such tuples $(c, p, X_{cp})$, 
-and $N$, the set of nodes representing all genomes in the ARG, together
-define this data structure.\\
+\comment{clarify what's different to McVean \& Cardin: unary nodes, maybe? and different to usual SMC stuff more generally}
 
-The ARG as described here records the impact of every possible event along each 
-inheritance path in full detail. Fig.\ref{fig:smc-unary} shows one such possible 
-ARG generated represented 
-as both a graph (top) as well as a series of local trees (bottom). 
-Note that both structures represent the same information. 
+Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
+in the generic sense, it is a labeled graph structure that records
+the inheritance of genetic material \citep{wong_general_2023}.
+An ARG of the sort we consider describes the history of inheritance
+of the genomic material of a focal set of individuals (that we call \emph{samples}).
+Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
+of the form ``$c$ inherited from $p$ on the genomic segment from $\ell$ to $r$'',
+where $c$ and $p$ are (haploid) genomes, and $\ell$ and $r$ are genetic coordinates.
+By ``describes the history'',
+we mean that each node in the ARG represents a specific (ancestral) genome,
+and that the ARG specifies the genealogical tree describing how the samples are related
+at each point on the genome.
+
+% % I don't think we need the notation here:
+% The process described in the previous paragraph details the transitions of the 
+% set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
+% $\mathcal{G} = (N, E)$ capturing the various inheritance paths along the genome 
+% is obtained by registering an edge represented by a tuple $(c, p, X_{cp})$ 
+% whenever a lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
+% replaced by one or two lineage(s), $p$ (and $p^{\prime}$), 
+% in case of a coalescence or recombination respectively.
+% The last element of the tuple is the extent of the edge. It is itself a  
+% disjoint set of intervals such that $X_{cp} \cup X_{cp^{\prime}} = X_c$. 
+% The set of edges $E$, containing all such tuples $(c, p, X_{cp})$, 
+% and $N$, the set of nodes representing all genomes in the ARG, together
+% define this data structure.
+
+So, an ARG can be thought of either as a collection of inheritance relationships
+between ancestral haplotypes,
+or as a sequence of genealogical trees that have common node labels.
+Fig.\ref{fig:smc-unary} shows a small ARG
+that describes relationships between four samples on a 100bp piece of genome,
+represented as both a graph (top) and a series of local trees (bottom). 
+Both structures represent the same information:
+for instance, the fact that the ancestral genome at $a$
+inherited the left half of its genome from ancestor 6
+and the right half from ancestor 5
+is seen in the fact that the first two trees have $a$ below 6,
+and the remaining have $a$ below 5.
+The blue dotted line from $a$ in the second tree to 5 in the third tree
+draws our attention to the fact that
+the path traced by $a$'s ancestry switches at that point.
+In fact, $a$ is in blue because it is a \emph{recombination} node --
+\comment{wait is it ``recombination node'' or ``recombinant node''?}
+an ancestor that has a segment of genome inherited by a sample,
+but that segment was inherited in two pieces from their parent.
+Other nodes (labeled by black numbers) are either samples
+or \emph{coalescence} nodes --
+ancestors that are the most recent common ancestor for two samples
+at some place along the genome.
+
 Crucial for the equivalence between both representations is the presence of
-nodes that (locally) only have a single child along one or more local 
-trees. Not all of these (locally) unary nodes are 
-(blue, fig.\ref{fig:smc-unary}) recombinant nodes. By recording edges along 
-the full extent of each lineage, coalescing nodes can be locally unary 
-along particular genome intervals. Furthermore, the unary nodes enable 
-us to uniquely identify all lineages that were hit by a recombination 
-event, even when we simplify the ARG by removing all 
-(blue, fig.\ref{fig:smc-unary}) recombinant nodes. 
-In the absence of recombination nodes, a recombination can be observed 
-easily enough on the local trees as a change in parent going from 
-one tree to the next. This simplification step 
-does mean however that we lose precise time information on the 
-recombination events. Yet we can constrain the uncertainty to 
-within a window defined by the age of the child node $c$ associated 
-with that lineage and the minimum age of the parents of the 
-edges connecting to $c$. This in turn is easier to visualise using the 
-graph representation (blue shaded area, fig.\ref{fig:smc-unary}).\\
+nodes that only have a single child along one or more local trees --
+i.e., are \emph{locally unary}.
+Recombination nodes (in blue) have a single child in all trees.
+However, coalescent nodes can be locally unary as well:
+for instance, sample 3 inherited a longer stretch of genome from ancestor 5
+than did sample 2.
+In the top plot, this is visible because the pink bar below node 5
+extends further to the left than does the blue bar,
+while in the bottom plot, this is visible because node 5 is parent to 3
+in the second tree, but not to node 2, and so is unary.
 
-ARG simplification is thus linked to the presence of unary nodes in 
-the local trees and simultaneously touches upon the question of what is 
-knowable. Many ARG reconstruction methods return a data structure that 
-is in essence a series of local tries [REFs]. 
-As highlighted, this does not need to be a contradiction. 
-All ARGs can be represented as a series of local trees \citep{wong_general_2023}.
-The important question is therefore is however about the extent by which any 
-ARG reconstruction methods incorporates the available information 
-on how nodes persist across these local trees.\\
+A key observation of \citet{mcvean_approximating_2005}
+was that the sequence of trees along the genome
+generated by the SMC was Markovian \emph{after removing all unary nodes}.
+In other words, if we remove recombination nodes entirely,
+and erase the unary portions of coalescent nodes,
+then the distribution of the next tree can be generated only knowing the current tree.
+(To see that the unary nodes makes the process non-Markov,
+consider that a node in a tree might be unary either because it's about to be a coalescent node
+or it already has been:
+looking at the second tree, we know node 5 will be coalescent further down the sequence
+because it was not present in the first tree.
+So, the first tree gives us information about subsequent trees
+that the second tree alone could not.)
+However, removing these unary portions of nodes seems odd
+when one considers the consequences in the haplotype inheritance graph
+of the top figure.
+As pictured, both figures tell us that
+3 inherits the middle bit of the genome from 7 through 5.
+However, if we erase 5 from the second tree,
+does this suggest that 3 inherited a single segment from 7 along two distinct paths?
+The haplotype inheritance graph above makes it clear that this is unlikely:
+it is in fact reasonable to leave in these unary sections.
 
-[TO DO] give a good explanation (using the represented haplotypes 
-in fig. \ref{fig:smc-unary}) on why these methods can infer the extent 
-of coalescing lineages.
-Currently, the following methods 
-return the full this information by default as part of their ARG reconstruction algorithm: 
-\argweaver \citep{rasmussen_genome-wide_2014} 
-\tsinfer \citep{kelleher_inferring_2019}, 
-\kwarg \citep{ignatieva_kwarg_2021}.\\
+Furthermore, these unary nodes enable 
+us to uniquely identify all lineages that were hit by a recombination event,
+even if we simplify the ARG by removing all recombinant nodes
+($a$, $b$, and $c$ in Fig.\ref{fig:smc-unary}). 
+In the absence of these recombination nodes, a recombination can be observed 
+easily enough as a change in parent going from one genomic region to the next.
+Without the recombination nodes we no longer know the precise time of each recombination event,
+but we do still have constraints on the time:
+for instance, we know that $c$ happened somewhere between node 5 and node 7.
+More generally, a recombination node must happen somewhere between the child node
+and the most recent of the parent nodes.
+This is again easier to visualise using the 
+graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
+Although these unary nodes are useful and to some degree clearly inferrable,
+they are not always represented by simulators or ARG reconstruction methods.
+However, to our understanding this is because
+the field is used to thinking in terms of the trees
+rather than in terms of relationships between haplotypes,
+not because of any question of knowability.
+Currently, 
+\argweaver{} \citep{rasmussen_genome-wide_2014} 
+\tsinfer{} \citep{kelleher_inferring_2019}, and
+\kwarg{} \citep{ignatieva_kwarg_2021}
+estimate ARGs including unary portions of coalescence nodes.
+However, it is unclear to what degree these are well-inferred,
+as most evaluation of inferred ARGs thus far has focused on marginal trees
+(e.g., \citet{brandt2022evaluation,kelleher_inferring_2019},
+but see \citet{deng2021distribution})
+and thus completely ignore such unary nodes
+(and their implications for inherited haplotypes).
+% ARG simplification is thus linked to the presence of unary nodes in 
+% the local trees and simultaneously touches upon the question of what is 
+% knowable. Many ARG reconstruction methods return a data structure that 
+% is in essence a series of local trees [REFs]. 
+% As highlighted, this does not need to be a contradiction. 
+% All ARGs can be represented as a series of local trees \citep{wong_general_2023}.
+% The important question is therefore is however about the extent by which any 
+% ARG reconstruction methods incorporates the available information 
+% on how nodes persist across these local trees.
+% 
+% [TO DO] give a good explanation (using the represented haplotypes 
+% in fig. \ref{fig:smc-unary}) on why these methods can infer the extent 
+% of coalescing lineages.
 
-[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.
-Is there more to say them the simple remark I made in \ref{fig:smc-unary}?\\
+\comment{[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.
+Is there more to say them the simple remark I made in \ref{fig:smc-unary}?}
+\comment{PLR: Not that I can think of?}
 
-In the next paragraph we will derive an expression to compute the 
+In the next section we will derive an expression to compute the 
 likelihood for an ARG under the SMC for which we have no information on the 
 time to any of the recombination events. We will then detail an algorithm 
-that allows us to compute this value given an ARG 
-represented as a series of local trees. To do this we will rely on the unary nodes 
+that allows us to efficiently compute this value by iterating over the local trees.
+To do this, we will rely on the unary nodes 
 to identify each lineage involved in a recombination event 
@@
 \includegraphics[width=\textwidth]{figures/smc_custom_2rows_area.png}
-\caption{ARG generated under the SMC and its representation as a 
+\caption{ARG (top) generated under the SMC and its representation as a 
 series of local trees (bottom). To ensure the one-to-one correspondance 
@@
 the parents of the edges connecting to $c$. 
+    \comment{The bits of genome that leave 5 by the left-hand path are present in 7's diagram but missing from 8's and 9's.}
  [TO DO: wanted to use the 
@@
 
+% % % % % % % % % % % % % % % % % % 
 \subsection{Calculating likelihoods} \label{par:liks}
@@
 the likelihood of all other edges, and in particular from the likelihood 
-of $(d, p, X_{dp})$.\\
+of $(d, p, X_{dp})$.
 
@@
 that terminates the edge. The indicator function is used to resolve 
-the non-reciprocal nature of coalescence events as defined in \ref{def:coal}.\\
+the non-reciprocal nature of coalescence events as defined in \ref{def:coal}.
 
@@
 length of non-ancestral material along which a recombination would have resulted  
-in the same observed local sequence of tree changes.\\
+in the same observed local sequence of tree changes.
 
-\begin{figure}[!ht] \label{fig:algo}
+Note that we are only considering the topology and branch length information of any ARG. 
+Extending this expression by adding the likelihood of observing a set of mutations 
+given an ARG is straightforward.
+
+\subsection{Algorithm} \label{par:algo}
+
+\begin{figure}[!ht]
 \centering
@@
 coalesce with. By moving along the genome in this way, each edge gets visited exactly once.}
+\label{fig:algo}
 \end{figure}
-
-Note that we are only considering the topology and branch length information of any ARG. 
-Extending this expression by adding the likelihood of observing a set of mutations 
-given an ARG is straightforward.
-
-\subsection{Tree sequence format} \label{par:algo}
 
@@
 will describe how to (efficiently) extract the information required to associate a
-tree sequence encoding of an ARG with a likelihood under the SMC (see fig. \ref{fig:smc-unary}).\\
+tree sequence encoding of an ARG with a likelihood under the SMC (see fig. \ref{fig:smc-unary}). 
+The tree sequence format allows us to sequentially recover the local trees 
+along the genome very efficiently, and in a way that allows us to reason about 
+the differences between those trees \citep{kelleher_efficient_2016, ralph_efficiently_2020}. 
+The order of the required edge deletions and insertions to move from one local tree $T_1$ to 
+the next, $T_2$ imposes the required strict total ordering 
+(see definition \ref{def:coal}) on all 
+lineages in the graph needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$. 
+Indeed, following the removal of edges that do not persist in $T_2$, 
+we can postulate that the lineage associated with each subsequently added edge 
+can only coalesce with the lineages represented by the subset of the 
+visited edges that make up the (partially) reconstructed local tree $T_2$.
 
-Formally, we represent a tree sequence using a set of tables \citep{kelleher_efficient_2018}. 
-This simple tabular representation of any ARG $\mathcal{G} = (N, E)$, can be 
-augmented with additional tables with information on sequence variation
-using a site and mutation table. The node table stores information on all 
-haplotypes $N$ in the tree sequence. Information about how nodes relate to 
-each other along the genome is defined in a tabular encoding of $E$.
-Each row in this edge table constitutes a tuple \texttt{(left, right, parent, child)}.
-Note that we defined the extent of any edge $\epsilon \in \mathcal{G}$ as the union of 
-disjoint intervals. In the tree sequence encoding however, $\epsilon$ will be represented 
-by a single entry for each contiguous interval. As remarked before, under the SMC, this 
-distinction only matters beyond the root of the genealogy. The distinction between both 
-types of edges will therefore not be made in the remainder of the paper.
-% Although it does requires us to keep track of some additional information to avoid
-% wrongfully assuming a recombination event happened.
-% Similarly, the distinction between the lineage associated with an edge, and the edge 
-% itself can render things very verbose.
-\\
+This ensures we can associate a likelihood 
+with all tree sequences in a single pass over all edges.
+More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks 
+the exact number of extant lineages in each internode interval (see fig \ref{fig:algo}).
+When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along the internode intervals 
+spanned by the time of the child and the parent of all edges that do not persist in $T_2$. 
+For each subsequently inserted edge, we first compute the likelihood by computing the integral 
+in eq \ref{eq:full-lik} using $I$ and then update the counts. We further keep track of 
+the last parent a child node was connected to to detect recombination events.
 
-The likelihood computation as described above can be done with a single pass 
-of the edge table.
-The central problem however is to efficiently maintain the state $I_e(t)$, which counts 
-the number of lineages (the lineage associated with) the focal edge $\epsilon$ can coalesce 
-with at time $t$. 
-The definition of $I_{kl}$ (\ref{def:coal}) is based on the assumption 
-that we can impose a strict total order on all (overlapping) lineages based on their 
-starting point. The tree sequence encoding provides us with a natural order on all lineages.  
-For each local tree with left coordinate $x$, lineages associated with an edge 
-starting at $x$ can only coalesce with the subset of lineages associated with one of the
-edges that make up that local tree. Each of these edges will either already have 
-been present in the previous local tree, or will have the same starting point $x$.
-In \texttt{tskit}, the order in which edges are inserted and removed 
-while moving from one tree to the next, is determined by the index 
-vectors $\textbf{i}$ and $\textbf{o}$ \citep{kelleher_efficient_2016}.
-The edge insertion vector $\textbf{i}$ gives the ordering of edges sorted by left 
-endpoint, and among edges with the same left endpoint, sorted so that edges closer 
-to the root appear later. The edge removal vector $\textbf{o}$ is similar, 
-except gives the ordering of edges by right endpoint and with edges closer to the 
-root appearing sooner. The variables $\texttt{i}$ and 
-$\texttt{o}$ are used to keep track of our position in the edge insertion and edge removal 
-indexes as we move along the tree sequence. This order imposes the 
-strong ordering on all edges as required by the definition of $I_{kl}$.\\
+% There is probably no need to potentially confuse people with this detail
+%Remark: better to describe edges here as edge-intervals, each describing a 
+%single contiguous interval of genome inheritance between a pair of nodes. 
+%Under the SMC and when recording unary nodes all edges can be represented 
+%by a single edge-interval.
 
-% edge diff algorithm
-The size of the subset of lineages in each internode interval each lineage 
-can coalesce with can thus be maintained by keeping track of a single vector $I$ 
-that is decremented or incremented as defined by
-$\textbf{o}[\texttt{o}]$ and $\textbf{i}[\texttt{i}]$, 
-Further keep track of a last-parent array to detect recombination events.
-We can thus compute a likelihood for the tree sequence, while visiting each edge only once.
-
-More concretely, prior to inserting an edge, $I$ represents the number of lineages 
-the lineage associated with 
-that edge can coalesce with. Then insert the edge and update the counts.
-
-To transition from one tree to the next, we first iterate across the indices of all edges 
-in $\textbf{o}$ from its current index up to $\texttt{o}$. For each edge in we decrement $I$ 
-from index \texttt{edge.child} to \texttt{edge.parent}. We subsequently iterate across the 
-edges in $\textbf{i}$ up to index $\texttt{i}$. For each new edge, we first compute the likelihood 
-associated with that edge and then 
-increment $I$ from \texttt{edge.child} to \texttt{edge.parent}.\\
+\comment{we can probably move this to the next section} 
+Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate) 
+through time by intersecting the internode intervals with the rate change intervals. 
+Rate changes along the genome can equally be incorporated (see \ref{sec:slicing}).
 
@@
 
-\subsection{Flexibility / general class of ARGs}
+\subsection{Slicing the ARG} \label{sec:slicing}
 
@@
 Any such subset of the tree sequence is itself a valid tree sequence 
-that can again be used to compute a likelihood with. Our approach therefore strenghtens 
+that can again be used to compute a likelihood with. Our approach therefore strengthens 
 the inherent ability of ARGs to provide us with temporal 
-resolution on the inferred genealogical history of a sample (see \ref{fig:3-arg-slices}). 
+resolution on the inferred genealogical history of a sample. 
 We can further readily incorporate changes in $N_e$ and $r$, 
@@
 or by splitting edges when/wherever the parameter set changes.
-[NOTE: these options have already been implemented in the current implementation]
+Figure \ref{fig:3-arg-slices} demonstrates this idea on a toy example. We simulated an ARG with 
+two population size changes and subsequently estimated the posterior distribution 
+using a uniform prior.
 
@@
 information on recombination events, our algorithm only implicitly assumes their presence.
-The algorithm can therefore also accomodate for 
-those inferred ARGs that lack unary nodes (e.g.\ \relate). 
 Without unary nodes though, our algorithm will detect a recombination event along 
@@
 edges would switch parent nodes. This can be mitigated by limiting the number of 
-'inferrable' recombination events per coalescence event to 1.
+'inferrable' recombination events per coalescence event to 1 (see fig \ref{sup:fig:rec-correction}).
 
-[TO DO]: improve the (scatter) plots that are referred to in this section such that not 
-all questions will be about why likelihood values do not exactly correspond. How to 
-compare these likelihoods in a more meaningful way?
-We simulated 1000 ARGs under the Hudson coalescent retaining all recombination information 
-and computed the corresponding likelihood using \texttt{msprime} \citep{baumdicker_efficient_2021}. 
-After removing the recombination nodes the likelihoods as defined here were computed both with and 
-without the unary nodes. Figure \ref{sup:fig:vs-hudson} shows that the proportion of variation 
-explained is higher when the unary nodes are retained.
-
-% polytomies
-[TO DO: polytomies] The above algorithm treats each edge independently which implies 
-we can deal with polytomies.
-[TO DO: discuss, what if the identity of internal nodes is not preserved across trees]
-[TO DO: recombination]: We do not *explicitly* assume a single recombination breakpoint to have 
-occurred at each breakpoint.
-[TO DO:] On a similar note and more general: we can compute the likelihood for any 
-valid tree sequence, however, future work to verify to what extent/in which scenarios 
-this is actually helpful.
+%We simulated 1000 ARGs under the Hudson coalescent retaining all recombination information and computed the corresponding likelihood using \texttt{msprime} \citep{baumdicker_efficient_2021}.
+After removing the recombination nodes the likelihoods as defined here were computed both with and without the unary nodes.
+Figure \ref{sup:fig:vs-hudson} shows that the proportion of variation explained is higher when the unary nodes are retained.
 
@@
 \section{Discussion}
+
+% a few details to be discussed
+[TO DO: polytomies] The above algorithm treats each edge independently which implies 
+we can deal with polytomies.
+[TO DO: discuss, what if the identity of internal nodes is not preserved across trees]
+[TO DO: recombination]: We do not *explicitly* assume a single recombination breakpoint to have 
+occurred at each breakpoint.
 
@@
 \bibliographystyle{plainnat}
-\bibliography{paper.bib, temp.bib}
+\bibliography{refs.bib,paper.bib}
 

### 2024-03-14 paper.tex
First pass at intro paragraph

@@
 
-\section{Abstract}
+\begin{abstract}
 There currently is no way to connect the output of the diverse
@@
 ARG space around the point estimate, or likelihood-based inference.
+\end{abstract}
 
 \section{Introduction}
+Following recent breakthroughs in simulation and inference methods,
+there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
+to a variety of questions in evolutionary
+biology~\citep{lewanski_era_2024,brandt_promise_2024}.
+ARGs describe the intricately linked paths
+of genetic inheritance resulting from
+recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
+and contain rich detail about ancestral processes.
+% TODO rephrase this a bit, would be a lot of duplication of references
+% in the current formulation.
+Applications using inferred ARGs as input have begun to
+appear~\citep{osmond_estimating_2021,
+fan_genealogical_2022,
+hejase_deep_2022,
+guo_recombination-aware_2022,
+zhang_biobank-scale_2023,
+nowbandegani_extremely_2023,
+ignatieva_distribution_2023,
+fan_likelihood_2023,
+huang_estimating_2024}.
+Additionally, there is a rich suite of tools growing around
+the \texttt{tskit} ARG library~\citep{ralph_efficiently_2020}
+including simulation~\citep{kelleher_efficient_2016,kelleher_efficient_2018,
+adrion_community_2020,terasaki_geonomics_2021,
+baumdicker_efficient_2021,
+korfmann_weak_2022, petr_slendr_2023,tagami_tstrait_2024},
+ARG inference~\citep{kelleher_inferring_2019,speidel_method_2019,
+wohns_unified_2022,mahmoudi_bayesian_2022,
+rasmussen_espalier_2022,
+zhang_biobank-scale_2023,zhan_towards_2023} and analysis
+packages
+% etc
+\citep{nowbandegani_extremely_2023,fan_likelihood_2023}.
+Infererence methods now tend to fall into two classes: those that
+sample from a statistical
+model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022}
+or parsimony~\citep{ignatieva_kwarg_2021} and that return
+a fully precise ARG estimate~\citep{wong_general_2023},
+and those based on
+heuristics~\citep{kelleher_inferring_2019,speidel_method_2019,
+zhang_biobank-scale_2023} that return an more approximate structure.
+Typically, the more rigorous approaches can handle 10-100 samples,
+while the approximate methods can scale to 10s to hundreds of
+thousands of samples. With the explosive grown in dataset
+sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
+all_genomic_2024} there is a pressing need for more statistical
+rigor in large scale ARG inference.
+
 \begin{itemize}
-    \item ARGs are cool: There are now lots of ways to infer ARGs. Large scale either using heuristics or approximations.
     \item Models are good/essential: CwR useful but terrible for inference -> SMC
@@
 \end{itemize}
-
-[inspiration] great point from Wong et al.:
-Note that it is important to distinguish here between structures and models: 
-whether an inference method is based on 
-heuristics or a rigorous mathematical model is orthogonal to the level 
-of detail provided in its estimate. One could heuristically estimate a 
-fully precise ARG, or statistically sample a partial, approximate ARG under 
-a model such as the coalescent.
 
@@
 Currently most/all ARG-based inference methods base their analysis on marginal
-trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
+trees \citep{hejase_summary_2020} [TO DO: add more refs]. The theory and associated algorithm
 provided here can function as a building block of [...]

### 2024-03-14 paper.tex
updates slicing

--- 2024-03-14 paper.tex
@@
 
-First, we describe in more detail what sort of information we have in estimated ARGs.
-Then, we define the SMC process in a ``backwards-in-time'' way,
+First, we define the SMC process in a ``backwards-in-time'' way,
 rather than the usual ``along-the-genome'' way. 
-This makes it relatively straightforward to compute the likelihood,
-as we explain in the next section.
+Note that this is a
+reiteration of the original work of \citet{mcvean_approximating_2005}.
+Then, we describe
+in more detail what sort of information we have in estimated ARGs.
+In the subsequent section we connect the two and
+show how the ``backwards-in-time'' description
+makes it relatively straightforward to compute the likelihood
+for a more general class of estimated ARGs.
 Finally, we close with a demonstration that the likelihood computation can be used
@@
 
-[TO DO: stress fact that this description does not add anything to McVean and Cardin although the SMC is mostly reasoned about in a left to right sense. We use the notation of McVean and Cardin here.]
 In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a
@@
 
-% % I don't think we need the notation here:
-% The process described in the previous paragraph details the transitions of the 
-% set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
-% $\mathcal{G} = (N, E)$ capturing the various inheritance paths along the genome 
-% is obtained by registering an edge represented by a tuple $(c, p, X_{cp})$ 
-% whenever a lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
-% replaced by one or two lineage(s), $p$ (and $p^{\prime}$), 
-% in case of a coalescence or recombination respectively.
-% The last element of the tuple is the extent of the edge. It is itself a  
-% disjoint set of intervals such that $X_{cp} \cup X_{cp^{\prime}} = X_c$. 
-% The set of edges $E$, containing all such tuples $(c, p, X_{cp})$, 
-% and $N$, the set of nodes representing all genomes in the ARG, together
-% define this data structure.
-
 So, an ARG can be thought of either as a collection of inheritance relationships
@@
 (and their implications for inherited haplotypes).
-% ARG simplification is thus linked to the presence of unary nodes in 
-% the local trees and simultaneously touches upon the question of what is 
-% knowable. Many ARG reconstruction methods return a data structure that 
-% is in essence a series of local trees [REFs]. 
-% As highlighted, this does not need to be a contradiction. 
-% All ARGs can be represented as a series of local trees \citep{wong_general_2023}.
-% The important question is therefore is however about the extent by which any 
-% ARG reconstruction methods incorporates the available information 
-% on how nodes persist across these local trees.
-% 
-% [TO DO] give a good explanation (using the represented haplotypes 
-% in fig. \ref{fig:smc-unary}) on why these methods can infer the extent 
-% of coalescing lineages.
-
-\comment{[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.
-Is there more to say them the simple remark I made in \ref{fig:smc-unary}?}
-\comment{PLR: Not that I can think of?}
 
@@
 the parents of the edges connecting to $c$.
-    \comment{The bits of genome that leave 5 by the left-hand path are present in 7's diagram but missing from 8's and 9's.}
- [TO DO: wanted to use the 
-added in haplotype information to explain why we can actually infer the 
-extent of each lineage, and thus create unary nodes in the first place.]
 }
@@
 
-% ADD IN RESULTS
-Figure \ref{sup:fig:vs-argweaver} shows the expected near-perfect correlation 
-between the coalescent prior as returned by \argweaver and the likelihood as computed here. 
-We simulated 1Mb of data for 10 diploid individuals 
-under the SMC using human-like parameters. 100 samples were taking from the posterior every for 
-a total of 1000 MCMC iterations.
+% ADD IN RESULTS: 
+% Figure \ref{sup:fig:vs-argweaver} shows the expected near-perfect correlation
+% between the coalescent prior as returned by \argweaver and the likelihood as computed here.
+% We simulated 1Mb of data for 10 diploid individuals
+% under the SMC using human-like parameters. 100 samples were taking from the posterior every for
+% a total of 1000 MCMC iterations.
 
@@
 % slicing of the ARG
-By explicitly associating a single likelihood with each edge, we can compute a likelihood 
+By explicitly associating a single likelihood with each edge,
+
+\comment{This makes it sound as if tskit is crucial for this observation.}
+we can compute a likelihood
 under the SMC for any valid tree sequence. The \tskit library provides a whole range
@@
 Any such subset of the tree sequence is itself a valid tree sequence
-that can again be used to compute a likelihood with. Our approach therefore strengthens 
+that can again be used to compute a likelihood with.
+
+Our approach therefore strengthens
 the inherent ability of ARGs to provide us with temporal
 resolution on the inferred genealogical history of a sample.
-We can further readily incorporate changes in $N_e$ and $r$, 
-both in time as well as along the genome by respectively adding intervals to $I$ 
-or by splitting edges when/wherever the parameter set changes.
-Figure \ref{fig:3-arg-slices} demonstrates this idea on a toy example. We simulated an ARG with 
-two population size changes and subsequently estimated the posterior distribution 
-using a uniform prior.
 
-Although we rely heavily on unary nodes to provide us with the necessary 
-information on recombination events, our algorithm only implicitly assumes their presence.
-Without unary nodes though, our algorithm will detect a recombination event along 
-both coalescing lineages in case of a (sub)tree height changes along the ARG as both 
-edges would switch parent nodes. This can be mitigated by limiting the number of 
-'inferrable' recombination events per coalescence event to 1 (see fig \ref{sup:fig:rec-correction}).
+Do this independently for any such slice
+Although we can readily incorporate changes in $N_e$ and $r$ 
+into the likelihood computation, both in time as well as along the genome,
+
+slicing of the ARG can be used to optimize the parameter of interest
+for each slice independently.
+
+Figure \ref{fig:3-arg-slices} demonstrates this idea on a toy example.
+We simulated an ARG under human-like parameters and with
+two population size changes and subsequently estimated the posterior 
+distribution using a uniform prior for each slice indepently.
+
+The observation that we can associate a likelihood with any valid tree sequences
+urges the need for some caveats. By treating each edge independently we can
+naturally deal with the various degrees of completeness to which inheritance
+is documented by the different ARG inference methods: e.g.\ polytomies, absence
+of locally unary coalescence nodes, unpreserved identify of internal nodes 
+across marginal trees, \dots However, these imprecisions do violate the implicit
+assumptions of the likelihood computation.
+\comment{Do we want these remarks in here, or as part of the Discussion?}
+For example, the described algorithm relies heavily on the
+presence of locally unary nodes in the tree sequence to correctly identify the 
+full extent of all lineages going back through time and therefore recombination
+events. Without unary nodes though, our algorithm will detect a recombination
+event along both coalescing lineages in case of a (sub)tree height changes along
+the ARG as both edges would switch parent nodes. This can be mitigated by limiting
+the number of 'inferrable' recombination events per coalescence event to 1 
+(see fig \ref{sup:fig:rec-correction}).
+
+\comment{this was previously part of the discussion, have grouped it here.}
+Although their presence is crucial to our ability to embed any local tree in the
+ARG, the concept and their importance are still relatively new (but see
+\citet{wong_general_2023}).
+We have briefly touched upon the observation that in
+the presence of unary nodes we seem to augment the left-to-right logic of the SMC
+with some additional information. Yet, we believe that the topic of locally 
+unary nodes in general requires further exploration.
+And more specifically, we see it as future work to fully quantify to what 
+extent the absence of these locally unary nodes will affect the shape of the
+likelihood surface given any set of parameters.
 
@@
 \caption{Simulated ARG under Hudson coalescent (left).
-$N_e$ varied was constant within each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000, (5000:\infty):10 000]$.
+$N_e$ is constant for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000, (5000:\infty):10 000]$.
 Remaining parameters: L=10kb, 20 diploid samples, r=2.5e-8, 148 edges in ARG).
@@
 The inferred posterior distribution (right) using the SMC-likelihood was obtained using a uniform prior.
-Number of edges per time slice: [68, 57, 23].
+Number of edges per time slice (from most recent slice to oldest): [68, 57, 23].
 Locally unary nodes were omitted for clarity.}
@@
 \section{Discussion}
-
-% a few details to be discussed
-[TO DO: polytomies] The above algorithm treats each edge independently which implies 
-we can deal with polytomies.
-[TO DO: discuss, what if the identity of internal nodes is not preserved across trees]
-[TO DO: recombination]: We do not *explicitly* assume a single recombination breakpoint to have 
-occurred at each breakpoint.
 
@@
 $N_e$-changes as well as multiple populations and migration can be readily incorporated
-in the current algorithm.\\
+in the current algorithm.
 
@@
 could be used for ARG-inference where one can rely on a heuristic inference method for
-the recent past, while using a model-based approach beyond that cutoff point.\\
+the recent past, while using a model-based approach beyond that cutoff point.
 
@@
 % can something similar be done for stacked recombinations?
-
-As a final remark, the described algorithm relies heavily on the presence of locally unary 
-nodes in the tree sequence to correctly identify the full extent of all lineages going back 
-through time. Although their presence is crucial to our ability embed any local tree in the 
-ARG, the concept and their importance are still relatively new (but see 
-\citet{wong_general_2023}). Although we have briefly touched upon the observation that in 
-the presence of unary nodes we seem to augment the left-to-right logic of the SMC 
-with some additional information. 
-We believe that the topic of locally unary nodes in general requires further exploration. 
-And more specifically, we see it as future work to fully quantify to what extent the 
-absence of these locally unary nodes will affect the shape of the likelihood surface given 
-any set of parameters.\\
-
-[ TO DO final conclusion?]
 

### 2024-03-14 paper.tex

@@
 \end{itemize}
-
-[inspiration] great point from Wong et al.:
-Note that it is important to distinguish here between structures and models:
-whether an inference method is based on
-heuristics or a rigorous mathematical model is orthogonal to the level
-of detail provided in its estimate. One could heuristically estimate a
-fully precise ARG, or statistically sample a partial, approximate ARG under
-a model such as the coalescent.
 
@@
 Currently most/all ARG-based inference methods base their analysis on marginal
-trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm
+trees \citep{hejase_summary_2020} [TO DO: add more refs]. The theory and associated algorithm
 provided here can function as a building block of [...]

### 2024-03-14 paper.tex
update

@@
 \centering
-\includegraphics[width=\textwidth]{figures/smc_custom_2rows_area.png}
+\includegraphics[width=\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
 \caption{ARG (top) generated under the SMC and its representation as a
@@
 
+Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate)
+through time by intersecting the internode intervals with the rate change intervals.
+Rate changes along the genome can equally be incorporated (see \ref{sec:slicing}).
+
 % There is probably no need to potentially confuse people with this detail
@@
 %by a single edge-interval.
-
-\comment{we can probably move this to the next section}
-Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate)
-through time by intersecting the internode intervals with the rate change intervals.
-Rate changes along the genome can equally be incorporated (see \ref{sec:slicing}).
 
@@
 By explicitly associating a single likelihood with each edge,
-
-\comment{This makes it sound as if tskit is crucial for this observation.}
-we can compute a likelihood
-under the SMC for any valid tree sequence. The \tskit library provides a whole range
-of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space,
-as well as reduce the number of tracked samples.
-Any such subset of the tree sequence is itself a valid tree sequence
-that can again be used to compute a likelihood with.
-
-Our approach therefore strengthens
+we can compute the likelihood for any slice of the ARG in both
+time as well as along the genome. Our approach therefore strengthens
 the inherent ability of ARGs to provide us with temporal
 resolution on the inferred genealogical history of a sample.
+It also opens up the opportunity for more efficient optimization 
+strategies during inference given that each slice of the ARG can be
+processed independently.
 
-Do this independently for any such slice
-Although we can readily incorporate changes in $N_e$ and $r$ 
-into the likelihood computation, both in time as well as along the genome,
-
-slicing of the ARG can be used to optimize the parameter of interest
-for each slice independently.
-
-Figure \ref{fig:3-arg-slices} demonstrates this idea on a toy example.
-We simulated an ARG under human-like parameters and with
+From a data structure perspective, we can compute a likelihood 
+under the SMC for any valid tree sequence. 
+The \tskit library provides a whole range
+of subsetting operations on tree sequences allowing us to slice
+the ARG in time and/or space, as well as reduce the number of
+tracked samples. Any such subset of the tree sequence is itself
+a valid tree sequence that can again be used to compute a
+likelihood for. Figure \ref{fig:3-arg-slices} demonstrates this idea 
+on a toy example. We simulated an ARG under human-like parameters and with
 two population size changes and subsequently estimated the posterior 
-distribution using a uniform prior for each slice indepently.
+distribution using a uniform prior for each slice independently.
 
@@
 \includegraphics[width=\textwidth]{figures/3-slices-whitespace.png}
-\caption{Simulated ARG under Hudson coalescent (left).
-$N_e$ is constant for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000, (5000:\infty):10 000]$.
-Remaining parameters: L=10kb, 20 diploid samples, r=2.5e-8, 148 edges in ARG).
-A single $N_e$ value was estimated for each slice of the ARG as defined by the true population size change time points.
-The inferred posterior distribution (right) using the SMC-likelihood was obtained using a uniform prior.
-Number of edges per time slice (from most recent slice to oldest): [68, 57, 23].
-Locally unary nodes were omitted for clarity.}
+\caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant
+for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000,
+(5000:\infty):10 000]$. Remaining parameters: $L=10$kb, $20$ diploid samples, 
+$r=2.5e-8$, $148$ edges in ARG). A single $N_e$ value was estimated for each
+slice of the ARG as defined by the true population size change time points.
+The inferred posterior distribution (right) using the SMC-likelihood was
+obtained using a uniform prior. Number of edges per time slice
+(from most recent slice to oldest): $[68, 57, 23]$. Locally unary nodes were
+omitted for clarity.}
 \label{fig:3-arg-slices}

### 2024-03-14 paper.tex
Remove whitespace

@@
 
+Model-based inference is a central element of population genetics.
+The coalescent~\citep{kingman_genealogy_1982,kingman_coalescent_1982,hudson_testing_1983,
+tajima_evolutionary_1983} has proved to be a useful model, and is the
+basis of a wide variety of inferences [some citations?]. The incorporation
+of recombination in to the coalescent process~\citep{hudson_properties_1983}
+has proved to be much more difficult.
+% FIXME rephrase, next two sentences, lifted from ARG paper
+Early work on inference under the coalescent with recombination
+focused on the problem of
+inferring the parameters of the
+stochastic process, where the ancestry was regarded as a
+latent parameter to be averaged out
+\citep[e.g.][]{griffiths_ancestral_1996,kuhner_maximum_2000,
+nielsen_estimation_2000,
+fearnhead_estimating_2001}.
+% TODO reuse some of the ideas from the SMC paper here, it has
+% some good explanations.
+These methods met with limited success
+because the state space of ARGs is overwhelmingly large, and
+lacks a simple geometry or neighbourhood structure for inference or
+sampling methods to  exploit.
+The introduction of the Sequentially Markovian Coalescent (SMC)
+approximation to the coalescent with recombination
+by~\cite{mcvean_approximating_2005} was a seminal contribution,
+and has made many different inferences
+% TODO add some more recent stuff
+feasible~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
+In particular, the SMC is the basis for ARG sampling methods
+like ARGweaver~\citep{rasmussen_genome-wide_2014}, which remains the
+% TODO list other places ARGweaver has come out on top in terms of accuracy.
+gold-standard in terms of accuracy~\citep{brandt2022evaluation}
+
 \begin{itemize}
-    \item Models are good/essential: CwR useful but terrible for inference -> SMC
     \item Problem: we can't use models on inferred ARGs (Relate, tsinfer, ARGneedle). These methods produce point estimates. What we can't do:

### 2024-03-21 paper.tex
Make Peter last author, remove Alison

@@
 \author[1, $\dagger$]{Gertjan Bisschop}
-
-% Middle Authors
-\author[2]{Alison Etheridge}
+\author[1]{Jerome Kelleher}
+% last author
 \author[3]{Peter Ralph}
 
-% last author
-\author[1]{Jerome Kelleher}
 \affil[$\dagger$]{Denotes corresponding author}
@@
 sample from a statistical
+% TODO add SINGER
 model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022}
-or parsimony~\citep{ignatieva_kwarg_2021} and that return
-a fully precise ARG estimate~\citep{wong_general_2023},
+or parsimony~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022},
 and those based on
 heuristics~\citep{kelleher_inferring_2019,speidel_method_2019,
-zhang_biobank-scale_2023} that return an more approximate structure.
+zhang_biobank-scale_2023}.
 Typically, the more rigorous approaches can handle 10-100 samples,
 while the approximate methods can scale to 10s to hundreds of
-thousands of samples. With the explosive grown in dataset
+thousands of samples. With the explosive growth in dataset
 sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
@@
 
-Model-based inference is a central element of population genetics.
+Perhaps the clearest route to statistical rigor is through scalable
+model-based inference, a central element of population genetics.
 The coalescent~\citep{kingman_genealogy_1982,kingman_coalescent_1982,hudson_testing_1983,
 tajima_evolutionary_1983} has proved to be a useful model, and is the
-basis of a wide variety of inferences [some citations?]. The incorporation
-of recombination in to the coalescent process~\citep{hudson_properties_1983}
+basis of a wide variety of inferences [some citations, e.g Hey IM model]. 
+The incorporation
+of recombination in the coalescent process~\citep{hudson_properties_1983}
 has proved to be much more difficult.
@@
 
-\begin{itemize}
-    \item Problem: we can't use models on inferred ARGs (Relate, tsinfer, ARGneedle). These methods produce point estimates. What we can't do:
-        \begin{itemize}
-            \item quantify uncertainty
-            \item explore ARG space systematically around the point estimates. We don't have a metric on what an improvement might be.
-            \item compare models of demography
-            \item can't evaluate goodness of fit of model estimates
-        \end{itemize}
-    \item what we do here
-\end{itemize}
+% THE PROBLEM!!! OMG it's terrible!!!
+Although major breakthroughs in scale have been achieved by hueristic
+ARG inference
+methods~\citep{kelleher_inferring_2019,speidel_method_2019,zhang_biobank-scale_2023},
+these methods are currently fundamentally incompatible with rigorous
+inference under population genetic models such as the SMC. This is 
+because these methods return approximate ARG
+\emph{structures}~\citep{wong_general_2023} [WEAK, FIXME]. Specifically,
+we cannot evaluate a likelihood of these ARGs in coalescent 
+models under the classical 
+\cite{kuhner_maximum_2000} formulation, as these formulations assume
+that complete information about recombination events is present.
+[NOTE the following is super weak, and needs a lot of work.]
+This inability to connect the results of large-scale inference methods
+with a generative model
+is a fundamental limitation, and limits both the utility of 
+current point estimates as well as the possibilities for further
+inference. Specifically, we are unable to quantify uncertainty 
+around heuristic point estimates. We cannot explore ARG space
+around the point estimates using (e.g. MCMC), as we have no way definition
+of improvement. We cannot compare models of demography, or evaluate the 
+goodness of fit for model estimates.
 
-[TL;DR]: way to think about genealogies is tied to the CwR. However,
-CwR does not enable us to simply compute a likelihood in the absence of a fully
-detailed history. In the past, this has given rise to approximations to the
-coalescent that had similar properties yet were significantly easier to use
-as a basis for inference.
-A similar process has taken place in ARG-based inference,
-where scalable ARG inference has been made possible due to both these approximations
-to the coalescent as well as the idea of ARGs without explicit, fully detailed
-recombination histories. What is left however is to connect the two.]
-
-%% ARGs are awesome, or general intro about approximations
-
-[FIX ME: introduction is very much work in progress.
-start is a mess. Need to find the right way in.]
-Knowledge of the exact genealogy of a set of samples would make certain problems
-in population genetics (e.g.\ inferring recombination rates) trivial.
-Recent advancements in large-scale inference methods of the intricately
-interwoven paths of inheritance referred to as ancestral recombination graphs (ARGs)
-\citep{rasmussen_genome-wide_2014, heine_bridging_2018,
-kelleher_inferring_2019, speidel_method_2019, rasmussen_espalier_2022, zhang_biobank-scale_2023},
-are revolutionising the way we look at large genomic datasets.
-
-Quite a number of studies have started to demonstrate the value of the recent ability to
-infer ARGs at scale by emphasizing the ability to rephrase many fundamental problems
-in genomics as questions on the transmission of genetic material from generation
-to generation [REFS].
-Yet even though the potential of ARGs in population and statistical genetics has
-been demonstrated, the fact that these inferred ARGs lack explicit detail on
-recombination events means that it remains unclear how to connect these
-reconstructed ARGs to the stochastic/generative models that have lead us
-to study these datastructures in the first place.
-
-For a very long time people have assumed that what is meant by an ARG is the
-fully specified genealogical history where all events are uniquely positioned
-both in time and along the genome [Wong et al. 2023].
-Yet, the ease by which such a fully detailed history can be simulated under the
-CwR, is not matched by the usability of the model for inference.
-In particular, given an ARG, there will always be an infinite amount of other
-ARGs that could have generated the observed genomic dataset
-[potentially with very different likelihoods under the CwR]
-\citep{mcvean_approximating_2005}.
-
-%% approximations have driven the rate of progress in inference
-Instead, approximations of the coalescence with recombination (CwR) have
-driven our ability to perform statistical genomic inference. Two approximations
-in particular have proven to be instrumental: the sequentially Markovian coalescent,
-and its extensions (SMC) on the one hand, and the Li and Stephens model on the other.
-The impact of both models is tightly linked to the fact that they have enabled
-the reformulation of inference problems as hidden Markov models (HMM) and thus
-the use of the well-known efficient HMM-algorithms.
-
-To improve the ability to simulate the CwR, McVean and Cardin
-provided an approximation of the CwR that is both Markovian backwards in time
-as well as left-to-right along the genome \citep{mcvean_approximating_2005}.
-By sacrificing precision on our ability to capture long range linkage
-information, the SMC became the engine of many powerful inference methods [PSMC].
-A later improvement, the SMC', closed the gap between the SMC and the
-CwR even further.
-
-The second key approximation is not a generative model. Rather than providing
-a mathematical description of the genealogies underlying the observed data
-the Li and Stephens or copying model pragmatically details the relationship
-of a single genome to a larger set of genomes as the result of an
-incomplete copying process \citep{li_modeling_2003}. This heuristic approach
-has been successfully applied for demographic inference [and other things: ADD]
-\citep{sheehan_estimating_2013, steinrucken_inference_2019}.
-
-Both approaches have further formed the basis (of many) of the recent advancements in
-ARG inference methods \citep{rasmussen_genome-wide_2014, heine_bridging_2018,
-kelleher_inferring_2019, speidel_method_2019, rasmussen_espalier_2022, zhang_biobank-scale_2023}.
-A further classification of these methods could be made based on the level of detail
-that is provided on recombination events. The first real breakthrough in large-scale
-ARG inference, \argweaver, heads the first group.
-The ARGs sampled by \argweaver from the posterior distribution
-given the observed variational data, are all fully precise and detailed in the sense
-that all events are explicitly inferred. Note however that the set of possible node times
-has to be limited to enable the usage of an HMM. The other more recent methods, although
-still SMC-based, have added heuristic pre- and/or post-processing steps
-that have allowed them to extend their approach to larger sample sizes. We should also
-mention \kwarg and \texttt{ARGinfer} in this category of methods producing fully detailed ARGs
-although these are not based on the SMC or the copying algorithm. In fact they represent
-two extremes along the approximation continuum with \kwarg being entirely
-parsimony-based while \texttt{ARGinfer} is a CwR-based MCMC sampler.
-
-%% Recent progress in ARGs through approximate structures}
-
-The second group of methods (\tsinfer, \relate, \argneedle) has marked a significant
-leap in scalability. These methods derive their scalability at least to a certain
-degree from the fact that the ARGs they infer are approximate structures. In particular,
-recombination is an emergent property of the inferred ARGs, explaining tree topology
-changes going from one local tree to the next. The recombination events themselves
-are not explicitly inferred and are thus not part of the resulting data structure.
-
-The computational gains from not having to pinpoint recombination events tie in
-with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022, ADD REFS].
-Some approaches have embraced the approximate nature of the resulting ARG by, for
-example, highlighting the inability to resolve certain splits and instead representing
-them as polytomies \citep{kelleher_inferring_2019}. Taken together this means that the
-degree of completeness in which genetic inheritance from ancestors to descendants is documented
-by each of these methods varies extensively.
-[TO ADD: these methods only provide point estimates]
-
-%% The Big problem: how do you connect such approximate structures to a generative model
-
-This has complicated the ability to connect the output of this diverse range of
-ARG inference methods to a model-based description that would allow us to perform
-probabilistic exploration of the ARG space around the point estimate,
-or likelihood-based inference.
-
-%% what we will do
-Here, the idea is to bridge this gap and use the backwards-in-time formulation of the SMC
-to compute likelihoods for for \textit{any} inferred ARG
-by integrating out the time to recombination.
-The application of the SMC in a way that resonates more
-with the true direction of flow of genetic information
-results in a concrete way to formulate an algorithm
-that is \textit{general} in the sense that it can deal with the various levels of
-completeness that characterise the current state of the field of ARG reconstruction.
+%% This is how we solve it
+[TODO figure out the gARG, tree sequence etc mess]
+Here, we bridge this gap between models and the products of approximate
+inference by defining a likelihood function for any 
+``genome ARG''~\citep{wong_general_2023} under the SMC. Using a novel
+formulation of the SMC that focuses on the contributions of nodes and 
+edges [FIXME] in a backwards-in-time way, we provide an efficient 
+algorithm that can compute a likelihood for any gARG under this model.
+We demonstrate the flexibility of this approach by XXX, and show
+[SOMETHING].
 
@@
 
+%% NOTE: this is some orphan text from the intro that can be rejiggered for the 
+% discussion. We can talk about precision of recombination estimates a bit 
+% here I think, as it is an important topic.
+
+% Both approaches have further formed the basis (of many) of the recent advancements in
+% ARG inference methods \citep{rasmussen_genome-wide_2014, heine_bridging_2018,
+% kelleher_inferring_2019, speidel_method_2019, rasmussen_espalier_2022, zhang_biobank-scale_2023}.
+% A further classification of these methods could be made based on the level of detail
+% that is provided on recombination events. The first real breakthrough in large-scale
+% ARG inference, \argweaver, heads the first group.
+% The ARGs sampled by \argweaver from the posterior distribution
+% given the observed variational data, are all fully precise and detailed in the sense
+% that all events are explicitly inferred. Note however that the set of possible node times
+% has to be limited to enable the usage of an HMM. The other more recent methods, although
+% still SMC-based, have added heuristic pre- and/or post-processing steps
+% that have allowed them to extend their approach to larger sample sizes. We should also
+% mention \kwarg and \texttt{ARGinfer} in this category of methods producing fully detailed ARGs
+% although these are not based on the SMC or the copying algorithm. In fact they represent
+% two extremes along the approximation continuum with \kwarg being entirely
+% parsimony-based while \texttt{ARGinfer} is a CwR-based MCMC sampler.
+
+%% Recent progress in ARGs through approximate structures}
+
+% The second group of methods (\tsinfer, \relate, \argneedle) has marked a significant
+% leap in scalability. These methods derive their scalability at least to a certain
+% degree from the fact that the ARGs they infer are approximate structures. In particular,
+% recombination is an emergent property of the inferred ARGs, explaining tree topology
+% changes going from one local tree to the next. The recombination events themselves
+% are not explicitly inferred and are thus not part of the resulting data structure.
+
+% The computational gains from not having to pinpoint recombination events tie in
+% with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022, ADD REFS].
+% Some approaches have embraced the approximate nature of the resulting ARG by, for
+% example, highlighting the inability to resolve certain splits and instead representing
+% them as polytomies \citep{kelleher_inferring_2019}. Taken together this means that the
+% degree of completeness in which genetic inheritance from ancestors to descendants is documented
+% by each of these methods varies extensively.
+% [TO ADD: these methods only provide point estimates]
+
 \section{Data availability}

### 2024-03-21 paper.tex
edits to likelihood section

@@
 \newcommand{\tskit}[0]{\texttt{tskit}}
+\renewcommand{\deg}{\mathrm{deg}}
 \newcommand{\comment}[1]{{\it \color{orange} (#1)}}
@@
 the resulting tree-valued process along the genome is also Markov.
+Although \citet{mcvean_approximating_2005} verbally described the SMC 
+using this backwards-in-time description, their analysis
+was almost entirely in left-to-right terms,
+and subsequent papers followed this trend
+(e.g., \comment{cite}).
 
@@
 the set of labeled lineages extant at time $t$,
-where each ``lineage'' is represented by a union of non-overlapping half-open intervals
-which represent the ancestral material carried by that lineage,
-and labels are integers.
+where each ``lineage'' is represented by a half-open interval
+which represents the ancestral material carried by that lineage,
+and labeled by an integer.
 So, we can write $L(t) = \{X_i(t)\}_{i=1}^{n(t)}$,
@@
 and is of the form
-$X_i(t) = ([x_{i0}, y_{i0}), \dotsc, [x_{im_i}, y_{im_i}))$
-for some interval endpoints $x_{i0} < y_{i0} < x_{i1} < \cdot < y_{im_i}$.
-The \emph{span} of this lineage is the interval $s(X_i(t)) = [x_{i0}, y_{im_i}$,
-and its \emph{size} is $|X_i(t)| = y_{im_i} - x_{i0}$.
+$X_i(t) = [x_{i}, y_{i})$,
+for some interval endpoints $x_{i} < y_{i}$.
+The \emph{size} of the interval is $|X_i(t)| = y_{i} - x_{i}$.
 Since we look backwards in time, $t=0$ is today, and so
@@
 A recombination to the left of $x$
-(i.e., between $x$ and $x-1$)
-with $x>x_{i0}$ and $x<y_{im}$ splits lineage $i$ into two new lineages,
+(i.e., between base pairs $x$ and $x-1$)
+with $x>x_{i}$ and $x<y_{i}$ splits lineage $i$ into two new lineages,
 which each are given new, unique labels
@@
 Coalescence occurs between
-two lineages with overlapping spans at rate $\lambda = 1/(2N_e)$
+two overlapping lineages at rate $\lambda = 1/(2N_e)$
 (if we assume a randomly mating, diploid population of constant size $N_e$).
@@
 i.e., $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$.
-(Therefore, $|C_i(t)| = I_{i}(t) = \sum_{j} I_{ij}(t)$.)
+(Therefore, $|C_i(t)| = I_{i}(t) = \sum_{j} I_{ij}(t)$,)
+and the total rate of coalescence at time $t$ is $\sum_{i} |C_i(t)|$.) 
 Because $I_{ij}$ is defined in terms of a total order, the sets $C_i(t)$ are disjoint,
@@
 \begin{equation}
-    N(t,z) = \#\{i : x \le z < y \text{ for some } [x,y) \in X_i(t) \} .
+    N(t,z) = \#\{i : z \in X_i(t) \} .
 \end{equation}
-Note that here we are not counting the ``gaps'' in a lineage,
-so that $N(t, z)$ might be less than the number of $i$ with $z \in s(X_i(t))$.
 
-The recombination and coalescence operations described above
-either split a lineage or merge two overlapping lineages,
-and so as described so far, each lineage will consist of only a single interval,
-rather than a collection of intervals.
-However, there is one more operation:
-when a sub-segment of a lineage is ancestral to all samples,
-that sub-segment is removed.
-Formally, define $Z(t)$ to be the segments on which the lineages have totally coalesced:
-i.e., $Z(t) = \{z : N(t,z) = 1$.
-Then, each coalescence event replaces the resulting lineage $X_i(t)$
-by $X_i(t) \setminus Z(t)$,
-the set of intervals that results from removing $Z(t)$ from all the intervals in $X_i(t)$.
+Finally, there is one more operation that allows the process to terminate:
+portions of the edges of lineages that have entirely coalesced
+(i.e., that are ancestral to all samples) are removed.
+Formally, after any coalescence or recombination event,
+the segment $[x,y)$ carried by the lineage
+is replaced by the largest segment whose endpoints have not coalesced,
+i.e., $x$ is replaced by $\min\{ z \ge x : N(t,z) > 1\}$,
+and $y$ is replaced by $\sup\{ z \le y : N(t,z) > 1\}$.
+\comment{explain the cases in which this happens: coalescence on one end,
+and recobmbination in the middle of a trapped ancestral segment}
 
 % note on the disjoint set of intervals for each lineage under the SMC
-Because of this last step, a single lineage can consist of more than one disjoint interval.
+\comment{I'm not sure we need this here now?}
+Note that 
+an alternative definition could instead remove any sub-segment of a lineage
+that has entirely coalesced.
+This would produce a process in which each lineage is a collection
+of disjoint intervals (instead of a single one).
 As noted by \cite{mcvean_approximating_2005},
@@
 SMC is slightly different from the left-to-right description,
-in which single ancestors only ever span a single segment of genome.
+which would usually not include these segments.
 This is depicted in fig.\ref{fig:smc-unary}:
 for instance, node 8 is ancestral to all samples on the middle two trees.
-However, this does not affect the distribution
-of marginal genealogies up to the roots.
-More precisely, all lineages will be associated with
-a single interval up to the point where one section of the
-genome has reached its most recent common ancestor but the neighbouring parts have not.
 
@@
 where $c$ and $p$ are (haploid) genomes, and $\ell$ and $r$ are genetic coordinates.
+We record each such statement in an ``edge'',
+often summarized as the tuple $(p,c,\ell,r)$.
 By ``describes the history'',
@@
 
-Retracing the ARG backwards in time, each coalescence event is associated with two child
-lineages, $c$ and $d$, and a single parent, $p$. By defining $C_i(t)$ asymmetrically,
-the history of each these two child lineages can be considered independently without
-risking to double count any of the coalescence events. The likelihood of
-observing the edge $(c, p, X_{cp})$ can thus be considered independently from
-the likelihood of all other edges, and in particular from the likelihood
-of $(d, p, X_{dp})$.
+The process of recombination only affects one lineage at a time
+(and does not even depend on the state of other lineages).
+On the other hand, coalescence events necessarily involve more than one lineage.
+However, the asymmetric way we define coalescence above
+allows us to uniquely associate each event with a single lineage,
+which allows us to cleanly factor the likelihood across edges and nodes.
+The data structure we will compute the likelihood of
+is a collection of \emph{edges} $E=\{e_i\}$,
+where each edge is of the form $e_i = (p_i,c_i,x_i,y_i)$.
+Recall that this means that node $c_i$ inherited genetic material from node $p_i$
+over a region with span from $x_i$ to $y_i$ along the genome
+(see Figure~\ref{fig:likelihood}).
+So, $x_i$ and $y_i$ are genomic coordinates,
+while $c_i$ and $p_i$ are in $N$, the set of \emph{nodes}
+(i.e., distinct sampled or ancestral haplotypes).
 
-% extends from here means min(X_{cp})=x, max(X_{cp})=y
-% improve this description here
-Assuming $(c, p, X_{cp})$ extends from $x$ to $y$ along the genome and $t_i$
-indicates the age of
-node $i$ in the ARG, then the probability of not observing a recombination across the
-entire area (span x depth) of this edge is
+We have defined the SMC process in terms of rates of various types of events.
+In general, the likelihood for a continuous-time Markov process specified in this way
+is $\exp(-\Lambda) \prod_i \lambda_i$,
+where $\lambda_i$ is the rate of the $i^\text{th}$ realized event,
+and $\Lambda$ is the sum of all rates of all possible events
+integrated over the entire process
+(also called the ``total hazard'').
 
-% full extent of segment
+The contributions to the likelihood from coalescent and recombination events
+is straightforward, since each possible coalescence is associated with rate $\lambda$,
+and each recombination with rate $r$.
+Under the SMC, we only ever have two lineages coalescing at a time;
+however, inferred ARGs may have more-than-binary mergers.
+We treat these as if there were zero-length edges
+between a sequence of mergers;
+the order of these unresolved mergers does not affect the likelihood.
+(Under the SMC the likelihood of a zero-length edge is zero;
+but this is not special: since edge lengths are continuously distributed,
+the likelihood of any particular value is also zero.)
+Let $\deg_o(n) = \#\{e_i \in E : p_i=n\}$ be the out-degree of a node $n$,
+i.e., the number of edges having $n$ as a parent.
+Since mergers are binary in the SMC,
+this must have been the result of $\deg(n) - 1$ coalescences,
+and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood.
+Similarly,
+let $\deg_i(n) = \#\{e_i \in E : c_i=n\}$ be the in-degree of a node $n$,
+i.e., the number of edges having $n$ as a child.
+Each additional edge that is ancestral to a given node
+implies one additional ancestral recombination event,
+and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood.
+Put together, these two contribute
+\begin{equation}\label{eq:coal}
+    \prod_{n \in N} \lambda^{\deg_o(n)-1} r^{\deg_i(n)-1}  
+    =
+    (\lambda r)^{|E|-|N|} 
+\end{equation}
+to the likelihood.
+(The simple form follows because each edge contributes exactly one
+to some node's out-degree and some node's in-degree.)
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{figures/likelihood-sketch}
+    \caption{
+        Proposed figure to explain the likelihood computation with.
+        TODO: change numbered labels to match the order in the text,
+        add $x$ and $y$.
+        \label{fig:likelihood}
+    }
+\end{figure}
+
+Now consider the ``hazard'' associated with recombination.
+Begin with an edge $e = (p,c,x,y)$.
+This implies that there was a sequence of lineages
+from the time of the child $c$ (call this time $t_c$)
+back to the time of the parent $p$ (again, $t_p$),
+that contained the segment $[x,y)$,
+and the lineage that directly inherited from $p$
+had the segment $[x,y)$.
+Under the SMC,
+we therefore know there was no recombination between $x$ and $y$
+on any of the lineages 
+over this time span,
+the probability of which is
+$\exp(-r \times A)$, where $r$ is the recombination rate
+and $A$ is the total area (span times depth) of the edge.
+So, the first contribution to the likelihood from this edge $e$ is
 \begin{equation}\label{eq:span}
-A_{(c, p)}(\theta) = e^{-r (y-x)(t_p - t_{c})}
+A_e(\theta) = \exp(-r (y-x)(t_p - t_{c})) .
 \end{equation}
-% also x and y not necessarily equal to x_{c_{1}0} and y_{c_{1}m}.
 
-What's left to compute the likelihood of the focal edge, is to take into account
-all events associated with $x$,
-the leftmost coordinate of $(c, p, X_{cp})$.
-This is either only the
-coalescence event involving
-$c$ and $d$. In that case $x=x_{c0}$ and $c$ did not coalesce until $t_p$.
-Alternatively, $x$ can be a new recombination break point that
-occurred on the lineage represented by $c$ at some time $s$ between $t_{c}$
-and $\min(t_p, t_{p^{\prime}})$, where $p^{\prime}$ is the parent of the
-edge $(c, p^{\prime}, X_{cp^{\prime}})$ ending at $x$. The new lineage formed
-by the recombination event and starting at $x$ can be given a temporary label
-$c_s$. In both cases, the cumulative hazard of a lineage $a$ coalescing
-between time $s$ and $t$ is given by $F(c, s, t) = \int_{s}^{t} I_{c}(u)du$
-such that:
+Finally, we turn to the hazard associated with coalescence.
+To do this, we need only consider what happens along the left side
+of each edge.
+The instantaneous rate of coalescence of a lineage whose left endpoint is at $x$
+at a given time is equal to $\lambda$ multiplied
+by the number of earlier lineages that overlap $x$ at that time.
+Recall that we define a total order on edges,
+by ordering by left endpoint and breaking ties by child
+(i.e., $(p,c,x,y) < (p',c',x',y')$
+if $x<x'$, or if $x=x'$ and $c<c'$).
+Using this total order, we can define
+$I_e(t)$ to be the total number of earlier edges that overlap edge $e$
+and are present at time $t$.
+Then, the cumulative hazard for coalescence of edge $e$
+between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(u) du$.
+If the edge $e = (p,c,x,y)$
+is the leftmost segment of material inherited by node $c$
+(e.g., the lefthand segment ancestral to node $c$
+in Figure~\ref{fig:likelihood}),
+then this edge simply contributes $\exp(-F(e,t_c,t_p))$ to the likelihood.
+However, if the edge is \emph{not} the leftmost segment,
+then it was initiated by a new recombination breakpoint,
+and we need to integrate over possible times the breakpoint occurred.
+If this is the case, there is another edge $(p',c,x',x)$ with the same child node,
+immediately adjacent to the focal edge $(p,c,x,y)$.
+The recombination event that split the two must have happened between $t_c$
+and the smaller of $t_p$ and $t_{p'}$;
+taking all this into consideration, the contribution to the likelihood here is
+\begin{equation}\label{eq:depth}
+B_e(\theta) = \begin{cases}
+    e^{-\lambda F(e, t_c, t_p)}
+        & \qquad \text{if } x=x_{c} \\
+    \int_{t_c}^{t_{p} \wedge t_{p'}} e^{-\lambda F(e, s, t_{p})} ds
+        & \qquad \text{if } x>x_{c} ,
+\end{cases}
+\end{equation}
+where $x_c$ is the leftmost position of the edges whose children are $c$.
 
-\begin{equation}\label{eq:depth}
+% In that case $x=x_{c0}$ and $c$ did not coalesce until $t_p$.
+% Alternatively, $x$ can be a new recombination break point that
+% occurred on the lineage represented by $c$ at some time $s$ between $t_{c}$
+% and $\min(t_p, t_{p^{\prime}})$, where $p^{\prime}$ is the parent of the
+% edge $(c, p^{\prime}, X_{cp^{\prime}})$ ending at $x$. The new lineage formed
+% by the recombination event and starting at $x$ can be given a temporary label
+% $c_s$. In both cases, the cumulative hazard of a lineage $a$ coalescing
+% between time $s$ and $t$ is given by $F(c, s, t) = \int_{s}^{t} I_{c}(u)du$
+% such that:
+
+\comment{This is the old equation; I think the $e^{-rs}$ should not be there,
+    since it's already taken into account by \eqref{eq:span}.
+    Plus, the $\lambda$ and $r$ factors are now elsewhere.}
+\begin{equation}\label{eq:depth_old}
 B_{(c, p)}(\theta) = \begin{cases}
@@
 \end{equation}
-
+\comment{Do we still need this bit?}
 Here, the (second) exponential term gives the probability of not observing a
@@
 gives the probability of observing a recombination
-at time $s$ given a Poisson process along position $x$ with rate $r$. The
-time of the recombination is integrated out.
+at time $s$ given a Poisson process along position $x$ with rate $r$.
+The time of the recombination is integrated out.
 The last term gives the point probability density of the coalescence event
@@
 
-The likelihood of ARG $\mathcal{G}$ defined by the set of edges $E$ and
+The likelihood of ARG defined by the set of edges $E$ and nodes $N$
 given parameters $\theta$ then is
-
 \begin{equation}\label{eq:full-lik}
-\mathcal{L}(\mathcal{G}|\theta) = \prod_{(c, p) \in E} A_{(c, p)}(\theta) * B_{(c, p)}(\theta)
+    \mathcal{L}(\mathcal{G}|\theta)
+    =
+    (r\lambda)^{|E|-|N|} \prod_{(c, p) \in E} A_e(\theta) B_e(\theta)
 \end{equation}
 
-% Does this hold for the case beyond the root of the genealogy?
-
-[TO DO: improve this remark: when recording an ARG we do record edges in this situation,
-however, msprime simulations will stop recording edges for local trees that have reached
-their local mrca.]
+\comment{TO DO: is this still true?}
 Because the likelihood is computed based on edges,
@@
 given an ARG is straightforward.
+\comment{Perhaps give the expression, for completeness?}
 
@@
 
+\subsection{Multiple segments description}
+
+\comment{This is copied from above in case it's helpful in the future.}
+
+We now define the process more formally,
+and following the notation of \citet{mcvean_approximating_2005}.
+At any point in time, the state of the process is $L(t)$,
+the set of labeled lineages extant at time $t$,
+where each ``lineage'' is represented by a union of non-overlapping half-open intervals
+which represent the ancestral material carried by that lineage,
+and labels are integers.
+So, we can write $L(t) = \{X_i(t)\}_{i=1}^{n(t)}$,
+where the $i^\text{th}$ lineage is $X_i(t)$
+and is of the form
+$X_i(t) = ([x_{i0}, y_{i0}), \dotsc, [x_{im_i}, y_{im_i}))$
+for some interval endpoints $x_{i0} < y_{i0} < x_{i1} < \cdot < y_{im_i}$.
+The \emph{span} of this lineage is the interval $s(X_i(t)) = [x_{i0}, y_{im_i}$,
+and its \emph{size} is $|X_i(t)| = y_{im_i} - x_{i0}$.
+Since we look backwards in time, $t=0$ is today, and so
+$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
+each represented by a single interval spanning the entire genome.
+Then, the process evolves by a succession of coalescence and recombination
+events until each segment of ancestral material is only present in one lineage.
+The waiting time to next event is determined by these two competing processes
+with exponentially distributed waiting times as outlined below.
+
+Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
+so, if $T(t) = \sum_i |X_i(t)|$ is the sum of the total sizes of all lineages,
+then recombination occurs at instaneous rate $rT(t)$.
+A recombination to the left of $x$
+(i.e., between $x$ and $x-1$)
+with $x>x_{i0}$ and $x<y_{im}$ splits lineage $i$ into two new lineages,
+which each are given new, unique labels
+(and lineage $i$ is removed).
+This operation keeps the total amount of ancestral material unchanged.
+
+Coalescence occurs between
+two lineages with overlapping spans at rate $\lambda = 1/(2N_e)$
+(if we assume a randomly mating, diploid population of constant size $N_e$).
+The newly formed lineage acquires the union of both ordered intervals, % L is updated accordingly
+and the original lineages are removed.
+Although coalescence is reciprocal, here we'll define a strict total order on lineages
+(see \ref{par:liks}) on all lineages in $L(t)$ based on their leftmost starting points.
+For any such strict total order,
+the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
+i.e., $\lambda \sum_{i \neq j} I_{ij}$,
+where
+\begin{equation} \label{def:coal}
+I_{ij} = \begin{cases}
+    1 & X_i > X_j \text{ and } s(X_i) \cap s(X_j) \neq \emptyset \\
+    0 & \text{otherwise.}
+\end{cases}
+\end{equation}
+We say that lineage $j$ can coalesce \emph{into} lineaege $i$ if $I_{ij} = 1$
+(and notice that if $I_{ij} = 1$ then $I_{ji} = 0$).
+
+For future use, define $C_i(t)$ to be the set of lineages that can coalesce into lineage $i$,
+i.e., $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$.
+(Therefore, $|C_i(t)| = I_{i}(t) = \sum_{j} I_{ij}(t)$.)
+Because $I_{ij}$ is defined in terms of a total order, the sets $C_i(t)$ are disjoint,
+which will simplify the likelihood computations.
+In particular,
+although recombination affects \emph{which} lineages can coalesce into $i$,
+it does not change the \emph{number} of such lineages:
+in other words, a recombination event changes $C_i(t)$ but not $I_{i}$.
+Another useful quantity to know is the number of lineages carrying material
+at each point on the genome:
+for a location $z$ and time $t$, this is
+\begin{equation}
+    N(t,z) = \#\{i : x \le z < y \text{ for some } [x,y) \in X_i(t) \} .
+\end{equation}
+Note that here we are not counting the ``gaps'' in a lineage,
+so that $N(t, z)$ might be less than the number of $i$ with $z \in s(X_i(t))$.
+
+The recombination and coalescence operations described above
+either split a lineage or merge two overlapping lineages,
+and so as described so far, each lineage will consist of only a single interval,
+rather than a collection of intervals.
+However, there is one more operation:
+when a sub-segment of a lineage is ancestral to all samples,
+that sub-segment is removed.
+Formally, define $Z(t)$ to be the segments on which the lineages have totally coalesced:
+i.e., $Z(t) = \{z : N(t,z) = 1$.
+Then, each coalescence event replaces the resulting lineage $X_i(t)$
+by $X_i(t) \setminus Z(t)$,
+the set of intervals that results from removing $Z(t)$ from all the intervals in $X_i(t)$.
+
+% note on the disjoint set of intervals for each lineage under the SMC
+Because of this last step, a single lineage can consist of more than one disjoint interval.
+As noted by \cite{mcvean_approximating_2005},
+this implies that the backwards-in-time description of the
+SMC is slightly different from the left-to-right description,
+in which single ancestors only ever span a single segment of genome.
+This is depicted in fig.\ref{fig:smc-unary}:
+for instance, node 8 is ancestral to all samples on the middle two trees.
+However, this does not affect the distribution
+of marginal genealogies up to the roots.
+More precisely, all lineages will be associated with
+a single interval up to the point where one section of the
+genome has reached its most recent common ancestor but the neighbouring parts have not.
+
 \end{document}

### 2024-03-22 paper.tex

@@
 this must have been the result of $\deg(n) - 1$ coalescences,
-and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood.
+and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood
+for nodes with any children.
 Similarly,
@@
 implies one additional ancestral recombination event,
-and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood.
+and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood,
+if the node has any parents.
 Put together, these two contribute
@@
     =
-    (\lambda r)^{|E|-|N|} 
+    r^{|E|-|N|+n_r} 
+    \lambda^{|E|-|N|+n_s} 
 \end{equation}
-to the likelihood.
+to the likelihood,
+where $n_r$ is the number of nodes with in-degree 0 (roots),
+and $n_s$ is the number of nodes with out-degree 0 (usually, samples).
 (The simple form follows because each edge contributes exactly one
@@
     =
-    (r\lambda)^{|E|-|N|} \prod_{(c, p) \in E} A_e(\theta) B_e(\theta)
+    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(c, p) \in E} A_e(\theta) B_e(\theta) .
 \end{equation}
+Although the first two factors are easily expressed in terms of numbers of edges and nodes,
+we will see below
+that in practice it is convenient to absorb these terms into per-edge contributions.
 
@@
 
-Although the argument outlined in the previous paragraph is valid for any ARG, here we
-will describe how to (efficiently) extract the information required to associate a
-tree sequence encoding of an ARG with a likelihood under the SMC (see fig. \ref{fig:smc-unary}).
-The tree sequence format allows us to sequentially recover the local trees
-along the genome very efficiently, and in a way that allows us to reason about
+Next, we describe how to efficiently compute the likelihood under the SMC
+from an ARG stored using the \emph{tree sequence} encoding. %  (see fig. \ref{fig:smc-unary}).
+The tree sequence format encodes relationships using \emph{edges}
+just as described above,
+along with additional structure that
+allows us to sequentially recover the local trees
+along the genome very efficiently, and to reason about
 the differences between those trees \citep{kelleher_efficient_2016, ralph_efficiently_2020}.
-The order of the required edge deletions and insertions to move from one local tree $T_1$ to
-the next, $T_2$ imposes the required strict total ordering
-(see definition \ref{def:coal}) on all
-lineages in the graph needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
+The order of the edge deletions and insertions required to move from one local tree $T_1$ to
+the next, $T_2$ imposes the required strict total ordering on edges
+(see definition \ref{def:coal})
+that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
 Indeed, following the removal of edges that do not persist in $T_2$,
-we can postulate that the lineage associated with each subsequently added edge
-can only coalesce with the lineages represented by the subset of the
-visited edges that make up the (partially) reconstructed local tree $T_2$.
-
-This ensures we can associate a likelihood
-with all tree sequences in a single pass over all edges.
+we simply declare that each subsequently added edge
+can only coalesce with the already-visited edges 
+that make up the (partially) reconstructed local tree $T_2$.
+
+\comment{This is \emph{very} nice!}
+This allows us to compute a likelihood in a single pass over all edges.
 More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
-the exact number of extant lineages in each internode interval (see fig \ref{fig:algo}).
+the number of extant lineages in each internode interval (see fig \ref{fig:algo}).
 When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along the internode intervals
 spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
-For each subsequently inserted edge, we first compute the likelihood by computing the integral
-in eq \ref{eq:full-lik} using $I$ and then update the counts. We further keep track of
+For each subsequently inserted edge, we first compute the likelihood contribution
+of that edge by computing the integral
+in equation \eqref{eq:full-lik} using $I$, and then update the counts in $I$.
+\comment{At first I thought we didn't need this next sentence now, but maybe
+assigning-$\lambda$-and-$r$ to each edge makes it easier to recompute the likelihood
+for slices?}
+We further keep track of
 the last parent a child node was connected to to detect recombination events.
@@
 through time by intersecting the internode intervals with the rate change intervals.
-Rate changes along the genome can equally be incorporated (see \ref{sec:slicing}).
+Rate changes along the genome can be incorporated similarly (see \ref{sec:slicing}).
 
@@
 By explicitly associating a single likelihood with each edge,
-we can compute the likelihood for any slice of the ARG in both
-time as well as along the genome. Our approach therefore strengthens
+we can compute the likelihood for any slice of the ARG taken in
+time and/or along the genome. This tool therefore strengthens
 the inherent ability of ARGs to provide us with temporal
-resolution on the inferred genealogical history of a sample.
+resolution and context to a dataset of genetic variation.
 It also opens up the opportunity for more efficient optimization
@@
 under the SMC for any valid tree sequence.
-The \tskit library provides a whole range
+The \tskit library provides a wide range
 of subsetting operations on tree sequences allowing us to slice
-the ARG in time and/or space, as well as reduce the number of
+the ARG in time and/or space, as well as to reduce the number of
 tracked samples. Any such subset of the tree sequence is itself
-a valid tree sequence that can again be used to compute a
-likelihood for. Figure \ref{fig:3-arg-slices} demonstrates this idea
-on a toy example. We simulated an ARG under human-like parameters and with
-two population size changes and subsequently estimated the posterior
-distribution using a uniform prior for each slice independently.
-
-The observation that we can associate a likelihood with any valid tree sequences
-urges the need for some caveats. By treating each edge independently we can
+a valid tree sequence for which we can again compute a likelihood.
+\comment{This next remark is totally not obvious, and a really nice observation;
+maybe it needs a bit more explanation?}
+\comment{Oh, I think it's not \emph{quite} true: if we slide at position $x$,
+then any recombinations that happened at $x$ are now invisible, right?
+So maybe you need to slice at non-integer positions, or do some bookeeping
+about the number of edges with starts exactly at $x$?}
+Remarkably, if we slice an ARG into disjoint pieces,
+the likelihood of the full ARG
+is the product of the likelihoods of each of the pieces.
+
+Figure \ref{fig:3-arg-slices} demonstrates this idea
+on a toy example. We simulated an ARG under human-like parameters with
+two effective population size changes, and subsequently computed the posterior
+distribution on the three effective population sizes
+using a uniform prior on each slice independently.
+The result is reasonably accurate,
+despite only using 10kb of sequence.
+
+Although we can associate a likelihood with any valid tree sequences,
+there are some caveats. By treating each edge independently we can
 naturally deal with the various degrees of completeness to which inheritance
-is documented by the different ARG inference methods: e.g.\ polytomies, absence
+is documented by the different ARG inference methods: e.g., polytomies, absence
 of locally unary coalescence nodes, unpreserved identify of internal nodes
-across marginal trees, \dots However, these imprecisions do violate the implicit
+across marginal trees, etcetera.
+However, many of these imprecisions violate the implicit
 assumptions of the likelihood computation.
 \comment{Do we want these remarks in here, or as part of the Discussion?}
-For example, the described algorithm relies heavily on the
-presence of locally unary nodes in the tree sequence to correctly identify the
-full extent of all lineages going back through time and therefore recombination
-events. Without unary nodes though, our algorithm will detect a recombination
+For example, the described algorithm assumes the
+presence of ``locally unary'' nodes to correctly identify the
+full extent of all lineages going back through time and therefore recombination events.
+Without unary nodes though, our algorithm will detect a recombination
 event along both coalescing lineages in case of a (sub)tree height changes along
@@
 (see fig \ref{sup:fig:rec-correction}).
+\comment{I'm not editing this in since I'm not sure I've got it right:
+I think that as written, our algorithm would count an extra recombination
+and and extra coalescence for any edge that got split up by removing the unary bits.
+This could be corrected for in the algorithm, but it would require extra bookkeeping,
+and make the likelihood not so easily a product over edges. (?)}
 

### 2024-04-03 paper.tex
Updates from Overleaf
review figures and likelihood section

@@
 and subsequent papers followed this trend 
-(e.g., \comment{cite}).
+\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,rasmussen_genome-wide_2014}.
 
@@
 and the original lineages are removed.
-Although coalescence is reciprocal, here we'll define a strict total order on lineages
-(see \ref{par:liks}) on all lineages in $L(t)$ based on their leftmost starting points.
+Although coalescence is reciprocal, here we'll define a strict total order
+on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
 For any such strict total order,
@@
 \end{equation}
-We say that lineage $j$ can coalesce \emph{into} lineaege $i$ if $I_{ij} = 1$
+We say that lineage $j$ can coalesce \emph{into} lineage $i$ if $I_{ij} = 1$
 (and notice that if $I_{ij} = 1$ then $I_{ji} = 0$).
@@
 (i.e., that are ancestral to all samples) are removed.
+
 Formally, after any coalescence or recombination event,
 the segment $[x,y)$ carried by the lineage
-is replaced by the largest segment whose endpoints have not coalesced,
-i.e., $x$ is replaced by $\min\{ z \ge x : N(t,z) > 1\}$,
+is replaced by the largest (pair of) segment(s) whose endpoints have not
+coalesced, i.e., $x$ is replaced by $\min\{ z \ge x : N(t,z) > 1\}$,
 and $y$ is replaced by $\sup\{ z \le y : N(t,z) > 1\}$.
-\comment{explain the cases in which this happens: coalescence on one end,
-and recobmbination in the middle of a trapped ancestral segment}
-
-% note on the disjoint set of intervals for each lineage under the SMC
-\comment{I'm not sure we need this here now?}
-Note that 
-an alternative definition could instead remove any sub-segment of a lineage
-that has entirely coalesced.
-This would produce a process in which each lineage is a collection
-of disjoint intervals (instead of a single one).
-As noted by \cite{mcvean_approximating_2005},
-this implies that the backwards-in-time description of the
-SMC is slightly different from the left-to-right description,
-which would usually not include these segments.
-This is depicted in fig.\ref{fig:smc-unary}:
-for instance, node 8 is ancestral to all samples on the middle two trees.
+By reasoning about the process in this way we can keep representing 
+all lineages by means of a single half-open interval throughout.
+
+\comment{Adapt figure to represent this}
 
@@
 \subsection{Estimated ARGs} \label{par:recording}
-
-\comment{clarify what's different to McVean \& Cardin: unary nodes, maybe? and different to usual SMC stuff more generally}
 
@@
 The data structure we will compute the likelihood of
-is a collection of \emph{edges} $E=\{e_i\}$,
-where each edge is of the form $e_i = (p_i,c_i,x_i,y_i)$.
-Recall that this means that node $c_i$ inherited genetic material from node $p_i$
-over a region with span from $x_i$ to $y_i$ along the genome
+is a collection of \emph{edges} $E$,
+where each edge $e$ is of the form $e = (p,c,x,y)$.
+Recall that this means that node $c$ inherited genetic material from node $p$
+over a region with span from $x$ to $y$ along the genome
 (see Figure~\ref{fig:likelihood}).
-So, $x_i$ and $y_i$ are genomic coordinates,
-while $c_i$ and $p_i$ are in $N$, the set of \emph{nodes}
+So, $x$ and $y$ are genomic coordinates,
+while $c$ and $p$ are in $N$, the set of \emph{nodes}
 (i.e., distinct sampled or ancestral haplotypes).
@@
 integrated over the entire process
-(also called the ``total hazard'').
+(also called the ``total hazard''). \comment{if we want to spell it out even further:
+Note that the time to the next event is distributed as $\Lambda \exp(-\Lambda)$. And
+the probablility that the next event is of type $i$ is $\lambda_i / \Lambda$.}
 
@@
 the likelihood of any particular value is also zero.)
-Let $\deg_o(n) = \#\{e_i \in E : p_i=n\}$ be the out-degree of a node $n$,
+Let $\deg_o(n) = \#\{e \in E : p_e=n\}$ be the out-degree of a node $n$,
 i.e., the number of edges having $n$ as a parent.
@@
 Similarly,
-let $\deg_i(n) = \#\{e_i \in E : c_i=n\}$ be the in-degree of a node $n$,
+let $\deg_i(n) = \#\{e \in E : c_e=n\}$ be the in-degree of a node $n$,
 i.e., the number of edges having $n$ as a child.
@@
 $\exp(-r \times A)$, where $r$ is the recombination rate
-and $A$ is the total area (span times depth) of the edge.
+and $A$ is the total area (eligbile links times depth) of the edge.
 So, the first contribution to the likelihood from this edge $e$ is
 \begin{equation}\label{eq:span}
-A_e(\theta) = \exp(-r (y-x)(t_p - t_{c})) .
+A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) .
 \end{equation}
+Note that if a recombination did occurr, then the number of links not
+affected by recombination is 1 less (see \ref{eq:depth}).
 
@@
 in Figure~\ref{fig:likelihood}),
-then this edge simply contributes $\exp(-F(e,t_c,t_p))$ to the likelihood.
+then no recombination occurred along this edge and 
+this edge simply contributes $\exp(-F(e,t_c,t_p))$ to the likelihood.
 However, if the edge is \emph{not} the leftmost segment,
@@
 The recombination event that split the two must have happened between $t_c$
-and the smaller of $t_p$ and $t_{p'}$;
-taking all this into consideration, the contribution to the likelihood here is
+and the smaller of $t_p$ and $t_{p'}$. Note that two events must have happened
+between $t_c$ and $\min(t_p, t_{p'})$; taking all this into consideration,
+the contribution to the likelihood here is
 \begin{equation}\label{eq:depth}
-B_e(\theta) = \begin{cases}
+B_e(r, \lambda) = \begin{cases}
     e^{-\lambda F(e, t_c, t_p)}
         & \qquad \text{if } x=x_{c} \\
-    \int_{t_c}^{t_{p} \wedge t_{p'}} e^{-\lambda F(e, s, t_{p})} ds
+    \int_{t_c}^{t_{p} \wedge t_{p'}} e^{- r s} e^{-\lambda F(e, s, t_{p})} ds
         & \qquad \text{if } x>x_{c} ,
@@
 
-\comment{This is the old equation; I think the $e^{-rs}$ should not be there,
-    since it's already taken into account by \eqref{eq:span}.
-    Plus, the $\lambda$ and $r$ factors are now elsewhere.}
-\begin{equation}\label{eq:depth_old}
-B_{(c, p)}(\theta) = \begin{cases}
-e^{-\lambda F(c, t_c, t_p)} \lambda^{I_{cd}(t_p)} & x=x_{c0} \\
-\int_{t_c}^{t_{p} \wedge t_{p^{\prime}}} r e^{-rs} e^{-\lambda F(c_s, s, t_{p})} ds \lambda^{I_{c_{s}d}(t_p)} & x=x_{c_{s}0}>x_{c0} \\
-\end{cases}
-\end{equation}
-\comment{Do we still need this bit?}
-Here, the (second) exponential term gives the probability of not observing a
-coalescence event before $t_p$. In case of a recombination, $re^{-rs}$ further
-gives the probability of observing a recombination
-at time $s$ given a Poisson process along position $x$ with rate $r$.
-The time of the recombination is integrated out.
-The last term gives the point probability density of the coalescence event
-that terminates the edge. The indicator function is used to resolve
-the non-reciprocal nature of coalescence events as defined in \ref{def:coal}.
+%\comment{This is the old equation; I think the $e^{-rs}$ should not be there,
+%    since it's already taken into account by \eqref{eq:span}.
+%    Plus, the $\lambda$ and $r$ factors are now elsewhere.}
+%\begin{equation}\label{eq:depth_old}
+%B_{(c, p)}(\theta) = \begin{cases}
+%e^{-\lambda F(c, t_c, t_p)} \lambda^{I_{cd}(t_p)} & x=x_{c0} \\
+%\int_{t_c}^{t_{p} \wedge t_{p^{\prime}}} r e^{-rs} e^{-\lambda F(c_s, s, t_{p})} ds \lambda^{I_{c_{s}d}(t_p)} & x=x_{c_{s}0}>x_{c0} \\
+%\end{cases}
+%\end{equation}
+%\comment{Do we still need this bit?}
+%Here, the (second) exponential term gives the probability of not observing a
+%coalescence event before $t_p$. In case of a recombination, $re^{-rs}$ further
+%gives the probability of observing a recombination
+%at time $s$ given a Poisson process along position $x$ with rate $r$.
+%The time of the recombination is integrated out.
+%The last term gives the point probability density of the coalescence event
+%that terminates the edge. The indicator function is used to resolve
+%the non-reciprocal nature of coalescence events as defined in \ref{def:coal}.
 
@@
 \begin{equation}\label{eq:full-lik}
-    \mathcal{L}(\mathcal{G}|\theta)
+    \mathcal{L}(\mathcal{G}|r, \lambda)
     =
-    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(c, p) \in E} A_e(\theta) B_e(\theta) .
+    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(c, p) \in E} A_e(r) B_e(r, \lambda) .
 \end{equation}
@@
 
-\comment{TO DO: is this still true?}
-Because the likelihood is computed based on edges,
-and edges are no longer recorded once the local most recent common ancestor
-has been reached, we will underestimate the number of positions where
-a recombination could have occurred in case a recombination hits a
-lineage with trapped non-ancestral material (see end \ref{par:description}).
-If we assume that only a single recombination could have
-occurred, then it suffices however to provide a correction factor $g$ in equation
-\ref{eq:depth}. Here $g$ represents the
-length of non-ancestral material along which a recombination would have resulted
-in the same observed local sequence of tree changes.
-
 Note that we are only considering the topology and branch length information of any ARG.
 Extending this expression by adding the likelihood of observing a set of mutations
-given an ARG is straightforward.
-\comment{Perhaps give the expression, for completeness?}
-
+given an ARG is given by
+\begin{equation}\label{eq:lik-mut}
+    \mathcal{L}(\mathcal{D}|\mathcal{G}, \mu)
+    =
+    \frac{1}{\mathcal{M}!} \prod_{(c, p) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}} .
+\end{equation}
+where $\mathcal{M}$ the total number of mutations and $m_e$
+the number of mutations along edge $(c,p)$ \citep{mahmoudi_bayesian_2022}.
+
+% % % % % % % % % % % % % % % % % %
 \subsection{Algorithm} \label{par:algo}
@@
 
-\comment{This is \emph{very} nice!}
 This allows us to compute a likelihood in a single pass over all edges.
@@
 in equation \eqref{eq:full-lik} using $I$, and then update the counts in $I$.
-\comment{At first I thought we didn't need this next sentence now, but maybe
-assigning-$\lambda$-and-$r$ to each edge makes it easier to recompute the likelihood
-for slices?}
 We further keep track of
 the last parent a child node was connected to to detect recombination events.
-
 Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate)
@@
 
+% % % % % % % % % % % % % % % % % %
 \subsection{Slicing the ARG} \label{sec:slicing}
@@
 we can compute the likelihood for any slice of the ARG taken in
-time and/or along the genome. This tool therefore strengthens
+time and/or along the genome.
+In the case of a fully precise ARG, with a known time to recombination,
+this is true as long as the slicing operation does not erase the
+information on whether an edge is associated with the lefthand or
+righthand segment generated by a recombination event. In the absence
+of the exact time to recombination, the integral in
+eq.\ \ref{eq:depth} means that the likelihood
+for the entire ARG does not factorize into the likelihoods of these
+individual slices. More precisely, the likelihood associated with any
+edge that is initiated by a recombination breakpoint cannot be factorized
+into the product of two integrals of the same form.
+
+However, as long as the individual slices of the ARG contain sufficient
+information we can approximate the true integral as the product of the
+integral of the slices. Figure \ref{fig:3-arg-slices} demonstrates this idea
+on a toy example. We simulated an ARG under human-like parameters with
+two effective population size changes, and subsequently computed the posterior
+distribution on the three effective population sizes
+using a uniform prior on each slice independently.
+The result is reasonably accurate, despite only using 10kb of sequence.
+This approach therefore strengthens
 the inherent ability of ARGs to provide us with temporal
@@
 strategies during inference given that each slice of the ARG can be
-processed independently.
-
-From a data structure perspective, we can compute a likelihood
-under the SMC for any valid tree sequence.
-The \tskit library provides a wide range
+processed independently. And it further highlights the flexibility
+of the algorithm. Simply stated and from a data structure perspective,
+we can compute a likelihood under the SMC for any valid tree sequence.
+The \texttt{tskit} library provides a wide range
 of subsetting operations on tree sequences allowing us to slice
@@
 a valid tree sequence for which we can again compute a likelihood.
-\comment{This next remark is totally not obvious, and a really nice observation;
-maybe it needs a bit more explanation?}
-\comment{Oh, I think it's not \emph{quite} true: if we slide at position $x$,
-then any recombinations that happened at $x$ are now invisible, right?
-So maybe you need to slice at non-integer positions, or do some bookeeping
-about the number of edges with starts exactly at $x$?}
-Remarkably, if we slice an ARG into disjoint pieces,
-the likelihood of the full ARG
-is the product of the likelihoods of each of the pieces.
-
-Figure \ref{fig:3-arg-slices} demonstrates this idea
-on a toy example. We simulated an ARG under human-like parameters with
-two effective population size changes, and subsequently computed the posterior
-distribution on the three effective population sizes
-using a uniform prior on each slice independently.
-The result is reasonably accurate,
-despite only using 10kb of sequence.
-
-Although we can associate a likelihood with any valid tree sequences,
-there are some caveats. By treating each edge independently we can
-naturally deal with the various degrees of completeness to which inheritance
-is documented by the different ARG inference methods: e.g., polytomies, absence
-of locally unary coalescence nodes, unpreserved identify of internal nodes
-across marginal trees, etcetera.
-However, many of these imprecisions violate the implicit
-assumptions of the likelihood computation.
-\comment{Do we want these remarks in here, or as part of the Discussion?}
-For example, the described algorithm assumes the
-presence of ``locally unary'' nodes to correctly identify the
-full extent of all lineages going back through time and therefore recombination events.
-Without unary nodes though, our algorithm will detect a recombination
-event along both coalescing lineages in case of a (sub)tree height changes along
-the ARG as both edges would switch parent nodes. This can be mitigated by limiting
-the number of 'inferrable' recombination events per coalescence event to 1
-(see fig \ref{sup:fig:rec-correction}).
-\comment{I'm not editing this in since I'm not sure I've got it right:
-I think that as written, our algorithm would count an extra recombination
-and and extra coalescence for any edge that got split up by removing the unary bits.
-This could be corrected for in the algorithm, but it would require extra bookkeeping,
-and make the likelihood not so easily a product over edges. (?)}
-
-\comment{this was previously part of the discussion, have grouped it here.}
-Although their presence is crucial to our ability to embed any local tree in the
-ARG, the concept and their importance are still relatively new (but see
-\citet{wong_general_2023}).
-We have briefly touched upon the observation that in
-the presence of unary nodes we seem to augment the left-to-right logic of the SMC
-with some additional information. Yet, we believe that the topic of locally
-unary nodes in general requires further exploration.
-And more specifically, we see it as future work to fully quantify to what
-extent the absence of these locally unary nodes will affect the shape of the
-likelihood surface given any set of parameters.
-
-%We simulated 1000 ARGs under the Hudson coalescent retaining all recombination information and computed the corresponding likelihood using \texttt{msprime} \citep{baumdicker_efficient_2021}.
-After removing the recombination nodes the likelihoods as defined here were computed both with and without the unary nodes.
-Figure \ref{sup:fig:vs-hudson} shows that the proportion of variation explained is higher when the unary nodes are retained.
+
+% Although we can associate a likelihood with any valid tree sequences,
+% there are some caveats. By treating each edge independently we can
+% naturally deal with the various degrees of completeness to which inheritance
+% is documented by the different ARG inference methods: e.g., polytomies, absence
+% of locally unary coalescence nodes, unpreserved identify of internal nodes
+% across marginal trees, etcetera.
+% However, many of these imprecisions violate the implicit
+% assumptions of the likelihood computation.
+% \comment{Do we want these remarks in here, or as part of the Discussion?}
+% 
+% For example, the described algorithm assumes the
+% presence of ``locally unary'' nodes to correctly identify the
+% full extent of all lineages going back through time and therefore recombination events.
+% Without unary nodes though, our algorithm will detect a recombination
+% event along both coalescing lineages in case of a (sub)tree height changes along
+% the ARG as both edges would switch parent nodes. This can be mitigated by limiting
+% the number of 'inferrable' recombination events per coalescence event to 1
+% (see fig \ref{sup:fig:rec-correction}).
+% \comment{I'm not editing this in since I'm not sure I've got it right:
+% I think that as written, our algorithm would count an extra recombination
+% and and extra coalescence for any edge that got split up by removing the unary bits.
+% This could be corrected for in the algorithm, but it would require extra bookkeeping,
+% and make the likelihood not so easily a product over edges. (?)}
+% 
+% \comment{this was previously part of the discussion, have grouped it here.}
+% Although their presence is crucial to our ability to embed any local tree in the
+% ARG, the concept and their importance are still relatively new (but see
+% \citet{wong_general_2023}).
+% We have briefly touched upon the observation that in
+% the presence of unary nodes we seem to augment the left-to-right logic of the SMC
+% with some additional information. Yet, we believe that the topic of locally
+% unary nodes in general requires further exploration.
+% And more specifically, we see it as future work to fully quantify to what
+% extent the absence of these locally unary nodes will affect the shape of the
+% likelihood surface given any set of parameters.
+
+%%% remarks on precision of information + useage of likelihood to either
+%%% estimate parameters (not goal here) or use as instrument to improve accuracy to infer ARGs
 
@@
     (5000:\infty):10 000]$. Remaining parameters: $L=10$kb, $20$ diploid samples,
-$r=2.5e-8$, $148$ edges in ARG). A single $N_e$ value was estimated for each
+    $r=2.5e-8$, $148$ edges in ARG. A single $N_e$ value was estimated for each
     slice of the ARG as defined by the true population size change time points.
@@
     \label{fig:3-arg-slices}
+\end{figure}
+
+\section{Likelihood surface for larger class of ARGs}
+
+% flexibility
+% presence of polytomies
+% lack of unary nodes
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/likelihood_surface.png}
+    \caption{Simulated ARG under SMC ($L=1$Mb, $100$ diploid samples,
+    $r=2.5e-8$, $N_e=1e4$). Likelihood-ratio contour plot. Currently two contours,
+    with and without locally unary nodes.}
+    \label{fig:lik-surface}
 \end{figure}
@@
 where
-\begin{equation} \label{def:coal}
+\begin{equation}
 I_{ij} = \begin{cases}

### 2024-04-03 paper.tex
lik-ratio section

@@
     \centering
-    \includegraphics[width=\textwidth]{figures/likelihood-sketch}
+    \includegraphics[width=\textwidth]{figures/single_event.png}
     \caption{
@@
 \end{equation}
-Although the first two factors are easily expressed in terms of numbers of edges and nodes,
-we will see below
-that in practice it is convenient to absorb these terms into per-edge contributions.
-
-Note that we are only considering the topology and branch length information of any ARG.
+
+% Although the first two factors are easily expressed in terms of numbers of edges and nodes,
+% we will see below
+% that in practice it is convenient to absorb these terms into per-edge contributions.
+
+% refer to the algorithm in the supplements, briefly mention scalability
+This likelihood expression is not only linear in the number of edges. An algorithm
+(see supplementary information \ref{par:algo}) can be devised to efficiently keep
+track of $I_e(t)$ by means of existing \texttt{tskit} functionality.
+Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate)
+through time by intersecting the internode intervals with the rate change intervals. 
+Rate changes along the genome can be incorporated similarly.
+Finally, we have only considered the topology and branch length information of any ARG.
 Extending this expression by adding the likelihood of observing a set of mutations
 given an ARG is given by
+
 \begin{equation}\label{eq:lik-mut}
@@
 % % % % % % % % % % % % % % % % % %
-\subsection{Algorithm} \label{par:algo}
+\subsection{Flexibility}
+
+% likelihood surface
+The last paragraph aims to highlight the flexibility of this approach 
+and the impact of the various
+degrees of completeness of the ARG on the likelihood.
+Note that the figures in this paragraph only serve this exact purpose,
+we are not claiming to have designed a new parameter estimation approach.
+
+In line with the observation that the presence of locally unary nodes 
+(see section \ref{par:recording}) is crucial for the equivalence between
+the graph or local trees view of an ARG, one can gradually simplify down
+a simulated ARG by gradually removing the various types of locally unary nodes.
+Figure \ref{fig:lik-surface} shows the impact of these simplification steps
+on the likelihood for a single simulated ARG.
+The log likelihood ratio is shown to improve comparability.
 
@@
     \centering
-\includegraphics[width=\textwidth]{figures/ts_algo_2rows.png}
-\caption{Description of algorithm: Moving along the genome we transition from
-local tree $T_1$ to the next, $T_2$ in three steps. 
-Edges that do not persist in $T_2$ are
-removed first. Edges starting at the left coordinate of $T_2$ are inserted.
-After deletion or insertion of an edge the count vector $I$ is updated along the
-corresponding internode intervals. Following an insertion, and prior to updating $I$,
-the likelihood for that edge is computed as $I$ then represents the total number of
-lineages the lineage associated with this new edge can
-coalesce with. By moving along the genome in this way, each edge gets visited exactly once.}
-\label{fig:algo}
+    \includegraphics[width=\textwidth]{figures/likratio.png}
+    \caption{Log likelihood-ratio curves for ARGs simulated under the SMC with
+    various degrees of completeness. The likelihood ratio is computed with
+    respect to the true recombination rate. In case of the full ARG the likelihood
+    under the SMC has been computed using the exact time to the recombination event.
+    Simulation performed with \texttt{msprime}, parameters: $L=1$Mb, $20$ diploid samples,
+    $r=2.5e-8$, $N_e=1e4$.}
+    \label{fig:lik-surface}
 \end{figure}
-
-Next, we describe how to efficiently compute the likelihood under the SMC
-from an ARG stored using the \emph{tree sequence} encoding. %  (see fig. \ref{fig:smc-unary}).
-The tree sequence format encodes relationships using \emph{edges}
-just as described above,
-along with additional structure that
-allows us to sequentially recover the local trees
-along the genome very efficiently, and to reason about
-the differences between those trees \citep{kelleher_efficient_2016, ralph_efficiently_2020}.
-The order of the edge deletions and insertions required to move from one local tree $T_1$ to
-the next, $T_2$ imposes the required strict total ordering on edges
-(see definition \ref{def:coal})
-that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
-Indeed, following the removal of edges that do not persist in $T_2$,
-we simply declare that each subsequently added edge
-can only coalesce with the already-visited edges 
-that make up the (partially) reconstructed local tree $T_2$.
-
-This allows us to compute a likelihood in a single pass over all edges.
-More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
-the number of extant lineages in each internode interval (see fig \ref{fig:algo}).
-When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along the internode intervals
-spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
-For each subsequently inserted edge, we first compute the likelihood contribution
-of that edge by computing the integral
-in equation \eqref{eq:full-lik} using $I$, and then update the counts in $I$.
-We further keep track of
-the last parent a child node was connected to to detect recombination events.
-Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate)
-through time by intersecting the internode intervals with the rate change intervals.
-Rate changes along the genome can be incorporated similarly (see \ref{sec:slicing}).
-
-% There is probably no need to potentially confuse people with this detail
-%Remark: better to describe edges here as edge-intervals, each describing a
-%single contiguous interval of genome inheritance between a pair of nodes.
-%Under the SMC and when recording unary nodes all edges can be represented
-%by a single edge-interval.
-
-% ADD IN RESULTS:
-% Figure \ref{sup:fig:vs-argweaver} shows the expected near-perfect correlation
-% between the coalescent prior as returned by \argweaver and the likelihood as computed here.
-% We simulated 1Mb of data for 10 diploid individuals
-% under the SMC using human-like parameters. 100 samples were taking from the posterior every for
-% a total of 1000 MCMC iterations.
-
-% % % % % % % % % % % % % % % % % %
-\subsection{Slicing the ARG} \label{sec:slicing}
 
@@
 
-\section{Likelihood surface for larger class of ARGs}
-
-% flexibility
-% presence of polytomies
-% lack of unary nodes
-
-\begin{figure}[!ht]
-    \centering
-    \includegraphics[width=\textwidth]{figures/likelihood_surface.png}
-    \caption{Simulated ARG under SMC ($L=1$Mb, $100$ diploid samples,
-    $r=2.5e-8$, $N_e=1e4$). Likelihood-ratio contour plot. Currently two contours,
-    with and without locally unary nodes.}
-    \label{fig:lik-surface}
-\end{figure}
-
 \section{Discussion}
@@
 and statistical genetics toolkit.
-In particular, we see three main applications for the work presented here.
+In particular, we see two main applications for the work presented here.
 Firstly, the ability to associate inferred ARGs with a likelihood given a demographic
@@
 past demography.
-Secondly, hybrid ARG-inference methods that combine the best of both
-Thirdly, inferred ARGs can be used as a good starting point for a future MCMC sampler.
-heuristic and model-based inference approaches are now possible.
-Finally, using a model we can quantify the uncertainty inherently associated
-with ARG-reconstruction.
-
-% 1/ statistical inference
-Currently most/all ARG-based inference methods base their analysis on marginal
-trees \citep{hejase_summary_2020} [TO DO: add more refs]. The theory and associated algorithm
-provided here can function as a building block of [...]
-$N_e$-changes as well as multiple populations and migration can be readily incorporated
-in the current algorithm.
-
-% 2/ hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
-Hybrid backwards-in-time simulations that combine Wright-Fisher dynamics in the
-very recent past and coalescent simulations for the remaining time are a tried and tested
-approach \citep{bhaskar_distortion_2014, nelson_accounting_2020} to enable large scale
-simulations while avoiding the documented biases of the coalescent relative to the
-Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. The same approach
-could be used for ARG-inference where one can rely on a heuristic inference method for
-the recent past, while using a model-based approach beyond that cutoff point.
-
-% 3/ great starting point for MCMC
+The second and probably the most important application of the ability to associate a more
+general class of ARGs to a principled model is the ability to boost the current efforts
+in ARG inference. 
+% great starting point for MCMC
 Any inferred ARG can be used a good starting point
@@
 efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
-
-% 4/ quantifying uncertainty [unfinished]:
-Acknowledging the uncertainty associated with ARG inference does not
+% hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
+To further improve scalability, hybrid ARG-inference methods could be designed which
+rely on heuristic approaches to bring the ARG down from the initial pedigree-dominated phase,
+and subsequently use a principled approach to perform ARG inference for the second phase.
+This idea resembles the tried and tested approach of hybrid backwards-in-time simulations 
+that combine Wright-Fisher dynamics in the very recent past and coalescent simulations 
+for the remaining time \citep{bhaskar_distortion_2014, nelson_accounting_2020}. This hybrid
+approach has enabled large scale simulations while avoiding the documented biases of the
+coalescent relative to the Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. 
+The same approach could be thus used for ARG-inference where one can rely on a heuristic 
+inference method for the recent past, while using a model-based approach beyond that cutoff point.
+
+Finally, acknowledging the uncertainty associated with ARG inference does not
 necessarily require a full-blown MCMC sampler. Instead, recognising those parts of
@@
 would already be a major step forward. 
-Furthermore, quantifying uncertainty would not necessarily always have to be associated
-with sampling from subset of all compatible genealogies. Uncertainty could also
+% this part is only a very rough idea, might be better to leave it out if we can't make it more concrete.
+Furthermore, we believe that quantifying uncertainty
+would not necessarily always have to be associated with sampling from the space of all
+compatible genealogies. Uncertainty could also
 be quantified and passed on as metadata to the nodes in the data format storing the ARG.
-% can something similar be done for stacked recombinations?
 
@@
 
+% % % % % % % % % % % % % % % % % %
+\subsection{Algorithm} \label{par:algo}
+
 \begin{figure}[!ht]
 \centering
-\includegraphics[width=0.75\textwidth]{figures/supplementary-figs/argweaver_vs_runsmc.png}
-\caption{Simulated 10kb sequence under human-like parameters, 10 diploid samples.
-Inferred ARG using ARGweaver.
-Scatterplot is comparison of prior (discretized SMC) as reported by ARGweaver (x-axis) and likelihood as defined here (y-axis).}
- \label{sup:fig:vs-argweaver}
+\includegraphics[width=\textwidth]{figures/ts_algo_2rows.png}
+\caption{Description of algorithm: Moving along the genome we transition from
+local tree $T_1$ to the next, $T_2$ in three steps. 
+Edges that do not persist in $T_2$ are
+removed first. Edges starting at the left coordinate of $T_2$ are inserted.
+After deletion or insertion of an edge the count vector $I$ is updated along the
+corresponding internode intervals. Following an insertion, and prior to updating $I$,
+the likelihood for that edge is computed as $I$ then represents the total number of
+lineages the lineage associated with this new edge can
+coalesce with. By moving along the genome in this way, each edge gets visited exactly once.}
+\label{fig:algo}
 \end{figure}
 
-\begin{figure}[!ht]
-\centering
-\includegraphics[width=0.75\textwidth]{figures/supplementary-figs/v_hudson_unary_simpl.png}
-\caption{Compare likelihoods Hudson (full ARG) and SMC with and without unary nodes (without recombination nodes): placeholder for final figure with similar content.}
-\label{sup:fig:vs-hudson}
-\end{figure}
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[width=0.75\textwidth]{figures/supplementary-figs/without_unary.png}
-\caption{Simulated 1Mb of sequence under human-like parameters.
-Computed likelihood as defined here with unary nodes (y-axis), and without, blue dots (x-axis).
-By limiting the number of recombinations (to 1 event) leading up to a single coalescent node we obtain the values in orange.}
- \label{sup:fig:rec-correction}
-\end{figure}
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[width=0.75\textwidth]{figures/supplementary-figs/hudson_liks_vs_tsinfer_smc.png}
-\caption{Simulated full ARG for 1Mb sequence under human-like parameters (100 diploid samples).
-Computed likelihood (x-axis) under Hudson.
-Then inferred ARG with \tsinfer following a random sprinkling of mutations and computed likelihood under SMC (y-axis)}
- \label{sup:fig:hudson-smc}
-\end{figure}
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[width=0.75\textwidth]{figures/supplementary-figs/single_rep_human_like_1gb.png}
-\caption{Single replicate from the dataset simulated above with a 100 random sprinklings of mutations.
-Likelihood under the SMC, mutation rate $\mu=1.25e-8$ (x-axis), and with mutation rate $4*\mu$ (y-axis).
-Likelihood for the full ARG under Hudson is indicated with the cross.}
- \label{sup:fig:single-tsinfer}
-\end{figure}
-
-\subsection{Multiple segments description}
-
-\comment{This is copied from above in case it's helpful in the future.}
-
-We now define the process more formally,
-and following the notation of \citet{mcvean_approximating_2005}.
-At any point in time, the state of the process is $L(t)$,
-the set of labeled lineages extant at time $t$,
-where each ``lineage'' is represented by a union of non-overlapping half-open intervals
-which represent the ancestral material carried by that lineage,
-and labels are integers.
-So, we can write $L(t) = \{X_i(t)\}_{i=1}^{n(t)}$,
-where the $i^\text{th}$ lineage is $X_i(t)$
-and is of the form
-$X_i(t) = ([x_{i0}, y_{i0}), \dotsc, [x_{im_i}, y_{im_i}))$
-for some interval endpoints $x_{i0} < y_{i0} < x_{i1} < \cdot < y_{im_i}$.
-The \emph{span} of this lineage is the interval $s(X_i(t)) = [x_{i0}, y_{im_i}$,
-and its \emph{size} is $|X_i(t)| = y_{im_i} - x_{i0}$.
-Since we look backwards in time, $t=0$ is today, and so
-$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
-each represented by a single interval spanning the entire genome.
-Then, the process evolves by a succession of coalescence and recombination
-events until each segment of ancestral material is only present in one lineage.
-The waiting time to next event is determined by these two competing processes
-with exponentially distributed waiting times as outlined below.
-
-Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
-so, if $T(t) = \sum_i |X_i(t)|$ is the sum of the total sizes of all lineages,
-then recombination occurs at instaneous rate $rT(t)$.
-A recombination to the left of $x$
-(i.e., between $x$ and $x-1$)
-with $x>x_{i0}$ and $x<y_{im}$ splits lineage $i$ into two new lineages,
-which each are given new, unique labels
-(and lineage $i$ is removed).
-This operation keeps the total amount of ancestral material unchanged.
-
-Coalescence occurs between
-two lineages with overlapping spans at rate $\lambda = 1/(2N_e)$
-(if we assume a randomly mating, diploid population of constant size $N_e$).
-The newly formed lineage acquires the union of both ordered intervals, % L is updated accordingly
-and the original lineages are removed.
-Although coalescence is reciprocal, here we'll define a strict total order
-on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
-For any such strict total order,
-the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
-i.e., $\lambda \sum_{i \neq j} I_{ij}$,
-where
-\begin{equation}
-I_{ij} = \begin{cases}
-    1 & X_i > X_j \text{ and } s(X_i) \cap s(X_j) \neq \emptyset \\
-    0 & \text{otherwise.}
-\end{cases}
-\end{equation}
-We say that lineage $j$ can coalesce \emph{into} lineage $i$ if $I_{ij} = 1$
-(and notice that if $I_{ij} = 1$ then $I_{ji} = 0$).
-
-For future use, define $C_i(t)$ to be the set of lineages that can coalesce into lineage $i$,
-i.e., $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$.
-(Therefore, $|C_i(t)| = I_{i}(t) = \sum_{j} I_{ij}(t)$.)
-Because $I_{ij}$ is defined in terms of a total order, the sets $C_i(t)$ are disjoint,
-which will simplify the likelihood computations.
-In particular,
-although recombination affects \emph{which} lineages can coalesce into $i$,
-it does not change the \emph{number} of such lineages:
-in other words, a recombination event changes $C_i(t)$ but not $I_{i}$.
-Another useful quantity to know is the number of lineages carrying material
-at each point on the genome:
-for a location $z$ and time $t$, this is
-\begin{equation}
-    N(t,z) = \#\{i : x \le z < y \text{ for some } [x,y) \in X_i(t) \} .
-\end{equation}
-Note that here we are not counting the ``gaps'' in a lineage,
-so that $N(t, z)$ might be less than the number of $i$ with $z \in s(X_i(t))$.
-
-The recombination and coalescence operations described above
-either split a lineage or merge two overlapping lineages,
-and so as described so far, each lineage will consist of only a single interval,
-rather than a collection of intervals.
-However, there is one more operation:
-when a sub-segment of a lineage is ancestral to all samples,
-that sub-segment is removed.
-Formally, define $Z(t)$ to be the segments on which the lineages have totally coalesced:
-i.e., $Z(t) = \{z : N(t,z) = 1$.
-Then, each coalescence event replaces the resulting lineage $X_i(t)$
-by $X_i(t) \setminus Z(t)$,
-the set of intervals that results from removing $Z(t)$ from all the intervals in $X_i(t)$.
-
-% note on the disjoint set of intervals for each lineage under the SMC
-Because of this last step, a single lineage can consist of more than one disjoint interval.
-As noted by \cite{mcvean_approximating_2005},
-this implies that the backwards-in-time description of the
-SMC is slightly different from the left-to-right description,
-in which single ancestors only ever span a single segment of genome.
-This is depicted in fig.\ref{fig:smc-unary}:
-for instance, node 8 is ancestral to all samples on the middle two trees.
-However, this does not affect the distribution
-of marginal genealogies up to the roots.
-More precisely, all lineages will be associated with
-a single interval up to the point where one section of the
-genome has reached its most recent common ancestor but the neighbouring parts have not.
+Next, we describe how to efficiently compute the likelihood under the SMC
+from an ARG stored using the \emph{tree sequence} encoding. %  (see fig. \ref{fig:smc-unary}).
+The tree sequence format encodes relationships using \emph{edges}
+just as described above,
+along with additional structure that
+allows us to sequentially recover the local trees
+along the genome very efficiently, and to reason about
+the differences between those trees \citep{kelleher_efficient_2016, ralph_efficiently_2020}.
+The order of the edge deletions and insertions required to move from one local tree $T_1$ to
+the next, $T_2$ imposes the required strict total ordering on edges
+(see definition \ref{def:coal})
+that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
+Indeed, following the removal of edges that do not persist in $T_2$,
+we simply declare that each subsequently added edge
+can only coalesce with the already-visited edges 
+that make up the (partially) reconstructed local tree $T_2$.
+
+This allows us to compute a likelihood in a single pass over all edges.
+More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
+the number of extant lineages in each internode interval (see fig \ref{fig:algo}).
+When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along the internode intervals
+spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
+For each subsequently inserted edge, we first compute the likelihood contribution
+of that edge by computing the integral
+in equation \eqref{eq:full-lik} using $I$, and then update the counts in $I$.
+We further keep track of
+the last parent a child node was connected to to detect recombination events.
+
+% There is probably no need to potentially confuse people with this detail
+%Remark: better to describe edges here as edge-intervals, each describing a
+%single contiguous interval of genome inheritance between a pair of nodes.
+%Under the SMC and when recording unary nodes all edges can be represented
+%by a single edge-interval.
+
+% ADD IN RESULTS:
+% Figure \ref{sup:fig:vs-argweaver} shows the expected near-perfect correlation
+% between the coalescent prior as returned by \argweaver and the likelihood as computed here.
+% We simulated 1Mb of data for 10 diploid individuals
+% under the SMC using human-like parameters. 100 samples were taking from the posterior every for
+% a total of 1000 MCMC iterations.
+
+% \subsection{Multiple segments description}
+% 
+% \comment{This is copied from above in case it's helpful in the future.}
+% 
+% We now define the process more formally,
+% and following the notation of \citet{mcvean_approximating_2005}.
+% At any point in time, the state of the process is $L(t)$,
+% the set of labeled lineages extant at time $t$,
+% where each ``lineage'' is represented by a union of non-overlapping half-open intervals
+% which represent the ancestral material carried by that lineage,
+% and labels are integers.
+% So, we can write $L(t) = \{X_i(t)\}_{i=1}^{n(t)}$,
+% where the $i^\text{th}$ lineage is $X_i(t)$
+% and is of the form
+% $X_i(t) = ([x_{i0}, y_{i0}), \dotsc, [x_{im_i}, y_{im_i}))$
+% for some interval endpoints $x_{i0} < y_{i0} < x_{i1} < \cdot < y_{im_i}$.
+% The \emph{span} of this lineage is the interval $s(X_i(t)) = [x_{i0}, y_{im_i}$,
+% and its \emph{size} is $|X_i(t)| = y_{im_i} - x_{i0}$.
+% Since we look backwards in time, $t=0$ is today, and so
+% $L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
+% each represented by a single interval spanning the entire genome.
+% Then, the process evolves by a succession of coalescence and recombination
+% events until each segment of ancestral material is only present in one lineage.
+% The waiting time to next event is determined by these two competing processes
+% with exponentially distributed waiting times as outlined below.
+% 
+% Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
+% so, if $T(t) = \sum_i |X_i(t)|$ is the sum of the total sizes of all lineages,
+% then recombination occurs at instaneous rate $rT(t)$.
+% A recombination to the left of $x$
+% (i.e., between $x$ and $x-1$)
+% with $x>x_{i0}$ and $x<y_{im}$ splits lineage $i$ into two new lineages,
+% which each are given new, unique labels
+% (and lineage $i$ is removed).
+% This operation keeps the total amount of ancestral material unchanged.
+% 
+% Coalescence occurs between
+% two lineages with overlapping spans at rate $\lambda = 1/(2N_e)$
+% (if we assume a randomly mating, diploid population of constant size $N_e$).
+% The newly formed lineage acquires the union of both ordered intervals, % L is updated accordingly
+% and the original lineages are removed.
+% Although coalescence is reciprocal, here we'll define a strict total order
+% on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
+% For any such strict total order,
+% the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
+% i.e., $\lambda \sum_{i \neq j} I_{ij}$,
+% where
+% \begin{equation}
+% I_{ij} = \begin{cases}
+%     1 & X_i > X_j \text{ and } s(X_i) \cap s(X_j) \neq \emptyset \\
+%     0 & \text{otherwise.}
+% \end{cases}
+% \end{equation}
+% We say that lineage $j$ can coalesce \emph{into} lineage $i$ if $I_{ij} = 1$
+% (and notice that if $I_{ij} = 1$ then $I_{ji} = 0$).
+% 
+% For future use, define $C_i(t)$ to be the set of lineages that can coalesce into lineage $i$,
+% i.e., $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$.
+% (Therefore, $|C_i(t)| = I_{i}(t) = \sum_{j} I_{ij}(t)$.)
+% Because $I_{ij}$ is defined in terms of a total order, the sets $C_i(t)$ are disjoint,
+% which will simplify the likelihood computations.
+% In particular,
+% although recombination affects \emph{which} lineages can coalesce into $i$,
+% it does not change the \emph{number} of such lineages:
+% in other words, a recombination event changes $C_i(t)$ but not $I_{i}$.
+% Another useful quantity to know is the number of lineages carrying material
+% at each point on the genome:
+% for a location $z$ and time $t$, this is
+% \begin{equation}
+%     N(t,z) = \#\{i : x \le z < y \text{ for some } [x,y) \in X_i(t) \} .
+% \end{equation}
+% Note that here we are not counting the ``gaps'' in a lineage,
+% so that $N(t, z)$ might be less than the number of $i$ with $z \in s(X_i(t))$.
+% 
+% The recombination and coalescence operations described above
+% either split a lineage or merge two overlapping lineages,
+% and so as described so far, each lineage will consist of only a single interval,
+% rather than a collection of intervals.
+% However, there is one more operation:
+% when a sub-segment of a lineage is ancestral to all samples,
+% that sub-segment is removed.
+% Formally, define $Z(t)$ to be the segments on which the lineages have totally coalesced:
+% i.e., $Z(t) = \{z : N(t,z) = 1$.
+% Then, each coalescence event replaces the resulting lineage $X_i(t)$
+% by $X_i(t) \setminus Z(t)$,
+% the set of intervals that results from removing $Z(t)$ from all the intervals in $X_i(t)$.
+% 
+% % note on the disjoint set of intervals for each lineage under the SMC
+% Because of this last step, a single lineage can consist of more than one disjoint interval.
+% As noted by \cite{mcvean_approximating_2005},
+% this implies that the backwards-in-time description of the
+% SMC is slightly different from the left-to-right description,
+% in which single ancestors only ever span a single segment of genome.
+% This is depicted in fig.\ref{fig:smc-unary}:
+% for instance, node 8 is ancestral to all samples on the middle two trees.
+% However, this does not affect the distribution
+% of marginal genealogies up to the roots.
+% More precisely, all lineages will be associated with
+% a single interval up to the point where one section of the
+% genome has reached its most recent common ancestor but the neighbouring parts have not.
 

### 2024-04-22 paper.tex
update figures

@@
 
-\comment{Adapt figure to represent this}
-
 % % % % % % % % % % % % % % % % % %
@@
 the path traced by $a$'s ancestry switches at that point.
-In fact, $a$ is in blue because it is a \emph{recombination} node --
-\comment{wait is it ``recombination node'' or ``recombinant node''?}
+In fact, $a$ is in blue because it is a \emph{recombinant} node --
 an ancestor that has a segment of genome inherited by a sample,
@@
 i.e., are \emph{locally unary}.
-Recombination nodes (in blue) have a single child in all trees.
+Recombinant nodes (in blue) have a single child in all trees.
 However, coalescent nodes can be locally unary as well:
@@
 generated by the SMC was Markovian \emph{after removing all unary nodes}.
-In other words, if we remove recombination nodes entirely,
+In other words, if we remove recombinant nodes entirely,
 and erase the unary portions of coalescent nodes,
@@
 ($a$, $b$, and $c$ in Fig.\ref{fig:smc-unary}).
-In the absence of these recombination nodes, a recombination can be observed
+In the absence of these recombinant nodes, a recombination can be observed
 easily enough as a change in parent going from one genomic region to the next.
-Without the recombination nodes we no longer know the precise time of each recombination event,
+Without the recombinant nodes we no longer know the precise time of each recombination event,
 but we do still have constraints on the time:
 for instance, we know that $c$ happened somewhere between node 5 and node 7.
-More generally, a recombination node must happen somewhere between the child node
+More generally, a recombinant node must happen somewhere between the child node
 and the most recent of the parent nodes.
@@
 would already be a major step forward.
+
 % this part is only a very rough idea, might be better to leave it out if we can't make it more concrete.
-Furthermore, we believe that quantifying uncertainty
-would not necessarily always have to be associated with sampling from the space of all
-compatible genealogies. Uncertainty could also
+We believe that quantifying uncertainty does not have to be associated with sampling
+from the space of all compatible genealogies. Uncertainty could also
 be quantified and passed on as metadata to the nodes in the data format storing the ARG.
@@
 
-Next, we describe how to efficiently compute the likelihood under the SMC
+Here, we describe how to efficiently compute the likelihood under the SMC
 from an ARG stored using the \emph{tree sequence} encoding. %  (see fig. \ref{fig:smc-unary}).

### 2024-04-24 paper.tex
address remarks smc and args

@@
 the set of labeled lineages extant at time $t$,
-where each ``lineage'' is represented by a half-open interval
-which represents the ancestral material carried by that lineage,
-and labeled by an integer.
-So, we can write $L(t) = \{X_i(t)\}_{i=1}^{n(t)}$,
-where the $i^\text{th}$ lineage is $X_i(t)$
-and is of the form
-$X_i(t) = [x_{i}, y_{i})$,
-for some interval endpoints $x_{i} < y_{i}$.
-The \emph{size} of the interval is $|X_i(t)| = y_{i} - x_{i}$.
+where each ``lineage'' $X_j(t)$ is labeled by an integer $j$
+and represented by a half-open interval delimiting the
+genetic material inherited by either one of a set of samples.
+This ancestral material \citep{wiuf_ancestry_1999} carried by a
+lineage can thus be summarized as $X_j(t) = [x_{j}, y_{j})$,
+for some interval endpoints $x_{j} < y_{j}$.
 Since we look backwards in time, $t=0$ is today, and so
@@
 each represented by a single interval spanning the entire genome.
-Then, the process evolves by a succession of coalescence and recombination
+Then, the process evolves by a succession of common ancestor and recombination
 events until each segment of ancestral material is only present in one lineage.
@@
 Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
-so, if $T(t) = \sum_i |X_i(t)|$ is the sum of the total sizes of all lineages,
-then recombination occurs at instaneous rate $rT(t)$.
+so, if $T(t) = \sum_i |X_j(t)|$ is the sum of the total sizes of all lineages,
+then recombination occurs at instantaneous rate $rT(t)$.
 A recombination to the left of $x$
 (i.e., between base pairs $x$ and $x-1$)
-with $x>x_{i}$ and $x<y_{i}$ splits lineage $i$ into two new lineages,
+with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
 which each are given new, unique labels
-(and lineage $i$ is removed).
+(and lineage $j$ is removed).
 This operation keeps the total amount of ancestral material unchanged.
 
-Coalescence occurs between
-two overlapping lineages at rate $\lambda = 1/(2N_e)$
-(if we assume a randomly mating, diploid population of constant size $N_e$).
-The newly formed lineage acquires the union of both ordered intervals, % L is updated accordingly
+Assuming a randomly mating, diploid population of constant size $N_e$,
+common ancestry occurs between
+two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
+The newly formed lineage acquires the union of both intervals, % L is updated accordingly
 and the original lineages are removed.
+As a consequence, under the SMC common ancestry events are always
+associated with coalescence events
+along some fraction of the merging half-open intervals.
+
 Although coalescence is reciprocal, here we'll define a strict total order
@@
 the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
-i.e., $\lambda \sum_{i \neq j} I_{ij}$,
+i.e., $\lambda \sum_{i \neq j} I_{jk}$,
 where
 \begin{equation} \label{def:coal}
-I_{ij} = \begin{cases}
-    1 & X_i > X_j \text{ and } s(X_i) \cap s(X_j) \neq \emptyset \\
+I_{jk} = \begin{cases}
+    1 & X_j > X_k \text{ and } y_k > x_j \\
     0 & \text{otherwise.}
@@
 \end{equation}
-We say that lineage $j$ can coalesce \emph{into} lineage $i$ if $I_{ij} = 1$
-(and notice that if $I_{ij} = 1$ then $I_{ji} = 0$).
-
-For future use, define $C_i(t)$ to be the set of lineages that can coalesce into lineage $i$,
-i.e., $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$.
-(Therefore, $|C_i(t)| = I_{i}(t) = \sum_{j} I_{ij}(t)$,)
-and the total rate of coalescence at time $t$ is $\sum_{i} |C_i(t)|$.) 
-Because $I_{ij}$ is defined in terms of a total order, the sets $C_i(t)$ are disjoint,
+We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$. From the
+definition of the strict total order it follows that
+if $I_{jk} = 1$ then $I_{kj} = 0$).
+
+For future use, define $C_j(t)$ to be the set of lineages that can coalesce into lineage $j$,
+i.e., $C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$. And therefore, 
+$|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
+and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$.
+Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
 which will simplify the likelihood computations.
 In particular,
-although recombination affects \emph{which} lineages can coalesce into $i$,
+although recombination affects \emph{which} lineages can coalesce into $j$,
 it does not change the \emph{number} of such lineages:
-in other words, a recombination event changes $C_i(t)$ but not $I_{i}$.
+in other words, a recombination event changes $C_j(t)$ but not $I_{j}(t)$.
 Another useful quantity to know is the number of lineages carrying material
@@
 \begin{equation}
-    N(t,z) = \#\{i : z \in X_i(t) \} .
+    N(t,z) = \vert\{j : z \in X_j(t) \} \vert
 \end{equation}
 
-Finally, there is one more operation that allows the process to terminate:
-portions of the edges of lineages that have entirely coalesced
-(i.e., that are ancestral to all samples) are removed.
-
-Formally, after any coalescence or recombination event,
-the segment $[x,y)$ carried by the lineage
-is replaced by the largest (pair of) segment(s) whose endpoints have not
-coalesced, i.e., $x$ is replaced by $\min\{ z \ge x : N(t,z) > 1\}$,
-and $y$ is replaced by $\sup\{ z \le y : N(t,z) > 1\}$.
+After any coalescence or recombination event,
+the segment(s) $[x_j,y_j)$ (and $[x_k, y_k]$) carried by the involved lineage(s)
+are replaced by a single segment $[x, y)$ whose endpoints have not fully
+coalesced, i.e., $x = \min\{ z \ge \min(x_j, x_k) : N(t,z) > 1\}$,
+and $y = \sup\{ z \le max(y_j, y_k) : N(t,z) > 1\}$.
 By reasoning about the process in this way we can keep representing 
-all lineages by means of a single half-open interval throughout.
+all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
+and the process terminates.
+
+[TO DO: remark on  beyond local roots of the genealogy]
+
+%The process of recombination and common ancestry events continues 
+%Finally, there is one more operation that allows the process to terminate:
+%portions of the edges of lineages that have entirely coalesced
+%(i.e., that are ancestral to all samples) are removed.
+
+%Formally, after any coalescence or recombination event,
+%the segment $[x,y)$ carried by the lineage
+%is replaced by the largest (pair of) segment(s) whose endpoints have not
+%coalesced, i.e., $x$ is replaced by $\min\{ z \ge x : N(t,z) > 1\}$,
+%and $y$ is replaced by $\sup\{ z \le y : N(t,z) > 1\}$.
+%By reasoning about the process in this way we can keep representing 
+%all lineages by means of a single half-open interval throughout.
 
 % % % % % % % % % % % % % % % % % %
-\subsection{Estimated ARGs} \label{par:recording}
+\subsection{ARGs} \label{par:recording}
 
@@
 Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
-of the form ``$c$ inherited from $p$ on the genomic segment from $\ell$ to $r$'',
-where $c$ and $p$ are (haploid) genomes, and $\ell$ and $r$ are genetic coordinates.
+of the form ``$c$ inherited from $p$ on the genomic segment from $x$ to $y$'',
+where $c$ and $p$ are (haploid) genomes, and $x$ and $y$ are genetic coordinates.
 We record each such statement in an ``edge'',
-often summarized as the tuple $(p,c,\ell,r)$.
+often summarized as the tuple $(p,c,x,y)$.
 By ``describes the history'',
@@
 
-So, an ARG can be thought of either as a collection of inheritance relationships
+\begin{figure}[!ht]
+\centering
+% JK: I put in 0.9 here because the fig is too big and takes up more than 
+% a page. This flushes all the figs to the end of the document. We should 
+% either make it a bit smaller or reduce the caption.
+\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
+\caption{ARG (top) generated under the SMC and its representation as a
+series of local trees (bottom). To ensure the one-to-one correspondance
+between both representations a node is recorded for each local tree whenever
+a node is encoundered along its ancestry path in the graph representation.
+This results in nodes that (locally) have only a single child, and are therefore
+(locally) unary. Blue (unary) nodes represent recombinant nodes.
+The blue arrows shows the familiar left-to-right logic of the SMC whereby
+the floating lineage coalesces randomly
+with a the remaining portion of the local tree following recombination.
+Note how the presence of the unary node on a local tree is
+indicative of either a past (to the left of the local tree)
+or future (to the right) coalescence event.
+Both ARG representations can be simplified by removing the recombinant nodes.
+In that case the information we have on the time to each of these events is
+limited to the windows (blue shaded array) delimited by by the age of
+the child node $c$ associated with that lineage and the minimum age of
+the parents of the edges connecting to $c$.
+}
+\label{fig:smc-unary}
+\end{figure}
+
+An ARG can therefore be thought of either as a collection of inheritance relationships
 between ancestral haplotypes,
@@
 Other nodes (labeled by black numbers) are either samples
-or \emph{coalescence} nodes --
+or \emph{common ancestry} nodes --
 ancestors that are the most recent common ancestor for two samples
@@
 nodes that only have a single child along one or more local trees --
-i.e., are \emph{locally unary}.
+i.e., are \emph{locally unary}~\citep{wong_general_2023}.
 Recombinant nodes (in blue) have a single child in all trees.
-However, coalescent nodes can be locally unary as well:
+However, common ancestry nodes can be locally unary as well:
 for instance, sample 3 inherited a longer stretch of genome from ancestor 5
@@
 or it already has been:
-looking at the second tree, we know node 5 will be coalescent further down the sequence
+looking at the second tree (Fig.\ref{fig:smc-unary}), we know node 5 will be coalescent further down the sequence
 because it was not present in the first tree.
@@
 However, removing these unary portions of nodes seems odd
-when one considers the consequences in the haplotype inheritance graph
-of the top figure.
+when one considers the consequences of the inheritance paths through the
+ARG in the top figure.
 As pictured, both figures tell us that
@@
 
-Furthermore, these unary nodes enable
+Furthermore, these locally unary nodes enable
 us to uniquely identify all lineages that were hit by a recombination event,
@@
 \kwarg{} \citep{ignatieva_kwarg_2021}
+[TODO ADD SINGER]
 estimate ARGs including unary portions of coalescence nodes.
@@
 
-In the next section we will derive an expression to compute the
-likelihood for an ARG under the SMC for which we have no information on the
-time to any of the recombination events. We will then detail an algorithm
-that allows us to efficiently compute this value by iterating over the local trees.
-To do this, we will rely on the unary nodes
-to identify each lineage involved in a recombination event
-as described in this section.
-
-\begin{figure}[!ht]
-\centering
-\includegraphics[width=\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
-\caption{ARG (top) generated under the SMC and its representation as a
-series of local trees (bottom). To ensure the one-to-one correspondance
-between both representations a node is recorded for each local tree whenever
-a node is encoundered along its ancestry path in the graph representation.
-This results in nodes that (locally) have only a single child, and are therefore
-(locally) unary. Blue (unary) nodes represent recombinant nodes.
-The blue arrows shows the familiar left-to-right logic of the SMC whereby
-the floating lineage coalesces randomly
-with a the remaining portion of the local tree following recombination.
-Note how the presence of the unary node on a local tree is
-indicative of either a past (to the left of the local tree)
-or future (to the right) coalescence event.
-Both ARG representations can be simplified by removing the recombinant nodes.
-In that case the information we have on the time to each of these events is
-limited to the windows (blue shaded array) delimited by by the age of
-the child node $c$ associated with that lineage and the minimum age of
-the parents of the edges connecting to $c$.
-}
-\label{fig:smc-unary}
-\end{figure}
-
 % % % % % % % % % % % % % % % % % %
@@
 
-The process of recombination only affects one lineage at a time
-(and does not even depend on the state of other lineages).
-On the other hand, coalescence events necessarily involve more than one lineage.
-However, the asymmetric way we define coalescence above
-allows us to uniquely associate each event with a single lineage,
-which allows us to cleanly factor the likelihood across edges and nodes.
-The data structure we will compute the likelihood of
-is a collection of \emph{edges} $E$,
-where each edge $e$ is of the form $e = (p,c,x,y)$.
-Recall that this means that node $c$ inherited genetic material from node $p$
-over a region with span from $x$ to $y$ along the genome
-(see Figure~\ref{fig:likelihood}).
-So, $x$ and $y$ are genomic coordinates,
-while $c$ and $p$ are in $N$, the set of \emph{nodes}
-(i.e., distinct sampled or ancestral haplotypes).
+Here we will compute the likelihood of an ARG in the absence of the
+exact time to recombination events. To do this, we will rely on the unary nodes
+to identify each lineage involved in a recombination event as described in \ref{par:recording}.
 
@@
 integrated over the entire process
-(also called the ``total hazard''). \comment{if we want to spell it out even further:
-Note that the time to the next event is distributed as $\Lambda \exp(-\Lambda)$. And
-the probablility that the next event is of type $i$ is $\lambda_i / \Lambda$.}
-
-The contributions to the likelihood from coalescent and recombination events
-is straightforward, since each possible coalescence is associated with rate $\lambda$,
-and each recombination with rate $r$.
-Under the SMC, we only ever have two lineages coalescing at a time;
-however, inferred ARGs may have more-than-binary mergers.
-We treat these as if there were zero-length edges
-between a sequence of mergers;
-the order of these unresolved mergers does not affect the likelihood.
-(Under the SMC the likelihood of a zero-length edge is zero;
-but this is not special: since edge lengths are continuously distributed,
-the likelihood of any particular value is also zero.)
-Let $\deg_o(n) = \#\{e \in E : p_e=n\}$ be the out-degree of a node $n$,
+(also called the ``total hazard''). 
+%\comment{if we want to spell it out even further:
+%Note that the time to the next event is distributed as $\Lambda \exp(-\Lambda)$. And
+%the probablility that the next event is of type $i$ is $\lambda_i / \Lambda$.}
+
+%where each edge $e$ is of the form $e = (p,c,x,y)$.
+%Recall that this means that node $c$ inherited genetic material from node $p$
+%over a region with span from $x$ to $y$ along the genome
+%(see Figure~\ref{fig:likelihood}).
+%So, $x$ and $y$ are genomic coordinates,
+%while $c$ and $p$ are in $N$ (i.e., distinct sampled or ancestral haplotypes).
+
+%Because the process of recombination only affects one lineage at a time
+%(and does not even depend on the state of other lineages).
+%On the other hand, common  events necessarily involve more than one lineage.
+%However, the asymmetric way we define common ancestry events above
+%allows us to uniquely associate each event with a single lineage,
+%which allows us to cleanly factor the likelihood across edges and nodes.
+% latter realized events, factor hazard across the edges.
+
+The structure of the ARG, as a collection of \emph{edges} $E$ and \emph{nodes} $N$,
+as well as the definition of both recombination and common ancestry events, allows us
+to cleanly factor the likelihood of the ARG across its edges and nodes.
+Not only does recombination
+naturally only affect one lineage at a time without even depending
+on the state of other lineages. Common ancestry events can be equally
+associated with a single edge given definition \ref{def:coal}.
+The next paragraphs detail precisely how the number and the nature of the realized events
+can be determined based on the number of
+parents and children of each of the nodes in $N$ as well as how the total hazard can be
+factorized into the individual contributions of each of the edges.
+
+More specifically,
+let $\deg_o(n) = \#\{e \in E : p_e=n\}$ be the out-degree of a node $n$,
 i.e., the number of edges having $n$ as a parent.
 Since mergers are binary in the SMC,
-this must have been the result of $\deg(n) - 1$ coalescences,
+this must have been the result of $\deg(n) - 1$ common ancestry events,
 and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood
 for nodes with any children.
+Note however that this description is general and holds for the case of
+more-than-binary mergers. Indeed, inferred ARGs (e.g.\ \texttt{tsinfer}) may
+contain polytomies. By treating these as if there were zero-length edges
+between a sequence of mergers; the order of these unresolved mergers does 
+not affect the likelihood. Under the SMC the likelihood of a zero-length edge is zero.
+This is not special: since edge lengths are continuously distributed,
+the likelihood of any particular value is also zero.
+
 Similarly,

### 2024-04-25 paper.tex
Update notation.

@@
 
-First, we define the SMC process in a ``backwards-in-time'' way,
+First, 
+we describe
+in more detail what sort of information we have in estimated ARGs.
+Then,
+we define the SMC process in a ``backwards-in-time'' way,
 rather than the usual ``along-the-genome'' way.
@@
 reiteration of the original work of \citet{mcvean_approximating_2005}.
-Then, we describe
-in more detail what sort of information we have in estimated ARGs.
+
 In the subsequent section we connect the two and
@@
 % % % % % % % % % % % % % % % % % %
-\subsection{SMC backwards-in-time}\label{par:description}
-
-In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a
-simple modification to the coalescent with recombination.
-Instead of allowing any pair of lineages to coalesce,
-the SMC is the process in which only lineage with \emph{overlapping} genomic intervals may coalesce.
-Surprisingly, although this is defined in terms of a backwards-in-time Markov process,
-the resulting tree-valued process along the genome is also Markov.
-Although \citet{mcvean_approximating_2005} verbally described the SMC 
-using this backwards-in-time description, their analysis
-was almost entirely in left-to-right terms,
-and subsequent papers followed this trend 
-\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,rasmussen_genome-wide_2014}.
-
-We now define the process more formally,
-and following the notation of \citet{mcvean_approximating_2005}.
-At any point in time, the state of the process is $L(t)$,
-the set of labeled lineages extant at time $t$,
-where each ``lineage'' $X_j(t)$ is labeled by an integer $j$
-and represented by a half-open interval delimiting the
-genetic material inherited by either one of a set of samples.
-This ancestral material \citep{wiuf_ancestry_1999} carried by a
-lineage can thus be summarized as $X_j(t) = [x_{j}, y_{j})$,
-for some interval endpoints $x_{j} < y_{j}$.
-Since we look backwards in time, $t=0$ is today, and so
-$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
-each represented by a single interval spanning the entire genome.
-Then, the process evolves by a succession of common ancestor and recombination
-events until each segment of ancestral material is only present in one lineage.
-The waiting time to next event is determined by these two competing processes
-with exponentially distributed waiting times as outlined below.
-
-Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
-so, if $T(t) = \sum_i |X_j(t)|$ is the sum of the total sizes of all lineages,
-then recombination occurs at instantaneous rate $rT(t)$.
-A recombination to the left of $x$
-(i.e., between base pairs $x$ and $x-1$)
-with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
-which each are given new, unique labels
-(and lineage $j$ is removed).
-This operation keeps the total amount of ancestral material unchanged.
-
-Assuming a randomly mating, diploid population of constant size $N_e$,
-common ancestry occurs between
-two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
-The newly formed lineage acquires the union of both intervals, % L is updated accordingly
-and the original lineages are removed.
-As a consequence, under the SMC common ancestry events are always
-associated with coalescence events
-along some fraction of the merging half-open intervals.
-
-Although coalescence is reciprocal, here we'll define a strict total order
-on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
-For any such strict total order,
-the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
-i.e., $\lambda \sum_{i \neq j} I_{jk}$,
-where
-\begin{equation} \label{def:coal}
-I_{jk} = \begin{cases}
-    1 & X_j > X_k \text{ and } y_k > x_j \\
-    0 & \text{otherwise.}
-\end{cases}
-\end{equation}
-We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$. From the
-definition of the strict total order it follows that
-if $I_{jk} = 1$ then $I_{kj} = 0$).
-
-For future use, define $C_j(t)$ to be the set of lineages that can coalesce into lineage $j$,
-i.e., $C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$. And therefore, 
-$|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
-and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$.
-Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
-which will simplify the likelihood computations.
-In particular,
-although recombination affects \emph{which} lineages can coalesce into $j$,
-it does not change the \emph{number} of such lineages:
-in other words, a recombination event changes $C_j(t)$ but not $I_{j}(t)$.
-Another useful quantity to know is the number of lineages carrying material
-at each point on the genome:
-for a location $z$ and time $t$, this is
-\begin{equation}
-    N(t,z) = \vert\{j : z \in X_j(t) \} \vert
-\end{equation}
-
-After any coalescence or recombination event,
-the segment(s) $[x_j,y_j)$ (and $[x_k, y_k]$) carried by the involved lineage(s)
-are replaced by a single segment $[x, y)$ whose endpoints have not fully
-coalesced, i.e., $x = \min\{ z \ge \min(x_j, x_k) : N(t,z) > 1\}$,
-and $y = \sup\{ z \le max(y_j, y_k) : N(t,z) > 1\}$.
-By reasoning about the process in this way we can keep representing 
-all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
-and the process terminates.
-
-[TO DO: remark on  beyond local roots of the genealogy]
-
-%The process of recombination and common ancestry events continues 
-%Finally, there is one more operation that allows the process to terminate:
-%portions of the edges of lineages that have entirely coalesced
-%(i.e., that are ancestral to all samples) are removed.
-
-%Formally, after any coalescence or recombination event,
-%the segment $[x,y)$ carried by the lineage
-%is replaced by the largest (pair of) segment(s) whose endpoints have not
-%coalesced, i.e., $x$ is replaced by $\min\{ z \ge x : N(t,z) > 1\}$,
-%and $y$ is replaced by $\sup\{ z \le y : N(t,z) > 1\}$.
-%By reasoning about the process in this way we can keep representing 
-%all lineages by means of a single half-open interval throughout.
-
-% % % % % % % % % % % % % % % % % %
 \subsection{ARGs} \label{par:recording}
@@
 An ARG of the sort we consider describes the history of inheritance
-of the genomic material of a focal set of individuals (that we call \emph{samples}).
+of the genomic material of a focal set of individuals, or \emph{samples}.
 Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
@@
 \centering
-% JK: I put in 0.9 here because the fig is too big and takes up more than 
-% a page. This flushes all the figs to the end of the document. We should 
-% either make it a bit smaller or reduce the caption.
 \includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
-\caption{ARG (top) generated under the SMC and its representation as a
-series of local trees (bottom). To ensure the one-to-one correspondance
+\caption{ARG (top), generated under the SMC, and its representation as a
+series of local trees (bottom). To ensure the one-to-one correspondence
 between both representations a node is recorded for each local tree whenever
-a node is encoundered along its ancestry path in the graph representation.
+a node is encountered along its ancestry path in the graph representation.
 This results in nodes that (locally) have only a single child, and are therefore
 (locally) unary. Blue (unary) nodes represent recombinant nodes.
-The blue arrows shows the familiar left-to-right logic of the SMC whereby
+The blue arrows show the familiar left-to-right logic of the SMC whereby
 the floating lineage coalesces randomly
 with a the remaining portion of the local tree following recombination.
-Note how the presence of the unary node on a local tree is
-indicative of either a past (to the left of the local tree)
-or future (to the right) coalescence event.
-Both ARG representations can be simplified by removing the recombinant nodes.
-In that case the information we have on the time to each of these events is
-limited to the windows (blue shaded array) delimited by by the age of
-the child node $c$ associated with that lineage and the minimum age of
-the parents of the edges connecting to $c$.
+By removing recombinant nodes, we limit the information on the time to each
+of these events to the windows (blue shaded array) delimited by by the age of
+the child node $c$ and the minimum age of the parents of the edges connecting to $c$.
 }
@@
 
-An ARG can therefore be thought of either as a collection of inheritance relationships
-between ancestral haplotypes,
+An ARG can therefore be thought of either as a collection of inheritance
+relationships between ancestral haplotypes,
 or as a sequence of genealogical trees that have common node labels.
@@
 
-A key observation of \citet{mcvean_approximating_2005}
-was that the sequence of trees along the genome
-generated by the SMC was Markovian \emph{after removing all unary nodes}.
-In other words, if we remove recombinant nodes entirely,
-and erase the unary portions of coalescent nodes,
-then the distribution of the next tree can be generated only knowing the current tree.
-(To see that the unary nodes makes the process non-Markov,
-consider that a node in a tree might be unary either because it's about to be a coalescent node
-or it already has been:
-looking at the second tree (Fig.\ref{fig:smc-unary}), we know node 5 will be coalescent further down the sequence
-because it was not present in the first tree.
-So, the first tree gives us information about subsequent trees
-that the second tree alone could not.)
-However, removing these unary portions of nodes seems odd
-when one considers the consequences of the inheritance paths through the
-ARG in the top figure.
-As pictured, both figures tell us that
-3 inherits the middle bit of the genome from 7 through 5.
-However, if we erase 5 from the second tree,
-does this suggest that 3 inherited a single segment from 7 along two distinct paths?
-The haplotype inheritance graph above makes it clear that this is unlikely:
-it is in fact reasonable to leave in these unary sections.
-
-Furthermore, these locally unary nodes enable
+%% Noved this section to section on smc
+% A key observation of \citet{mcvean_approximating_2005}
+% was that the sequence of trees along the genome
+% generated by the SMC was Markovian \emph{after removing all unary nodes}.
+% In other words, if we remove recombinant nodes entirely,
+% and erase the unary portions of coalescent nodes,
+% then the distribution of the next tree can be generated only knowing the current tree.
+% (To see that the unary nodes makes the process non-Markov,
+% consider that a node in a tree might be unary either because it's about to be a coalescent node
+% or it already has been:
+% looking at the second tree (Fig.\ref{fig:smc-unary}), we know node 5 will be coalescent further down the sequence
+% because it was not present in the first tree.
+% So, the first tree gives us information about subsequent trees
+% that the second tree alone could not.)
+% However, removing these unary portions of nodes seems odd
+% when one considers the consequences of the inheritance paths through the
+% ARG in the top figure.
+% As pictured, both figures tell us that
+% 3 inherits the middle bit of the genome from 7 through 5.
+% However, if we erase 5 from the second tree,
+% does this suggest that 3 inherited a single segment from 7 along two distinct paths?
+% The graph-view of the ARG (above) makes it clear that this is unlikely:
+% it is in fact reasonable to leave in these unary sections.
+
+These locally unary nodes enable
 us to uniquely identify all lineages that were hit by a recombination event,
@@
 Currently,
-\argweaver{} \citep{rasmussen_genome-wide_2014}
-\tsinfer{} \citep{kelleher_inferring_2019}, and
-\kwarg{} \citep{ignatieva_kwarg_2021}
-[TODO ADD SINGER]
-estimate ARGs including unary portions of coalescence nodes.
+\argweaver{} \citep{rasmussen_genome-wide_2014},
+\tsinfer{} \citep{kelleher_inferring_2019},
+\kwarg{} \citep{ignatieva_kwarg_2021}, and
+\texttt{SINGER} \citep{deng2024robust} estimate ARGs 
+including unary portions of coalescence nodes.
 However, it is unclear to what degree these are well-inferred,
 as most evaluation of inferred ARGs thus far has focused on marginal trees
-(e.g., \citet{brandt2022evaluation,kelleher_inferring_2019},
+(e.g., \citet{brandt2022evaluation, kelleher_inferring_2019},
 but see \citet{deng2021distribution})
@@
 % % % % % % % % % % % % % % % % % %
+\subsection{SMC backwards-in-time}\label{par:description}
+
+In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a
+simple modification to the coalescent with recombination.
+Instead of allowing any pair of lineages to coalesce,
+the SMC is the process in which only lineages with \emph{overlapping} genomic intervals may coalesce.
+By limiting common ancestry events to overlapping pairs of lineages, we are
+guaranteed that all nodes in the ARG only carry ancestral material, this is
+genetic material that is inherited by either one of a set of samples \citep{wiuf_ancestry_1999}.
+As such, we only require a single half-open interval to describe the ancestral
+material associated with each common ancestor node in the ARG.
+Surprisingly, although the SMC can be defined
+in terms of a backwards-in-time Markov process,
+the resulting tree-valued process along the genome is also Markov.
+Linking this back to relationship between a graph and local trees view on
+the ARG, then the distribution of the next tree within the series
+of local trees can be generated only knowing the current tree,
+provided that we remove recombinant nodes entirely,
+and erase the unary portions of coalescent nodes.
+(To see that the locally unary nodes make the process non-Markov,
+consider that a node in a tree might be unary either because it's about to be a coalescent node
+or it already has been:
+looking at the second tree (Fig.\ref{fig:smc-unary}), we know node 5 will be coalescent further down the sequence
+because it was not present in the first tree.
+So, the first tree gives us information about subsequent trees
+that the second tree alone could not.)
+Although \citet{mcvean_approximating_2005} verbally described the SMC
+using this backwards-in-time description, their analysis
+was almost entirely in left-to-right terms,
+and subsequent papers followed this trend
+\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,rasmussen_genome-wide_2014}.
+
+We now define the process more formally,
+and following the notation of \citet{mcvean_approximating_2005}.
+At any point in time, the state of the process is $L(t)$,
+the set of labeled lineages extant at time $t$,
+where each lineage $X_j(t)$ is labeled by an integer $j$
+and represented by a half-open interval delimiting the
+ancestral material. Any lineage can thus be summarized as $X_j(t) = [x_{j}, y_{j})$,
+for some interval endpoints $x_{j} < y_{j}$.
+Since we look backwards in time, $t=0$ is today, and so
+$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
+each represented by a single interval spanning the entire genome.
+Then, the process evolves by a succession of common ancestor and recombination
+events until each segment of ancestral material is only present in one lineage.
+The waiting time to next event is determined by these two competing processes
+with exponentially distributed waiting times as outlined below.
+
+Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
+so, if $T(t) = \sum_i |X_j(t)|$ is the sum of the total sizes of all lineages,
+then recombination occurs at instantaneous rate $rT(t)$.
+A recombination to the left of $x$
+(i.e., between base pairs $x$ and $x-1$)
+with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
+which each are given new, unique labels and added to $L(t)$
+(and lineage $j$ is removed).
+This operation keeps the total amount of ancestral material unchanged.
+
+Assuming a randomly mating, diploid population of constant size $N_e$,
+common ancestry occurs between
+two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
+The newly formed lineage acquires the union of both intervals,
+the original lineages are removed, and $L(t)$ is updated accordingly.
+
+Although coalescence is reciprocal, here we'll define a strict total order
+on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
+For any such strict total order,
+the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
+i.e., $\lambda \sum_{i \neq j} I_{jk}$,
+where
+\begin{equation} \label{def:coal}
+I_{jk} = \begin{cases}
+    1 & X_j > X_k \text{ and } y_k > x_j \\
+    0 & \text{otherwise.}
+\end{cases}
+\end{equation}
+We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$. From the
+definition of the strict total order it follows that
+if $I_{jk} = 1$ then $I_{kj} = 0$.
+
+For future use, define $C_j(t)$ to be the set of lineages that can coalesce into lineage $j$,
+i.e., $C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$. And therefore, 
+$|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
+and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$.
+Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
+which will simplify the likelihood computations.
+In particular,
+although recombination affects \emph{which} lineages can coalesce into $j$,
+it does not change the \emph{number} of such lineages:
+in other words, a recombination event changes $C_j(t)$ but not $I_{j}(t)$.
+
+%% I have updated the description of the process below to avoid having to
+% define yet another symbol.
+%Another useful quantity to know is the number of lineages carrying material
+%at each point on the genome:
+%for a location $z$ and time $t$, this is
+%\begin{equation}
+%    N(t,z) = \vert\{j : z \in X_j(t) \} \vert
+%\end{equation}
+
+To ensure that the ancestral material carried by a single lineage can at
+all time be represented by a single half-open interval, we only update the
+endpoints of the involved lineages following common ancestry or recombination
+such that only the minimal half-open interval whose endpoints have not fully coalesced
+is used to update $L(t)$. Note that empty intervals are not added to the state.
+%By reasoning about the process in this way we can keep representing 
+% all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
+% and the process terminates.
+
+%After any coalescence or recombination event,
+%the segment(s) $[x_j,y_j)$ (and $[x_k, y_k]$) carried by the involved lineage(s)
+%are replaced by a single segment $[x, y)$ whose endpoints have not fully
+%coalesced, i.e., $x = \min\{ z \ge \min(x_j, x_k) : N(t,z) > 1\}$,
+%and $y = \sup\{ z \le max(y_j, y_k) : N(t,z) > 1\}$. Note that empty intervals
+%are not added back to $L(t)$.
+
+Finally, by describing the SMC
+backwards in time, some intervals contained within the half-open
+interval representing a lineage might have fully coalesced before all the genetic
+material carried by that lineage has. As noted by \citet{mcvean_approximating_2005},
+such intervals can be observed as the stalks on the marginal genealogies (e.g.\ marginal trees
+for $[25, 81)$ Fig. \ref{fig:smc-unary}). Recombination events occurring within these smaller
+intervals however do not affect the marginal genealogies.
+
+%The process of recombination and common ancestry events continues 
+%Finally, there is one more operation that allows the process to terminate:
+%portions of the edges of lineages that have entirely coalesced
+%(i.e., that are ancestral to all samples) are removed.
+
+%Formally, after any coalescence or recombination event,
+%the segment $[x,y)$ carried by the lineage
+%is replaced by the largest (pair of) segment(s) whose endpoints have not
+%coalesced, i.e., $x$ is replaced by $\min\{ z \ge x : N(t,z) > 1\}$,
+%and $y$ is replaced by $\sup\{ z \le y : N(t,z) > 1\}$.
+%By reasoning about the process in this way we can keep representing 
+%all lineages by means of a single half-open interval throughout.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Calculating likelihoods} \label{par:liks}
@@
 More specifically,
-let $\deg_o(n) = \#\{e \in E : p_e=n\}$ be the out-degree of a node $n$,
+let $\deg_o(n) = \vert\{e \in E : p_e=n\}\vert$ be the out-degree of a node $n$,
 i.e., the number of edges having $n$ as a parent.
@@
 Similarly,
-let $\deg_i(n) = \#\{e \in E : c_e=n\}$ be the in-degree of a node $n$,
+let $\deg_i(n) = \vert\{e \in E : c_e=n\}\vert$ be the in-degree of a node $n$,
 i.e., the number of edges having $n$ as a child.
@@
     \centering
-    \includegraphics[width=\textwidth]{figures/single_event.png}
+    \includegraphics[width=0.65\textwidth]{figures/single_event.png}
     \caption{
-        Proposed figure to explain the likelihood computation with.
-        TODO: change numbered labels to match the order in the text,
-        add $x$ and $y$.
+        Graphical representation of the events associated with an edge $e(p,c,x_c,y)$.
+        The lineage carried by $c$ spans $[x,y)$ until a recombination event splits it
+        into two segments. The right hand segment coalesces at time $t_p$. The time to
+        recombination is thus bounded by $(c. \min(t_p, t_{p'})$.
         \label{fig:likelihood}
@@
 the probability of which is
-$\exp(-r \times A)$, where $r$ is the recombination rate
-and $A$ is the total area (eligbile links times depth) of the edge.
+$\exp(-r \times S)$, where $r$ is the recombination rate
+and $\mathcal{A}$ is the total area (eligbile links times depth) of the edge.
 So, the first contribution to the likelihood from this edge $e$ is

### 2024-04-25 paper.tex
update process description

@@
 \includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
-\caption{ARG (top), generated under the SMC, and its representation as a
+\caption{ARG generated under the SMC, represented as a graph (top) and as a
 series of local trees (bottom). To ensure the one-to-one correspondence
@@
 This results in nodes that (locally) have only a single child, and are therefore
-(locally) unary. Blue (unary) nodes represent recombinant nodes.
-The blue arrows show the familiar left-to-right logic of the SMC whereby
+(locally) unary. By removing recombinant nodes, we limit the information on the time to each
+of these events to the windows (blue shaded array) delimited by by the age of
+the child node $c$ and the minimum age of the parents of the edges connecting to $c$.
+Blue (unary) nodes represent recombinant nodes. The blue arrows show the
+familiar left-to-right logic of the SMC whereby
 the floating lineage coalesces randomly
 with a the remaining portion of the local tree following recombination.
-By removing recombinant nodes, we limit the information on the time to each
-of these events to the windows (blue shaded array) delimited by by the age of
-the child node $c$ and the minimum age of the parents of the edges connecting to $c$.
 }
@@
 By limiting common ancestry events to overlapping pairs of lineages, we are
-guaranteed that all nodes in the ARG only carry ancestral material, this is
+guaranteed that all nodes in the ARG only carry ancestral material, this is,
 genetic material that is inherited by either one of a set of samples \citep{wiuf_ancestry_1999}.
@@
 material associated with each common ancestor node in the ARG.
-Surprisingly, although the SMC can be defined
-in terms of a backwards-in-time Markov process,
-the resulting tree-valued process along the genome is also Markov.
-Linking this back to relationship between a graph and local trees view on
-the ARG, then the distribution of the next tree within the series
-of local trees can be generated only knowing the current tree,
-provided that we remove recombinant nodes entirely,
-and erase the unary portions of coalescent nodes.
-(To see that the locally unary nodes make the process non-Markov,
-consider that a node in a tree might be unary either because it's about to be a coalescent node
-or it already has been:
-looking at the second tree (Fig.\ref{fig:smc-unary}), we know node 5 will be coalescent further down the sequence
-because it was not present in the first tree.
-So, the first tree gives us information about subsequent trees
-that the second tree alone could not.)
 Although \citet{mcvean_approximating_2005} verbally described the SMC
@@
 \citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,rasmussen_genome-wide_2014}.
+Indeed, the resulting tree-valued process along the genome is also Markov.
+Linking this back to the relationship between the ``graph'' and
+``local trees'' representation of
+the ARG, this can be formulated as follows: the distribution of the
+next tree within the series
+of local trees can be generated only knowing the current tree,
+provided that we remove recombinant nodes entirely,
+and erase the unary portions of coalescent nodes.
+To see that the locally unary nodes make the process non-Markov,
+consider that a node in a tree might be unary either because
+it's about to be a coalescent node or it already has been:
+looking at the second tree (Fig.\ref{fig:smc-unary}),
+we know node 5 will be coalescent further down the sequence
+because it was not present in the first tree.
+So, the first tree gives us information about subsequent trees
+that the second tree alone could not.
 
@@
 all time be represented by a single half-open interval, we only update the
-endpoints of the involved lineages following common ancestry or recombination
+endpoints of the involved lineages following common ancestry or recombination events
 such that only the minimal half-open interval whose endpoints have not fully coalesced
@@
     \caption{
-        Graphical representation of the events associated with an edge $e(p,c,x_c,y)$.
-        The lineage carried by $c$ spans $[x,y)$ until a recombination event splits it
+        Graphical representation of the events associated with an edge $(p,c,x,y)$.
+        The lineage carried by $c$ spans $[x',y)$ until a recombination event splits it
         into two segments. The right hand segment coalesces at time $t_p$. The time to
@@
 the probability of which is
-$\exp(-r \times S)$, where $r$ is the recombination rate
+$\exp(-r \times \mathcal{A})$, where $r$ is the recombination rate
 and $\mathcal{A}$ is the total area (eligbile links times depth) of the edge.

### 2024-04-25 paper.tex
update figure

@@
     \caption{
-        Graphical representation of the events associated with an edge $(p,c,x,y)$.
-        The lineage carried by $c$ spans $[x',y)$ until a recombination event splits it
+        Graphical representation of the events associated with the edge $(p,c,x,y)$.
+        The lineage carried by $c$ spans $[x',y)$ until a recombination event (blue dotted line) splits it
         into two segments. The right hand segment coalesces at time $t_p$. The time to
@@
 and the lineage that directly inherited from $p$
-had the segment $[x,y)$.
+had the segment $[x,y)$ (see fig.\ \ref{fig:likelihood}).
 Under the SMC,
@@
 \end{equation}
-Note that if a recombination did occurr, then the number of links not
-affected by recombination is 1 less (see \ref{eq:depth}).
-
-Finally, we turn to the hazard associated with coalescence.
-To do this, we need only consider what happens along the left side
-of each edge.
-The instantaneous rate of coalescence of a lineage whose left endpoint is at $x$
-at a given time is equal to $\lambda$ multiplied
-by the number of earlier lineages that overlap $x$ at that time.
+Note that if a recombination did occur, then the number of links not
+affected by recombination is 1 less (see eq.\ \ref{eq:depth}).
+
+Finally, we turn to the hazard associated with common ancestry events.
+To do this,
+we need only consider what happens along the left side of each edge.
 Recall that we define a total order on edges,
@@
 if $x<x'$, or if $x=x'$ and $c<c'$).
+The instantaneous rate of coalescence of a lineage whose left endpoint is at $x$
+at a given time is equal to $\lambda$ multiplied
+by the number of earlier lineages that overlap $x$ at that time.
 Using this total order, we can define
@@
 and are present at time $t$.
-Then, the cumulative hazard for coalescence of edge $e$
+Then, the cumulative hazard for common ancestry events of edge $e$
 between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(u) du$.
@@
 is the leftmost segment of material inherited by node $c$
-(e.g., the lefthand segment ancestral to node $c$
-in Figure~\ref{fig:likelihood}),
+(e.g., the left hand segment ancestral to node $c$
+in fig.\ ~\ref{fig:likelihood}),
 then no recombination occurred along this edge and 
 this edge simply contributes $\exp(-F(e,t_c,t_p))$ to the likelihood.
-However, if the edge is \emph{not} the leftmost segment,
+However, if the edge is \emph{not} the leftmost segment
+(as shown in fig.\ \ref{fig:likelihood}),
 then it was initiated by a new recombination breakpoint,
@@
 track of $I_e(t)$ by means of existing \texttt{tskit} functionality.
-Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate)
+Note that we can even accommodate for stepwise changes in the coalescence rate (or recombination rate)
 through time by intersecting the internode intervals with the rate change intervals. 
@@
     =
-    \frac{1}{\mathcal{M}!} \prod_{(c, p) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}} .
+    \frac{1}{\mathcal{M}!} \prod_{(c, p) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
 \end{equation}

### 2024-04-26 paper.tex
update smc section

@@
         Graphical representation of the events associated with the edge $(p,c,x,y)$.
-        The lineage carried by $c$ spans $[x',y)$ until a recombination event (blue dotted line) splits it
-        into two segments. The right hand segment coalesces at time $t_p$. The time to
-        recombination is thus bounded by $(c. \min(t_p, t_{p'})$.
+        The lineage carried by $c$ spans $[x_c,y)$ until a recombination event (blue dotted line) splits it
+        into two segments. Left and right hand segments coalesce at times $t_p'$ and $t_p$ respectively,
+        bounding the time to recombination by $(c. \min(t_p, t_{p'})$.
+        Note that the contribution to the recombination hazard by edge $(p,c,x,y)$ only covers ``half'' of the
+        green shaded area. The other half is contributed by edge $(p', c, x_c, x)$. 
         \label{fig:likelihood}

### 2024-05-09 paper.tex
soft transition to SMC

@@
 
-In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a
+% coalescent introduction: backwards-in-time
+% left-to-right: Wiuf-Hein (difficult)
+% SMC: although easy to reason about left-to-right sense
+% 
+The coalescent process describes the ancestry of a set of sampled genomes.
+The original algorithm 
+to simulate the coalescent with
+recombination was formulated backwards in time
+by Hudson \citep{hudson_properties_1983}. 
+Wiuf and Hein later described
+a tree-by-tree method to simulate the same stochastic process 
+\citep{wiuf_recombination_1999}.
+Such an along-the-genome approach is considerably more complex as the next 
+tree to be simulated depends on all previous trees.
+This idea however provided the basis of the 
+SMC \citep{mcvean_approximating_2005}, restricting the set of all state
+space transitions possible under the Hudson coalescent to
+obtain a process that is Markovian both backwards-in-time, 
+as well as along-the-genome.
+In its backwards-in-time formulation, the SMC only requires a
 simple modification to the coalescent with recombination.
 Instead of allowing any pair of lineages to coalesce,
-the SMC is the process in which only lineages with \emph{overlapping} genomic intervals may coalesce.
-By limiting common ancestry events to overlapping pairs of lineages, we are
-guaranteed that all nodes in the ARG only carry ancestral material, this is,
-genetic material that is inherited by either one of a set of samples \citep{wiuf_ancestry_1999}.
-As such, we only require a single half-open interval to describe the ancestral
-material associated with each common ancestor node in the ARG.
+the SMC is the process in which only lineages with \emph{overlapping} 
+genomic intervals may coalesce. 
 Although \citet{mcvean_approximating_2005} verbally described the SMC
@@
 \citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,rasmussen_genome-wide_2014}.
-Indeed, the resulting tree-valued process along the genome is also Markov.
-Linking this back to the relationship between the ``graph'' and
+Linking the resulting tree-valued process along the genome back to
+the relationship between the ``graph'' and
 ``local trees'' representation of
-the ARG, this can be formulated as follows: the distribution of the
-next tree within the series
+the ARG, the Markovian property can be formulated as follows: 
+the distribution of the next tree within the series
 of local trees can be generated only knowing the current tree,
@@
 that the second tree alone could not.
-
-We now define the process more formally,
-and following the notation of \citet{mcvean_approximating_2005}.
+Further note that by limiting common ancestry events to overlapping 
+pairs of lineages, we are guaranteed that all nodes in the ARG 
+only carry ancestral material, this is,
+genetic material that is inherited by either one of a set of samples
+\citep{wiuf_ancestry_1999}.
+As such, we only require a single half-open interval to describe
+the ancestral
+material associated with each common ancestor node in the ARG.
+
+We now define the process more formally, by reintroducing
+the notation of \citet{mcvean_approximating_2005} required to describe
+the likelihood computation. This section
+does not deviate from the original description of the SMC apart from
+\textit{how} coalescable pairs are counted. This subtle nuance
+greatly simplifies the ability to keep track of coalescable pairs going
+backwards in time.
 At any point in time, the state of the process is $L(t)$,

### 2024-05-09 paper.tex
arg nodes to letters

@@
 of these events to the windows (blue shaded array) delimited by by the age of
-the child node $c$ and the minimum age of the parents of the edges connecting to $c$.
+the child node and the minimum age of the parents of the edges connecting to that child node.
 Blue (unary) nodes represent recombinant nodes. The blue arrows show the
@@
 Both structures represent the same information:
-for instance, the fact that the ancestral genome at $a$
-inherited the left half of its genome from ancestor 6
-and the right half from ancestor 5
-is seen in the fact that the first two trees have $a$ below 6,
-and the remaining have $a$ below 5.
-The blue dotted line from $a$ in the second tree to 5 in the third tree
+for instance, the fact that the ancestral genome at $A$
+inherited the left half of its genome from ancestor $g$
+and the right half from ancestor $f$
+is seen in the fact that the first two trees have $A$ below $g$,
+and the remaining have $A$ below $f$.
+The blue dotted line from $A$ in the second tree to $f$ in the third tree
 draws our attention to the fact that
-the path traced by $a$'s ancestry switches at that point.
-In fact, $a$ is in blue because it is a \emph{recombinant} node --
+the path traced by $A$'s ancestry switches at that point.
+In fact, $A$ is in blue because it is a \emph{recombinant} node --
 an ancestor that has a segment of genome inherited by a sample,
@@
 However, common ancestry nodes can be locally unary as well:
-for instance, sample 3 inherited a longer stretch of genome from ancestor 5
-than did sample 2.
-In the top plot, this is visible because the pink bar below node 5
+for instance, sample $d$ inherited a longer stretch of genome from ancestor $f$
+than did sample $c$.
+In the top plot, this is visible because the pink bar below node $f$
 extends further to the left than does the blue bar,
-while in the bottom plot, this is visible because node 5 is parent to 3
-in the second tree, but not to node 2, and so is unary.
+while in the bottom plot, this is visible because node $f$ is parent to $d$
+in the second tree, but not to node $c$, and so is unary.
 
@@
 even if we simplify the ARG by removing all recombinant nodes
-($a$, $b$, and $c$ in Fig.\ref{fig:smc-unary}).
+($A$, $B$, and $C$ in Fig.\ref{fig:smc-unary}).
 In the absence of these recombinant nodes, a recombination can be observed
@@
 but we do still have constraints on the time:
-for instance, we know that $c$ happened somewhere between node 5 and node 7.
+for instance, we know that $C$ happened somewhere between node $f$ and node $h$.
 More generally, a recombinant node must happen somewhere between the child node
@@
 looking at the second tree (Fig.\ref{fig:smc-unary}),
-we know node 5 will be coalescent further down the sequence
+we know node $f$ will be coalescent further down the sequence
 because it was not present in the first tree.

### 2024-07-05 paper.tex
Add in application refs

@@
 Applications using inferred ARGs as input have begun to
-appear~\citep{osmond_estimating_2021,
+appear~\citep{
+stern_approximate_2019,
+osmond_estimating_2021,
 fan_genealogical_2022,
@@
 guo_recombination-aware_2022,
+ignatieva_ongoing_2022,
+wang_complex_2022,
 zhang_biobank-scale_2023,
@@
 fan_likelihood_2023,
-huang_estimating_2024}.
+huang_estimating_2024,
+grundler_geographic_2024,
+korfmann_simultaneous_2024,
+deraje_inferring_2024}.
 Additionally, there is a rich suite of tools growing around
@@
 sample from a statistical
-% TODO add SINGER
-model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022}
+model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022,
+deng_robust_2024}
 or parsimony~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022},

### 2024-07-05 paper.tex
More updates on the intro

@@
 \section{Introduction}
+% TODO need to rewrite this as its too tskit-centric. We don't need to 
+% emphasise tskit/msprime so much here, it's not necessary and will 
+% annoy people. 
 Following recent breakthroughs in simulation and inference methods,
@@
 approximation to the coalescent with recombination
-by~\cite{mcvean_approximating_2005} was a seminal contribution,
-and has made many different inferences
+\citep{mcvean_approximating_2005} 
+and its refinement in the SMC'~\citep{marjoram_fast_2006},
+are the basis of many powerful inference 
 % TODO add some more recent stuff
-feasible~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
+methods~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
 In particular, the SMC is the basis for ARG sampling methods
-like ARGweaver~\citep{rasmussen_genome-wide_2014}, which remains the
-% TODO list other places ARGweaver has come out on top in terms of accuracy.
-gold-standard in terms of accuracy~\citep{brandt2022evaluation}
+like ARGweaver~\citep{rasmussen_genome-wide_2014} and 
+SINGER~\citep{deng_robust_2024}, which are the gold-standard 
+in terms of inference accuracy.
+ARGweaver has consistently out-performed other methods in terms 
+of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
+and, while SINGER has not been evaluated as thoroughly yet, it promises
+to be even more accurate as well as substantially more scalable 
+computationally~\citep{deng_robust_2024}.
 
 % THE PROBLEM!!! OMG it's terrible!!!
+Despite these recent advances in statistically rigorous ARG 
+inference, however, there is still a gap of some three orders 
+of magnitude between the sample-sizes that can be handled 
+(hundreds) compared to the large-scale hueristic methods
+like tsinfer and ARGneedle (hundreds of thousands).
+[TODO working through a more positive narrative here, based on
+the idea of combining the strengths of the approaches]
 Although major breakthroughs in scale have been achieved by hueristic

### 2024-07-05 paper.tex
Minor tweaks to notation

@@
 
-\begin{figure}[!ht]
+\begin{figure}
 \centering
@@
 or \emph{common ancestry} nodes --
-ancestors that are the most recent common ancestor for two samples
+ancestors that are the most recent common ancestor for two 
+or more samples
 at some place along the genome.
@@
 \kwarg{} \citep{ignatieva_kwarg_2021}, and
-\texttt{SINGER} \citep{deng2024robust} estimate ARGs 
+\texttt{SINGER} \citep{deng_robust_2024} estimate ARGs 
 including unary portions of coalescence nodes.
@@
 recombination was formulated backwards in time
-by Hudson \citep{hudson_properties_1983}. 
+by Hudson \citeyearpar{hudson_properties_1983}. 
 Wiuf and Hein later described
@@
 the SMC is the process in which only lineages with \emph{overlapping} 
-genomic intervals may coalesce. 
+genomic intervals may coalesce. (The SMC' additionally allows 
+lineages with abutting genomic intervals to coalesce.)
 Although \citet{mcvean_approximating_2005} verbally described the SMC
@@
 and subsequent papers followed this trend
-\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,rasmussen_genome-wide_2014}.
+\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,
+rasmussen_genome-wide_2014}.
 Linking the resulting tree-valued process along the genome back to
@@
 consider that a node in a tree might be unary either because
-it's about to be a coalescent node or it already has been:
-looking at the second tree (Fig.\ref{fig:smc-unary}),
-we know node $f$ will be coalescent further down the sequence
+it's about to be (in the left-to-right sense) a coalescent node 
+or it already has been:
+looking at the second tree in Fig.\ref{fig:smc-unary},
+we know node $f$ will be coalescent further along the sequence
 because it was not present in the first tree.
@@
 does not deviate from the original description of the SMC apart from
-\textit{how} coalescable pairs are counted. This subtle nuance
+how coalescable pairs are counted. This subtle nuance
 greatly simplifies the ability to keep track of coalescable pairs going
@@
 \end{equation}
-We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$. From the
-definition of the strict total order it follows that
-if $I_{jk} = 1$ then $I_{kj} = 0$.
-
-For future use, define $C_j(t)$ to be the set of lineages that can coalesce into lineage $j$,
-i.e., $C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$. And therefore, 
-$|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
+We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$.
+From the definition of the strict total order it follows that if $I_{jk} = 1$
+then $I_{kj} = 0$. For later use, also define $C_j(t)$ to be the set of
+lineages that can coalesce into lineage $j$, i.e., 
+$C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$; 
+thus, $|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
 and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$. 
 Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
-which will simplify the likelihood computations.
-In particular,
-although recombination affects \emph{which} lineages can coalesce into $j$,
-it does not change the \emph{number} of such lineages:
-in other words, a recombination event changes $C_j(t)$ but not $I_{j}(t)$.
+which will simplify the likelihood computations. In particular, although
+recombination affects \emph{which} lineages can coalesce into $j$, it does not
+change the \emph{number} of such lineages: in other words, a recombination
+event changes $C_j(t)$ but not $I_{j}(t)$.
 
@@
 
+[Do we need to get into this above-the-root subtelty here?]
 To ensure that the ancestral material carried by a single lineage can at
@@
 on the state of other lineages. Common ancestry events can be equally
-associated with a single edge given definition \ref{def:coal}.
+associated with a single edge given definition \eqref{def:coal}.
 The next paragraphs detail precisely how the number and the nature of the realized events
@@
 More specifically,
-let $\deg_o(n) = \vert\{e \in E : p_e=n\}\vert$ be the out-degree of a node $n$,
-i.e., the number of edges having $n$ as a parent.
+let $\deg_o(u) = \vert\{e \in E : p_e=u\}\vert$ be the out-degree of a node $u$,
+i.e., the number of edges having $u$ as a parent.
 Since mergers are binary in the SMC,
-this must have been the result of $\deg(n) - 1$ common ancestry events,
-and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood
+this must have been the result of $\deg(u) - 1$ common ancestry events,
+and so this contributes a factor of $\lambda^{\deg_o(u)-1}$ to the likelihood
 for nodes with any children.
@@
 contain polytomies. By treating these as if there were zero-length edges
-between a sequence of mergers; the order of these unresolved mergers does 
+between a sequence of mergers, the order of these unresolved mergers does 
 not affect the likelihood. Under the SMC the likelihood of a zero-length edge is zero.
@@
 Similarly,
-let $\deg_i(n) = \vert\{e \in E : c_e=n\}\vert$ be the in-degree of a node $n$,
-i.e., the number of edges having $n$ as a child.
+let $\deg_i(u) = \vert\{e \in E : c_e=u\}\vert$ be the in-degree of a node $u$,
+i.e., the number of edges having $u$ as a child.
 Each additional edge that is ancestral to a given node
 implies one additional ancestral recombination event,
-and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood,
+and thus a factor of  $r^{\deg_i(u)-1}$ in the likelihood,
 if the node has any parents.
@@
 \begin{equation}\label{eq:coal}
-    \prod_{n \in N} \lambda^{\deg_o(n)-1} r^{\deg_i(n)-1}  
+    \prod_{u \in N} \lambda^{\deg_o(u)-1} r^{\deg_i(u)-1}  
     =
@@
 and the lineage that directly inherited from $p$
-had the segment $[x,y)$ (see fig.\ \ref{fig:likelihood}).
+had the segment $[x,y)$ (see Fig.~\ref{fig:likelihood}).
 Under the SMC,
@@
 $\exp(-r \times \mathcal{A})$, where $r$ is the recombination rate
-and $\mathcal{A}$ is the total area (eligbile links times depth) of the edge.
+and $\mathcal{A}$ is the total area (eligible links times depth) of the edge.
 So, the first contribution to the likelihood from this edge $e$ is
@@
 Note that if a recombination did occur, then the number of links not
-affected by recombination is 1 less (see eq.\ \ref{eq:depth}).
+affected by recombination is 1 less (see eq.~\eqref{eq:depth}).
 
@@
 Then, the cumulative hazard for common ancestry events of edge $e$
-between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(u) du$.
+between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(z) dz$.
 If the edge $e = (p,c,x,y)$
@@
 However, if the edge is \emph{not} the leftmost segment
-(as shown in fig.\ \ref{fig:likelihood}),
+(as shown in Fig.~\ref{fig:likelihood}),
 then it was initiated by a new recombination breakpoint,

### 2024-07-05 paper.tex
Minor tweaks

@@
 \end{equation}
-
+One significant advantage of this formulation is that the computation time
+is linear in the number of edges, and an algorithm can be devised to 
+to efficiently keep track of $I_e(t)$ (see supplementary information 
+\ref{par:algo}).
 % Although the first two factors are easily expressed in terms of numbers of edges and nodes,
@@
 % that in practice it is convenient to absorb these terms into per-edge contributions.
-
-% refer to the algorithm in the supplements, briefly mention scalability
-This likelihood expression is not only linear in the number of edges. An algorithm
-(see supplementary information \ref{par:algo}) can be devised to efficiently keep
-track of $I_e(t)$ by means of existing \texttt{tskit} functionality.
-Note that we can even accommodate for stepwise changes in the coalescence rate (or recombination rate)
-through time by intersecting the internode intervals with the rate change intervals. 
-Rate changes along the genome can be incorporated similarly.
+We can also accommodate stepwise changes in the 
+coalescence rate (or recombination rate)
+through time by intersecting the internode intervals with the rate change
+intervals; rate changes along the genome can be incorporated similarly.
 Finally, we have only considered the topology and branch length information of any ARG.
@@
 given an ARG is given by
-
 \begin{equation}\label{eq:lik-mut}
@@
 \subsection{Flexibility}
-
 % likelihood surface
@@
 of the exact time to recombination, the integral in
-eq.\ \ref{eq:depth} means that the likelihood
+eq.~\eqref{eq:depth} means that the likelihood
 for the entire ARG does not factorize into the likelihoods of these
@@
 information we can approximate the true integral as the product of the
-integral of the slices. Figure \ref{fig:3-arg-slices} demonstrates this idea
+integral of the slices. Figure~\ref{fig:3-arg-slices} demonstrates this idea
 on a toy example. We simulated an ARG under human-like parameters with

### 2024-07-05 paper.tex
Change supplement to Appendix

@@
 
-% deal with supplementary items
-\newcommand{\supplementarysection}{%
-  \setcounter{figure}{0}% Reset figure counter
-  \let\oldthefigure\thefigure% Capture figure numbering scheme
-  \renewcommand{\thefigure}{S\oldthefigure}% Prefix figure number with S
-  \section{Supplementary section}% Set supplementary section
-}
-
 % adjust inter-paragraph spacing - this one's for you, Gertjan
-\setlength{\parskip}{0.5\baselineskip}
+% \setlength{\parskip}{0.5\baselineskip}
 
@@
 is linear in the number of edges, and an algorithm can be devised to 
-to efficiently keep track of $I_e(t)$ (see supplementary information 
-\ref{par:algo}).
+to efficiently keep track of $I_e(t)$ (see Appendix).
 % Although the first two factors are easily expressed in terms of numbers of edges and nodes,
@@
 
-\pagebreak
-
-\supplementarysection
-\section*{Supplementary Information}
+\setcounter{secnumdepth}{2} % Print out appendix section numbers
+
+\section*{Appendix}
+\appendix
+
+\setcounter{table}{0}
+\setcounter{figure}{0}
+\renewcommand{\thetable}{A\arabic{table}}
+\renewcommand{\thefigure}{A\arabic{figure}}
 
 % % % % % % % % % % % % % % % % % %
-\subsection{Algorithm} \label{par:algo}
+\subsection*{Algorithm} \label{par:algo}
+In this section we describe how to efficiently compute the likelihood under the SMC
+from an ARG stored using the \texttt{tskit} tree sequence encoding. 
+The tree sequence format encodes relationships using \emph{edges},
+along with additional information that
+allows us to sequentially recover (and reason about 
+their differences) 
+efficiently~\citep{kelleher_efficient_2016, ralph_efficiently_2020}.
+The order of the edge deletions and insertions required to move from one local tree $T_1$ to
+the next, $T_2$ imposes the strict total ordering on edges
+(see definition \eqref{def:coal})
+that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
+Indeed, following the removal of edges that do not persist in $T_2$,
+we simply declare that each subsequently added edge
+can only coalesce with the already-visited edges 
+that make up the (partially) reconstructed local tree $T_2$.
 
@@
 \includegraphics[width=\textwidth]{figures/ts_algo_2rows.png}
-\caption{Description of algorithm: Moving along the genome we transition from
+\caption{Likelihood computation algorithm. 
+Moving along the genome we transition from
 local tree $T_1$ to the next, $T_2$ in three steps. 
@@
 
-Here, we describe how to efficiently compute the likelihood under the SMC
-from an ARG stored using the \emph{tree sequence} encoding. %  (see fig. \ref{fig:smc-unary}).
-The tree sequence format encodes relationships using \emph{edges}
-just as described above,
-along with additional structure that
-allows us to sequentially recover the local trees
-along the genome very efficiently, and to reason about
-the differences between those trees \citep{kelleher_efficient_2016, ralph_efficiently_2020}.
-The order of the edge deletions and insertions required to move from one local tree $T_1$ to
-the next, $T_2$ imposes the required strict total ordering on edges
-(see definition \ref{def:coal})
-that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
-Indeed, following the removal of edges that do not persist in $T_2$,
-we simply declare that each subsequently added edge
-can only coalesce with the already-visited edges 
-that make up the (partially) reconstructed local tree $T_2$.
-
 This allows us to compute a likelihood in a single pass over all edges.
 More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
-the number of extant lineages in each internode interval (see fig \ref{fig:algo}).
-When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along the internode intervals
+the number of extant lineages in each internode interval (see Fig.~\ref{fig:algo}).
+When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along 
+the internode intervals
 spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
@@
 the last parent a child node was connected to to detect recombination events.
+
+[TODO another paragraph summarising what you actually do]
 

### 2024-07-05 paper.tex
A pass at the abstract

@@
 \begin{abstract}
-There currently is no way to connect the output of the diverse
-range of ARG inference methods back
-to a model that would allow us to perform a probabilistic exploration of the
-ARG space around the point estimate, or likelihood-based inference.
+Ancestral recombination graphs (ARGs) are the focus of much ongoing research
+interest. Recent progress in inference has made ARG-based approaches feasible
+across of range of applications, and many new methods using inferred ARGs as
+input have appeared. This progress on the long-standing problem of ARG
+inference has proceeded in two distinct directions. The first
+class of method in which dramatic improvements have been made is in the
+statistically rigorous inference of ARGs under the Sequentially Markov
+Coalescent (SMC) model, which are now practical for tens-to-hundreds of
+samples. The second class of recently-developed inference methods are based on
+approximate models and heuristics which scale to samples sizes roughly three
+orders of magnitude larger. Although these heuristic methods are reasonably
+accurate under many metrics, one major problem is that the ARGs they 
+estimate cannot easily be connected with population genetic theory.
+Specifically, because the estimated ARGs lack detail about recombination
+events, there is currently no way to compute their likelihood
+under a population genetics model. In this paper we provide a 
+backwards-time formulation of the SMC based on a total ordering of 
+lineages, and show how this leads to straightforward description 
+of the likelihood of a general class of ARG under the SMC. We show that this 
+formulation does not require precise details of recombination events
+to be estimated, and is robust to other topological issues such as 
+polytomies. This opens up many future directions, most excitingly the possibility
+of combining the strengths of both inference approaches. Applying
+these ideas may allow us to 
+make rigorous inference more scalable, and scalable inference 
+more rigorous.
 \end{abstract}
@@
 because these methods return approximate ARG
-\emph{structures}~\citep{wong_general_2023} [WEAK, FIXME]. Specifically,
+\emph{structures}~\citep{wong_general_2023} [WEAK, FIXME]. 
+Specifically,
 we cannot evaluate a likelihood of these ARGs in coalescent 

### 2024-07-10 paper.tex
Rework first paragraph

@@
 under a population genetics model. In this paper we provide a 
-backwards-time formulation of the SMC based on a total ordering of 
-lineages, and show how this leads to straightforward description 
-of the likelihood of a general class of ARG under the SMC. We show that this 
-formulation does not require precise details of recombination events
-to be estimated, and is robust to other topological issues such as 
+novel backwards-time formulation of the SMC 
+and derive a straightforward definition of the likelihood
+of the likelihood of a general class of ARG under the SMC. 
+We show that this formulation does not require precise details of recombination events
+to be estimated, and is robust to other topological inconveniences such as 
 polytomies. This opens up many future directions, most excitingly the possibility
@@
 \section{Introduction}
-% TODO need to rewrite this as its too tskit-centric. We don't need to 
-% emphasise tskit/msprime so much here, it's not necessary and will 
-% annoy people. 
 Following recent breakthroughs in simulation and inference methods,
@@
 % in the current formulation.
-Applications using inferred ARGs as input have begun to
-appear~\citep{
+Numerous applications taking advantage of this detailed history
+in inferred ARGs are now appearing~\citep{
 stern_approximate_2019,
@@
 korfmann_simultaneous_2024,
-deraje_inferring_2024}.
-Additionally, there is a rich suite of tools growing around
-the \texttt{tskit} ARG library~\citep{ralph_efficiently_2020}
-including simulation~\citep{kelleher_efficient_2016,kelleher_efficient_2018,
-adrion_community_2020,terasaki_geonomics_2021,
-baumdicker_efficient_2021,
-korfmann_weak_2022, petr_slendr_2023,tagami_tstrait_2024},
-ARG inference~\citep{kelleher_inferring_2019,speidel_method_2019,
-wohns_unified_2022,mahmoudi_bayesian_2022,
-rasmussen_espalier_2022,
-zhang_biobank-scale_2023,zhan_towards_2023} and analysis
-packages
-% etc
-\citep{nowbandegani_extremely_2023,fan_likelihood_2023}.
-Infererence methods now tend to fall into two classes: those that
-sample from a statistical
-model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022,
-deng_robust_2024}
-or parsimony~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022},
-and those based on
-heuristics~\citep{kelleher_inferring_2019,speidel_method_2019,
-zhang_biobank-scale_2023}.
+deraje_inferring_2024}
+and it seems likely that many more will
+follow~\citep{harris_database_2019,harris_using_2023}.
+Three classes of ARG inference method have emerged as the focus of recent
+work, and have all been successfully applied in different contexts. 
+Parsimony
+based methods~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022}
+seek to minimise some aspect of the inferred ARG
+(usually the number number of recombinations required), and work well for 
+answering specific detailed questions about 
+recombination~\citep[e.g.][]{ignatieva_ongoing_2022} when the expected
+number of ancestral recombinations is small (tens to hundreds).
+The second class of inference method is based on sampling ARGs statistically 
+from a population genetics model~\citep{rasmussen_genome-wide_2014,
+hubisz_mapping_2020,mahmoudi_bayesian_2022, deng_robust_2024}, and have 
+been applied in several
+contexts~\citep[e.g.][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
+stankowski_genetic_2024}. These methods are well-suited to applications 
+in which population genetics models are a good fit [TODO: Say this differently, 
+I just want to distinguish from e.g. SARS-CoV-2 where pop gen models are 
+basically meaningless]
+in cases in which the expected number of recombinations is in the 
+thousands to tens-of-thousands range (e.g., human datasets in which
+we have X samples and a segment of Y Mb of chromosome [TODO quantify
+from SINGER paper). 
+The third class of inference method and based on heuristics
+and approximate models and focus on large scale inference,
+where the expected number of recombinations is in the millions or
+tens of millions [TODO check numbers here]~\citep{kelleher_inferring_2019,
+speidel_method_2019,zhang_biobank-scale_2023}.
+Other methods specialised to particular applications have also 
+recently been proposed~\citep{zhan_towards_2023,schaefer_ancestral_2021}.
+See~\cite{wong_general_2023} for a detailed review of ARG inference 
+methods.
+
 Typically, the more rigorous approaches can handle 10-100 samples,
@@
 \cite{kuhner_maximum_2000} formulation, as these formulations assume
-that complete information about recombination events is present.
+that complete information about recombination events is present
+[Also: \cite{guo_recombination-aware_2022}]
 [NOTE the following is super weak, and needs a lot of work.]
@@
 
+% Additionally, there is a rich suite of tools growing around
+% the \texttt{tskit} ARG library~\citep{ralph_efficiently_2020}
+% including simulation~\citep{kelleher_efficient_2016,kelleher_efficient_2018,
+% adrion_community_2020,terasaki_geonomics_2021,
+% baumdicker_efficient_2021,
+% korfmann_weak_2022, petr_slendr_2023,tagami_tstrait_2024},
+% ARG inference~\citep{kelleher_inferring_2019,speidel_method_2019,
+% wohns_unified_2022,mahmoudi_bayesian_2022,
+% rasmussen_espalier_2022,
+% zhang_biobank-scale_2023,zhan_towards_2023} and analysis
+% packages
+% % etc
+% \citep{nowbandegani_extremely_2023,fan_likelihood_2023}.
 %% NOTE: this is some orphan text from the intro that can be rejiggered for the 

### 2024-07-10 paper.tex
Update the intro

@@
 follow~\citep{harris_database_2019,harris_using_2023}.
+
 Three classes of ARG inference method have emerged as the focus of recent
@@
 from SINGER paper). 
-The third class of inference method and based on heuristics
+The third class of inference method is based on heuristics
 and approximate models and focus on large scale inference,
@@
 Other methods specialised to particular applications have also 
-recently been proposed~\citep{zhan_towards_2023,schaefer_ancestral_2021}.
-See~\cite{wong_general_2023} for a detailed review of ARG inference 
+recently been proposed~\citep{zhan_towards_2023,schaefer_ancestral_2021};
+see~\cite{wong_general_2023} for a detailed review of ARG inference 
 methods.
+
+Statistically rigorous ARG sampling methods like
+ARGweaver~\citep{rasmussen_genome-wide_2014} and
+SINGER~\citep{deng_robust_2024} have many advantages. 
+Firstly, ARGweaver has consistently out-performed other methods in terms 
+of accuracy in a variety of benchmarks~\citep{brandt2022evaluation}[MORE
+citations,anywhere ARGweaver has been tested for accuracy]
+and, while SINGER has not been evaluated as thoroughly yet, it promises
+to be even more accurate~\citep{deng_robust_2024}. 
+Secondly, the direct connection with
+population genetic models greatly aids interpretability, and allows for
+natural extensions like the joint estimation of ARGs and
+demography~\citep{hubisz_mapping_2020}. Thirdly, sampling ARGs from a
+posterior distribution provides a means of quantifying the uncertainty 
+around estimates, which can be used in downstream applications [citation
+of this?]. Heuristic, approximate inference methods like
+Relate~\citep{speidel_method_2019}, tsinfer~\citep{kelleher_inferring_2019}
+and ARG-Needle~\citep{zhang_biobank-scale_2023} on the other hand 
+tend to be less accurate
 
@@
 
-Perhaps the clearest route to statistical rigor is through scalable
-model-based inference, a central element of population genetics.
-The coalescent~\citep{kingman_genealogy_1982,kingman_coalescent_1982,hudson_testing_1983,
-tajima_evolutionary_1983} has proved to be a useful model, and is the
-basis of a wide variety of inferences [some citations, e.g Hey IM model]. 
-The incorporation
-of recombination in the coalescent process~\citep{hudson_properties_1983}
-has proved to be much more difficult.
-% FIXME rephrase, next two sentences, lifted from ARG paper
-Early work on inference under the coalescent with recombination
-focused on the problem of
-inferring the parameters of the
-stochastic process, where the ancestry was regarded as a
-latent parameter to be averaged out
-\citep[e.g.][]{griffiths_ancestral_1996,kuhner_maximum_2000,
-nielsen_estimation_2000,
-fearnhead_estimating_2001}.
-% TODO reuse some of the ideas from the SMC paper here, it has
-% some good explanations.
-These methods met with limited success
-because the state space of ARGs is overwhelmingly large, and
-lacks a simple geometry or neighbourhood structure for inference or
-sampling methods to  exploit.
-The introduction of the Sequentially Markovian Coalescent (SMC)
-approximation to the coalescent with recombination
-\citep{mcvean_approximating_2005} 
-and its refinement in the SMC'~\citep{marjoram_fast_2006},
-are the basis of many powerful inference 
-% TODO add some more recent stuff
-methods~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
-In particular, the SMC is the basis for ARG sampling methods
-like ARGweaver~\citep{rasmussen_genome-wide_2014} and 
-SINGER~\citep{deng_robust_2024}, which are the gold-standard 
-in terms of inference accuracy.
-ARGweaver has consistently out-performed other methods in terms 
-of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
-and, while SINGER has not been evaluated as thoroughly yet, it promises
-to be even more accurate as well as substantially more scalable 
-computationally~\citep{deng_robust_2024}.
+% Perhaps the clearest route to statistical rigor is through scalable
+% model-based inference, a central element of population genetics.
+% The coalescent~\citep{kingman_genealogy_1982,kingman_coalescent_1982,hudson_testing_1983,
+% tajima_evolutionary_1983} has proved to be a useful model, and is the
+% basis of a wide variety of inferences [some citations, e.g Hey IM model]. 
+% The incorporation
+% of recombination in the coalescent process~\citep{hudson_properties_1983}
+% has proved to be much more difficult.
+% % FIXME rephrase, next two sentences, lifted from ARG paper
+% Early work on inference under the coalescent with recombination
+% focused on the problem of
+% inferring the parameters of the
+% stochastic process, where the ancestry was regarded as a
+% latent parameter to be averaged out
+% \citep[e.g.][]{griffiths_ancestral_1996,kuhner_maximum_2000,
+% nielsen_estimation_2000,
+% fearnhead_estimating_2001}.
+% % TODO reuse some of the ideas from the SMC paper here, it has
+% % some good explanations.
+% These methods met with limited success
+% because the state space of ARGs is overwhelmingly large, and
+% lacks a simple geometry or neighbourhood structure for inference or
+% sampling methods to  exploit.
+% The introduction of the Sequentially Markovian Coalescent (SMC)
+% approximation to the coalescent with recombination
+% \citep{mcvean_approximating_2005} 
+% and its refinement in the SMC'~\citep{marjoram_fast_2006},
+% are the basis of many powerful inference 
+% % TODO add some more recent stuff
+% methods~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
+% In particular, the SMC is the basis for ARG sampling methods
+% like ARGweaver~\citep{rasmussen_genome-wide_2014} and 
+% SINGER~\citep{deng_robust_2024}, which are the gold-standard 
+% in terms of inference accuracy.
+% ARGweaver has consistently out-performed other methods in terms 
+% of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
+% and, while SINGER has not been evaluated as thoroughly yet, it promises
+% to be even more accurate as well as substantially more scalable 
+% computationally~\citep{deng_robust_2024}.
 

### 2024-07-10 paper.tex
Pass at intro. Muddled narrative, but improving.

--- 2024-07-10 paper.tex
@@
 follow~\citep{harris_database_2019,harris_using_2023}.
-Three classes of ARG inference method have emerged as the focus of recent
-work, and have all been successfully applied in different contexts. 
-Parsimony
-based methods~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022}
-seek to minimise some aspect of the inferred ARG
-(usually the number number of recombinations required), and work well for 
-answering specific detailed questions about 
-recombination~\citep[e.g.][]{ignatieva_ongoing_2022} when the expected
-number of ancestral recombinations is small (tens to hundreds).
-The second class of inference method is based on sampling ARGs statistically 
-from a population genetics model~\citep{rasmussen_genome-wide_2014,
-hubisz_mapping_2020,mahmoudi_bayesian_2022, deng_robust_2024}, and have 
-been applied in several
+Although there are many different approaches to ARG
+inference~\citep{wong_general_2023}, two broad classes of method
+have been the focus of most recent interest.
+
+% TODO Maybe swap the order of these two paragraphs around, so it's 
+% more natrual to talk about the advantages of model-driven stuff next.
+The first class of inference methods are statistically rigorous approaches
+that sample ARGs under a population genetics model such as the
+coalescent with recombination~\citep{hudson_properties_1983},
+and its approximatation, the Sequentially Markovian
+Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
+ARGweaver~\citep{rasmussen_genome-wide_2014} is the most popular,
+consistently out-performed other methods in terms 
+of accuracy in a variety of benchmarks~\citep{brandt2022evaluation}[MORE
+citations,anywhere ARGweaver has been tested for accuracy]
+and has succesfully been applied in several
 contexts~\citep[e.g.][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
-stankowski_genetic_2024}. These methods are well-suited to applications 
-in which population genetics models are a good fit [TODO: Say this differently, 
-I just want to distinguish from e.g. SARS-CoV-2 where pop gen models are 
-basically meaningless]
-in cases in which the expected number of recombinations is in the 
-thousands to tens-of-thousands range (e.g., human datasets in which
-we have X samples and a segment of Y Mb of chromosome [TODO quantify
-from SINGER paper). 
-The third class of inference method and based on heuristics
-and approximate models and focus on large scale inference,
-where the expected number of recombinations is in the millions or
-tens of millions [TODO check numbers here]~\citep{kelleher_inferring_2019,
-speidel_method_2019,zhang_biobank-scale_2023}.
-Other methods specialised to particular applications have also 
-recently been proposed~\citep{zhan_towards_2023,schaefer_ancestral_2021}.
-See~\cite{wong_general_2023} for a detailed review of ARG inference 
-methods.
-
-Typically, the more rigorous approaches can handle 10-100 samples,
-while the approximate methods can scale to 10s to hundreds of
-thousands of samples. With the explosive growth in dataset
+stankowski_genetic_2024}. 
+ARGweaver works [by explanation.]
+The recently-introduced SINGER method takes a similar approach [doing something 
+slightly different]~\citep{deng_robust_2024}.
+While SINGER has not been evaluated as thoroughly as ARGweaver yet, it promises
+to be even more accurate while being substantially more
+scalable~\citep{deng_robust_2024}.
+
+The second class of inference method that has been of recent interest 
+is based on heuristics and approximate models.
+Relate~\citep{speidel_method_2019},
+tsinfer~\citep{kelleher_inferring_2019} and 
+ARG-Needle~\citep{zhang_biobank-scale_2023} work on quite different 
+principles, but share some common properties. Firstly, they are all 
+hueristic and approximation driven, favouring computational
+scalability over statistical rigor, and lacking a direct 
+connection to population genetic theory. 
+Secondly, they all produce a single
+ARG as output, inferring a deterministic point estimate with no means 
+of quantifying uncertainty. Thirdly (and most importantly for the 
+purposes of this paper) the ARGs they produce lack substantial
+detail about ancestral recombination events: the estimates are
+rough sketches, hopefully capturing large-scale properties, but lacking
+in the finer detail of outputs of ARGweaver or SINGER.
+
+There are substantial advantages to the statistically rigorous sampling 
+approaches of ARGweaver and SINGER. Firstly, they tend to be more 
+accurate; ARGweaver has consistently out-performed other methods in terms 
+of accuracy in a variety of benchmarks~\citep{brandt2022evaluation}[MORE
+citations,anywhere ARGweaver has been tested for accuracy]
+and, while SINGER has not been evaluated as thoroughly yet, it promises
+to be even more accurate~\citep{deng_robust_2024}. 
+Secondly, the direct connection with
+population genetic models greatly aids interpretability, and allows for
+natural extensions like the joint estimation of ARGs and
+demography~\citep{hubisz_mapping_2020}. Thirdly, sampling ARGs from a
+posterior distribution provides a means of quantifying the uncertainty 
+around estimates, which can be used in downstream applications [citation
+of this?].
+
+[Needs more work]
+Despite these significant advantages over hueristic methods, however,
+hueristic methods remain of interest because of the profound difference
+in computational scalability between these approaches. Even with the 
+major advances made in this direction by SINGER, ARG inference is 
+only feasible for hundreds of samples. Tsinfer and ARG-Needle have 
+been applied to datasets of hundreds of \emph{thousands} of samples.
+With the explosive growth in dataset
 sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
@@
 rigor in large scale ARG inference. 
-
-Perhaps the clearest route to statistical rigor is through scalable
-model-based inference, a central element of population genetics.
-The coalescent~\citep{kingman_genealogy_1982,kingman_coalescent_1982,hudson_testing_1983,
-tajima_evolutionary_1983} has proved to be a useful model, and is the
-basis of a wide variety of inferences [some citations, e.g Hey IM model]. 
-The incorporation
-of recombination in the coalescent process~\citep{hudson_properties_1983}
-has proved to be much more difficult.
-% FIXME rephrase, next two sentences, lifted from ARG paper
-Early work on inference under the coalescent with recombination
-focused on the problem of
-inferring the parameters of the
-stochastic process, where the ancestry was regarded as a
-latent parameter to be averaged out
-\citep[e.g.][]{griffiths_ancestral_1996,kuhner_maximum_2000,
-nielsen_estimation_2000,
-fearnhead_estimating_2001}.
-% TODO reuse some of the ideas from the SMC paper here, it has
-% some good explanations.
-These methods met with limited success
-because the state space of ARGs is overwhelmingly large, and
-lacks a simple geometry or neighbourhood structure for inference or
-sampling methods to  exploit.
-The introduction of the Sequentially Markovian Coalescent (SMC)
-approximation to the coalescent with recombination
-\citep{mcvean_approximating_2005} 
-and its refinement in the SMC'~\citep{marjoram_fast_2006},
-are the basis of many powerful inference 
-% TODO add some more recent stuff
-methods~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
-In particular, the SMC is the basis for ARG sampling methods
-like ARGweaver~\citep{rasmussen_genome-wide_2014} and 
-SINGER~\citep{deng_robust_2024}, which are the gold-standard 
-in terms of inference accuracy.
-ARGweaver has consistently out-performed other methods in terms 
-of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
-and, while SINGER has not been evaluated as thoroughly yet, it promises
-to be even more accurate as well as substantially more scalable 
-computationally~\citep{deng_robust_2024}.
-
-% THE PROBLEM!!! OMG it's terrible!!!
-Despite these recent advances in statistically rigorous ARG 
-inference, however, there is still a gap of some three orders 
-of magnitude between the sample-sizes that can be handled 
-(hundreds) compared to the large-scale hueristic methods
-like tsinfer and ARGneedle (hundreds of thousands).
-[TODO working through a more positive narrative here, based on
-the idea of combining the strengths of the approaches]
-Although major breakthroughs in scale have been achieved by hueristic
-ARG inference
-methods~\citep{kelleher_inferring_2019,speidel_method_2019,zhang_biobank-scale_2023},
-these methods are currently fundamentally incompatible with rigorous
-inference under population genetic models such as the SMC. This is 
-because these methods return approximate ARG
-\emph{structures}~\citep{wong_general_2023} [WEAK, FIXME]. 
-Specifically,
-we cannot evaluate a likelihood of these ARGs in coalescent 
-models under the classical 
-\cite{kuhner_maximum_2000} formulation, as these formulations assume
-that complete information about recombination events is present
-[Also: \cite{guo_recombination-aware_2022}]
-[NOTE the following is super weak, and needs a lot of work.]
-This inability to connect the results of large-scale inference methods
-with a generative model
-is a fundamental limitation, and limits both the utility of 
-current point estimates as well as the possibilities for further
-inference. Specifically, we are unable to quantify uncertainty 
-around heuristic point estimates. We cannot explore ARG space
-around the point estimates using (e.g. MCMC), as we have no way definition
-of improvement. We cannot compare models of demography, or evaluate the 
-goodness of fit for model estimates.
-
-%% This is how we solve it
-[TODO figure out the gARG, tree sequence etc mess]
-Here, we bridge this gap between models and the products of approximate
-inference by defining a likelihood function for any 
-``genome ARG''~\citep{wong_general_2023} under the SMC. Using a novel
-formulation of the SMC that focuses on the contributions of nodes and 
-edges [FIXME] in a backwards-in-time way, we provide an efficient 
-algorithm that can compute a likelihood for any gARG under this model.
-We demonstrate the flexibility of this approach by XXX, and show
-[SOMETHING].
+% This inability to connect the results of large-scale inference methods
+% with a generative model
+% is a fundamental limitation, and limits both the utility of 
+% current point estimates as well as the possibilities for further
+% inference. Specifically, we are unable to quantify uncertainty 
+% around heuristic point estimates. We cannot explore ARG space
+% around the point estimates using (e.g. MCMC), as we have no way definition
+% of improvement. We cannot compare models of demography, or evaluate the 
+% goodness of fit for model estimates.
+A key problem facing the field is that the ARGs estimated by heuristic large-scale 
+methods cannot be connected to population genetics theory because 
+we cannot compute their likelihood under the SMC. The classical 
+method~\citep{kuhner_maximum_2000} assumes complete knowledge about
+recombination events, and there is no clear way to compute a likelihood for 
+the approximate structures output by hueristic methods.
+
+In this paper we provide a necessary first step towards this goal
+of statistically rigorous large-scale ARG inference, by defining a function
+to compute the likelihood of a general class of ARG under the SMC.
+To do this, we rederive the SMC under a backwards-time formulation,
+modelling the process in terms of lineages and [something] rather
+than events. Using this novel formulation, we then show how to 
+easily compute the likelihood of any ARG under the SMC and outline 
+an efficient algorithm to do so. We illustrate the flexibility of this
+approach via [some examples] and conclude by discussing avenues
+for future research opened up.
 
@@
 \subsection{Calculating likelihoods} \label{par:liks}
+[Paragraph setting the scene on likelihood calculations]
+\cite{kuhner_maximum_2000} formulation, as these formulations assume
+that complete information about recombination events is present
+[Also: \cite{guo_recombination-aware_2022}]
+[NOTE the following is super weak, and needs a lot of work.]
 
@@
 be quantified and passed on as metadata to the nodes in the data format storing the ARG.
+
+[Note: a point to make here is that the idea of combining hueristic large-scale 
+inference with model based inference is analogous to the combination of 
+coalescent simulation and forwards time simulation. We can also draw lessons
+from developers of different methods cooperating and using the tskit library
+to enable this.]
 

### 2024-07-10 paper.tex

@@
 
+[Things to include: locally unary nodes rather than explicit events.
+edge-centric formulation vs event centric. Graceful degradation of
+performance with incomplete ARGs vs complete inability to compute.]
+
 In this paper we provide a necessary first step towards this goal
@@
 the SMC is the process in which only lineages with \emph{overlapping} 
-genomic intervals may coalesce. (The SMC' additionally allows 
-lineages with abutting genomic intervals to coalesce.)
+genomic intervals may coalesce. The SMC' is a refinement 
+of the SMC which also allows lineages with abutting genomic intervals to 
+coalesce~\citep{marjoram_fast_2006}.
 Although \citet{mcvean_approximating_2005} verbally described the SMC
@@
 rasmussen_genome-wide_2014}.
+
 Linking the resulting tree-valued process along the genome back to
 the relationship between the ``graph'' and
-``local trees'' representation of
-the ARG, the Markovian property can be formulated as follows: 
+``local trees'' representation of ARGs, 
+the Markovian property can be formulated as follows: 
 the distribution of the next tree within the series
@@
 pairs of lineages, we are guaranteed that all nodes in the ARG 
-only carry ancestral material, this is,
+only carry ancestral material, that is,
 genetic material that is inherited by either one of a set of samples
 \citep{wiuf_ancestry_1999}.
-As such, we only require a single half-open interval to describe
+We require a single half-open interval to describe
 the ancestral
@@
 for some interval endpoints $x_{j} < y_{j}$.
-Since we look backwards in time, $t=0$ is today, and so
+Because we look backwards in time, $t=0$ is today, and 
 $L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
@@
 with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
-which each are given new, unique labels and added to $L(t)$
+given new, unique labels and added to $L(t)$
 (and lineage $j$ is removed).
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Calculating likelihoods} \label{par:liks}
-[Paragraph setting the scene on likelihood calculations]
-\cite{kuhner_maximum_2000} formulation, as these formulations assume
-that complete information about recombination events is present
-[Also: \cite{guo_recombination-aware_2022}]
-[NOTE the following is super weak, and needs a lot of work.]
+\subsection{Likelihoods} \label{par:liks}
+
+Computing the likelihood of a putative ARG under a model is a fundamental element of
+probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
+\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
+decomposes the likelihood of an ARG under the coalescent with recombination
+into a product of the probablility of 
+each common ancestor event, each recombination event and the probablility of
+the inter-event waiting times. In essence, computing the KYF likelihood requires
+``replaying'' the full state of Hudson's simulation
+algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
+compute the probability of the configuration of lineages and distribution
+of ancestral material at each time point. It therefore requires complete
+detail about each of these events to be stored. 
+Such complete detail is not present in many estimated
+ARGs~\citep{wong_general_2023} or even in most simulations.
+Indeed, supporting likelihood calculations originally motivated the addition of
+the \texttt{record\_full\_arg}
+option to the \texttt{msprime} simulator~\citep{baumdicker_efficient_2021}.
+More recent approaches to defining likelihoods on such models require
+similarly exhaustive input\citep[e.g.][]{guo_recombination-aware_2022} (MORE?
+Anything for GC in bacteria, etc?)
 

### 2024-07-11 paper.tex
Minor tweak to caption

@@
 This results in nodes that (locally) have only a single child, and are therefore
-(locally) unary. By removing recombinant nodes, we limit the information on the time to each
-of these events to the windows (blue shaded array) delimited by by the age of
-the child node and the minimum age of the parents of the edges connecting to that child node.
-Blue (unary) nodes represent recombinant nodes. The blue arrows show the
-familiar left-to-right logic of the SMC whereby
+(locally) unary. 
+Nodes A, B and C are recombinants, i.e, have two parents in the graph
+and are unary in the local trees. The blue shaded regions indicate
+the range of possible times associated with these recombinants.
+The blue arrows show the
+standard left-to-right logic of the SMC whereby
 the floating lineage coalesces randomly
 with a the remaining portion of the local tree following recombination.
+% By removing recombinant nodes, we limit the information on the time to each
+% of these events to the windows (blue shaded array) delimited by the age of
+% the child node and the minimum age of its parents.
 }

### 2024-07-11 paper.tex
Add some background motivational stuff on ARGs

@@
 \citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
-decomposes the likelihood of an ARG under the coalescent with recombination
-into a product of the probablility of 
-each common ancestor event, each recombination event and the probablility of
-the inter-event waiting times. In essence, computing the KYF likelihood requires
-``replaying'' the full state of Hudson's simulation
+works by assigning a probablility to each recombination and common
+ancestor event, and to the inter-event  waiting times. The likelihood of an 
+ARG is then a product over all these three classes of term.
+In essence, computing the KYF likelihood requires
+``replaying'' the  state of Hudson's simulation
 algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
-compute the probability of the configuration of lineages and distribution
-of ancestral material at each time point. It therefore requires complete
+compute the exact configuration of lineages and distribution
+of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
+It therefore requires complete
 detail about each of these events to be stored. 
-Such complete detail is not present in many estimated
-ARGs~\citep{wong_general_2023} or even in most simulations.
-Indeed, supporting likelihood calculations originally motivated the addition of
+Other approaches to defining the likelihood of an ARG under a coalescent-like 
+model follow the same basic event-by-event 
+strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
+[Are there others?  The bacterial stuff from Didelot etc seem to focus on
+a tree-by-tree likelihood, rather than global graph, so not clear it's the 
+same thing]
+Computing the likelihood under the SMC rather than the Hudson coalescent 
+simplifies bookkeeping somewhat, 
+but a complete event-by-event description of the 
+input ARG is still required.
+
+Although the KYF formulation is a natural and elegant way to describe 
+the likelihood of an ARG, such complete detail is not present 
+in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
+Simulators such as the classical \texttt{ms}
+program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
+which contain no information about recombination events (other than their
+breakpoints) and where internal nodes marking coalesces are unlabelled.
+The \texttt{msprime}
+simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
+by default outputs a more complete and efficient representation of 
+the simulated ancestry, but still does not provide the exhaustive
+detail required.
+Indeed, providing the information required to 
+support likelihood calculations originally motivated the addition of
 the \texttt{record\_full\_arg}
-option to the \texttt{msprime} simulator~\citep{baumdicker_efficient_2021}.
-More recent approaches to defining likelihoods on such models require
-similarly exhaustive input\citep[e.g.][]{guo_recombination-aware_2022} (MORE?
-Anything for GC in bacteria, etc?)
+option to \texttt{msprime}~\citep{baumdicker_efficient_2021}.
+Forwards-time simulations that output 
+ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
+contain the level of detail required to compute a likelihood under
+the KYF approach.
 

### 2024-07-11 paper.tex

@@
 
+Computing the likelihood of a putative ARG under a model is a fundamental element of
+probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
+\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
+works by assigning a probablility to each recombination and common
+ancestor event, and to the inter-event  waiting times. The likelihood of an 
+ARG is then a product over all these three classes of term.
+In essence, computing the KYF likelihood requires
+``replaying'' the  state of Hudson's simulation
+algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
+compute the exact configuration of lineages and distribution
+of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
+It therefore requires complete
+detail about each of these events to be stored. 
+Other approaches to defining the likelihood of an ARG under a coalescent-like 
+model follow the same basic event-by-event 
+strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
+[Are there others?  The bacterial stuff from Didelot etc seem to focus on
+a tree-by-tree likelihood, rather than global graph, so not clear it's the 
+same thing]
+Computing the likelihood under the SMC rather than the Hudson coalescent 
+simplifies bookkeeping somewhat, 
+but a complete event-by-event description of the 
+input ARG is still required.
+
+Although the KYF formulation is a natural and elegant way to describe 
+the likelihood of an ARG, such complete detail is not present 
+in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
+Simulators such as the classical \texttt{ms}
+program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
+which contain no information about recombination events (other than their
+breakpoints) and where internal nodes marking coalesces are unlabelled.
+The \texttt{msprime}
+simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
+by default outputs a more complete and efficient representation of 
+the simulated ancestry, but still does not provide the exhaustive
+detail required.
+Indeed, providing the information required to 
+support likelihood calculations originally motivated the addition of
+the \texttt{record\_full\_arg}
+option to \texttt{msprime}~\citep{baumdicker_efficient_2021}.
+Forwards-time simulations that output 
+ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
+contain the level of detail required to compute a likelihood under
+the KYF approach.
+
 [Things to include: locally unary nodes rather than explicit events.
@@
 \citep{wiuf_ancestry_1999}.
-We require a single half-open interval to describe
+We therefore require a single half-open interval to describe
 the ancestral
-material associated with each common ancestor node in the ARG.
+material associated with each node in the ARG.
 
@@
 backwards in time.
-At any point in time, the state of the process is $L(t)$,
+At any point in time, the state of the process is 
 the set of labeled lineages extant at time $t$,
-where each lineage $X_j(t)$ is labeled by an integer $j$
-and represented by a half-open interval delimiting the
-ancestral material. Any lineage can thus be summarized as $X_j(t) = [x_{j}, y_{j})$,
-for some interval endpoints $x_{j} < y_{j}$.
+$L(t) = \{X_j(t)\}$.
+A lineage $X_j(t) = [x_{j}, y_{j})$ is labeled by 
+an integer $j$ and defines the 
+half-open genomic interval of its ancestral material.
+% Ea
+% where each lineage $X_j(t)$ is labeled by an integer $j$
+% and represented by a 
+% ancestral material. 
+% Any lineage can thus be summarized as 
+% for some interval endpoints $x_{j} < y_{j}$.
 Because we look backwards in time, $t=0$ is today, and 
@@
 
-Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
-so, if $T(t) = \sum_i |X_j(t)|$ is the sum of the total sizes of all lineages,
+Recombination is described by a Poisson process of rate $r$ per unit 
+of genomic length and time:
+so, if $T(t) = \sum_j |X_j(t)|$ is total length of ancestral material
+carried by lineages at time $t$,
 then recombination occurs at instantaneous rate $rT(t)$.
@@
 with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
-given new, unique labels and added to $L(t)$
-(and lineage $j$ is removed).
+given new, unique labels. We then remove lineage $j$
+and add the two new lineages to the state $L(t)$.
 This operation keeps the total amount of ancestral material unchanged.
@@
 two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
-The newly formed lineage acquires the union of both intervals,
-the original lineages are removed, and $L(t)$ is updated accordingly.
-
-Although coalescence is reciprocal, here we'll define a strict total order
-on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
-For any such strict total order,
-the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
+Here we deviate from the McVean and Cardin formulation,
+and define a strict total order on all lineages in $L(t)$:
+for each lineage $X_j(t) = [x_{j}, y_{j})$ we define its 
+order by the left coordinate $x_j$ and label $j$ (to break ties).
+The instantaneous rate of coalescence then equals $\lambda$ 
+multiplied by the number of overlapping pairs,
 i.e., $\lambda \sum_{i \neq j} I_{jk}$,
@@
 From the definition of the strict total order it follows that if $I_{jk} = 1$
-then $I_{kj} = 0$. For later use, also define $C_j(t)$ to be the set of
+then $I_{kj} = 0$. 
+The newly formed lineage acquires the union of both intervals,
+the original lineages are removed, and $L(t)$ is updated accordingly.
+[JK notes: need to explain here that this deliberately introduced 
+asymmetry doesn't affect the actual process. Also need to comment on 
+how the process terminates. Basically has the same subtleties as McVean and 
+Cardin, and not important for our purposes.]
+
+For later use, also define $C_j(t)$ to be the set of
 lineages that can coalesce into lineage $j$, i.e., 
@@
 Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
-which will simplify the likelihood computations. In particular, although
+which will simplify likelihood computations later. In particular, although
 recombination affects \emph{which} lineages can coalesce into $j$, it does not
@@
 
-[Do we need to get into this above-the-root subtelty here?]
-To ensure that the ancestral material carried by a single lineage can at
-all time be represented by a single half-open interval, we only update the
-endpoints of the involved lineages following common ancestry or recombination events
-such that only the minimal half-open interval whose endpoints have not fully coalesced
-is used to update $L(t)$. Note that empty intervals are not added to the state.
-%By reasoning about the process in this way we can keep representing 
-% all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
-% and the process terminates.
-
-%After any coalescence or recombination event,
-%the segment(s) $[x_j,y_j)$ (and $[x_k, y_k]$) carried by the involved lineage(s)
-%are replaced by a single segment $[x, y)$ whose endpoints have not fully
-%coalesced, i.e., $x = \min\{ z \ge \min(x_j, x_k) : N(t,z) > 1\}$,
-%and $y = \sup\{ z \le max(y_j, y_k) : N(t,z) > 1\}$. Note that empty intervals
-%are not added back to $L(t)$.
-
-Finally, by describing the SMC
-backwards in time, some intervals contained within the half-open
-interval representing a lineage might have fully coalesced before all the genetic
-material carried by that lineage has. As noted by \citet{mcvean_approximating_2005},
-such intervals can be observed as the stalks on the marginal genealogies (e.g.\ marginal trees
-for $[25, 81)$ Fig. \ref{fig:smc-unary}). Recombination events occurring within these smaller
-intervals however do not affect the marginal genealogies.
+% [Do we need to get into this above-the-root subtelty here?]
+% To ensure that the ancestral material carried by a single lineage can at
+% all time be represented by a single half-open interval, we only update the
+% endpoints of the involved lineages following common ancestry or recombination events
+% such that only the minimal half-open interval whose endpoints have not fully coalesced
+% is used to update $L(t)$. Note that empty intervals are not added to the state.
+% %By reasoning about the process in this way we can keep representing 
+% % all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
+% % and the process terminates.
+
+% %After any coalescence or recombination event,
+% %the segment(s) $[x_j,y_j)$ (and $[x_k, y_k]$) carried by the involved lineage(s)
+% %are replaced by a single segment $[x, y)$ whose endpoints have not fully
+% %coalesced, i.e., $x = \min\{ z \ge \min(x_j, x_k) : N(t,z) > 1\}$,
+% %and $y = \sup\{ z \le max(y_j, y_k) : N(t,z) > 1\}$. Note that empty intervals
+% %are not added back to $L(t)$.
+
+% Finally, by describing the SMC
+% backwards in time, some intervals contained within the half-open
+% interval representing a lineage might have fully coalesced before all the genetic
+% material carried by that lineage has. As noted by \citet{mcvean_approximating_2005},
+% such intervals can be observed as the stalks on the marginal genealogies (e.g.\ marginal trees
+% for $[25, 81)$ Fig. \ref{fig:smc-unary}). Recombination events occurring within these smaller
+% intervals however do not affect the marginal genealogies.
 
@@
 \subsection{Likelihoods} \label{par:liks}
-
-Computing the likelihood of a putative ARG under a model is a fundamental element of
-probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
-\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
-works by assigning a probablility to each recombination and common
-ancestor event, and to the inter-event  waiting times. The likelihood of an 
-ARG is then a product over all these three classes of term.
-In essence, computing the KYF likelihood requires
-``replaying'' the  state of Hudson's simulation
-algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
-compute the exact configuration of lineages and distribution
-of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
-It therefore requires complete
-detail about each of these events to be stored. 
-Other approaches to defining the likelihood of an ARG under a coalescent-like 
-model follow the same basic event-by-event 
-strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
-[Are there others?  The bacterial stuff from Didelot etc seem to focus on
-a tree-by-tree likelihood, rather than global graph, so not clear it's the 
-same thing]
-Computing the likelihood under the SMC rather than the Hudson coalescent 
-simplifies bookkeeping somewhat, 
-but a complete event-by-event description of the 
-input ARG is still required.
-
-Although the KYF formulation is a natural and elegant way to describe 
-the likelihood of an ARG, such complete detail is not present 
-in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
-Simulators such as the classical \texttt{ms}
-program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
-which contain no information about recombination events (other than their
-breakpoints) and where internal nodes marking coalesces are unlabelled.
-The \texttt{msprime}
-simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
-by default outputs a more complete and efficient representation of 
-the simulated ancestry, but still does not provide the exhaustive
-detail required.
-Indeed, providing the information required to 
-support likelihood calculations originally motivated the addition of
-the \texttt{record\_full\_arg}
-option to \texttt{msprime}~\citep{baumdicker_efficient_2021}.
-Forwards-time simulations that output 
-ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
-contain the level of detail required to compute a likelihood under
-the KYF approach.
 
@@
     \includegraphics[width=0.65\textwidth]{figures/single_event.png}
-    \caption{
-        Graphical representation of the events associated with the edge $(p,c,x,y)$.
-        The lineage carried by $c$ spans $[x_c,y)$ until a recombination event (blue dotted line) splits it
-        into two segments. Left and right hand segments coalesce at times $t_p'$ and $t_p$ respectively,
-        bounding the time to recombination by $(c. \min(t_p, t_{p'})$.
-        Note that the contribution to the recombination hazard by edge $(p,c,x,y)$ only covers ``half'' of the
-        green shaded area. The other half is contributed by edge $(p', c, x_c, x)$. 
-        \label{fig:likelihood}
-    }
+    % FIXME commenting this caption out because it mucks up syntax highlighting for me.
+    % \caption{
+    % Graphical representation of the events associated with the edge $(p,c,x,y)$.
+    % The lineage carried by $c$ spans $[x_c,y)$ until 
+% a recombination event (blue dotted line) splits it
+    %     into two segments. Left and right hand segments coalesce at times $t_p'$ and $t_p$ respectively,
+    %     bounding the time to recombination by $(c. \min(t_p, t_{p'})$.
+    %     Note that the contribution to the recombination hazard by edge $(p,c,x,y)$ only covers ``half'' of the
+    %     green shaded area. The other half is contributed by edge $(p', c, x_c, x)$. 
+    %     \label{fig:likelihood}
+    % % }
 \end{figure}
@@
 had the segment $[x,y)$ (see Fig.~\ref{fig:likelihood}).
-Under the SMC,
-we therefore know there was no recombination between $x$ and $y$
-on any of the lineages 
+We therefore know there was no recombination between $x$ and $y$
+on any of those lineages 
 over this time span,
@@
     =
-    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(c, p) \in E} A_e(r) B_e(r, \lambda) .
+    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(p, c) \in E} A_e(r) B_e(r, \lambda) .
 \end{equation}
-One significant advantage of this formulation is that the computation time
-is linear in the number of edges, and an algorithm can be devised to 
-to efficiently keep track of $I_e(t)$ (see Appendix).
+% One significant advantage of this formulation is that the computation time
+% is linear in the number of edges, and an algorithm can be devised to 
+% to efficiently keep track of $I_e(t)$ (see Appendix).
 % Although the first two factors are easily expressed in terms of numbers of edges and nodes,
@@
 % that in practice it is convenient to absorb these terms into per-edge contributions.
-We can also accommodate stepwise changes in the 
-coalescence rate (or recombination rate)
-through time by intersecting the internode intervals with the rate change
-intervals; rate changes along the genome can be incorporated similarly.
+% We can also accommodate stepwise changes in the 
+% coalescence rate (or recombination rate)
+% through time by intersecting the internode intervals with the rate change
+% intervals; rate changes along the genome can be incorporated similarly.
 Finally, we have only considered the topology and branch length information of any ARG.
@@
     =
-    \frac{1}{\mathcal{M}!} \prod_{(c, p) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
+    \frac{1}{\mathcal{M}!} \prod_{(p, c, x, y) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
 \end{equation}

### 2024-07-12 paper.tex
Some text on introducing likelihood calculations

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methods and Results}
-
-First, 
-we describe
-in more detail what sort of information we have in estimated ARGs.
-Then,
-we define the SMC process in a ``backwards-in-time'' way,
-rather than the usual ``along-the-genome'' way.
-Note that this is a
-reiteration of the original work of \citet{mcvean_approximating_2005}.
-
-In the subsequent section we connect the two and
-show how the ``backwards-in-time'' description
-makes it relatively straightforward to compute the likelihood
-for a more general class of estimated ARGs.
-Finally, we close with a demonstration that the likelihood computation can be used
-to do Bayesian inference on ancestral population sizes.
+\section*{Results}
+
+\section*{ARG likelihoods}
+
+In one of the first approaches to probilistic ARG inference that 
+regarded the ARG itself as an output rather than a latent variable,
+\cite{kuhner_maximum_2000} provided an explicit formula for 
+the likelihood of a realised ARG under the coalescent with
+recombination:
+\[
+\mathbb{P}[G|N, r] = \left(\frac{1}{2N}\right)^K
+r^X \exp\left(-\sum \left( \frac{1}{2N} \binom{k_i}{2} + r\ell_i\right)
+t_i\right),
+\]
+where [explain notation].
+Under this KYF formulation, the likelihood is computed by taking a 
+product of the probability of each realised event, and their 
+inter-event waiting times. While it is straightforward to write 
+this down, computation is not so easy, and in essence requires
+In essence, computing the KYF likelihood requires
+``replaying'' the  state of Hudson's simulation
+algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
+compute the exact configuration of lineages and distribution
+of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
+
+Computing the likelihood of a realised ARG under the SMC simplifies the 
+bookkeeping somewhat, because we no long have the possibility of 
+``trapped'' ancestral material, and each lineage can be represented 
+by a single interval representing their ancestral material. Nonetheless,
+we still require complete knowledge of the graph to compute the 
+required probablilities.
+
+% \section{Methods and Results}
+% First, 
+% we describe
+% in more detail what sort of information we have in estimated ARGs.
+% Then,
+% we define the SMC process in a ``backwards-in-time'' way,
+% rather than the usual ``along-the-genome'' way.
+% Note that this is a
+% reiteration of the original work of \citet{mcvean_approximating_2005}.
+
+% In the subsequent section we connect the two and
+% show how the ``backwards-in-time'' description
+% makes it relatively straightforward to compute the likelihood
+% for a more general class of estimated ARGs.
+% Finally, we close with a demonstration that the likelihood computation can be used
+% to do Bayesian inference on ancestral population sizes.
 
@@
     % FIXME commenting this caption out because it mucks up syntax highlighting for me.
-    % \caption{
+    \caption{
     % Graphical representation of the events associated with the edge $(p,c,x,y)$.
@@
     %     green shaded area. The other half is contributed by edge $(p', c, x_c, x)$. 
-    %     \label{fig:likelihood}
-    % % }
+        \label{fig:likelihood}
+    }
 \end{figure}

### 2025-02-14 paper.tex
Pass through abstract

@@
 inference has proceeded in two distinct directions. The first
-class of method in which dramatic improvements have been made is in the
+class of method in which dramatic progress has been made is in the
 statistically rigorous inference of ARGs under the Sequentially Markov
-Coalescent (SMC) model, which are now practical for tens-to-hundreds of
-samples. The second class of recently-developed inference methods are based on
-approximate models and heuristics which scale to samples sizes roughly three
+Coalescent (SMC), which are now practical for tens-to-hundreds of samples. 
+The second class of recently-developed inference methods are based on
+approximate models and heuristics which scale to sample sizes two to three
 orders of magnitude larger. Although these heuristic methods are reasonably
-accurate under many metrics, one major problem is that the ARGs they 
-estimate cannot easily be connected with population genetic theory.
-Specifically, because the estimated ARGs lack detail about recombination
-events, there is currently no way to compute their likelihood
-under a population genetics model. In this paper we provide a 
+accurate under many metrics, one significant drawback is that the ARGs they 
+estimate do not have the topological properties required to compute a 
+likelihood under models such as the SMC under present-day formulations.
+In particular, hueristic inference methods typically do not estimate
+precise details about recombination events, which are currently
+required to compute a likelihood.
+In this paper we present a 
 novel backwards-time formulation of the SMC 
-and derive a straightforward definition of the likelihood
-of the likelihood of a general class of ARG under the SMC. 
+and derive a straightforward definition 
+of the likelihood of a general class of ARG under this model. 
 We show that this formulation does not require precise details of recombination events
-to be estimated, and is robust to other topological inconveniences such as 
-polytomies. This opens up many future directions, most excitingly the possibility
-of combining the strengths of both inference approaches. Applying
-these ideas may allow us to 
-make rigorous inference more scalable, and scalable inference 
-more rigorous.
+to be estimated, and is robust to the presence of polytomies. 
+We discuss the possibilities for inference that this opens.
 \end{abstract}
+
+% This opens up many future directions, most excitingly the possibility
+% of combining the strengths of both inference approaches. Applying
+% these ideas may allow us to 
+% make rigorous inference more scalable, and scalable inference 
+% more rigorous.
 

### 2025-02-14 paper.tex
Pass through the intro

@@
 to a variety of questions in evolutionary
-biology~\citep{lewanski_era_2024,brandt_promise_2024}.
+biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}.
 ARGs describe the intricately linked paths
@@
 korfmann_simultaneous_2024,
-deraje_inferring_2024}
+deraje_inferring_2024,
+speidel2025high}
 and it seems likely that many more will
@@
 Relate~\citep{speidel_method_2019},
-tsinfer~\citep{kelleher_inferring_2019} and 
-ARG-Needle~\citep{zhang_biobank-scale_2023} work on quite different 
-principles, but share some common properties. Firstly, they are all 
+tsinfer~\citep{kelleher_inferring_2019},
+ARG-Needle~\citep{zhang_biobank-scale_2023} 
+and Threads~\citep{gunnarsson_scalable_2024}
+work on quite different 
+principles, but share some common properties. 
+Firstly, they are all 
 hueristic and approximation driven, favouring computational
-scalability over statistical rigor, and lacking a direct 
-connection to population genetic theory. 
+scalability over statistical rigor.
 Secondly, they all produce a single
@@
 
-[Needs more work]
-Despite these significant advantages over hueristic methods, however,
+Despite these significant advantages of rigorous inference, however,
 hueristic methods remain of interest because of the profound difference
-in computational scalability between these approaches. Even with the 
-major advances made in this direction by SINGER, ARG inference is 
-only feasible for hundreds of samples. Tsinfer and ARG-Needle have 
+in computational scalability. Although SINGER is a major step forward
+in terms of scalability over ARGweaver and other methods, inference is still 
+only feasible for hundreds of samples. 
+Tsinfer, ARG-Needle and Threads have 
 been applied to datasets of hundreds of \emph{thousands} of samples.
 With the explosive growth in dataset
-sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
-all_genomic_2024} there is a pressing need for more statistical
+sizes in recent years~\citep{caulfield2017national,turnbull2018100,
+bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole,
+all2024genomic,stark2024call,cook2025our}
+there is a pressing need for more statistical
 rigor in large scale ARG inference.
@@
 A key problem facing the field is that the ARGs estimated by heuristic large-scale 
-methods cannot be connected to population genetics theory because 
-we cannot compute their likelihood under the SMC. The classical 
-method~\citep{kuhner_maximum_2000} assumes complete knowledge about
-recombination events, and there is no clear way to compute a likelihood for 
-the approximate structures output by hueristic methods.
-
-Computing the likelihood of a putative ARG under a model is a fundamental element of
+methods lack a concrete connection to population genetic theory because
+we cannot compute their likelihood under the SMC or similar models. 
+% The classical method~\citep{kuhner_maximum_2000} assumes complete knowledge about
+% recombination events, and there is no clear way to compute a likelihood for 
+% the approximate structures output by hueristic methods.
+
+Computing the likelihood of an ARG under a model is a fundamental element of
 probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
@@
 strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
-[Are there others?  The bacterial stuff from Didelot etc seem to focus on
-a tree-by-tree likelihood, rather than global graph, so not clear it's the 
-same thing]
+% [Are there others?  The bacterial stuff from Didelot etc seem to focus on
+% a tree-by-tree likelihood, rather than global graph, so not clear it's the 
+% same thing]
 Computing the likelihood under the SMC rather than the Hudson coalescent 
-simplifies bookkeeping somewhat, 
+simplifies bookkeeping somewhat,  % Dunno what I'm talking about here
 but a complete event-by-event description of the 
@@
 program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
-which contain no information about recombination events (other than their
-breakpoints) and where internal nodes marking coalesces are unlabelled.
+which does not contain sufficient detail for likelihood computation.
 The \texttt{msprime}
@@
 ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
-contain the level of detail required to compute a likelihood under
+typically contain the level of detail required to compute a likelihood under
 the KYF approach.
 
-[Things to include: locally unary nodes rather than explicit events.
-edge-centric formulation vs event centric. Graceful degradation of
-performance with incomplete ARGs vs complete inability to compute.]
-
-In this paper we provide a necessary first step towards this goal
-of statistically rigorous large-scale ARG inference, by defining a function
-to compute the likelihood of a general class of ARG under the SMC.
+% [Things to include: locally unary nodes rather than explicit events.
+% edge-centric formulation vs event centric. Graceful degradation of
+% performance with incomplete ARGs vs complete inability to compute.]
+
+In this paper we provide a new formulation of ARG likelihood which 
+has much more lenient input requirements than KYF, and gracefully handles
+incomplete and imprecise ARGs.
+% In this paper we provide a necessary first step towards this goal
+% of statistically rigorous large-scale ARG inference, by defining a function
+% to compute the likelihood of a general class of ARG under the SMC.
 To do this, we rederive the SMC under a backwards-time formulation,

### 2025-02-14 paper.tex
Pass through ARG section

@@
 ARG is then a product over all these three classes of term.
-In essence, computing the KYF likelihood requires
-``replaying'' the  state of Hudson's simulation
-algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
-compute the exact configuration of lineages and distribution
-of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
-It therefore requires complete
-detail about each of these events to be stored. 
-Other approaches to defining the likelihood of an ARG under a coalescent-like 
-model follow the same basic event-by-event 
-strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
-% [Are there others?  The bacterial stuff from Didelot etc seem to focus on
-% a tree-by-tree likelihood, rather than global graph, so not clear it's the 
-% same thing]
-Computing the likelihood under the SMC rather than the Hudson coalescent 
-simplifies bookkeeping somewhat,  % Dunno what I'm talking about here
-but a complete event-by-event description of the 
-input ARG is still required.
-
+% In essence, computing the KYF likelihood requires
+% ``replaying'' the  state of Hudson's simulation
+% algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
+% compute the exact configuration of lineages and distribution
+% of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
+% It therefore requires complete
+% detail about each of these events to be stored. 
+% Other approaches to defining the likelihood of an ARG under a coalescent-like 
+% model follow the same basic event-by-event 
+% strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
+% % [Are there others?  The bacterial stuff from Didelot etc seem to focus on
+% % a tree-by-tree likelihood, rather than global graph, so not clear it's the 
+% % same thing]
+% Computing the likelihood under the SMC rather than the Hudson coalescent 
+% simplifies bookkeeping somewhat,  % Dunno what I'm talking about here
+% but a complete event-by-event description of the 
+% input ARG is still required.
 Although the KYF formulation is a natural and elegant way to describe 
@@
 
-\section*{ARG likelihoods}
-
-In one of the first approaches to probilistic ARG inference that 
-regarded the ARG itself as an output rather than a latent variable,
-\cite{kuhner_maximum_2000} provided an explicit formula for 
-the likelihood of a realised ARG under the coalescent with
-recombination:
-\[
-\mathbb{P}[G|N, r] = \left(\frac{1}{2N}\right)^K
-r^X \exp\left(-\sum \left( \frac{1}{2N} \binom{k_i}{2} + r\ell_i\right)
-t_i\right),
-\]
-where [explain notation].
-Under this KYF formulation, the likelihood is computed by taking a 
-product of the probability of each realised event, and their 
-inter-event waiting times. While it is straightforward to write 
-this down, computation is not so easy, and in essence requires
-In essence, computing the KYF likelihood requires
-``replaying'' the  state of Hudson's simulation
-algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
-compute the exact configuration of lineages and distribution
-of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
-
-Computing the likelihood of a realised ARG under the SMC simplifies the 
-bookkeeping somewhat, because we no long have the possibility of 
-``trapped'' ancestral material, and each lineage can be represented 
-by a single interval representing their ancestral material. Nonetheless,
-we still require complete knowledge of the graph to compute the 
-required probablilities.
+% \section*{ARG likelihoods}
+
+% In one of the first approaches to probilistic ARG inference that 
+% regarded the ARG itself as an output rather than a latent variable,
+% \cite{kuhner_maximum_2000} provided an explicit formula for 
+% the likelihood of a realised ARG under the coalescent with
+% recombination:
+% \[
+% \mathbb{P}[G|N, r] = \left(\frac{1}{2N}\right)^K
+% r^X \exp\left(-\sum \left( \frac{1}{2N} \binom{k_i}{2} + r\ell_i\right)
+% t_i\right),
+% \]
+% where [explain notation].
+% Under this KYF formulation, the likelihood is computed by taking a 
+% product of the probability of each realised event, and their 
+% inter-event waiting times. While it is straightforward to write 
+% this down, computation is not so easy, and in essence requires
+% ``replaying'' the  state of Hudson's simulation
+% algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
+% compute the exact configuration of lineages and distribution
+% of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
+
+% % Computing the likelihood of a realised ARG under the SMC simplifies the 
+% % bookkeeping somewhat, because we no long have the possibility of 
+% % ``trapped'' ancestral material, and each lineage can be represented 
+% % by a single interval representing their ancestral material. Nonetheless,
+% % we still require complete knowledge of the graph to compute the 
+% % required probablilities.
 
@@
 % % % % % % % % % % % % % % % % % %
-\subsection{ARGs} \label{par:recording}
-
+\subsection*{ARGs} \label{par:recording}
 Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
@@
 
-Crucial for the equivalence between both representations is the presence of
+Crucial for the equivalence between the two representations is the presence of
 nodes that only have a single child along one or more local trees --
@@
 for instance, sample $d$ inherited a longer stretch of genome from ancestor $f$
-than did sample $c$.
+than sample $c$.
 In the top plot, this is visible because the pink bar below node $f$
@@
 us to uniquely identify all lineages that were hit by a recombination event,
-even if we simplify the ARG by removing all recombinant nodes
+even if we simplify %~\citep{wong_general_2023,kelleher_efficient_2018} 
+the ARG by removing all recombinant nodes
 ($A$, $B$, and $C$ in Fig.\ref{fig:smc-unary}).
@@
 Without the recombinant nodes we no longer know the precise time of each recombination event,
-but we do still have constraints on the time:
+but we do still have constraints on it;
 for instance, we know that $C$ happened somewhere between node $f$ and node $h$.
@@
 This is again easier to visualise using the
-graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
+graph representation (blue shaded area, Fig.\ref{fig:smc-unary}).
 Although these unary nodes are useful and to some degree clearly inferrable,
-they are not always represented by simulators or ARG reconstruction methods.
+they are not always represented by simulators or ARG inference methods.
 However, to our understanding this is because
@@
 (and their implications for inherited haplotypes).
+See~\cite{fritze2024forest} for more discussion on unary nodes
+and haplotype aware ARG comparison metrics.
 

### 2025-02-14 paper.tex
Pass trhough the SMC backwards in time section

@@
 not because of any question of knowability.
-Currently,
-\argweaver{} \citep{rasmussen_genome-wide_2014},
-\tsinfer{} \citep{kelleher_inferring_2019},
-\kwarg{} \citep{ignatieva_kwarg_2021}, and
-\texttt{SINGER} \citep{deng_robust_2024} estimate ARGs 
-including unary portions of coalescence nodes.
-However, it is unclear to what degree these are well-inferred,
-as most evaluation of inferred ARGs thus far has focused on marginal trees
-(e.g., \citet{brandt2022evaluation, kelleher_inferring_2019},
-but see \citet{deng2021distribution})
-and thus completely ignore such unary nodes
-(and their implications for inherited haplotypes).
+% JK I think we can but this out without-loss now given Halley's paper?
+% Currently,
+% \argweaver{} \citep{rasmussen_genome-wide_2014},
+% \tsinfer{} \citep{kelleher_inferring_2019},
+% \kwarg{} \citep{ignatieva_kwarg_2021}, and
+% \texttt{SINGER} \citep{deng_robust_2024} estimate ARGs 
+% including unary portions of coalescence nodes.
+% However, it is unclear to what degree these are well-inferred,
+% as most evaluation of inferred ARGs thus far has focused on marginal trees
+% (e.g., \citet{brandt2022evaluation, kelleher_inferring_2019},
+% but see \citet{deng2021distribution})
+% and thus completely ignore such unary nodes
+% (and their implications for inherited haplotypes).
 See~\cite{fritze2024forest} for more discussion on unary nodes
@@
 \citep{wiuf_recombination_1999}.
-Such an along-the-genome approach is considerably more complex as the next 
+This along-the-genome approach is considerably more complex as the next 
 tree to be simulated depends on all previous trees.
@@
 ``local trees'' representation of ARGs, 
-the Markovian property can be formulated as follows: 
-the distribution of the next tree within the series
+the Markovian property can be formulated as follows.
+The distribution of the next tree within the series
 of local trees can be generated only knowing the current tree,
@@
 the original lineages are removed, and $L(t)$ is updated accordingly.
-[JK notes: need to explain here that this deliberately introduced 
-asymmetry doesn't affect the actual process. Also need to comment on 
-how the process terminates. Basically has the same subtleties as McVean and 
-Cardin, and not important for our purposes.]
-
+% [JK notes: need to explain here that this deliberately introduced 
+% asymmetry doesn't affect the actual process. Also need to comment on 
+% how the process terminates. Basically has the same subtleties as McVean and 
+% Cardin, and not important for our purposes.]
 For later use, also define $C_j(t)$ to be the set of

### 2025-02-14 paper.tex
Pass through likelihood calculations

@@
 \subsection{Likelihoods} \label{par:liks}
-
-Here we will compute the likelihood of an ARG in the absence of the
-exact time to recombination events. To do this, we will rely on the unary nodes
-to identify each lineage involved in a recombination event as described in \ref{par:recording}.
-
 We have defined the SMC process in terms of rates of various types of events.
@@
 (also called the ``total hazard''). 
+As discussed in the Introduction, in many cases these events are not what is 
+estimated by inference methods, which produce, more fundamentally, a collection
+of inheritance relationships (edges) between ancestral haplotypes (nodes).
+Our goal, therefore, is to decompose the overall likelihood into the
+per-node and edge contributions to enable convenient and efficient 
+calculation. 
+% , as well as graceful degradation in the presence of incomplete information.
+
 %\comment{if we want to spell it out even further:
@@
 
-The structure of the ARG, as a collection of \emph{edges} $E$ and \emph{nodes} $N$,
-as well as the definition of both recombination and common ancestry events, allows us
-to cleanly factor the likelihood of the ARG across its edges and nodes.
-Not only does recombination
-naturally only affect one lineage at a time without even depending
-on the state of other lineages. Common ancestry events can be equally
-associated with a single edge given definition \eqref{def:coal}.
-The next paragraphs detail precisely how the number and the nature of the realized events
-can be determined based on the number of
-parents and children of each of the nodes in $N$ as well as how the total hazard can be
-factorized into the individual contributions of each of the edges.
-
-More specifically,
-let $\deg_o(u) = \vert\{e \in E : p_e=u\}\vert$ be the out-degree of a node $u$,
+% The structure of the ARG, as a collection of \emph{edges} $E$ and \emph{nodes} $N$,
+% as well as the definition of both recombination and common ancestry events, allows us
+% to cleanly factor the likelihood of the ARG across its edges and nodes.
+% Not only does recombination
+% naturally only affect one lineage at a time without even depending
+% on the state of other lineages. Common ancestry events can be equally
+% associated with a single edge given definition \eqref{def:coal}.
+% The next paragraphs detail precisely how the number and the nature of the realized events
+% can be determined based on the number of
+% parents and children of each of the nodes in $N$ as well as how the total hazard can be
+% factorized into the individual contributions of each of the edges.
+
+Let $\deg_o(u) = \vert\{e \in E : p_e=u\}\vert$ be the out-degree of a node $u$,
 i.e., the number of edges having $u$ as a parent.
@@
     \caption{
-    % Graphical representation of the events associated with the edge $(p,c,x,y)$.
-    % The lineage carried by $c$ spans $[x_c,y)$ until 
-% a recombination event (blue dotted line) splits it
-    %     into two segments. Left and right hand segments coalesce at times $t_p'$ and $t_p$ respectively,
-    %     bounding the time to recombination by $(c. \min(t_p, t_{p'})$.
-    %     Note that the contribution to the recombination hazard by edge $(p,c,x,y)$ only covers ``half'' of the
-    %     green shaded area. The other half is contributed by edge $(p', c, x_c, x)$. 
+    Graphical representation of the events associated with the edge $(p,c,x,y)$.
+    The lineage carried by $c$ spans $[x_c,y)$ until 
+a recombination event (blue dotted line) splits it
+        into two segments. Left and right hand segments coalesce at times $t_p'$ and $t_p$ respectively,
+        bounding the time to recombination by $(c. \min(t_p, t_{p'})$.
+        Note that the contribution to the recombination hazard by edge $(p,c,x,y)$ only covers ``half'' of the
+        green shaded area. The other half is contributed by edge $(p', c, x_c, x)$. 
         \label{fig:likelihood}
@@
 (e.g., the left hand segment ancestral to node $c$
-in fig.\ ~\ref{fig:likelihood}),
+in Fig.~\ref{fig:likelihood}),
 then no recombination occurred along this edge and 
@@
 %the non-reciprocal nature of coalescence events as defined in \ref{def:coal}.
-
-The likelihood of ARG defined by the set of edges $E$ and nodes $N$
-given parameters $\theta$ then is
+The overall likelihood of an ARG defined by the set of edges $E$ and nodes $N$
+given parameters $\theta$ is then given by
 \begin{equation}\label{eq:full-lik}
@@
 % intervals; rate changes along the genome can be incorporated similarly.
-Finally, we have only considered the topology and branch length information of any ARG.
-Extending this expression by adding the likelihood of observing a set of mutations
-given an ARG is given by
+Note that we have only considered the topology and branch length information here
+an not considered mutational information.
+The likelihood of a set of mutations given an ARG is
 \begin{equation}\label{eq:lik-mut}
@@
 information we can approximate the true integral as the product of the
-integral of the slices. Figure~\ref{fig:3-arg-slices} demonstrates this idea
+integral of the slices. Fig.~\ref{fig:3-arg-slices} demonstrates this idea
 on a toy example. We simulated an ARG under human-like parameters with

### 2025-02-14 paper.tex

@@
 \end{equation}
-% One significant advantage of this formulation is that the computation time
-% is linear in the number of edges, and an algorithm can be devised to 
-% to efficiently keep track of $I_e(t)$ (see Appendix).
+One significant advantage of this formulation is that the computation time
+is linear in the number of edges, and an efficient algorithm exists
+to keep track of $I_e(t)$ (see Appendix).
 % Although the first two factors are easily expressed in terms of numbers of edges and nodes,
@@
 % % % % % % % % % % % % % % % % % %
-\subsection{Flexibility}
+\subsection*{Flexibility}
 % likelihood surface
-The last paragraph aims to highlight the flexibility of this approach 
-and the impact of the various
-degrees of completeness of the ARG on the likelihood.
-Note that the figures in this paragraph only serve this exact purpose,
-we are not claiming to have designed a new parameter estimation approach.
-
+% The last paragraph aims to highlight the flexibility of this approach 
+% and the impact of the various
+% degrees of completeness of the ARG on the likelihood.
+% Note that the figures in this paragraph only serve this exact purpose,
+% we are not claiming to have designed a new parameter estimation approach.
 In line with the observation that the presence of locally unary nodes 
@@
 
-\begin{figure}[!ht]
+\begin{figure}
     \centering
@@
 
-\section{Discussion}
-
+\section*{Discussion}
+[JK: this needs significant work]
 Here, we have introduced a scalable algorithm to compute likelihoods
@@
 
-[TODO another paragraph summarising what you actually do]
+% [TODO another paragraph summarising what you actually do]
 

### 2025-02-14 paper.tex
Light resectioning

@@
 
-\section{Introduction}
+\section*{Introduction}
 Following recent breakthroughs in simulation and inference methods,
@@
 % % % % % % % % % % % % % % % % % %
-\subsection{SMC backwards-in-time}\label{par:description}
+\subsection*{SMC backwards-in-time}\label{par:description}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Likelihoods} \label{par:liks}
+\subsection*{ARG likelihood} \label{par:liks}
 We have defined the SMC process in terms of rates of various types of events.
@@
 % % % % % % % % % % % % % % % % % %
-\subsection*{Flexibility}
+\subsection*{Illustrative examples}
 % likelihood surface
@@
 % we are not claiming to have designed a new parameter estimation approach.
+[JK: Need a bit more narraitive here to feed back into the setup 
+in the intro. Computing a likelihood on any ARG, graceful degradation
+of performance with incomplete input..]
+
 In line with the observation that the presence of locally unary nodes 

### 2025-02-17 paper.tex
Struggles to understand the polytomy stuff

@@
 % Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
+% % Captions will be left justified
+% \usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+% \renewcommand{\figurename}{Fig}
 
@@
 estimated by inference methods, which produce, more fundamentally, a collection
-of inheritance relationships (edges) between ancestral haplotypes (nodes).
+of genetic inheritance relationships (edges) between ancestral haplotypes (nodes).
 Our goal, therefore, is to decompose the overall likelihood into the
-per-node and edge contributions to enable convenient and efficient 
-calculation. 
+per-node and edge contributions, allowing us to compute likelihoods
+for the incomplete ancestries estimated by current methods.
 % , as well as graceful degradation in the presence of incomplete information.
@@
 i.e., the number of edges having $u$ as a parent.
-Since mergers are binary in the SMC,
+Because mergers are binary in the SMC,
 this must have been the result of $\deg(u) - 1$ common ancestry events,
@@
 Note however that this description is general and holds for the case of
-more-than-binary mergers. Indeed, inferred ARGs (e.g.\ \texttt{tsinfer}) may
-contain polytomies. By treating these as if there were zero-length edges
+more-than-binary mergers. By treating such polytomies as if there were zero-length edges
 between a sequence of mergers, the order of these unresolved mergers does 
@@
     \includegraphics[width=0.65\textwidth]{figures/single_event.png}
-    % FIXME commenting this caption out because it mucks up syntax highlighting for me.
     \caption{
     Graphical representation of the events associated with the edge $(p,c,x,y)$.
-    The lineage carried by $c$ spans $[x_c,y)$ until 
-a recombination event (blue dotted line) splits it
-        into two segments. Left and right hand segments coalesce at times $t_p'$ and $t_p$ respectively,
-        bounding the time to recombination by $(c. \min(t_p, t_{p'})$.
-        Note that the contribution to the recombination hazard by edge $(p,c,x,y)$ only covers ``half'' of the
-        green shaded area. The other half is contributed by edge $(p', c, x_c, x)$. 
+    The lineage carried by $c$ (dark green bar) spans $[x_c,y)$ 
+    until a recombination event (blue
+    dotted line) splits it into two segments. Left and right hand segments coalesce
+    (with the pink and yellow lineages, respectively)
+    at times $t_{p'}$ and $t_p$ respectively, bounding the time to recombination by
+    % JK used to be: $(c. \min(t_p, t_{p'})$
+    $\min(t_p, t_{p'}$.
+    % JK: not sure this helps? 
+    % Note that the contribution to the recombination hazard
+    % by edge $(p,c,x,y)$ only covers ``half'' of the green shaded area. The other
+    % half is contributed by edge $(p', c, x_c, x)$. 
     \label{fig:likelihood}
@@
 where $x_c$ is the leftmost position of the edges whose children are $c$.
-
 % In that case $x=x_{c0}$ and $c$ did not coalesce until $t_p$.
@@
 % such that:
-
 %\comment{This is the old equation; I think the $e^{-rs}$ should not be there,
@@
 % we are not claiming to have designed a new parameter estimation approach.
-[JK: Need a bit more narraitive here to feed back into the setup 
-in the intro. Computing a likelihood on any ARG, graceful degradation
-of performance with incomplete input..]
-
-In line with the observation that the presence of locally unary nodes 
-(see section \ref{par:recording}) is crucial for the equivalence between
-the graph or local trees view of an ARG, one can gradually simplify down
-a simulated ARG by gradually removing the various types of locally unary nodes.
-Figure \ref{fig:lik-surface} shows the impact of these simplification steps
-on the likelihood for a single simulated ARG.
-The log likelihood ratio is shown to improve comparability.
+
+Equation~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
+collection of nodes (ancestral haplotypes that lived at some time 
+$t$ in the past) and edges (a record that child node $c$ 
+inherited the genomic interval $[x, y)$ from parent node $p$)
+under the SMC. Other than some basic consistency constraints (e.g.\ parents
+must be older than children) there are no requirements on the topology
+described, and in particular, no requirement that the events from the 
+model (the SMC) should be explicitly identified in the graph. Our 
+formulation reasons about the events that \emph{could have resulted in}
+the observed nodes and edges, and therefore frees us from the requirement
+of having to explicitly write those down. This gives a great deal 
+of flexibility, which we illustrate in this section via some examples.
+
+The first example concerns the simplest possible polytomy (see
+the Discussion for why we consider polytomies important).
+Suppose we have a graph consisting of four nodes, $c_1,c_2,c_3,p$ and three edges
+${(p, c_j, 0, 1): 1 \leq j \leq 3}$. The child nodes are samples at time
+zero, and the parent has time $1$. The likelihood of this ARG is undefined
+under HKY because the three-way star topology described cannot be interpreted as 
+a sequence of binary coalescence events. However, it is clearly 
+\emph{consistent with} such a sequence of events, which we can reason
+about probabilistically. In our formulation we [Peter/Gertjan, help me 
+out any hold my hand a bit here. How exactly to we assign a likelihood
+to such a polytomy? Why do I get a larger likelihood for the ternary
+tree described above than an example in which I break the polytomy? 
+Naively, I would have thought that the likelihood should have been 
+lower, but I clearly don't understand likelihoods...]
+
+Our second example shows a different aspect of this robustness to missing 
+data. In Fig~\ref{fig:lik-surface} we can see the likelihood-ratio curves
+obtained when we compute the likelihood of an example simulation
+at varying degrees of completeness.[JK: WIP]
 
@@
 \end{figure}
+
+In line with the observation that the presence of locally unary nodes 
+(see section \ref{par:recording}) is crucial for the equivalence between
+the graph or local trees view of an ARG, one can gradually simplify down
+a simulated ARG by gradually removing the various types of locally unary nodes.
+Figure \ref{fig:lik-surface} shows the impact of these simplification steps
+on the likelihood for a single simulated ARG.
+The log likelihood ratio is shown to improve comparability.
 
@@
 \section*{Discussion}
+
+The key contribution of this paper is to rederive the classical KYF
+ARG likelihood function in a way that supports partial and imprecise
+input. 
+The underlying process that we reason about probabilistically is still
+composed of events affecting lineages, but, acknowledging that we 
+do not necessarily observe or estimate these events, we calculate 
+the possible contributions of various type of event to the inheritance
+relationships that are provided as input.
+
 [JK: this needs significant work]

### 2025-02-18 paper.tex
Draft of explaining example

@@
 \usepackage{placeins}
-\usepackage[hidelinks]{hyperref}
+% \usepackage[hidelinks]{hyperref}
 
@@
 us to uniquely identify all lineages that were hit by a recombination event,
-even if we simplify %~\citep{wong_general_2023,kelleher_efficient_2018} 
+even if we simplify 
 the ARG by removing all recombinant nodes
@@
 
-Our second example shows a different aspect of this robustness to missing 
-data. In Fig~\ref{fig:lik-surface} we can see the likelihood-ratio curves
-obtained when we compute the likelihood of an example simulation
-at varying degrees of completeness.[JK: WIP]
-
 \begin{figure}
@@
     \includegraphics[width=\textwidth]{figures/likratio.png}
-    \caption{Log likelihood-ratio curves for ARGs simulated under the SMC with
+    \caption{TODO add A/B and change ``unary regions''.
+    Log likelihood-ratio curves for ARGs simulated under the SMC with
     various degrees of completeness. The likelihood ratio is computed with
@@
 
-In line with the observation that the presence of locally unary nodes 
-(see section \ref{par:recording}) is crucial for the equivalence between
-the graph or local trees view of an ARG, one can gradually simplify down
-a simulated ARG by gradually removing the various types of locally unary nodes.
-Figure \ref{fig:lik-surface} shows the impact of these simplification steps
-on the likelihood for a single simulated ARG.
-The log likelihood ratio is shown to improve comparability.
+Our second example shows a different aspect of this robustness to missing 
+data. In Fig~\ref{fig:lik-surface}A we can see the likelihood-ratio curves
+obtained when we compute the likelihood of an example simulation,
+in which the ARG is at three different levels of completeness
+(illustrated in Fig~\ref{fig:lik-surface}B).
+We first run a simulation under the SMC
+using the \texttt{record\_full\_arg} option in \texttt{msprime}. This 
+ARG explicitly identifies all of the recombination and common 
+ancestor events that occurred in the simulation, and 
+we can therefore compute a likelihood under the HKY formulation
+(for example, using the \texttt{msprime.log\_arg\_likelihood} 
+function). The example full ARG in 
+Fig~\ref{fig:lik-surface}B has three nodes that are ``locally
+unary''~\citep{wong_general_2023}. Nodes 6 and 7 correspond to the 
+left and right parents of the recombination event causing the 
+breakpoint at position 536~\cite[we need two nodes to uniquely identify
+ the passage of ancestral material through recombination 
+events; see][]{baumdicker_efficient_2021}, and therefore present
+in only those trees. 
+Node 8 is a coalescent node in the right-hand tree, but unary
+in the left-hand tree, marking the passage of the lineage through
+this ancestral haplotype (similar to node g in Fig~\ref{fig:smc-unary}).
+Fig~\ref{fig:lik-surface}A shows that, remarkably, when we 
+remove the recombination nodes (e.g., 6/7) but keep the locally unary nodes (e.g., 8)
+and compute the likelihood using equation~\eqref{eq:full-lik},
+there is very little difference in the likelihood ratio
+curve from the exact value with complete input. 
+Most importantly, the minimum occurs at the true value of the 
+recombination parameter.
+When we further simplify
+the ARG to remove all unary nodes~\citep{wong_general_2023,kelleher_efficient_2018} 
+we obtain an ARG in the standard format used by \texttt{msprime}.
+Although there is substantial information about the correlation
+structure of trees in this representation, we lose important
+information about recombination. This is demonstrated in 
+Fig~\ref{fig:lik-surface}A, where the corresponding likelihood ratio curve
+is clearly biased.
 

### 2025-02-18 paper.tex
Pass on final example

@@
 
-Equation~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
+Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
 collection of nodes (ancestral haplotypes that lived at some time 
@@
     \centering
-    \includegraphics[width=\textwidth]{figures/likratio.png}
+    \includegraphics[width=0.8\textwidth]{figures/likratio.png}
     \caption{TODO add A/B and change ``unary regions''.
@@
 remove the recombination nodes (e.g., 6/7) but keep the locally unary nodes (e.g., 8)
-and compute the likelihood using equation~\eqref{eq:full-lik},
+and compute the likelihood using eq.~\eqref{eq:full-lik},
 there is very little difference in the likelihood ratio
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=0.8\textwidth]{figures/3-slices-whitespace.png}
+    \caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant
+    for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000,
+    (5000:\infty):10 000]$. Remaining parameters: $L=10$kb, $20$ diploid samples,
+    $r=2.5e-8$, $148$ edges in ARG. A single $N_e$ value was estimated for each
+    slice of the ARG as defined by the true population size change time points.
+    The inferred posterior distribution (right) using the SMC-likelihood was
+    obtained using a uniform prior. Number of edges per time slice
+    (from most recent slice to oldest): $[68, 57, 23]$. Locally unary nodes were
+    omitted for clarity.}
+    \label{fig:3-arg-slices}
+\end{figure}
+
+Our final example concerns [TODO motivate this example; why is 
+this interesting/useful?]
 % slicing of the ARG
@@
 into the product of two integrals of the same form.
-
 However, as long as the individual slices of the ARG contain sufficient
@@
 strategies during inference given that each slice of the ARG can be
-processed independently. And it further highlights the flexibility
-of the algorithm. Simply stated and from a data structure perspective,
-we can compute a likelihood under the SMC for any valid tree sequence.
-The \texttt{tskit} library provides a wide range
-of subsetting operations on tree sequences allowing us to slice
-the ARG in time and/or space, as well as to reduce the number of
-tracked samples. Any such subset of the tree sequence is itself
-a valid tree sequence for which we can again compute a likelihood.
+processed independently.
+
+% And it further highlights the flexibility
+% of the algorithm. Simply stated and from a data structure perspective,
+% we can compute a likelihood under the SMC for any valid tree sequence.
+% The \texttt{tskit} library provides a wide range
+% of subsetting operations on tree sequences allowing us to slice
+% the ARG in time and/or space, as well as to reduce the number of
+% tracked samples. Any such subset of the tree sequence is itself
+% a valid tree sequence for which we can again compute a likelihood.
 
@@
 %%% estimate parameters (not goal here) or use as instrument to improve accuracy to infer ARGs
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{figures/3-slices-whitespace.png}
-    \caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant
-    for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000,
-    (5000:\infty):10 000]$. Remaining parameters: $L=10$kb, $20$ diploid samples,
-    $r=2.5e-8$, $148$ edges in ARG. A single $N_e$ value was estimated for each
-    slice of the ARG as defined by the true population size change time points.
-    The inferred posterior distribution (right) using the SMC-likelihood was
-    obtained using a uniform prior. Number of edges per time slice
-    (from most recent slice to oldest): $[68, 57, 23]$. Locally unary nodes were
-    omitted for clarity.}
-    \label{fig:3-arg-slices}
-\end{figure}
 

### 2025-02-18 paper.tex
Rough pass on Discussion

@@
 Our final example concerns [TODO motivate this example; why is 
-this interesting/useful?]
+this interesting/useful? Presumably we don't have to jointly estimate
+ the ARG with demography, but can do so post-hoc? But, couldn't 
+you do that anyway if you had a full ARG using HKY?]
 % slicing of the ARG
@@
 this is true as long as the slicing operation does not erase the
-information on whether an edge is associated with the lefthand or
-righthand segment generated by a recombination event. In the absence
+information on whether an edge is associated with the left or
+right hand segment generated by a recombination event. In the absence
 of the exact time to recombination, the integral in
@@
 
-% Although we can associate a likelihood with any valid tree sequences,
-% there are some caveats. By treating each edge independently we can
-% naturally deal with the various degrees of completeness to which inheritance
-% is documented by the different ARG inference methods: e.g., polytomies, absence
-% of locally unary coalescence nodes, unpreserved identify of internal nodes
-% across marginal trees, etcetera.
-% However, many of these imprecisions violate the implicit
-% assumptions of the likelihood computation.
-% \comment{Do we want these remarks in here, or as part of the Discussion?}
-% 
-% For example, the described algorithm assumes the
-% presence of ``locally unary'' nodes to correctly identify the
-% full extent of all lineages going back through time and therefore recombination events.
-% Without unary nodes though, our algorithm will detect a recombination
-% event along both coalescing lineages in case of a (sub)tree height changes along
-% the ARG as both edges would switch parent nodes. This can be mitigated by limiting
-% the number of 'inferrable' recombination events per coalescence event to 1
+\section*{Discussion}
+
+The key contribution of this paper is to rederive the classical KYF
+ARG likelihood function in a way that supports partial and imprecise
+input, using a novel backwards-time derivation of the SMC.
+The underlying process that we reason about probabilistically is still
+composed of events affecting lineages, but, acknowledging that we 
+do not necessarily observe or estimate these events, we calculate 
+the possible contributions of various type of event to the inheritance
+relationships that are provided as input.
+Although we can associate a likelihood with any ARG in \texttt{tskit}
+format, there are some caveats. By treating each edge independently we can
+naturally deal with the various degrees of completeness to which inheritance
+is documented by the different ARG inference methods: e.g., polytomies, absence
+of locally unary coalescence nodes, unpreserved identify of internal nodes
+across local trees, etc.
+However, many of these imprecisions violate the implicit
+assumptions of the likelihood computation.
+For example, we assume the 
+presence of locally unary nodes to correctly identify the
+full extent of all lineages going back through time and therefore recombination events.
+Without these unary nodes, our algorithm will detect a recombination
+event along both coalescing lineages in case of a (sub)tree height changes along
+the ARG as both edges would switch parent nodes. 
+Locally unary nodes are not a common aspect of current ARG research, but 
+are estimated by \texttt{tsinfer}, and can be imputed to high
+degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
+
+[A paragraph about possililties for inference]
+
+[A paragraph about why what we've presented is useful, and a good starting
+point for other research. We provide an efficient algorithm and 
+reference implementation]
+
+% This can be mitigated by limiting
+% the number of `inferrable' recombination events per coalescence event to 1
 % (see fig \ref{sup:fig:rec-correction}).
@@
 
-\section*{Discussion}
-
-The key contribution of this paper is to rederive the classical KYF
-ARG likelihood function in a way that supports partial and imprecise
-input. 
-The underlying process that we reason about probabilistically is still
-composed of events affecting lineages, but, acknowledging that we 
-do not necessarily observe or estimate these events, we calculate 
-the possible contributions of various type of event to the inheritance
-relationships that are provided as input.
-
-[JK: this needs significant work]
-Here, we have introduced a scalable algorithm to compute likelihoods
-under the SMC for a general class of ARGs. By formulating the SMC backwards in time
-we were able to associate a single likelihood value with each edge in the ARG.
-The compatibility of the algorithm with the \texttt{tskit} library allows us to
-exploit its associated efficiencies both for the current work and its future
-applications. This work will contribute significantly
-to ARG-based analysis becoming a standard part of the population
-and statistical genetics toolkit.
-In particular, we see two main applications for the work presented here.
-Firstly, the ability to associate inferred ARGs with a likelihood given a demographic
-model and an associated set of parameters, will help bridge the gap between the tremendous recent
-advances in ARG inference and the current state of the art statistical inference of
-past demography.
-The second and probably the most important application of the ability to associate a more
-general class of ARGs to a principled model is the ability to boost the current efforts
-in ARG inference. 
-% great starting point for MCMC
-Any inferred ARG can be used a good starting point
-for a future MCMC-sampler. This potentially implies a massive runtime reduction by
-reducing the necessary burn-in period. Furthermore, by integrating out the exact time
-to the recombination event we further restrict the size of the ARG space that needs to
-be explored. This MCMC-sampler would again rely on the tree sequence data structure to
-efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
-% hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
-To further improve scalability, hybrid ARG-inference methods could be designed which
-rely on heuristic approaches to bring the ARG down from the initial pedigree-dominated phase,
-and subsequently use a principled approach to perform ARG inference for the second phase.
-This idea resembles the tried and tested approach of hybrid backwards-in-time simulations 
-that combine Wright-Fisher dynamics in the very recent past and coalescent simulations 
-for the remaining time \citep{bhaskar_distortion_2014, nelson_accounting_2020}. This hybrid
-approach has enabled large scale simulations while avoiding the documented biases of the
-coalescent relative to the Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. 
-The same approach could be thus used for ARG-inference where one can rely on a heuristic 
-inference method for the recent past, while using a model-based approach beyond that cutoff point.
-
-Finally, acknowledging the uncertainty associated with ARG inference does not
-necessarily require a full-blown MCMC sampler. Instead, recognising those parts of
-the ARG for which we lack information for confident reconstruction and resolving
-them stochastically by sampling from the coalescent rather than completely arbitrarily
-would already be a major step forward.
-
-% this part is only a very rough idea, might be better to leave it out if we can't make it more concrete.
-We believe that quantifying uncertainty does not have to be associated with sampling
-from the space of all compatible genealogies. Uncertainty could also
-be quantified and passed on as metadata to the nodes in the data format storing the ARG.
-
-[Note: a point to make here is that the idea of combining hueristic large-scale 
-inference with model based inference is analogous to the combination of 
-coalescent simulation and forwards time simulation. We can also draw lessons
-from developers of different methods cooperating and using the tskit library
-to enable this.]
+% [JK: this needs significant work]
+% Here, we have introduced a scalable algorithm to compute likelihoods
+% under the SMC for a general class of ARGs. By formulating the SMC backwards in time
+% we were able to associate a single likelihood value with each edge in the ARG.
+% The compatibility of the algorithm with the \texttt{tskit} library allows us to
+% exploit its associated efficiencies both for the current work and its future
+% applications. This work will contribute significantly
+% to ARG-based analysis becoming a standard part of the population
+% and statistical genetics toolkit.
+% In particular, we see two main applications for the work presented here.
+% Firstly, the ability to associate inferred ARGs with a likelihood given a demographic
+% model and an associated set of parameters, will help bridge the gap between the tremendous recent
+% advances in ARG inference and the current state of the art statistical inference of
+% past demography.
+
+% The second and probably the most important application of the ability to associate a more
+% general class of ARGs to a principled model is the ability to boost the current efforts
+% in ARG inference. 
+% % great starting point for MCMC
+% Any inferred ARG can be used a good starting point
+% for a future MCMC-sampler. This potentially implies a massive runtime reduction by
+% reducing the necessary burn-in period. Furthermore, by integrating out the exact time
+% to the recombination event we further restrict the size of the ARG space that needs to
+% be explored. This MCMC-sampler would again rely on the tree sequence data structure to
+% efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
+% % hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
+% To further improve scalability, hybrid ARG-inference methods could be designed which
+% rely on heuristic approaches to bring the ARG down from the initial pedigree-dominated phase,
+% and subsequently use a principled approach to perform ARG inference for the second phase.
+% This idea resembles the tried and tested approach of hybrid backwards-in-time simulations 
+% that combine Wright-Fisher dynamics in the very recent past and coalescent simulations 
+% for the remaining time \citep{bhaskar_distortion_2014, nelson_accounting_2020}. This hybrid
+% approach has enabled large scale simulations while avoiding the documented biases of the
+% coalescent relative to the Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. 
+% The same approach could be thus used for ARG-inference where one can rely on a heuristic 
+% inference method for the recent past, while using a model-based approach beyond that cutoff point.
+
+% Finally, acknowledging the uncertainty associated with ARG inference does not
+% necessarily require a full-blown MCMC sampler. Instead, recognising those parts of
+% the ARG for which we lack information for confident reconstruction and resolving
+% them stochastically by sampling from the coalescent rather than completely arbitrarily
+% would already be a major step forward.
+
+% % this part is only a very rough idea, might be better to leave it out if we can't make it more concrete.
+% We believe that quantifying uncertainty does not have to be associated with sampling
+% from the space of all compatible genealogies. Uncertainty could also
+% be quantified and passed on as metadata to the nodes in the data format storing the ARG.
+
+% [Note: a point to make here is that the idea of combining hueristic large-scale 
+% inference with model based inference is analogous to the combination of 
+% coalescent simulation and forwards time simulation. We can also draw lessons
+% from developers of different methods cooperating and using the tskit library
+% to enable this.]
 

### 2025-02-19 paper.tex

@@
 input have appeared. This progress on the long-standing problem of ARG
-inference has proceeded in two distinct directions. The first
-class of method in which dramatic progress has been made is in the
-statistically rigorous inference of ARGs under the Sequentially Markov
-Coalescent (SMC), which are now practical for tens-to-hundreds of samples. 
-The second class of recently-developed inference methods are based on
-approximate models and heuristics which scale to sample sizes two to three
+inference has proceeded in two distinct directions. First,
+the Bayesian inference of ARGs under the Sequentially Markov
+Coalescent (SMC), is now practical for tens-to-hundreds of samples. 
+Second, approximate models and heuristics can now scale to sample sizes two to three
 orders of magnitude larger. Although these heuristic methods are reasonably
@@
 In this paper we present a 
-novel backwards-time formulation of the SMC 
+backwards-time formulation of the SMC 
 and derive a straightforward definition 
@@
 Following recent breakthroughs in simulation and inference methods,
-there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
+there is now strong interest in applying Ancestral Recombination Graphs (ARGs)
 to a variety of questions in evolutionary
@@
 % more natrual to talk about the advantages of model-driven stuff next.
-The first class of inference methods are statistically rigorous approaches
+The first class of inference methods are Bayesian approaches
 that sample ARGs under a population genetics model such as the
@@
 Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
-ARGweaver~\citep{rasmussen_genome-wide_2014} is the most popular,
-consistently out-performed other methods in terms 
-of accuracy in a variety of benchmarks~\citep{brandt2022evaluation}[MORE
-citations,anywhere ARGweaver has been tested for accuracy]
+ARGweaver~\citep{rasmussen_genome-wide_2014} has been the most widely used and studied,
+and until recently consistently out-performed other methods in terms 
+of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
 and has succesfully been applied in several
-contexts~\citep[e.g.][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
+contexts~\citep[e.g.,][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
 stankowski_genetic_2024}. 
-ARGweaver works [by explanation.]
-The recently-introduced SINGER method takes a similar approach [doing something 
-slightly different]~\citep{deng_robust_2024}.
-While SINGER has not been evaluated as thoroughly as ARGweaver yet, it promises
-to be even more accurate while being substantially more
+The recently-introduced method SINGER takes a similar approach,
+and by using an improved Monte Carlo algorithm
+promises to be even more accurate while being substantially more
 scalable~\citep{deng_robust_2024}.
@@
 hueristic and approximation driven, favouring computational
-scalability over statistical rigor.
+scalability over a basis in a statistical model.
 Secondly, they all produce a single
@@
 of quantifying uncertainty. Thirdly (and most importantly for the 
-purposes of this paper) the ARGs they produce lack substantial
-detail about ancestral recombination events: the estimates are
-rough sketches, hopefully capturing large-scale properties, but lacking
-in the finer detail of outputs of ARGweaver or SINGER.
+purposes of this paper) the ARGs they produce 
+lack some of the information present in ARGweaver or SINGER. 
+This occurs, roughly speaking, because of a fundamental difference in approach:
+not all details of the ARG can be reliably inferred
+(e.g., the timing of recombination events, as discussed below);
+and the heuristic methods deal with this uncertainty by omitting these events,
+while ARGweaver and SINGER provide a sample from the posterior.
+(So, some aspects of the resulting ARGs should be interpreted slightly differently;
+see \citet{wong_general_2023} for more discussion.)
 
 There are substantial advantages to the statistically rigorous sampling 
-approaches of ARGweaver and SINGER. Firstly, they tend to be more 
-accurate; ARGweaver has consistently out-performed other methods in terms 
-of accuracy in a variety of benchmarks~\citep{brandt2022evaluation}[MORE
-citations,anywhere ARGweaver has been tested for accuracy]
-and, while SINGER has not been evaluated as thoroughly yet, it promises
-to be even more accurate~\citep{deng_robust_2024}. 
-Secondly, the direct connection with
-population genetic models greatly aids interpretability, and allows for
+approaches of ARGweaver and SINGER.
+Firstly, they tend to be more accurate when compared on the same data set.
+Secondly, the direct connection with population genetic models allows for
 natural extensions like the joint estimation of ARGs and
-demography~\citep{hubisz_mapping_2020}. Thirdly, sampling ARGs from a
+demography~\citep{hubisz_mapping_2020}.
+Thirdly, sampling ARGs from a
 posterior distribution provides a means of quantifying the uncertainty 
-around estimates, which can be used in downstream applications [citation
-of this?].
-
-Despite these significant advantages of rigorous inference, however,
-hueristic methods remain of interest because of the profound difference
-in computational scalability. Although SINGER is a major step forward
+around estimates, which can be used in downstream applications.
+
+On the other hand, heuristic methods have the potential
+to utilize the greater amount of information in large datasets,
+thus possibly producing even more accurate estimates
+(although this has not, to our knowledge, been tested).
+Although SINGER is a major step forward
 in terms of scalability over ARGweaver and other methods, inference is still 
 only feasible for hundreds of samples. 
-Tsinfer, ARG-Needle and Threads have 
+Meanwhile, tsinfer, ARG-Needle and Threads have 
 been applied to datasets of hundreds of \emph{thousands} of samples.
@@
 % around heuristic point estimates. We cannot explore ARG space
-% around the point estimates using (e.g. MCMC), as we have no way definition
+% around the point estimates using (e.g., MCMC), as we have no way definition
 % of improvement. We cannot compare models of demography, or evaluate the 
@@
 % model follow the same basic event-by-event 
-% strategy~\citep[e.g.][]{guo_recombination-aware_2022}.
+% strategy~\citep[e.g.,][]{guo_recombination-aware_2022}.
 % % [Are there others?  The bacterial stuff from Didelot etc seem to focus on
@@
 % to compute the likelihood of a general class of ARG under the SMC.
-To do this, we rederive the SMC under a backwards-time formulation,
-modelling the process in terms of lineages and [something] rather
-than events. Using this novel formulation, we then show how to 
-easily compute the likelihood of any ARG under the SMC and outline 
-an efficient algorithm to do so. We illustrate the flexibility of this
-approach via [some examples] and conclude by discussing avenues
-for future research opened up.
+To do this, we give a backwards-in-time description of the SMC
+which naturally leads to a straightforward method for computing the likelihood
+of any ARG, and describe an efficient algorithm to do so.
+We also illustrate the utility of this approach via some simple examples.
 
@@
 \includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
-\caption{ARG generated under the SMC, represented as a graph (top) and as a
+\caption{An ARG generated under the SMC, represented as a graph (top) and as a
 series of local trees (bottom). To ensure the one-to-one correspondence
@@
 consider that a node in a tree might be unary either because
-it's about to be (in the left-to-right sense) a coalescent node 
-or it already has been:
+it's \emph{about to be} (in the left-to-right sense) a coalescent node 
+or it \emph{already has been}:
 looking at the second tree in Fig.\ref{fig:smc-unary},
@@
 I_{jk} = \begin{cases}
-    1 & X_j > X_k \text{ and } y_k > x_j \\
+    1 & X_j > X_k \text{ and } x_j < y_k \\
     0 & \text{otherwise.}
@@
 We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$.
+Note that since $X_j > X_k$ implies that $x_j \ge x_k$,
+this condition implies that the left endpoint $x_j$ of $X_j$
+falls in the interval defined by $X_k$.
 From the definition of the strict total order it follows that if $I_{jk} = 1$
@@
 % material carried by that lineage has. As noted by \citet{mcvean_approximating_2005},
-% such intervals can be observed as the stalks on the marginal genealogies (e.g.\ marginal trees
+% such intervals can be observed as the stalks on the marginal genealogies (e.g., marginal trees
 % for $[25, 81)$ Fig. \ref{fig:smc-unary}). Recombination events occurring within these smaller
@@
     % JK used to be: $(c. \min(t_p, t_{p'})$
-    $\min(t_p, t_{p'}$.
+    $\min(t_p, t_{p'})$.
     % JK: not sure this helps? 
@@
 the probability of which is
-$\exp(-r \times \mathcal{A})$, where $r$ is the recombination rate
-and $\mathcal{A}$ is the total area (eligible links times depth) of the edge.
+$\exp(-r \mathcal{A}_e)$, where $r$ is the recombination rate
+and $\mathcal{A}_e$ is the total area (eligible links multiplied by length in time) of the edge.
+The length in time of the edge is just $t_p - t_c$,
+while the number of ``eleigible links'' in the edge is the number of adjacent base pairs in the edge
+on which recombination did not occur, which is either $y-x$
+(if this edge is the leftmost of all edges above node $c$)
+or $y-x-1$ (otherwise).
 So, the first contribution to the likelihood from this edge $e$ is
 \begin{equation}\label{eq:span}
-A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) .
+A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) ,
 \end{equation}
-Note that if a recombination did occur, then the number of links not
-affected by recombination is 1 less (see eq.~\eqref{eq:depth}).
+where as for lineages, $x_c$ is the leftmost position of the edges whose children are $c$.
+% Note that if a recombination did occur, then the number of links not
+% affected by recombination is 1 less (see eq.~\eqref{eq:depth}).
 
@@
 we need only consider what happens along the left side of each edge.
-Recall that we define a total order on edges,
+Following what we did for lineages, we define a total order on edges
 by ordering by left endpoint and breaking ties by child
@@
 The recombination event that split the two must have happened between $t_c$
-and the smaller of $t_p$ and $t_{p'}$. Note that two events must have happened
-between $t_c$ and $\min(t_p, t_{p'})$; taking all this into consideration,
+and the smaller of $t_p$ and $t_{p'}$.
+Taking all this into consideration,
 the contribution to the likelihood here is
@@
     \int_{t_c}^{t_{p} \wedge t_{p'}} e^{- r s} e^{-\lambda F(e, s, t_{p})} ds
-        & \qquad \text{if } x>x_{c} ,
+        & \qquad \text{if } x>x_{c} .
 \end{cases}
 \end{equation}
-where $x_c$ is the leftmost position of the edges whose children are $c$.
 % In that case $x=x_{c0}$ and $c$ did not coalesce until $t_p$.
@@
 %the non-reciprocal nature of coalescence events as defined in \ref{def:coal}.
-The overall likelihood of an ARG defined by the set of edges $E$ and nodes $N$
+The overall likelihood of an ARG $\mathcal{G}$ defined by the set of edges $E$ and nodes $N$
 given parameters $\theta$ is then given by
@@
     =
-    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(p, c) \in E} A_e(r) B_e(r, \lambda) .
+    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{e \in E} A_e(r) B_e(r, \lambda) .
 \end{equation}
@@
 an not considered mutational information.
-The likelihood of a set of mutations given an ARG is
+The likelihood of a set of mutations given an ARG $\mathcal{G}$ given mutation rate $\mu$ is
 \begin{equation}\label{eq:lik-mut}
@@
     =
-    \frac{1}{\mathcal{M}!} \prod_{(p, c, x, y) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
+    \prod_{e \in E} \frac{(\mu(t_p-t_c)(y-x))^{m_e}}{m_e!} e^{-\mu (t_p-t_c)(y-x)} ,
+    % I don't think the following is correct:
+    % \frac{1}{\mathcal{M}!} \prod_{(p, c, x, y) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
 \end{equation}
-where $\mathcal{M}$ the total number of mutations and $m_e$
-the number of mutations along edge $(c,p)$ \citep{mahmoudi_bayesian_2022}.
+where 
+% $\mathcal{M}$ the total number of mutations and 
+$m_e$ is the number of mutations along edge $e$ 
+and as before $e = (p,c,x,y)$
+\citep{mahmoudi_bayesian_2022}.
 
@@
 inherited the genomic interval $[x, y)$ from parent node $p$)
-under the SMC. Other than some basic consistency constraints (e.g.\ parents
+under the SMC. Other than some basic consistency constraints (e.g., parents
 must be older than children) there are no requirements on the topology
@@
 lower, but I clearly don't understand likelihoods...]
+\comment{The likelihood of the tree described is $e^{-3rt_p-3\lambda t_p} \lambda^2$;
+if the polytomy was broken by adding another node at $(1-u)t_p/2$ then
+the likelihood would be $e^{-(3-u)rt_p - (3-2u)\lambda t_p} \lambda^2$.
+The latter is larger because there's less total area in the tree
+for recombination to have not happened and less time over which coalescence hasn't happened.
+However, these are different trees, so I'm not sure what the point is?}
 
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{figures/likratio.png}
+    \includegraphics[width=0.8\textwidth]{figures/likelihood_surface}
     \caption{TODO add A/B and change ``unary regions''.
@@
     under the SMC has been computed using the exact time to the recombination event.
-    Simulation performed with \texttt{msprime}, parameters: $L=1$Mb, $20$ diploid samples,
-    $r=2.5e-8$, $N_e=1e4$.}
+    Simulation performed with \texttt{msprime}, parameters:
+    genome length 1Mb, 20 diploid samples,
+    $r=2.5\times10^{-8}$, $N_e=1\times10^4$.}
     \label{fig:lik-surface}
@@
 Fig~\ref{fig:lik-surface}B has three nodes that are ``locally
-unary''~\citep{wong_general_2023}. Nodes 6 and 7 correspond to the 
+unary''~\citep{wong_general_2023,fritze2024forest}. Nodes 6 and 7 correspond to the 
 left and right parents of the recombination event causing the 
@@
     \includegraphics[width=0.8\textwidth]{figures/3-slices-whitespace.png}
-    \caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant
-    for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000,
-    (5000:\infty):10 000]$. Remaining parameters: $L=10$kb, $20$ diploid samples,
-    $r=2.5e-8$, $148$ edges in ARG. A single $N_e$ value was estimated for each
-    slice of the ARG as defined by the true population size change time points.
+    \caption{Simulated ARG under the Hudson coalescent (left). $N_e$ is constant
+    for each of the 3 time slices:
+    from 0 to 1,000 generations ago $N_e$ is 10,000;
+    from 1,000 to 5,000 generations ago it is 5,000;
+    and past that it is 10,000.
+    Remaining parameters: genome length is 10kb, 20 diploid samples, and
+    $r=2.5\times10^{-8}$. 
+    There are 148 edges in the resulting ARG.
+    A single $N_e$ value was estimated for each
+    of the three slices of the ARG separated by
+    the true population size change time points.
     The inferred posterior distribution (right) using the SMC-likelihood was
-    obtained using a uniform prior. Number of edges per time slice
-    (from most recent slice to oldest): $[68, 57, 23]$. Locally unary nodes were
+    obtained using a uniform prior. The number of edges per time slice
+    (from most recent to oldest) was: 68, 57, and 23. Locally unary nodes were
     omitted for clarity.}
@@
 
-Our final example concerns [TODO motivate this example; why is 
-this interesting/useful? Presumably we don't have to jointly estimate
- the ARG with demography, but can do so post-hoc? But, couldn't 
-you do that anyway if you had a full ARG using HKY?]
 % slicing of the ARG
+Our final example explores decomposition of the likelihood
+into separate components for different segments of time.
 By explicitly associating a single likelihood with each edge,
@@
 The result is reasonably accurate, despite only using 10kb of sequence.
-This approach therefore strengthens
-the inherent ability of ARGs to provide us with temporal
-resolution and context to a dataset of genetic variation.
-It also opens up the opportunity for more efficient optimization
-strategies during inference given that each slice of the ARG can be
-processed independently.
+% This approach therefore strengthens
+% the inherent ability of ARGs to provide us with temporal
+% resolution and context to a dataset of genetic variation.
+% It also opens up the opportunity for more efficient optimization
+% strategies during inference given that each slice of the ARG can be
+% processed independently.
 
@@
 
-The key contribution of this paper is to rederive the classical KYF
-ARG likelihood function in a way that supports partial and imprecise
-input, using a novel backwards-time derivation of the SMC.
+The key contribution of this paper is to rederive a classical
+ARG likelihood function in a way that supports the sort of partial information
+provided by some modern ARG inference tools,
+using a novel backwards-time derivation of the SMC.
 The underlying process that we reason about probabilistically is still
-composed of events affecting lineages, but, acknowledging that we 
-do not necessarily observe or estimate these events, we calculate 
-the possible contributions of various type of event to the inheritance
-relationships that are provided as input.
-Although we can associate a likelihood with any ARG in \texttt{tskit}
-format, there are some caveats. By treating each edge independently we can
-naturally deal with the various degrees of completeness to which inheritance
+composed of events affecting lineages, but, because we
+do not necessarily observe or estimate these events,
+we integrate over the possible timings implied by the information
+in the input ARG.
+Although we can associate a likelihood with any ARG in \texttt{tskit} format,
+there are some caveats. By treating each edge independently,
+our algorithm will still produce a value for ARGs with
+the various degrees of completeness to which inheritance
 is documented by the different ARG inference methods: e.g., polytomies, absence
@@
 full extent of all lineages going back through time and therefore recombination events.
-Without these unary nodes, our algorithm will detect a recombination
-event along both coalescing lineages in case of a (sub)tree height changes along
-the ARG as both edges would switch parent nodes. 
+% Without these unary nodes, our algorithm will detect a recombination
+% event along both coalescing lineages in case of a (sub)tree height changes along
+% the ARG as both edges would switch parent nodes. 
 Locally unary nodes are not a common aspect of current ARG research, but 
@@
 reference implementation]
+
+[A paragraph about another major sort of uncertainty: polytomies]
 
@@
 
-All scripts and data used for this manuscript are available on https://github.com/gertjanbisschop/smc-bit-paper
-An implementation of the algorithm is available on https://github.com/gertjanbisschop/runsmc.
+All scripts and data used for this manuscript are available at
+\url{https://github.com/gertjanbisschop/smc-bit-paper},
+and an implementation of the algorithm is available at
+\url{https://github.com/gertjanbisschop/runsmc}.
 \FloatBarrier

### 2025-02-19 paper.tex
Some bibliography tweaks

@@
 \usepackage{placeins}
-% \usepackage[hidelinks]{hyperref}
+\usepackage[hidelinks]{hyperref}
 
@@
 likelihood under models such as the SMC under present-day formulations.
-In particular, hueristic inference methods typically do not estimate
+In particular, heuristic inference methods typically do not estimate
 precise details about recombination events, which are currently
@@
 Firstly, they are all 
-hueristic and approximation driven, favouring computational
+heuristic and approximation driven, favouring computational
 scalability over a basis in a statistical model.
@@
 % recombination events, and there is no clear way to compute a likelihood for 
-% the approximate structures output by hueristic methods.
+% the approximate structures output by heuristic methods.
 
@@
 for the incomplete ancestries estimated by current methods.
+% Feel like we need a single sentence here that ties the first and second paragraph together
+Using the out- and in-degree of the nodes we'll first reason about the number 
+of common ancestor events and recombination events that could have resulted in 
+the observed ARG. Next the information contained in the edges is leveraged to 
+compute the total hazard associated with both types of events.
 % , as well as graceful degradation in the presence of incomplete information.
@@
 
-% [Note: a point to make here is that the idea of combining hueristic large-scale 
+% [Note: a point to make here is that the idea of combining heuristic large-scale 
 % inference with model based inference is analogous to the combination of 

### 2025-02-20 figures/generate-figs/plot_likelihood_surface.py
Rejig the likelihood figure

@@ -4,64 +4,78 @@ import matplotlib.pyplot as plt
 import matplotlib.lines as mlines
 import pandas as pd
 
-REC = 1.25e-8
+def plot(df_rat, path):
 
-df_rat = pd.read_csv("likelihood_surface.csv").set_index("rho")
+    REC = 1.25e-8
 
-fig, ax = plt.subplots(1, 1)
 
-colours = {
-    "full_arg_smc": "tab:red",
-    "unary_coal": "tab:blue",
-    "fully_simplified": "tab:orange",
-    "full_arg_hudson": "tab:purple",
-}
-titles = {
-    "full_arg_smc": "Full ARG (SMC)",
-    "unary_coal": "RE nodes removed",
-    "fully_simplified": "Fully simplified",
-    "full_arg_hudson": "Full ARG (HKY)",
-}
+    fig, ax = plt.subplots(1, 1)
 
-line_styles = {
-    "0": "solid",
-    "0.05": "dashed",
-    "0.25": "dotted",
-}
+    colours = {
+        "full_arg_smc": "tab:red",
+        "unary_coal": "tab:blue",
+        "fully_simplified": "tab:orange",
+        "full_arg_hudson": "tab:purple",
+    }
+    titles = {
+        "full_arg_smc": "Full ARG (SMC)",
+        "unary_coal": "RE nodes removed",
+        "fully_simplified": "Fully simplified",
+        "full_arg_hudson": "Full ARG (HKY)",
+    }
 
-ls_patches = [
-    mlines.Line2D([0], [0], color="black", lw=1, ls=ls, label=f"{float(label):.0%}")
-    for label, ls in line_styles.items()
-]
+    line_styles = {
+        "0": "solid",
+        "0.05": "dashed",
+        "0.25": "dotted",
+    }
 
-x = df_rat.index
-for col in df_rat:
-    if col.startswith("poly"):
-        splits = col.split("_")
-        arg = "_".join(splits[1:-1])
-        poly = splits[-1]
-    else:
-        arg = col
-        poly = "0"
+    ls_patches = [
+        mlines.Line2D([0], [0], color="black", lw=1, ls=ls, label=f"{float(label):.0%}")
+        for label, ls in line_styles.items()
+    ]
 
-    y = df_rat[col]  # / df_rat[arg]
-    line = ax.semilogx(
-        x,
-        y,
-        color=colours[arg],
-        ls=line_styles[poly],
-        label="_nolegend_" if poly != "0" else titles[arg],
[…]

### 2025-02-20 paper.tex
Pass through the Robustness section

@@
 % % % % % % % % % % % % % % % % % %
-\subsection*{Illustrative examples}
-% likelihood surface
-% The last paragraph aims to highlight the flexibility of this approach 
-% and the impact of the various
-% degrees of completeness of the ARG on the likelihood.
-% Note that the figures in this paragraph only serve this exact purpose,
-% we are not claiming to have designed a new parameter estimation approach.
-
+\subsection*{Robustness to partial information}
 Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
@@
 the observed nodes and edges, and therefore frees us from the requirement
-of having to explicitly write those down. This gives a great deal 
-of flexibility, which we illustrate in this section via some examples.
-
-The first example concerns the simplest possible polytomy (see
-the Discussion for why we consider polytomies important).
-Suppose we have a graph consisting of four nodes, $c_1,c_2,c_3,p$ and three edges
-${(p, c_j, 0, 1): 1 \leq j \leq 3}$. The child nodes are samples at time
-zero, and the parent has time $1$. The likelihood of this ARG is undefined
-under HKY because the three-way star topology described cannot be interpreted as 
-a sequence of binary coalescence events. However, it is clearly 
-\emph{consistent with} such a sequence of events, which we can reason
-about probabilistically. In our formulation we [Peter/Gertjan, help me 
-out any hold my hand a bit here. How exactly to we assign a likelihood
-to such a polytomy? Why do I get a larger likelihood for the ternary
-tree described above than an example in which I break the polytomy? 
-Naively, I would have thought that the likelihood should have been 
-lower, but I clearly don't understand likelihoods...]
-\comment{The likelihood of the tree described is $e^{-3rt_p-3\lambda t_p} \lambda^2$;
-if the polytomy was broken by adding another node at $(1-u)t_p/2$ then
-the likelihood would be $e^{-(3-u)rt_p - (3-2u)\lambda t_p} \lambda^2$.
-The latter is larger because there's less total area in the tree
-for recombination to have not happened and less time over which coalescence hasn't happened.
-However, these are different trees, so I'm not sure what the point is?}
+of having to explicitly write those down. In this section we 
+illustrate the robustness
+of this approach to incomplete information about recombination and 
+coalescent events via a proof-of-concept example, in which we 
+remove various levels of detail from a simulated ARG.
 
@@
     \includegraphics[width=0.8\textwidth]{figures/likelihood_surface}
-    \caption{TODO add A/B and change ``unary regions''.
-    Log likelihood-ratio curves for ARGs simulated under the SMC with
-    various degrees of completeness. The likelihood ratio is computed with
-    respect to the true recombination rate. In case of the full ARG the likelihood
-    under the SMC has been computed using the exact time to the recombination event.
-    Simulation performed with \texttt{msprime}, parameters:
-    genome length 1Mb, 20 diploid samples,
-    $r=2.5\times10^{-8}$, $N_e=1\times10^4$.}
-    \label{fig:lik-surface}
+    \caption{Log likelihood-ratio curves for ARGs containing
+     varying degrees of information about the realised simulation.
+    The ``Full ARG'' here is the result of an 
+    \texttt{msprime} simulation of the SMC (100 diploid samples, genome length 1Mb, 
+    $r=2.5\times10^{-8}$, $N_e=1\times10^4$) which contains complete 
+    information about the simulated events.
+    with various degrees of completeness. The likelihood ratio is computed with
+    respect to the true recombination rate. 
+    \label{fig:lik-surface}}
 \end{figure}
 
-Our second example shows a different aspect of this robustness to missing 
-data. In Fig~\ref{fig:lik-surface}A we can see the likelihood-ratio curves
-obtained when we compute the likelihood of an example simulation,
-in which the ARG is at three different levels of completeness
-(illustrated in Fig~\ref{fig:lik-surface}B).
-We first run a simulation under the SMC
-using the \texttt{record\_full\_arg} option in \texttt{msprime}. This 
-ARG explicitly identifies all of the recombination and common 
-ancestor events that occurred in the simulation, and 
-we can therefore compute a likelihood under the HKY formulation
-(for example, using the \texttt{msprime.log\_arg\_likelihood} 
-function). The example full ARG in 
-Fig~\ref{fig:lik-surface}B has three nodes that are ``locally
-unary''~\citep{wong_general_2023,fritze2024forest}. Nodes 6 and 7 correspond to the 
-left and right parents of the recombination event causing the 
-breakpoint at position 536~\cite[we need two nodes to uniquely identify
- the passage of ancestral material through recombination 
-events; see][]{baumdicker_efficient_2021}, and therefore present
-in only those trees. 
-Node 8 is a coalescent node in the right-hand tree, but unary
-in the left-hand tree, marking the passage of the lineage through
-this ancestral haplotype (similar to node g in Fig~\ref{fig:smc-unary}).
-Fig~\ref{fig:lik-surface}A shows that, remarkably, when we 
-remove the recombination nodes (e.g., 6/7) but keep the locally unary nodes (e.g., 8)
-and compute the likelihood using eq.~\eqref{eq:full-lik},
-there is very little difference in the likelihood ratio
-curve from the exact value with complete input. 
-Most importantly, the minimum occurs at the true value of the 
-recombination parameter.
-When we further simplify
-the ARG to remove all unary nodes~\citep{wong_general_2023,kelleher_efficient_2018} 
-we obtain an ARG in the standard format used by \texttt{msprime}.
-Although there is substantial information about the correlation
-structure of trees in this representation, we lose important
-information about recombination. This is demonstrated in 
-Fig~\ref{fig:lik-surface}A, where the corresponding likelihood ratio curve
-is clearly biased.
+In Fig~\ref{fig:lik-surface} we begin with a realisation of the SMC as a ``full
+ARG'', that is, containing complete information about every simulated event.
+We then plot the likelihood computed by Eq.~\eqref{eq:full-lik} 
+as we vary the recombination rate parameter $r$ (as a ratio
+of the likelihood computed at the true value of $r$). Unsurprisingly, we can
+see that the minimum of the likelihood-ratio curve when using the 
+full simulated ARG is at the true value of the parameter.
+What is surprising is that we obtain an indistinguishable curve when 
+we remove all the recombination nodes from this ARG, and therefore 
+lose information about the precise timing of recombination events.
+In this case, the locally unary portions of coalescent nodes have given
+all the information that we need about recombination events.
+Removing all locally unary nodes in the fully simplified ARG does result
+in a loss of information, and we can see that the minimum of the 
+likelihood ratio surface is biased away from its true value.
+
+% Our second example shows a different aspect of this robustness to missing 
+% data. In Fig~\ref{fig:lik-surface}A we can see the likelihood-ratio curves
+% obtained when we compute the likelihood of an example simulation,
+% in which the ARG is at three different levels of completeness
+% (illustrated in Fig~\ref{fig:lik-surface}B).
+% We first run a simulation under the SMC
+% using the \texttt{record\_full\_arg} option in \texttt{msprime}. This 
+% ARG explicitly identifies all of the recombination and common 
+% ancestor events that occurred in the simulation, and 
+% we can therefore compute a likelihood under the HKY formulation
+% (for example, using the \texttt{msprime.log\_arg\_likelihood} 
+% function). The example full ARG in 
+% Fig~\ref{fig:lik-surface}B has three nodes that are ``locally
+% unary''~\citep{wong_general_2023,fritze2024forest}. Nodes 6 and 7 correspond to the 
+% left and right parents of the recombination event causing the 
+% breakpoint at position 536~\cite[we need two nodes to uniquely identify
+%  the passage of ancestral material through recombination 
+% events; see][]{baumdicker_efficient_2021}, and therefore present
+% in only those trees. 
+% Node 8 is a coalescent node in the right-hand tree, but unary
+% in the left-hand tree, marking the passage of the lineage through
+% this ancestral haplotype (similar to node g in Fig~\ref{fig:smc-unary}).
+% Fig~\ref{fig:lik-surface}A shows that, remarkably, when we 
+% remove the recombination nodes (e.g., 6/7) but keep the locally unary nodes (e.g., 8)
+% and compute the likelihood using eq.~\eqref{eq:full-lik},
+% there is very little difference in the likelihood ratio
+% curve from the exact value with complete input. 
+% Most importantly, the minimum occurs at the true value of the 
+% recombination parameter.
+% When we further simplify
+% the ARG to remove all unary nodes~\citep{wong_general_2023,kelleher_efficient_2018} 
+% we obtain an ARG in the standard format used by \texttt{msprime}.
+% Although there is substantial information about the correlation
+% structure of trees in this representation, we lose important
+% information about recombination. This is demonstrated in 
+% Fig~\ref{fig:lik-surface}A, where the corresponding likelihood ratio curve
+% is clearly biased.
 
@@
 \bibliographystyle{plainnat}
-\bibliography{refs.bib,paper.bib}
+\bibliography{paper.bib}
 

### 2025-02-20 figures/generate-figs/plot_likelihood_surface.py

@@ -18,7 +18,7 @@ def plot(df_rat, path):
         "full_arg_hudson": "tab:purple",
     }
     titles = {
-        "full_arg_smc": "Full ARG (SMC)",
+        "full_arg_smc": "Full ARG",
         "unary_coal": "RE nodes removed",
         "fully_simplified": "Fully simplified",
         "full_arg_hudson": "Full ARG (HKY)",
@@ -63,7 +63,7 @@ def plot(df_rat, path):
     )
     ax.add_artist(legend1)
 
-    ax.set_ylabel("Relative log likelihood")
+    ax.set_ylabel("Log likelihood ratio")
     ax.axvline(REC, linestyle="dashed", color="grey")
     ax.set_xlabel("Recombination rate")
 

### 2025-02-20 paper.tex
remove equation for mutations
Closes #20

@@
 Note that we have only considered the topology and branch length information here
-an not considered mutational information.
-The likelihood of a set of mutations given an ARG $\mathcal{G}$ given mutation rate $\mu$ is
-\begin{equation}\label{eq:lik-mut}
-    \mathcal{L}(\mathcal{D}|\mathcal{G}, \mu)
-    =
-    \prod_{e \in E} \frac{(\mu(t_p-t_c)(y-x))^{m_e}}{m_e!} e^{-\mu (t_p-t_c)(y-x)} ,
-    % I don't think the following is correct:
-    % \frac{1}{\mathcal{M}!} \prod_{(p, c, x, y) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
-\end{equation}
-where 
-% $\mathcal{M}$ the total number of mutations and 
-$m_e$ is the number of mutations along edge $e$ 
-and as before $e = (p,c,x,y)$
-\citep{mahmoudi_bayesian_2022}.
+and not considered mutational information. 
+Assuming the infinite sites model, computing the likelihood of 
+an ARG given a mutation rate $\mu$ and set of mutations is straightforward, 
+as the overall likelihood can be directly decomposed into the per-edge
+contributions~\citep{baumdicker_efficient_2021,mahmoudi_bayesian_2022}.
+
+% The likelihood of a set of mutations given an ARG $\mathcal{G}$ given mutation rate $\mu$ is
+% \begin{equation}\label{eq:lik-mut}
+%     \mathcal{L}(\mathcal{D}|\mathcal{G}, \mu)
+%     =
+%     \prod_{e \in E} \frac{(\mu(t_p-t_c)(y-x))^{m_e}}{m_e!} e^{-\mu (t_p-t_c)(y-x)} ,
+%     % I don't think the following is correct:
+%     % \frac{1}{\mathcal{M}!} \prod_{(p, c, x, y) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
+% \end{equation}
+% where 
+% % $\mathcal{M}$ the total number of mutations and 
+% $m_e$ is the number of mutations along edge $e$ 
+% and as before $e = (p,c,x,y)$
+% \citep{mahmoudi_bayesian_2022}.
 

### 2025-02-20 paper.tex
Fix some issues with polytomies

@@
 
-% Our second example shows a different aspect of this robustness to missing 
-% data. In Fig~\ref{fig:lik-surface}A we can see the likelihood-ratio curves
-% obtained when we compute the likelihood of an example simulation,
-% in which the ARG is at three different levels of completeness
-% (illustrated in Fig~\ref{fig:lik-surface}B).
-% We first run a simulation under the SMC
-% using the \texttt{record\_full\_arg} option in \texttt{msprime}. This 
-% ARG explicitly identifies all of the recombination and common 
-% ancestor events that occurred in the simulation, and 
-% we can therefore compute a likelihood under the HKY formulation
-% (for example, using the \texttt{msprime.log\_arg\_likelihood} 
-% function). The example full ARG in 
-% Fig~\ref{fig:lik-surface}B has three nodes that are ``locally
-% unary''~\citep{wong_general_2023,fritze2024forest}. Nodes 6 and 7 correspond to the 
-% left and right parents of the recombination event causing the 
-% breakpoint at position 536~\cite[we need two nodes to uniquely identify
-%  the passage of ancestral material through recombination 
-% events; see][]{baumdicker_efficient_2021}, and therefore present
-% in only those trees. 
-% Node 8 is a coalescent node in the right-hand tree, but unary
-% in the left-hand tree, marking the passage of the lineage through
-% this ancestral haplotype (similar to node g in Fig~\ref{fig:smc-unary}).
-% Fig~\ref{fig:lik-surface}A shows that, remarkably, when we 
-% remove the recombination nodes (e.g., 6/7) but keep the locally unary nodes (e.g., 8)
-% and compute the likelihood using eq.~\eqref{eq:full-lik},
-% there is very little difference in the likelihood ratio
-% curve from the exact value with complete input. 
-% Most importantly, the minimum occurs at the true value of the 
-% recombination parameter.
-% When we further simplify
-% the ARG to remove all unary nodes~\citep{wong_general_2023,kelleher_efficient_2018} 
-% we obtain an ARG in the standard format used by \texttt{msprime}.
-% Although there is substantial information about the correlation
-% structure of trees in this representation, we lose important
-% information about recombination. This is demonstrated in 
-% Fig~\ref{fig:lik-surface}A, where the corresponding likelihood ratio curve
-% is clearly biased.
-
+Fig~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably
+robust to the presents of polytomies in this example. For each of the three
+ARGs discussed in the previous paragraph, we also evaluated the likelihood
+for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and 
+edges adjusted accordingly.
+This introduces polytomies at varying degrees in the different ARGs because 
+deleting a node in a tree (and connecting its children directly to its
+parent) will only create a polytomy if the parent 
+is not unary. Thus, while deleting 25\% of the internal nodes 
+in the fully simplified ARG created an average of 33 polytomies
+over the 2,933 trees along the sequence (each tree having 
+200 leaf nodes), it resulted in an average of 21 polytomies
+per tree in the ARG where recombination nodes have been removed,
+and only 10 in the full ARG (which has an average of 1026 unary 
+nodes per tree).
+%                     name    leaves  unary       binary      poly
+% 0   poly_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
+% 1   poly_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
+% 2   poly_fully_sfiied_0.05  200.0   0.000000    185.691468  6.393174
+% 3   poly_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
+% 4   poly_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
+% 5   poly_fully_sified_0.25  200.0   0.000000    117.096174  33.241986
+Given this disparity in the number of polytomies introduced, the 
+effects on the likelihood across the three examples are not entirely
+comparable. It is remarkable, however, how little effect introducing 
+these polytomies has on the likelihood surface in this example. For 
+the ARGs containing unary nodes, removing 25\% of the internal 
+nodes has negligible effect, and even on the fully simplified 
+ARG has a very minor effect on the minimum.
+
+\subsection*{Piecewise decomposition over time}
 \begin{figure}
@@
 
-% slicing of the ARG
 Our final example explores decomposition of the likelihood
@@
 time and/or along the genome.
-In the case of a fully precise ARG, with a known time to recombination,
+In the case of a fully precise ARG, in which each recombination
+event has an exact time,
 this is true as long as the slicing operation does not erase the
@@
 The result is reasonably accurate, despite only using 10kb of sequence.
+
 % This approach therefore strengthens

### 2025-02-20 paper.tex
Stuff

@@
 % the ARG as both edges would switch parent nodes. 
-Locally unary nodes are not a common aspect of current ARG research, but 
+Locally unary nodes are not a common aspect of current ARG 
+research~\citep{wong_general_2023}, but 
 are estimated by \texttt{tsinfer}, and can be imputed to high
 degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
+
+Our results, while preliminary, suggest that ``full ARGs``, in which 
+each recombination event is precisely pinpointed in time, 
+contain a significant degree of redundancy in terms of conveying 
+information about the generative process. 
+It is possible, therefore, that explicitly estimating these 
+events is not necessary for either inference under the SMC or 
+for downstream applications, and we can integrate over the 
+plausible times as we do here.
+Further research is required to fully understand the
 

### 2025-02-20 paper.tex
Pass through the discussion

--- 2025-02-20 paper.tex
@@
 Computing the likelihood of an ARG under a model is a fundamental element of
-probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
+probabilistic inference. The classical Kuhner-Yamato-Felsenstein
 \citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
@@
 in the input ARG.
+Our method can compute a likelihood under the SMC for a very general class of 
+ARG~\citep{wong_general_2023}, 
+a significant increase in flexibility over the 
+strict input requirements for the classical Kuhner-Yamato-Felsenstein 
+formulation. 
+We have shown via some illustrative examples that the formulation is 
+remarkably robust to incomplete information, tolerating the loss 
+of a significant fraction of the internal nodes in the ARG to minimal effect.
+
 Although we can associate a likelihood with any ARG in \texttt{tskit} format,
@@
 the various degrees of completeness to which inheritance
-is documented by the different ARG inference methods: e.g., polytomies, absence
-of locally unary coalescence nodes, unpreserved identify of internal nodes
-across local trees, etc.
-However, many of these imprecisions violate the implicit
-assumptions of the likelihood computation.
+is documented by the different ARG inference methods.
+% : e.g., polytomies, absence
+% of locally unary coalescence nodes, unpreserved identify of internal nodes
+% across local trees, etc.
+% However, many of these imprecisions violate the implicit
+% assumptions of the likelihood computation.
 For example, we assume the 
 presence of locally unary nodes to correctly identify the
-full extent of all lineages going back through time and therefore recombination events.
+full extent of all lineages going back through time 
+and therefore recombination events, but this information
+may be incomplete (e.g., the full simplifed ARG in Fig.~\ref{fig:lik-surface})
+or even incorrect.
 % Without these unary nodes, our algorithm will detect a recombination
@@
 % the ARG as both edges would switch parent nodes. 
-Locally unary nodes are not a common aspect of current ARG research, but 
+Locally unary nodes are not a common aspect of current ARG 
+research~\citep{wong_general_2023}, but 
 are estimated by \texttt{tsinfer}, and can be imputed to high
 degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
-
-[A paragraph about possililties for inference]
-
-[A paragraph about why what we've presented is useful, and a good starting
-point for other research. We provide an efficient algorithm and 
-reference implementation]
-
-[A paragraph about another major sort of uncertainty: polytomies]
+Determining the effects of various forms of \emph{incorrect} topology on
+these likelihoods requires further research.
+
+[Something about polytomies]
+
+Our results, while preliminary, suggest that ``full ARGs'', in which 
+each recombination event is precisely pinpointed in time, 
+contain a significant degree of redundancy in terms of capturing 
+information about the generative process. 
+It is possible, therefore, that explicitly estimating recombination 
+events is not necessary to sample ARGs under the SMC (as ARGweaver
+and SINGER do), potentially
+leading to further improvements in the scalability of such methods.
+Another exciting prospect for ARG inference is the development of 
+``hybrid'' inference methods, where hueristic methods such as 
+\texttt{tsinfer} or \texttt{ARG-Needle} are used to infer
+the details of the recent past in huge datasets such as 
+UK Biobank, and passing over to a rigorous model-based
+approach for the ancient past.
+This is similar in spirit to the widely-used approach 
+of simulating complex dynamics of the recent past using 
+detailed but slow models and using faster coalescent simulations
+in the more distant past, when the assumptions required 
+are more justified~\citep{bhaskar_distortion_2014,
+kelleher_efficient_2018,haller_tree_2018,nelson_accounting_2020}.
+% Maybe we shouldn't bother saying this, but I can't resist the 
+% opportunity to say "hey, you should be using tskit!"
+% It feels like some PIs aren't getting this message, so need to have
+% it spelled out.
+Just as forward and backward simulations are useful in different contexts
+and can be combined to great effect,
+large-scale hueristic and MCMC sampling inference methods have their 
+applications, and there may be synergy in their combination. 
+Practical combination of methods will require cooperation at the 
+data structure level, and the interoperability provided by the 
+\texttt{tskit} library [something good].
 

### 2025-02-20 paper.tex

@@
 \linenumbers
-\title{Computing likelihoods under the SMC for a general class of ARGs.}
+\title{Likelihoods for a general class of ARGs under the SMC}
 
@@
     $r=2.5\times10^{-8}$, $N_e=1\times10^4$) which contains complete 
-    information about the simulated events.
-    with various degrees of completeness. The likelihood ratio is computed with
-    respect to the true recombination rate. 
+    information about the simulated events 
+    (9,378 nodes; 13,221 edges; 2,933 trees). 
+    The ``RE nodes removed'' ARG is the 
+    result of simplifying out all of the recombination nodes (3,332 nodes;
+    6,901 edges; note that \texttt{msprime} uses distinct nodes for the 
+    left and right parent of each recombination). The ``Fully simplified''
+    ARG has all locally unary nodes removed from trees (3,332 nodes; 10,674
+    edges). Each of these ARGs additionally has 5\% and 25\% of the 
+    internal nodes removed to create polytomies.
+    See the text for details on how the likelihood ratio values are computed.
+    The ``Full ARG'' and ``RE nodes deleted'' lines coincide.
     \label{fig:lik-surface}}

### 2025-02-21 paper.tex
Switch to abbrvnat
Intro para on uncertainty

@@
 stern_approximate_2019,
-osmond_estimating_2021,
 fan_genealogical_2022,
@@
 fan_likelihood_2023,
+osmond_estimating_2021,
 huang_estimating_2024,
@@
 and by using an improved Monte Carlo algorithm
-promises to be even more accurate while being substantially more
+promises to be even more accurate and substantially more 
 scalable~\citep{deng_robust_2024}. 
-
-The second class of inference method that has been of recent interest 
-is based on heuristics and approximate models.
+The second class of inference methods that has been of recent interest 
+are based on heuristics and approximate models.
 Relate~\citep{speidel_method_2019},
@@
 Secondly, they all produce a single
-ARG as output, inferring a deterministic point estimate with no means 
-of quantifying uncertainty. Thirdly (and most importantly for the 
-purposes of this paper) the ARGs they produce 
-lack some of the information present in ARGweaver or SINGER. 
-This occurs, roughly speaking, because of a fundamental difference in approach:
-not all details of the ARG can be reliably inferred
-(e.g., the timing of recombination events, as discussed below);
+ARG as output, inferring a deterministic point estimate.
+Thirdly, they regard the precise timing of nodes as a separate
+problem, focusing only on the relative ordering of nodes when 
+producing the initial ARG. General purpose ARG dating methods are 
+appearing~\citep{wohns_unified_2022,deng2025general} and can also improve
+dating performance in ARGs sampled from the SMC~\citep{deng2025general}.
+Fourthly (and most importantly for the 
+purposes of this paper) the ARGs that hueristic methods produce 
+lack some of the topological information present in the output of ARGweaver or SINGER. 
+This occurs, roughly speaking, because of a basic difference in their
+approach to uncertainty.
+
+There is considerable uncertainty in ARG inference, and often a fundamental
+lack of information in the sequence data to distinguish different
+possibilities. Consider, for example, a case in which we have three
+ancestral lineages with identical sequences. We know that there
+must have been two coalescence events, but there is no mutational information to 
+distinguish their relative ordering. Sampling methods overcome this problem
+by randomly choosing an ordering, and uncertainty communicated by the order
+changing in the different ARGs sampled from the posterior.
+The other approach (used by methods such as tsinfer) is to communicate
+the uncertainty structurally by means of a polytomy: we have no
+information about the intermediate coalescent event, and so we omit it.
+Similarly, there is often a fundamental lack of information about the 
+details of recombination events, and heuristic methods (in different ways)
+omit these details. In many ways, the significant progress made in
+scalability by these heuristic methods is \emph{because} they don't
+attempt to precisely reconstruct recombination events.
+
+As there is no mutational
+information
+
+(e.g., the timing of recombination events, as discussed below).
+
 and the heuristic methods deal with this uncertainty by omitting these events,
@@
 works by assigning a probablility to each recombination and common
-ancestor event, and to the inter-event  waiting times. The likelihood of an 
+ancestor event, and to the inter-event  waiting times. 
+The likelihood of an 
 ARG is then a product over all these three classes of term.
-% In essence, computing the KYF likelihood requires
-% ``replaying'' the  state of Hudson's simulation
-% algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
-% compute the exact configuration of lineages and distribution
-% of ancestral material at each time point~\citep{baumdicker_efficient_2021}.
-% It therefore requires complete
-% detail about each of these events to be stored. 
-% Other approaches to defining the likelihood of an ARG under a coalescent-like 
-% model follow the same basic event-by-event 
-% strategy~\citep[e.g.,][]{guo_recombination-aware_2022}.
-% % [Are there others?  The bacterial stuff from Didelot etc seem to focus on
-% % a tree-by-tree likelihood, rather than global graph, so not clear it's the 
-% % same thing]
-% Computing the likelihood under the SMC rather than the Hudson coalescent 
-% simplifies bookkeeping somewhat,  % Dunno what I'm talking about here
-% but a complete event-by-event description of the 
-% input ARG is still required.
 Although the KYF formulation is a natural and elegant way to describe 
-the likelihood of an ARG, such complete detail is not present 
+the likelihood of an ARG given a complete description of
+
+, such complete detail is not present 
 in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
-Simulators such as the classical \texttt{ms}
+Simulators such as the classical ms
 program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
 which does not contain sufficient detail for likelihood computation.
-The \texttt{msprime}
+The msprime
 simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
@@
 the \texttt{record\_full\_arg}
-option to \texttt{msprime}~\citep{baumdicker_efficient_2021}.
+option to msprime~\citep{baumdicker_efficient_2021}.
 Forwards-time simulations that output 
@@
 research~\citep{wong_general_2023}, but 
-are estimated by \texttt{tsinfer}, and can be imputed to high
+are estimated by tsinfer, and can be imputed to high
 degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
@@
 ``hybrid'' inference methods, where hueristic methods such as 
-\texttt{tsinfer} or \texttt{ARG-Needle} are used to infer
+tsinfer or ARG-Needle are used to infer
 the details of the recent past in huge datasets such as 
@@
 \FloatBarrier
-\bibliographystyle{plainnat}
+\bibliographystyle{abbrvnat}
 \bibliography{paper.bib}

### 2025-02-21 paper.tex
More updates on the intro

@@
 Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
-ARGweaver~\citep{rasmussen_genome-wide_2014} has been the most widely used and studied,
+ARGweaver~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020} 
+has been the most widely used and studied,
 and until recently consistently out-performed other methods in terms 
@@
 distinguish their relative ordering. Sampling methods overcome this problem
-by randomly choosing an ordering, and uncertainty communicated by the order
+by randomly choosing an ordering, with the uncertainty communicated by the order
 changing in the different ARGs sampled from the posterior.
@@
 scalability by these heuristic methods is \emph{because} they don't
-attempt to precisely reconstruct recombination events.
-
-As there is no mutational
-information
-
-(e.g., the timing of recombination events, as discussed below).
-
-and the heuristic methods deal with this uncertainty by omitting these events,
-while ARGweaver and SINGER provide a sample from the posterior.
-(So, some aspects of the resulting ARGs should be interpreted slightly differently;
-see \citet{wong_general_2023} for more discussion.)
+attempt to precisely reconstruct recombination events
+(this should be taken into account when interpreting the resulting ARGs;
+see \citet{wong_general_2023} for more discussion).
 
@@
 approaches of ARGweaver and SINGER.
-Firstly, they tend to be more accurate when compared on the same data set.
-Secondly, the direct connection with population genetic models allows for
-natural extensions like the joint estimation of ARGs and
-demography~\citep{hubisz_mapping_2020}.
-Thirdly, sampling ARGs from a
+In particular, they tend to be more accurate (at the 
+scale at which they can be compared),
+and sampling ARGs from a
 posterior distribution provides a means of quantifying the uncertainty 
-around estimates, which can be used in downstream applications.
-
+around estimates.
 On the other hand, heuristic methods have the potential
-to utilize the greater amount of information in large datasets,
-thus possibly producing even more accurate estimates
-(although this has not, to our knowledge, been tested).
+to use the greater amount of information in large datasets,
+capturing fine-scale details about the recent past.
+% thus possibly producing even more accurate estimates
+% (although this has not, to our knowledge, been tested).
 Although SINGER is a major step forward
@@
 % goodness of fit for model estimates.
+
 A key problem facing the field is that the ARGs estimated by heuristic large-scale 
@@
 we cannot compute their likelihood under the SMC or similar models. 
-% The classical method~\citep{kuhner_maximum_2000} assumes complete knowledge about
-% recombination events, and there is no clear way to compute a likelihood for 
-% the approximate structures output by heuristic methods.
-
 Computing the likelihood of an ARG under a model is a fundamental element of
@@
 ancestor event, and to the inter-event  waiting times. 
-The likelihood of an 
-ARG is then a product over all these three classes of term.
 Although the KYF formulation is a natural and elegant way to describe 
-the likelihood of an ARG given a complete description of
-
-, such complete detail is not present 
+the likelihood of an ARG, the high-level of detail about the underlying
+events that is required is not present 
 in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
@@
 Forwards-time simulations that output 
-ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
+ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} also do not
 typically contain the level of detail required to compute a likelihood under

### 2025-02-21 paper.tex
Full pass through intro

@@
 of any ARG, and describe an efficient algorithm to do so.
-We also illustrate the utility of this approach via some simple examples.
+We show that explicitly identifying recombination events via nodes 
+in the graph is not necessary as long as the ``locally
+unary''~\cite{wong_general_2023,fritze2024forest} sections of common ancestor
+nodes are retained, from the perspective of computing the likelihood.
+We also show via a empirical examples that the likelihood is remarkably
+robust to the presence of polytomies, and illustrate how it can be
+decomposed into separately evaluated time slices.
+Finally, we discuss how these results may be applied and extended.
 
@@
 of the exact time to recombination, the integral in
-eq.~\eqref{eq:depth} means that the likelihood
+Eq.~\eqref{eq:depth} means that the likelihood
 for the entire ARG does not factorize into the likelihoods of these
@@
 the next, $T_2$ imposes the strict total ordering on edges
-(see definition \eqref{def:coal})
+(see definition~\eqref{def:coal})
 that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
@@
 of that edge by computing the integral
-in equation \eqref{eq:full-lik} using $I$, and then update the counts in $I$.
+in Eq.~\eqref{eq:full-lik} using $I$, and then update the counts in $I$.
 We further keep track of

### 2025-02-21 paper.tex
Add results for tsinfer+tsdate

@@
 % % % % % % % % % % % % % % % % % %
-\subsection*{Robustness to partial information}
+\subsection*{Robustness to partial and incorrect information}
 Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
@@
 illustrate the robustness
-of this approach to incomplete information about recombination and 
-coalescent events via a proof-of-concept example, in which we 
-remove various levels of detail from a simulated ARG.
+of this approach to incomplete and potentially incorrect
+information about recombination and 
+coalescent events via a proof-of-concept example. 
+% in which we 
+% remove various levels of detail from a simulated ARG.
 
@@
     \includegraphics[width=0.8\textwidth]{figures/likelihood_surface}
-    \caption{Log likelihood-ratio curves for ARGs containing
-     varying degrees of information about the realised simulation.
+    \caption{Log likelihood-ratio curves for ARGs with various properties.
     The ``Full ARG'' here is the result of an 
-    \texttt{msprime} simulation of the SMC (100 diploid samples, genome length 1Mb, 
-    $r=2.5\times10^{-8}$, $N_e=1\times10^4$) which contains complete 
-    information about the simulated events 
-    (9,378 nodes; 13,221 edges; 2,933 trees). 
+    msprime simulation of the SMC (100 diploid samples, genome length 1Mb, 
+    $r=\mu=2.5\times10^{-8}$, $N_e=1\times10^4$).
     The ``RE nodes removed'' ARG is the 
-    result of simplifying out all of the recombination nodes (3,332 nodes;
-    6,901 edges; note that \texttt{msprime} uses distinct nodes for the 
-    left and right parent of each recombination). The ``Fully simplified''
-    ARG has all locally unary nodes removed from trees (3,332 nodes; 10,674
-    edges). Each of these ARGs additionally has 5\% and 25\% of the 
+    result of simplifying out all of the recombination nodes,
+    and the ``Fully simplified'' ARG has all locally unary nodes removed. 
+    Each of these simulated ARGs additionally has 5\% and 25\% of the 
     internal nodes removed to create polytomies.
-    See the text for details on how the likelihood ratio values are computed.
-    The ``Full ARG'' and ``RE nodes deleted'' lines coincide.
+    We also show the results for ARGs inferred from the original simulation
+    data using tsinfer and tsdate 
+    (see text for details), in which we use the default output of tsinfer
+    including unary nodes (tsinfer+tsdate), and in which we fully simplify 
+    before passing to tsdate (tsinfer+tsdate FS). 
+    See the text for details on how the likelihood ratio 
+    values are computed.
+    The ``Full ARG'' and ``RE nodes removed'' lines coincide.
     \label{fig:lik-surface}}
@@
 
-In Fig~\ref{fig:lik-surface} we begin with a realisation of the SMC as a ``full
-ARG'', that is, containing complete information about every simulated event.
+In Fig~\ref{fig:lik-surface} we begin with a realisation of the SMC 
+containing complete information about every simulated event
+(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). 
 We then plot the likelihood computed by Eq.~\eqref{eq:full-lik} 
-as we vary the recombination rate parameter $r$ (as a ratio
-of the likelihood computed at the true value of $r$). Unsurprisingly, we can
+as we vary the recombination rate parameter $r$, as a ratio
+of the likelihood at the true value of $r$. 
+Reassuringly, we can
 see that the minimum of the likelihood-ratio curve when using the 
-full simulated ARG is at the true value of the parameter.
-What is surprising is that we obtain an indistinguishable curve when 
-we remove all the recombination nodes from this ARG, and therefore 
-lose information about the precise timing of recombination events.
-In this case, the locally unary portions of coalescent nodes have given
-all the information that we need about recombination events.
-Removing all locally unary nodes in the fully simplified ARG does result
-in a loss of information, and we can see that the minimum of the 
+full simulated ARG is close to the true value of the parameter.
+(The likelihood ratio curve obtained when computing the likelihood
+under the full coalescent with recombination model using the KYF
+formula is identical; not shown.)
+We next compute the likelihood ratio for the ARG obtained when we 
+simplify out the recombination nodes from the full ARG
+(RE nodes removed; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime 
+uses distinct nodes for the left and right parent of each recombination).
+The curve we obtain is indistinguishable from that of the full
+ARG, demonstrating that the locally unary portions of coalescent nodes 
+contain all the information that we need about recombination events
+for this application.
+When we fully simplify the simulated ARG
+to remove all locally unary nodes from trees 
+(Fully simplified; 3,332 nodes; 10,674 edges, 2,933 trees) 
+we can see there is a significant loss of information,
+and the minimum of the 
 likelihood ratio surface is biased away from its true value.
+(See~\cite{wong_general_2023} for more discussion on varying 
+degrees of ARG simplification.)
 
 Fig~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably
-robust to the presents of polytomies in this example. For each of the three
+robust to the presences of polytomies in this example. For each of the three
 ARGs discussed in the previous paragraph, we also evaluated the likelihood
@@
 %     name                  leaves  unary       binary  poly
-% 0   poly_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
-% 1   poly_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
-% 2   poly_fully_sfiied_0.05  200.0   0.000000    185.691468  6.393174
-% 3   poly_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
-% 4   poly_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
-% 5   poly_fully_sified_0.25  200.0   0.000000    117.096174  33.241986
+% 0   full_arg_smc          200.0   1094.417661 199.000000  0.000000
+% 1   unary_coal            200.0   456.396522  199.000000  0.000000
+% 2   fully_simplified      200.0   0.000000    199.000000  0.000000
+% 3   tsinfer_tsdate        200.0   192.447590  119.754399  32.270849
+% 4   tsinfer_fs_tsdate     200.0   0.000000    119.563898  32.322684
+% 5   py_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
+% 6   py_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
+% 7   py_fully_simied_0.05  200.0   0.000000    185.691468  6.393174
+% 8   py_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
+% 9   py_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
+% 10  py_fully_simied_0.25  200.0   0.000000    117.096174  33.241986
 Given this disparity in the number of polytomies introduced, the 
 effects on the likelihood across the three examples are not entirely
-comparable. It is remarkable, however, how little effect introducing 
-these polytomies has on the likelihood surface in this example. For 
-the ARGs containing unary nodes, removing 25\% of the internal 
+comparable. For the ARGs containing unary nodes, 
+removing 25\% of the internal 
 nodes has negligible effect, and even on the fully simplified 
-ARG has a very minor effect on the minimum.
+ARG has a very minor effect on location of the minimum.
+
+Fig~\ref{fig:lik-surface} also shows results for ARGs inferred from the 
+simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
+using default parameters for tsinfer, and the true values of mutation
+rate and $N_e$ for tsdate.
+In the first, we use the default output of tsinfer 
+(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations).
+This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
+with three or more children per tree. Tsinfer does not directly estimate
+recombination events; the breakpoints between trees are the consequence of
+switches between parental haplotypes in the 
+\cite{li2003modeling} copying process. Information about 
+recombination events in the returned ARG is therefore quite imprecise,
+and likely to contain incorrect and contradictory details if one were to 
+attempt to reconstruct the exact subtree-prune-and-regraft 
+moves~\citep[e.g.][]{rasmussen2023espalier}. Nonetheless, in
+this toy example at least, our likelihood function performs well
+on the output of tsinfer, with the minimum close to the true
+parameter value. We also show the likelihood curve obtained when we 
+fully simplify the output of tsinfer 
+(1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
+before dating. We can see that, as with the case of the true data,
+removing all the unary nodes results in a significant bias
+away from the true value of the recombination rate parameter.
 
@@
 of a significant fraction of the internal nodes in the ARG to minimal effect.
-
-Although we can associate a likelihood with any ARG in \texttt{tskit} format,
+% I think we have to grasp this nettle or we're inviting the reviewers to 
+% give us a bunch of "what if you tried this" bits of extra homework.
+While there are many different potential applications to concrete 
+inferential problems, the details are nontrivial and best consigned to future work.
+
+A major advantage of our approach is that it does not require
+us to explicitly identify common ancestor and recombination events 
+in the input. Indeed, 
+we can compute a likelihood for any ARG in \texttt{tskit} format,
+potentially bridging the gap between population genetic theory and the 
+varied structures inferred by current methods~\citep{wong_general_2023}.
+
 there are some caveats. By treating each edge independently,

### 2025-02-21 figures/generate-figs/plot_likelihood_surface.py

@@ -4,10 +4,10 @@ import matplotlib.pyplot as plt
 import matplotlib.lines as mlines
 import pandas as pd
 
+
 def plot(df_rat, path):
 
     REC = 1.25e-8
-
 
     fig, ax = plt.subplots(1, 1)
 
@@ -15,6 +15,8 @@ def plot(df_rat, path):
         "full_arg_smc": "tab:red",
         "unary_coal": "tab:blue",
         "fully_simplified": "tab:orange",
+        "tsinfer_tsdate": "tab:olive",
+        "tsinfer_fs_tsdate": "tab:cyan",
         "full_arg_hudson": "tab:purple",
     }
     titles = {
@@ -22,6 +24,8 @@ def plot(df_rat, path):
         "unary_coal": "RE nodes removed",
         "fully_simplified": "Fully simplified",
         "full_arg_hudson": "Full ARG (HKY)",
+        "tsinfer_tsdate": "tsinfer+tsdate",
+        "tsinfer_fs_tsdate": "tsinfer+tsdate (FS)",
     }
 
     line_styles = {
@@ -68,6 +72,7 @@ def plot(df_rat, path):
     ax.set_xlabel("Recombination rate")
 
     plt.savefig(path)
+
 
 df = pd.read_csv("likelihood_surface.csv").set_index("rho")
 

### 2025-02-21 paper.tex
Pass through disco

@@
 (1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
-before dating. We can see that, as with the case of the true data,
+before dating. We can see that, as with the true data,
 removing all the unary nodes results in a significant bias
@@
 \section*{Discussion}
-
 The key contribution of this paper is to rederive a classical
-ARG likelihood function in a way that supports the sort of partial information
+ARG likelihood function in a way that supports the sort of partial 
+and imprecise information
 provided by some modern ARG inference tools,
@@
 We have shown via some illustrative examples that the formulation is 
-remarkably robust to incomplete information, tolerating the loss 
+robust to incomplete and imprecise information, tolerating the loss 
 of a significant fraction of the internal nodes in the ARG to minimal effect.
+We have also shown that the likelihood function performs well 
+(at least in our toy example) on ARGs inferred by tsinfer+tsdate,
+successfully capturing information about latent recombination events
+from imprecise and noisy data.
+Locally unary nodes play a key role, and although not
+currently a common focus of ARG research~\citep{wong_general_2023}, 
+they are estimated by tsinfer and can be imputed to high a
+degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
 % I think we have to grasp this nettle or we're inviting the reviewers to 
 % give us a bunch of "what if you tried this" bits of extra homework.
-While there are many different potential applications to concrete 
-inferential problems, the details are nontrivial and best consigned to future work.
-
-A major advantage of our approach is that it does not require
-us to explicitly identify common ancestor and recombination events 
-in the input. Indeed, 
-we can compute a likelihood for any ARG in \texttt{tskit} format,
-potentially bridging the gap between population genetic theory and the 
-varied structures inferred by current methods~\citep{wong_general_2023}.
-
-there are some caveats. By treating each edge independently,
-our algorithm will still produce a value for ARGs with
-the various degrees of completeness to which inheritance
-is documented by the different ARG inference methods.
-% : e.g., polytomies, absence
-% of locally unary coalescence nodes, unpreserved identify of internal nodes
-% across local trees, etc.
-% However, many of these imprecisions violate the implicit
-% assumptions of the likelihood computation.
-For example, we assume the 
-presence of locally unary nodes to correctly identify the
-full extent of all lineages going back through time 
-and therefore recombination events, but this information
-may be incomplete (e.g., the full simplifed ARG in Fig.~\ref{fig:lik-surface})
-or even incorrect.
-% Without these unary nodes, our algorithm will detect a recombination
-% event along both coalescing lineages in case of a (sub)tree height changes along
-% the ARG as both edges would switch parent nodes. 
-Locally unary nodes are not a common aspect of current ARG 
-research~\citep{wong_general_2023}, but 
-are estimated by tsinfer, and can be imputed to high
-degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
-Determining the effects of various forms of \emph{incorrect} topology on
-these likelihoods requires further research.
-
-[Something about polytomies]
+Fully characterising the properties of the likelihood on the output 
+of tsinfer and other methods is beyond the scope of this work.
+Similarly, while there are many different potential applications to concrete 
+inferential problems, the details are nontrivial and must be
+consigned to future work.
+
+% Structural uncertainty
+
+% A major advantage of our approach is that it does not require
+% us to explicitly identify common ancestor and recombination events 
+% in the input. Indeed, 
+% we can compute a likelihood for any ARG in \texttt{tskit} format,
+% potentially bridging the gap between population genetic theory and the 
+% varied structures inferred by current methods~\citep{wong_general_2023}.
+
+% there are some caveats. By treating each edge independently,
+% our algorithm will still produce a value for ARGs with
+% the various degrees of completeness to which inheritance
+% is documented by the different ARG inference methods.
+% % : e.g., polytomies, absence
+% % of locally unary coalescence nodes, unpreserved identify of internal nodes
+% % across local trees, etc.
+% % However, many of these imprecisions violate the implicit
+% % assumptions of the likelihood computation.
+% For example, we assume the 
+% presence of locally unary nodes to correctly identify the
+% full extent of all lineages going back through time 
+% and therefore recombination events, but this information
+% may be incomplete (e.g., the full simplifed ARG in Fig.~\ref{fig:lik-surface})
+% or even incorrect.
+% % Without these unary nodes, our algorithm will detect a recombination
+% % event along both coalescing lineages in case of a (sub)tree height changes along
+% % the ARG as both edges would switch parent nodes. 
+% Locally unary nodes are not a common aspect of current ARG 
+% research~\citep{wong_general_2023}, but 
+% are estimated by tsinfer, and can be imputed to high
+% degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
+% Determining the effects of various forms of \emph{incorrect} topology on
+% these likelihoods requires further research.
+
+% [Something about polytomies]
 
 Our results, while preliminary, suggest that ``full ARGs'', in which 
-each recombination event is precisely pinpointed in time, 
+each recombination and common ancestor event is precisely estimated,
 contain a significant degree of redundancy in terms of capturing 
 information about the generative process. 
-It is possible, therefore, that explicitly estimating recombination 
-events is not necessary to sample ARGs under the SMC (as ARGweaver
-and SINGER do), potentially
-leading to further improvements in the scalability of such methods.
+It is possible, therefore, that explicitly estimating 
+\emph{all} such events is not necessary to sample ARGs under the 
+SMC (as ARGweaver and SINGER do). 
+A combined approach, in which the uncertainty around the
+ordering of events that are 
+weakly informed by the data is encoded structurally in the 
+graph and other forms of uncertainty captured by sampling from 
+the posterior, may yield further improvements in 
+scalability for statistically rigorous inference.
+% Signicant work would be needed, however, to make 
+% downstream analysis robust to such stuctural uncertainty;
+% polytomies, for example, are often not
 Another exciting prospect for ARG inference is the development of 
@@
 the details of the recent past in huge datasets such as 
-UK Biobank, and passing over to a rigorous model-based
+UK Biobank, passing over to a rigorous model-based
 approach for the ancient past.
@@
 detailed but slow models and using faster coalescent simulations
-in the more distant past, when the assumptions required 
+in the distant past when the assumptions required 
 are more justified~\citep{bhaskar_distortion_2014,
@@
 applications, and there may be synergy in their combination. 
-Practical combination of methods will require cooperation at the 
-data structure level, and the interoperability provided by the 
-\texttt{tskit} library [something good].
+% This is pretty weak, tbh.
+The practical details of ``handing over'' between methods
+will require cooperation and precise communication;
+the tskit library, with its well-defined data model and flexible
+metadata support, is the ideal basis for this.
 

### 2025-02-21 paper.tex
Another pass

@@
 scalability by these heuristic methods is \emph{because} they don't
-attempt to precisely reconstruct recombination events
-(this should be taken into account when interpreting the resulting ARGs;
-see \citet{wong_general_2023} for more discussion).
+attempt to precisely reconstruct recombination events.
 
@@
 Although the KYF formulation is a natural and elegant way to describe 
-the likelihood of an ARG, the high-level of detail about the underlying
+the likelihood of an ARG, the high level of detail about the underlying
 events that is required is not present 
@@
 We show that explicitly identifying recombination events via nodes 
-in the graph is not necessary as long as the ``locally
-unary''~\cite{wong_general_2023,fritze2024forest} sections of common ancestor
-nodes are retained, from the perspective of computing the likelihood.
-We also show via a empirical examples that the likelihood is remarkably
-robust to the presence of polytomies, and illustrate how it can be
+in the graph is not necessary 
+(from the perspective of computing the likelihood)
+as long as the ``locally
+unary''~\citep{wong_general_2023,fritze2024forest} sections of common ancestor
+nodes are retained.
+We also show via an empirical example that the likelihood is 
+robust to the presence of polytomies and other imprecise 
+information, and illustrate how it can be
 decomposed into separately evaluated time slices.
@@
 formulation. 
+Locally unary nodes play a key role, and although not
+currently a common focus of ARG research~\citep{wong_general_2023}, 
+they are estimated by tsinfer and can be imputed to high a
+degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
 We have shown via some illustrative examples that the formulation is 
@@
 from imprecise and noisy data.
-Locally unary nodes play a key role, and although not
-currently a common focus of ARG research~\citep{wong_general_2023}, 
-they are estimated by tsinfer and can be imputed to high a
-degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
 % I think we have to grasp this nettle or we're inviting the reviewers to 
 % give us a bunch of "what if you tried this" bits of extra homework.
-Fully characterising the properties of the likelihood on the output 
-of tsinfer and other methods is beyond the scope of this work.
-Similarly, while there are many different potential applications to concrete 
+Theoretical explanations of this apparent robustness,
+and fully characterising the properties of the likelihood on the output 
+of tsinfer and other methods is an important avenue for future work.
+Similarly, while there are many potential applications to concrete 
 inferential problems, the details are nontrivial and must be
@@
 and can be combined to great effect,
-large-scale hueristic and MCMC sampling inference methods have their 
+large-scale hueristic and Bayesian sampling inference methods have their 
 applications, and there may be synergy in their combination. 

### 2025-02-21 paper.tex

@@
 coalescent with recombination~\citep{hudson_properties_1983},
-and its approximatation, the Sequentially Markovian
+and its approximation, the Sequentially Markovian
 Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
@@
 of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
-and has succesfully been applied in several
+and has successfully been applied in several
 contexts~\citep[e.g.,][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
@@
 Fourthly (and most importantly for the 
-purposes of this paper) the ARGs that hueristic methods produce 
+purposes of this paper) the ARGs that heuristic methods produce 
 lack some of the topological information present in the output of ARGweaver or SINGER. 
@@
 there is a pressing need for more statistical
-rigor in large scale ARG inference.
+rigour in large scale ARG inference.
 % This inability to connect the results of large-scale inference methods
@@
 \citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
-works by assigning a probablility to each recombination and common
+works by assigning a probability to each recombination and common
 ancestor event, and to the inter-event  waiting times. 
@@
 Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
-in the generic sense, it is a labeled graph structure that records
+in the generic sense, it is a labelled graph structure that records
 the inheritance of genetic material \citep{wong_general_2023}.
@@
 We record each such statement in an ``edge'',
-often summarized as the tuple $(p,c,x,y)$.
+often summarised as the tuple $(p,c,x,y)$.
 By ``describes the history'',
@@
 but that segment was inherited in two pieces from their parent.
-Other nodes (labeled by black numbers) are either samples
+Other nodes (labelled by black numbers) are either samples
 or \emph{common ancestry} nodes --
@@
 At any point in time, the state of the process is 
-the set of labeled lineages extant at time $t$,
+the set of labelled lineages extant at time $t$,
 $L(t) = \{X_j(t)\}$.
-A lineage $X_j(t) = [x_{j}, y_{j})$ is labeled by 
+A lineage $X_j(t) = [x_{j}, y_{j})$ is labelled by 
 an integer $j$ and defines the 
@@
 is $\exp(-\Lambda) \prod_i \lambda_i$,
-where $\lambda_i$ is the rate of the $i^\text{th}$ realized event,
+where $\lambda_i$ is the rate of the $i^\text{th}$ realised event,
 and $\Lambda$ is the sum of all rates of all possible events
@@
 The length in time of the edge is just $t_p - t_c$,
-while the number of ``eleigible links'' in the edge is the number of adjacent base pairs in the edge
+while the number of ``eligible links'' in the edge is the number of adjacent base pairs in the edge
 on which recombination did not occur, which is either $y-x$
@@
 Eq.~\eqref{eq:depth} means that the likelihood
-for the entire ARG does not factorize into the likelihoods of these
+for the entire ARG does not factorise into the likelihoods of these
 individual slices. More precisely, the likelihood associated with any
-edge that is initiated by a recombination breakpoint cannot be factorized
+edge that is initiated by a recombination breakpoint cannot be factorised
 into the product of two integrals of the same form.
@@
 Another exciting prospect for ARG inference is the development of 
-``hybrid'' inference methods, where hueristic methods such as 
+``hybrid'' inference methods, where heuristic methods such as 
 tsinfer or ARG-Needle are used to infer
@@
 and can be combined to great effect,
-large-scale hueristic and Bayesian sampling inference methods have their 
+large-scale heuristic and Bayesian sampling inference methods have their 
 applications, and there may be synergy in their combination. 

### 2025-02-21 paper.tex
edits

@@
 simplify out the recombination nodes from the full ARG
-(RE nodes removed; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime 
+(``RE nodes removed''; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime 
 uses distinct nodes for the left and right parent of each recombination).
@@
 When we fully simplify the simulated ARG
-to remove all locally unary nodes from trees 
-(Fully simplified; 3,332 nodes; 10,674 edges, 2,933 trees) 
+to remove all locally unary portions of nodes from trees 
+(``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees) 
 we can see there is a significant loss of information,
 and the minimum of the 
-likelihood ratio surface is biased away from its true value.
+likelihood ratio surface is biased away from its true value,
+as expected, because removing the unary portions of nodes has removed some of the area
+of the ARG in which recombination did not occur.
 (See~\cite{wong_general_2023} for more discussion on varying 
@@
 Given this disparity in the number of polytomies introduced, the 
-effects on the likelihood across the three examples are not entirely
-comparable. For the ARGs containing unary nodes, 
+effects on the likelihood across the three examples are not entirely comparable.
+For the ARGs containing unary nodes, 
 removing 25\% of the internal 
-nodes has negligible effect, and even on the fully simplified 
-ARG has a very minor effect on location of the minimum.
+nodes has negligible effect on the likelihood curve in Fig~\ref{fig:lik-surface},
+and even on the fully simplified 
+ARG has only a very minor effect on location of the minimum.
 
@@
 (2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations).
-This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
+This ARG has 192 unary nodes, 120 binary nodes, and 32 nodes
 with three or more children per tree. Tsinfer does not directly estimate
@@
 attempt to reconstruct the exact subtree-prune-and-regraft 
-moves~\citep[e.g.][]{rasmussen2023espalier}. Nonetheless, in
+moves~\citep[e.g.,][]{rasmussen2023espalier}. Nonetheless, in
 this toy example at least, our likelihood function performs well
@@
 currently a common focus of ARG research~\citep{wong_general_2023}, 
-they are estimated by tsinfer and can be imputed to high a
+they are estimated by tsinfer and may be imputed to a high
 degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
@@
 from imprecise and noisy data.
+However, there are many more aspects of the likelihood function
+that could be examined, beyond coarse parameters like recombination rate
+or effective populations size.
 % I think we have to grasp this nettle or we're inviting the reviewers to 

### 2025-04-18 review-responses.tex

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it Genetics}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Gertjan Bisschop  \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript,
XXX
titled 
``XXX''.
We appreciate the thoughtful comments by the reviewers.
We've done our best to address all the comments,
as detailed in our Response to Reviewers,
XXX

In addition to a Response to Reviewers (in which page/line numbers refer to the revised manuscript file),
we also provide a pdf with differences to the previously submitted version highlighted.

\vskip 2em

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
   Gertjan Bisschop, Peter Ralph, and Jerome Kelleher
}\\
\end{minipage}

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
The manuscript by Bisschop et al. presents a dense yet concise derivation of a novel approach for computing the likelihood of a broad class of Ancestral Recombination Graphs (ARGs) under the Sequentially Markovian Coalescent (SMC) model. This is a significant advance that enables the use of output from heuristic ARG inference tools such as tsinfer, ARG-Needle, and Threads, for rigorous statistical inference of population genetic parameters. These recent ARG inference methods can infer local gene trees across genomes for hundreds to thousands of samples, but they do not estimate recombination events directly in their data models, which prevented the application of previous formulations of the likelihood.
\end{quote}

\begin{quote}
The new likelihood framework has the potential to unlock parameter inference-such as recombination rates and historical population sizes-from large-scale genomic datasets using highly efficient methods. The manuscript provides a compelling proof-of-concept for such applications, while appropriately leaving more extensive exploration for future work.
\end{quote}

\begin{quote}
The subject matter is conceptually challenging due to the inherent complexity of the topic, but the authors provide rigorous and accessible explanations. Figures 1 and 2 are especially helpful.
\end{quote}

\begin{point}{}
A Python implementation of the method is available on GitHub. However, due to the algorithm's complexity and the very limited documentation of the available code, it would be beneficial to include pseudocode in the appendix. This pseudocode should lay out the full sequence of steps required to compute the likelihood, ideally supplemented with additional explanatory figures. Doing so would not only enhance the clarity and completeness of the manuscript, but also facilitate implementation in other programming languages.
\end{point}

\reply{
}

\begin{point}{\revref} %line 212
 Since you talk about the rate of coalescence, I think a ``lambda'' is missing in front of the sum (i.e., multiplying by 1/2/N).
\end{point}

\reply{
}

\begin{point}{\revref} % line 218
 Could you provide more detail or intuition for this formula of the likelihood? E.g., what's the cumulative hazard in the context of the coalescent or the ARG.
\end{point}

\reply{
}

\begin{point}{\revref} % line 262
 Again, I think there is a ``lambda'' missing in front of the integral. Alternatively, lambda is meant to be multiplied with the F function, like in equation 4, but then it also needs to be added to the formula in line 265.
\end{point}

\reply{
}

\begin{point}{\revref} % line 577
 I don't understand ``, along with additional information that allows us to sequentially recover (...) efficiently.'' Recover what?
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{2}
\begin{quote}
This paper does a couple of interrelated things.
\begin{enumerate}
    \item The authors develop a new backward-in-time formulation of the Sequentially Markov Coalescent (SMC) model.
    \item This backward-in-time formulation of the SMC model provides a way to efficiently calculate likelihood for a sample of genomes from this model, provided that we have information on how these genomes are related to each other. The latter is done by a data structure called Ancestral Recombination Graph (ARG). The authors provide an algorithm and Python code for this likelihood calculation (is this emphasised enough across the paper!?).
    \item The authors also evaluate the likelihood under different levels of information available in the ARG. This enables them to connect challenges when comparing methods and outputs from the rigorous, yet ``scale-limited'', probabilistic ARG inference methods with scaleable, yet less-rigorous, heuristic ARG inference methods.
\end{enumerate}
I welcome this paper and quite like it's succinctness. I am not able to follow all the maths, but the rationale and presentation is clear to me and I only have minor comments.
\end{quote}

% ABSTRACT

\begin{point}{\revref} % \&75
 Isn't it possible to run SINGER on up to thousands of samples? I know that several hundreds is also a thousand!
\end{point}

\reply{
}

\begin{point}{}
Mention that the standard SMC formulation was left-to-right
\end{point}

\reply{
}

\begin{point}{\revref} % 21:
 for ARG inference that this new formulation opens
\end{point}

\reply{
}

% INTRO

\begin{point}{\revref} % 24:
 evol biol $\to$ genetics?
\end{point}

\reply{
}

\begin{point}{\revref} % 27:
 ancestral processes $\to$ genetic processes?
\end{point}

\reply{
}

\begin{point}{\revref} % 35:
 Bayesian $\to$ probabilistic (I think this is more general and would encompas any model-based approach, say max likelihood etc.; alternatively consider using model-based, but probabilistic/statistical assumes we have an underlying model)
\end{point}

\reply{
}

\begin{point}{\revref} % 47:
 statistical $\to$ probabilistic
\end{point}

\reply{
}

\begin{point}{\revref} % 66:
 and heuristic methods omit these details in different ways. Notablym the sig....
\end{point}

\reply{
}

\begin{point}{\revref} % 72:
 uncertainty of (point) estimates.
\end{point}

\reply{
}

\begin{point}{\revref} % 82:
 heuristic methods
\end{point}

\reply{
}

\begin{point}{\revref} % 84:
 likelihood of a dataset? under a model (to make it more general)
\end{point}

\reply{
}

\begin{point}{\revref}
I would appreciate if you here make it clear which information is missing in these different tools. Sorry for nitpicking.
\end{point}

\reply{
}

\begin{point}{\revref} % 98:
 a new formulation of SMC model and corresponding ARG likelihood
\end{point}

\reply{
}

% RESULTS

\begin{point}{\revref} % 117:
 specific genome (node can also be a sample so best to remove ancestral)
\end{point}

\reply{
}

\begin{point}{\revref} % 118:
 sampled genomes are related at each base pair.
\end{point}

\reply{
}

\begin{point}{\revref} % 121:
 sampled and ancestral haplotypes (haploid genomes), or ...
\end{point}

\reply{
}

\begin{point}{Fig \ref{fig:smc-unary}:}
 it would be cool if you shade regions (in blue) also in the local trees? That's effectively uncertain time around the A, B, and C node!
\end{point}

\reply{
}

\begin{point}{Fig \ref{fig:smc-unary} caption:}
An ARG ``generated under the SMC'' - do you need this last part in my quotes?
\end{point}

\reply{
}

\begin{point}{Figure~\ref{fig:smc-unary} caption:}
with a the remaining $\to$ with the remai....
\end{point}

\reply{
    Thanks! Done.
}

\begin{point}{\revref} % 130:
 either sample nodes or common ancestry nodes
\end{point}

\reply{
}

\begin{point}{\revref} % 138:
 is unary in the second tree.
\end{point}

\reply{
}

\begin{point}{\revref} % 140:
 all lineages with recombination event
\end{point}

\reply{
}

\begin{point}{\revref} % 144:
 sometime between the times of nodes f and h; similarlyt in next line % 145
\end{point}

\reply{
}

\begin{point}{\revref{} and a few lines later} % 156/160:
 consolidate how you write backwards in time
\end{point}

\reply{
}

\begin{point}{\revref{} and elsewhere}
similarly for left-to-right and along-the-genome
\end{point}

\reply{
}

\begin{point}{\revref} % 164:
 had to Google what ``abutting'' means! Consider using bordering/adjoning
\end{point}

\reply{
}

\begin{point}{\revref} % 169:
 I don't fully grasp the meaning of ``tree-valued process'' - tree-generating process or tree-structured process ...
\end{point}

\reply{
}

% Here starts the part that I am not sufficiently-skilled to judge the text, though I could read and follow what is says and means!

\begin{point}{\revref} % 249:
 define ``eligible links'' sooner? Not critical.
\end{point}

\reply{
}

\begin{point}{\revref} % 280:
 ARG likelihood formulation in Eq. (5) ... of nodes (haplotypes that existed at some time t)
\end{point}

\reply{
}

\begin{point}{\revref} % 289:
 with a sample from the SMC model
\end{point}

\reply{
}

\begin{point}{\revref} % 300:
 cite what simplification does (Wang et al., YYYY; Kelleher et al. - your PLOS Genetics paper). Do you need to cite this sooner in the manuscript?
\end{point}

\reply{
}

\begin{point}{\revref} % 300:
 for this application, despite retaininig fewer nodes and edges.
\end{point}

\reply{
}

\begin{point}{\revref{} and subsequent paragraph} % 306-316:
 Is this number of polytomies too small and is then the impact on the likelihood small? It would be good to do one more extreme example for supplement to ensure this. This is the point where you can strengthen your paper the most.
\end{point}

\reply{
}

\begin{point}{Fig \ref{fig:lik-surface} caption:}
 lines overlap hence we cannot see the full red line.
\end{point}

\reply{
}

\begin{point}{\revref{} and elsewhere} % 319:
 Consolidate how you ref to figs (Fig 1. or Fig 3).
\end{point}

\reply{
}

\begin{point}{\revref} % 321:
 you have not mentioned how many mutations the simulation produced. I understand that these are not important for ARG likelihood (emphasise this), but good to tie up things together.
\end{point}

\reply{
}

\begin{point}{\revref} % 328:
 It's not that likelihood function performs well, right!? It's how absence of complete information gives close to the full information likelihood value!
\end{point}

\reply{
}

\begin{point}{\revref} % 332:
 remove significant? It does not look that bad!
\end{point}

\reply{
}

% DISCUSSION

\begin{point}{\revref} % 352:
 modern heuristic ARG ...
\end{point}

\reply{
}

\begin{point}{\revref} % 358:
 consider Kuhner-Yamato-Felsenstein $\to$ KYF (for consistency)
\end{point}

\reply{
}

\begin{point}{\revref} % 381:
 probabilistic approach
\end{point}

\reply{
}

\begin{point}{\revref} % 387:
 probabilistic inference
\end{point}

\reply{
}

\begin{point}{\revref} % 392-394:
 I checked the code and is indeed there! Thanks for sharing! The algorithm has an acronym RUNSMC standing for ``Recombination (time) UNaware SMC''. Should this be emphasised in the manuscript? Hmm, but it's not fully unaware, is it? You do consider topology, so think about a better acronym.
\end{point}

\reply{
}

### 2025-04-18 paper.tex
.

@@
 \newcommand{\comment}[1]{{\it \color{orange} (#1)}}
+
+% Responses to reviews
+\input{review-response-commands}
+% set this to show line numbers and include responses to reviews or not
+\newif\ifreviewresponses
+\reviewresponsestrue  % include them
+% \reviewresponsesfalse  % don't include them
+\newcommand{\responsefile}{review-responses}  % name of the review responses file
 
@@
 to be estimated, and is robust to the presence of polytomies. 
-We discuss the possibilities for inference that this opens.
+We discuss the possibilities for inference that this opens. \revpoint{2}{3}
 \end{abstract}
@@
 there is now strong interest in applying Ancestral Recombination Graphs (ARGs)
-to a variety of questions in evolutionary
-biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}.
+to a variety of questions in evolutionary biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}. \revpoint{2}{4}
 ARGs describe the intricately linked paths
@@
 recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
-and contain rich detail about ancestral processes.
+and contain rich detail about ancestral processes. \revpoint{2}{5}
 % TODO rephrase this a bit, would be a lot of duplication of references
@@
 % more natrual to talk about the advantages of model-driven stuff next.
-The first class of inference methods are Bayesian approaches
+The first class of inference methods are Bayesian approaches \revpoint{2}{6}
 that sample ARGs under a population genetics model such as the
@@
 heuristic and approximation driven, favouring computational
-scalability over a basis in a statistical model.
+scalability over a basis in a statistical model. \revpoint{2}{7}
 Secondly, they all produce a single
@@
 details of recombination events, and heuristic methods (in different ways)
-omit these details. In many ways, the significant progress made in
+omit these details. \revpoint{2}{8}
+In many ways, the significant progress made in
 scalability by these heuristic methods is \emph{because} they don't
@@
 posterior distribution provides a means of quantifying the uncertainty 
-around estimates.
+around estimates. \revpoint{2}{9}
 On the other hand, heuristic methods have the potential
@@
 in terms of scalability over ARGweaver and other methods, inference is still 
-only feasible for hundreds of samples. 
+only feasible for hundreds of samples.  \revpoint{2}{1}
 Meanwhile, tsinfer, ARG-Needle and Threads have 
@@
 
-A key problem facing the field is that the ARGs estimated by heuristic large-scale 
+A key problem facing the field is that the ARGs estimated by heuristic large-scale \revpoint{2}{10}
 methods lack a concrete connection to population genetic theory because
-we cannot compute their likelihood under the SMC or similar models. 
+we cannot compute their likelihood under the SMC or similar models. \revpoint{2}{11}
 Computing the likelihood of an ARG under a model is a fundamental element of
@@
 program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
-which does not contain sufficient detail for likelihood computation.
+which does not contain sufficient detail for likelihood computation. \revpoint{2}{12}
 The msprime
@@
 has much more lenient input requirements than KYF, and gracefully handles
-incomplete and imprecise ARGs.
+incomplete and imprecise ARGs. \revpoint{2}{13}
 % In this paper we provide a necessary first step towards this goal
@@
 By ``describes the history'',
-we mean that each node in the ARG represents a specific (ancestral) genome,
+we mean that each node in the ARG represents a specific (ancestral) genome, \revpoint{2}{14}
 and that the ARG specifies the genealogical tree describing how the samples are related
-at each point on the genome.
+at each point on the genome. \revpoint{2}{15}
 
@@
 the floating lineage coalesces randomly
-with a the remaining portion of the local tree following recombination.
+    with the remaining portion of the local tree following recombination.
 % By removing recombinant nodes, we limit the information on the time to each
@@
 An ARG can therefore be thought of either as a collection of inheritance
-relationships between ancestral haplotypes,
+relationships between ancestral haplotypes, \revpoint{2}{16}
 or as a sequence of genealogical trees that have common node labels.
@@
 Other nodes (labelled by black numbers) are either samples
-or \emph{common ancestry} nodes --
+or \emph{common ancestry} nodes -- \revpoint{2}{20}
 ancestors that are the most recent common ancestor for two 
@@
 while in the bottom plot, this is visible because node $f$ is parent to $d$
-in the second tree, but not to node $c$, and so is unary.
+in the second tree, but not to node $c$, and so is unary. \revpoint{2}{21}
 
@@
 These locally unary nodes enable
-us to uniquely identify all lineages that were hit by a recombination event,
+us to uniquely identify all lineages that were hit by a recombination event, \revpoint{2}{22}
 even if we simplify 
@@
 but we do still have constraints on it;
-for instance, we know that $C$ happened somewhere between node $f$ and node $h$.
+for instance, we know that $C$ happened somewhere between node $f$ and node $h$. \revpoint{2}{23}
 More generally, a recombinant node must happen somewhere between the child node
@@
 to simulate the coalescent with
-recombination was formulated backwards in time
+recombination was formulated backwards in time \revpoint{2}{24}
 by Hudson \citeyearpar{hudson_properties_1983}. 
@@
 obtain a process that is Markovian both backwards-in-time, 
-as well as along-the-genome.
+as well as along-the-genome. \revpoint{2}{25}
 In its backwards-in-time formulation, the SMC only requires a
@@
 genomic intervals may coalesce. The SMC' is a refinement 
-of the SMC which also allows lineages with abutting genomic intervals to 
-coalesce~\citep{marjoram_fast_2006}.
+of the SMC which also allows lineages with abutting genomic intervals \revpoint{2}{26}
+to coalesce~\citep{marjoram_fast_2006}.
 Although \citet{mcvean_approximating_2005} verbally described the SMC
@@
 
-Linking the resulting tree-valued process along the genome back to
-the relationship between the ``graph'' and
+Linking the resulting tree-valued process \revpoint{2}{27}
+along the genome back to the relationship between the ``graph'' and
 ``local trees'' representation of ARGs, 
@@
 thus, $|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
-and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$. 
+and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$.  \revpoint{1}{2}
 Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
@@
 In general, the likelihood for a continuous-time Markov process specified in this way
-is $\exp(-\Lambda) \prod_i \lambda_i$,
+is $\exp(-\Lambda) \prod_i \lambda_i$, \revpoint{1}{3}
 where $\lambda_i$ is the rate of the $i^\text{th}$ realised event,
@@
 $\exp(-r \mathcal{A}_e)$, where $r$ is the recombination rate
-and $\mathcal{A}_e$ is the total area (eligible links multiplied by length in time) of the edge.
+and $\mathcal{A}_e$ is the total area (eligible links multiplied by length in time) \revpoint{2}{28}
+of the edge.
 The length in time of the edge is just $t_p - t_c$,
@@
 Then, the cumulative hazard for common ancestry events of edge $e$
-between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(z) dz$.
+between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(z) dz$. \revpoint{1}{4}
 If the edge $e = (p,c,x,y)$
@@
 \subsection*{Robustness to partial and incorrect information}
-Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
+
+Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any} \revpoint{2}{29}
 collection of nodes (ancestral haplotypes that lived at some time 
@@
 containing complete information about every simulated event
-(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). 
+(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). \revpoint{2}{30}
 We then plot the likelihood computed by Eq.~\eqref{eq:full-lik} 
@@
 to remove all locally unary portions of nodes from trees 
-(``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees) 
-we can see there is a significant loss of information,
+(``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees)  \revpoint{2}{31}
+we can see there is a significant loss of information, \revpoint{2}{32}
 and the minimum of the 
@@
 
-Fig~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably
+Fig~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably \revpoint{2}{33}
 robust to the presences of polytomies in this example. For each of the three
@@
 
-Fig~\ref{fig:lik-surface} also shows results for ARGs inferred from the 
+Fig~\ref{fig:lik-surface} also shows results for ARGs inferred from the  \revpoint{2}{35}
 simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
@@
 In the first, we use the default output of tsinfer 
-(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations).
-This ARG has 192 unary nodes, 120 binary nodes, and 32 nodes
+(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations). \revpoint{2}{36}
+This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
 with three or more children per tree. Tsinfer does not directly estimate
@@
 moves~\citep[e.g.,][]{rasmussen2023espalier}. Nonetheless, in
-this toy example at least, our likelihood function performs well
+this toy example at least, our likelihood function performs well \revpoint{2}{37}
 on the output of tsinfer, with the minimum close to the true
@@
 before dating. We can see that, as with the true data,
-removing all the unary nodes results in a significant bias
+removing all the unary nodes results in a significant bias \revpoint{2}{38}
 away from the true value of the recombination rate parameter.
@@
 and imprecise information
-provided by some modern ARG inference tools,
+provided by some modern ARG inference tools, \revpoint{2}{39}
 using a novel backwards-time derivation of the SMC.
@@
 a significant increase in flexibility over the 
-strict input requirements for the classical Kuhner-Yamato-Felsenstein 
+strict input requirements for the classical Kuhner-Yamato-Felsenstein  \revpoint{2}{40}
 formulation. 
@@
 the details of the recent past in huge datasets such as 
-UK Biobank, passing over to a rigorous model-based
-approach for the ancient past.
+UK Biobank, passing over to a rigorous model-based approach \revpoint{2}{41}
+for the ancient past.
 This is similar in spirit to the widely-used approach 
@@
 and can be combined to great effect,
-large-scale heuristic and Bayesian sampling inference methods have their 
-applications, and there may be synergy in their combination. 
+large-scale heuristic and Bayesian sampling inference  \revpoint{2}{42}
+methods have their applications, and there may be synergy in their combination. 
 % This is pretty weak, tbh.
@@
 
-\section{Data availability}
+\section*{Data availability}
 
@@
 \url{https://github.com/gertjanbisschop/runsmc}.
+\revpoint{2}{43}
+
+\section*{Acknowledgements}
+JK acknowledges EPSRC (research grant EP/X024881/1)
+and the Robertson Foundation.
+JK and PR acknowledge support from 
+NIH (research grants HG011395 and HG012473).
+
 \FloatBarrier
@@
 along with additional information that
-allows us to sequentially recover (and reason about 
-their differences) 
+allows us to sequentially recover (and reason about their differences) \revpoint{1}{5}
 efficiently~\citep{kelleher_efficient_2016, ralph_efficiently_2020}.
@@
 
+\includereviews
+
 \end{document}

### 2025-04-23 paper.tex
Address point

@@
 along with additional information that
-allows us to sequentially recover (and reason about their differences) \revpoint{1}{5}
+allows us to sequentially recover local trees (and reason about their differences) \revpoint{1}{5}
 efficiently~\citep{kelleher_efficient_2016, ralph_efficiently_2020}.

### 2025-04-23 review-responses.tex
nuke whitespace
Address point
Address some comments

@@
 {\bf
-   Gertjan Bisschop, Peter Ralph, and Jerome Kelleher
+   Gertjan Bisschop, Jerome Kelleher, and Peter Ralph
 }\\
@@
 \reply{
+We were missing ``the local trees here'', thanks.
 }
@@
 \reply{
+We inserted a parenthetical comment stating that the SMC
+is ``conventionally regarded as an along-the-genome process''.
 }
@@
 \reply{
+Done.
 }

### 2025-04-23 paper.tex
Address some comments

@@
 In this paper we present a
-backwards-time formulation of the SMC
+backwards-time formulation of the SMC (conventionally regarded as
+an along-the-genome process)
 and derive a straightforward definition
@@
 to be estimated, and is robust to the presence of polytomies.
-We discuss the possibilities for inference that this opens. \revpoint{2}{3}
+We discuss the possibilities
+for ARG inference that this new formulation opens. \revpoint{2}{3}
 \end{abstract}

### 2025-04-23 paper.tex
ADdress a few points

@@
 there is now strong interest in applying Ancestral Recombination Graphs (ARGs)
-to a variety of questions in evolutionary biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}. \revpoint{2}{4}
+to a variety of questions in genetics~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}. \revpoint{2}{4}
 ARGs describe the intricately linked paths
@@
 % more natrual to talk about the advantages of model-driven stuff next.
-The first class of inference methods are Bayesian approaches \revpoint{2}{6}
+The first class of inference methods are probabilistic approaches \revpoint{2}{6}
 that sample ARGs under a population genetics model such as the
@@
 heuristic and approximation driven, favouring computational
-scalability over a basis in a statistical model. \revpoint{2}{7}
+scalability over a basis in a probabilistic model. \revpoint{2}{7}
 Secondly, they all produce a single

### 2025-04-23 review-responses.tex
ADdress a few points
Address some more points

@@
 \reply{
+We would prefer to keep ancestral processes here as ARGs contain
+information about, for example, population structure which is not
+a genetic process, but something that we gains insights into
+through genetics.
 }
@@
 \reply{
+Done (but slightly rephrased next sentence).
 }
@@
 \reply{
+We would prefer to keep this as it is to make it clear that it
+is the ARGs themselves that we want to compute the likelihood of.
 }
@@
 \reply{
+This is quite a subtle point which we feel would distract from the main
+message here by expanding. We have attempted to clarify by stating
+that a sequence of local trees does not have enough information to
+reconstruct a full ARG (required for likelihood calculation under KYF)
+and referred to Wong et al for the details, where there is a full
+section dedicated to the correspondence between an  ARG and the
+sequence of local trees.
 }

### 2025-04-23 paper.tex
Address some more points

@@
 Similarly, there is often a fundamental lack of information about the
-details of recombination events, and heuristic methods (in different ways)
-omit these details. \revpoint{2}{8}
-In many ways, the significant progress made in
-scalability by these heuristic methods is \emph{because} they don't
+details of recombination events, and heuristic methods
+omit these details in different ways. \revpoint{2}{8}
+The significant progress made in
+scalability by these heuristic methods is in part \emph{because} they don't
 attempt to precisely reconstruct recombination events.
@@
 posterior distribution provides a means of quantifying the uncertainty
-around estimates. \revpoint{2}{9}
+around point estimates. \revpoint{2}{9}
 On the other hand, heuristic methods have the potential
@@
 
-A key problem facing the field is that the ARGs estimated by heuristic large-scale \revpoint{2}{10}
+A key problem facing the field is that the ARGs estimated by heuristic \revpoint{2}{10}
 methods lack a concrete connection to population genetic theory because
@@
 program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
-which does not contain sufficient detail for likelihood computation. \revpoint{2}{12}
+which does not contain sufficient detail to reconstruct
+the ``full ARG'' required for likelihood
+computation~\citep{wong_general_2023}. \revpoint{2}{12}
 The msprime

### 2025-04-23 review-responses.tex
Some responses

@@
 \reply{
+We are hesitant to call this a ``new formulation of the SMC''
+because we are very much building on the foundations laid by
+McVean and Cardin. They developed the SMC initially in a
+backwards-time context, but the field has subsequently
+focused on the left-to-right perspective. Thus, we wish
+to keep this as-is.
 }

### 2025-04-23 paper.tex
Some responses

@@
 By ``describes the history'',
-we mean that each node in the ARG represents a specific (ancestral) genome, \revpoint{2}{14}
-and that the ARG specifies the genealogical tree describing how the samples are related
-at each point on the genome. \revpoint{2}{15}
+we mean that each node in the ARG represents a specific genome, \revpoint{2}{14}
+and that the ARG specifies the genealogical tree describing how the
+sampled genomes are related
+at each base pair. \revpoint{2}{15}
 
@@
 An ARG can therefore be thought of either as a collection of inheritance
-relationships between ancestral haplotypes, \revpoint{2}{16}
+relationships between
+sampled and ancestral haplotypes (haploid genomes), \revpoint{2}{16}
 or as a sequence of genealogical trees that have common node labels.

### 2025-04-23 paper.tex
Some updates

@@
 \includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
-\caption{An ARG generated under the SMC, represented as a graph (top) and as a
+\caption{An ARG represented as a graph (top) and as a
 series of local trees (bottom). To ensure the one-to-one correspondence
@@
 but that segment was inherited in two pieces from their parent.
-Other nodes (labelled by black numbers) are either samples
+Other nodes (labelled by black numbers) are either sample nodes
 or \emph{common ancestry} nodes -- \revpoint{2}{20}
@@
 while in the bottom plot, this is visible because node $f$ is parent to $d$
-in the second tree, but not to node $c$, and so is unary. \revpoint{2}{21}
+in the second tree, but not to node $c$, and so is unary
+in the second tree. \revpoint{2}{21}
 
@@
 but we do still have constraints on it;
-for instance, we know that $C$ happened somewhere between node $f$ and node $h$. \revpoint{2}{23}
-More generally, a recombinant node must happen somewhere between the child node
+for instance, we know that $C$ happened sometime between the
+times of node $f$ and node $h$. \revpoint{2}{23}
+More generally, a recombinant node must happen in the time interval between the child node
 and the most recent of the parent nodes.

### 2025-04-23 paper.tex
Backwards in time

@@
 % to compute the likelihood of a general class of ARG under the SMC.
-To do this, we give a backwards-in-time description of the SMC
+To do this, we give a backwards in time description of the SMC
 which naturally leads to a straightforward method for computing the likelihood
@@
 % Then,
-% we define the SMC process in a ``backwards-in-time'' way,
+% we define the SMC process in a ``backwards in time'' way,
 % rather than the usual ``along-the-genome'' way.
@@
 % In the subsequent section we connect the two and
-% show how the ``backwards-in-time'' description
+% show how the ``backwards in time'' description
 % makes it relatively straightforward to compute the likelihood
@@
 % % % % % % % % % % % % % % % % % %
-\subsection*{SMC backwards-in-time}\label{par:description}
-
-% coalescent introduction: backwards-in-time
+\subsection*{SMC backwards in time}\label{par:description}
+
+% coalescent introduction: backwards in time
 % left-to-right: Wiuf-Hein (difficult)
@@
 space transitions possible under the Hudson coalescent to
-obtain a process that is Markovian both backwards-in-time,
+obtain a process that is Markovian both backwards in time,
 as well as along-the-genome. \revpoint{2}{25}
-In its backwards-in-time formulation, the SMC only requires a
+In its backwards in time formulation, the SMC only requires a
 simple modification to the coalescent with recombination.
@@
 Although \citet{mcvean_approximating_2005} verbally described the SMC
-using this backwards-in-time description, their analysis
+using this backwards in time description, their analysis
 was almost entirely in left-to-right terms,
@@
 % and subsequently use a principled approach to perform ARG inference for the second phase.
-% This idea resembles the tried and tested approach of hybrid backwards-in-time simulations
+% This idea resembles the tried and tested approach of hybrid backwards in time simulations
 % that combine Wright-Fisher dynamics in the very recent past and coalescent simulations
@@
 % As noted by \cite{mcvean_approximating_2005},
-% this implies that the backwards-in-time description of the
+% this implies that the backwards in time description of the
 % SMC is slightly different from the left-to-right description,

### 2025-04-23 paper.tex
Changed my mind, backwards-in-time

--- 2025-04-23 paper.tex
@@
 to simulate the coalescent with
-recombination was formulated backwards in time \revpoint{2}{24}
+recombination was formulated backwards-in-time \revpoint{2}{24}
 by Hudson \citeyearpar{hudson_properties_1983}.
@@
 greatly simplifies the ability to keep track of coalescable pairs going
-backwards in time.
+backwards-in-time.
 At any point in time, the state of the process is
@@
 % for some interval endpoints $x_{j} < y_{j}$.
-Because we look backwards in time, $t=0$ is today, and
+Because we look backwards-in-time, $t=0$ is today, and
 $L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
@@
 % Finally, by describing the SMC
-% backwards in time, some intervals contained within the half-open
+% backwards-in-time, some intervals contained within the half-open
 % interval representing a lineage might have fully coalesced before all the genetic
@@
 % Here, we have introduced a scalable algorithm to compute likelihoods
-% under the SMC for a general class of ARGs. By formulating the SMC backwards in time
+% under the SMC for a general class of ARGs. By formulating the SMC backwards-in-time
 % we were able to associate a single likelihood value with each edge in the ARG.
@@
 % and its \emph{size} is $|X_i(t)| = y_{im_i} - x_{i0}$.
-% Since we look backwards in time, $t=0$ is today, and so
+% Since we look backwards-in-time, $t=0$ is today, and so
 % $L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,

### 2025-04-23 paper.tex
Update

@@
 genomic intervals may coalesce. The SMC' is a refinement
-of the SMC which also allows lineages with abutting genomic intervals \revpoint{2}{26}
+of the SMC which also allows lineages with adjoining genomic intervals \revpoint{2}{26}
 to coalesce~\citep{marjoram_fast_2006}.

### 2025-04-23 review-responses.tex
Some updates
Backwards in time
Update
More review points

@@
 \reply{
+Deleted, thanks!
 }
@@
 \reply{
+We  prefer "hit by a recombination event" as it's a little more concrete,
+and in keeping with standard phrasing in the field.
 }
@@
 \reply{
+Done - using "backwards-in-time" now.
 }
@@
 \reply{
+Done (although we do have ``along the genome'' in places, where
+we're not referring to the actual process).
 }
@@
 \reply{
+We prefer to keep this as ``modern ARG inference'' as we feel
+that the fact they are heuristic is not really the point---one
+could define rigorous methods that also omit full details
+about recombination.
 }

### 2025-04-23 paper.tex
More review points

@@
 a significant increase in flexibility over the
-strict input requirements for the classical Kuhner-Yamato-Felsenstein  \revpoint{2}{40}
+strict input requirements for the classical KYF  \revpoint{2}{40}
 formulation.
@@
 the details of the recent past in huge datasets such as
-UK Biobank, passing over to a rigorous model-based approach \revpoint{2}{41}
+UK Biobank, passing over to a rigorous probabilistic approach \revpoint{2}{41}
 for the ancient past.
@@
 and can be combined to great effect,
-large-scale heuristic and Bayesian sampling inference  \revpoint{2}{42}
+large-scale heuristic and probabilistic inference  \revpoint{2}{42}
 methods have their applications, and there may be synergy in their combination.

### 2025-04-28 paper.tex

--- 2025-04-23 paper.tex
@@
 In this paper we present a
-backwards-time formulation of the SMC (conventionally regarded as
-an along-the-genome process)
+backwards-time formulation of the SMC
 and derive a straightforward definition
@@
 to be estimated, and is robust to the presence of polytomies.
-We discuss the possibilities
-for ARG inference that this new formulation opens. \revpoint{2}{3}
+We discuss the possibilities for inference that this opens. \revpoint{2}{3}
 \end{abstract}
@@
 there is now strong interest in applying Ancestral Recombination Graphs (ARGs)
-to a variety of questions in genetics~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}. \revpoint{2}{4}
+to a variety of questions in evolutionary biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}. \revpoint{2}{4}
 ARGs describe the intricately linked paths
@@
 % more natrual to talk about the advantages of model-driven stuff next.
-The first class of inference methods are probabilistic approaches \revpoint{2}{6}
+The first class of inference methods are Bayesian approaches \revpoint{2}{6}
 that sample ARGs under a population genetics model such as the
@@
 heuristic and approximation driven, favouring computational
-scalability over a basis in a probabilistic model. \revpoint{2}{7}
+scalability over a basis in a statistical model. \revpoint{2}{7}
 Secondly, they all produce a single
@@
 Similarly, there is often a fundamental lack of information about the
-details of recombination events, and heuristic methods
-omit these details in different ways. \revpoint{2}{8}
-The significant progress made in
-scalability by these heuristic methods is in part \emph{because} they don't
+details of recombination events, and heuristic methods (in different ways)
+omit these details. \revpoint{2}{8}
+In many ways, the significant progress made in
+scalability by these heuristic methods is \emph{because} they don't
 attempt to precisely reconstruct recombination events.
@@
 posterior distribution provides a means of quantifying the uncertainty
-around point estimates. \revpoint{2}{9}
+around estimates. \revpoint{2}{9}
 On the other hand, heuristic methods have the potential
@@
 
-A key problem facing the field is that the ARGs estimated by heuristic \revpoint{2}{10}
+A key problem facing the field is that the ARGs estimated by heuristic large-scale \revpoint{2}{10}
 methods lack a concrete connection to population genetic theory because
@@
 program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
-which does not contain sufficient detail to reconstruct
-the ``full ARG'' required for likelihood
-computation~\citep{wong_general_2023}. \revpoint{2}{12}
+which does not contain sufficient detail for likelihood computation. \revpoint{2}{12}
 The msprime
@@
 By ``describes the history'',
-we mean that each node in the ARG represents a specific genome, \revpoint{2}{14}
-and that the ARG specifies the genealogical tree describing how the
-sampled genomes are related
-at each base pair. \revpoint{2}{15}
+we mean that each node in the ARG represents a specific (ancestral) genome, \revpoint{2}{14}
+and that the ARG specifies the genealogical tree describing how the samples are related
+at each point on the genome. \revpoint{2}{15}
 
@@
 \includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
-\caption{An ARG represented as a graph (top) and as a
+\caption{An ARG generated under the SMC, represented as a graph (top) and as a
 series of local trees (bottom). To ensure the one-to-one correspondence
@@
 An ARG can therefore be thought of either as a collection of inheritance
-relationships between
-sampled and ancestral haplotypes (haploid genomes), \revpoint{2}{16}
+relationships between ancestral haplotypes, \revpoint{2}{16}
 or as a sequence of genealogical trees that have common node labels.
@@
 but that segment was inherited in two pieces from their parent.
-Other nodes (labelled by black numbers) are either sample nodes
+Other nodes (labelled by black numbers) are either samples
 or \emph{common ancestry} nodes -- \revpoint{2}{20}
@@
 while in the bottom plot, this is visible because node $f$ is parent to $d$
-in the second tree, but not to node $c$, and so is unary
-in the second tree. \revpoint{2}{21}
+in the second tree, but not to node $c$, and so is unary. \revpoint{2}{21}
 
@@
 but we do still have constraints on it;
-for instance, we know that $C$ happened sometime between the
-times of node $f$ and node $h$. \revpoint{2}{23}
-More generally, a recombinant node must happen in the time interval between the child node
+for instance, we know that $C$ happened somewhere between node $f$ and node $h$. \revpoint{2}{23}
+More generally, a recombinant node must happen somewhere between the child node
 and the most recent of the parent nodes.
@@
 to simulate the coalescent with
-recombination was formulated backwards-in-time \revpoint{2}{24}
+recombination was formulated backwards in time \revpoint{2}{24}
 by Hudson \citeyearpar{hudson_properties_1983}.
@@
 greatly simplifies the ability to keep track of coalescable pairs going
-backwards-in-time.
+backwards in time.
 At any point in time, the state of the process is
@@
 % for some interval endpoints $x_{j} < y_{j}$.
-Because we look backwards-in-time, $t=0$ is today, and
+Because we look backwards in time, $t=0$ is today, and
 $L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
@@
 % Finally, by describing the SMC
-% backwards-in-time, some intervals contained within the half-open
+% backwards in time, some intervals contained within the half-open
 % interval representing a lineage might have fully coalesced before all the genetic
@@
 % Here, we have introduced a scalable algorithm to compute likelihoods
-% under the SMC for a general class of ARGs. By formulating the SMC backwards-in-time
+% under the SMC for a general class of ARGs. By formulating the SMC backwards in time
 % we were able to associate a single likelihood value with each edge in the ARG.
@@
 along with additional information that
-allows us to sequentially recover local trees (and reason about their differences) \revpoint{1}{5}
+allows us to sequentially recover (and reason about their differences) \revpoint{1}{5}
 efficiently~\citep{kelleher_efficient_2016, ralph_efficiently_2020}.
@@
 % and its \emph{size} is $|X_i(t)| = y_{im_i} - x_{i0}$.
-% Since we look backwards-in-time, $t=0$ is today, and so
+% Since we look backwards in time, $t=0$ is today, and so
 % $L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,

### 2025-04-28 review-responses.tex

@@
 information about, for example, population structure which is not
-a genetic process, but something that we gains insights into
+a genetic process, but something that we gain insights into
 through genetics.
@@
 \reply{
-We  prefer "hit by a recombination event" as it's a little more concrete,
+We  prefer ``hit by a recombination event" as it's a little more concrete,
 and in keeping with standard phrasing in the field.
@@
 \reply{
-Done - using "backwards-in-time" now.
+Done - using ``backwards-in-time" now.
 }

### 2025-04-28 paper.tex

--- 2025-04-23 paper.tex
@@
 The significant progress made in
-scalability by these heuristic methods is in part \emph{because} they don't
+scalability by these heuristic methods is in part because they don't
 attempt to precisely reconstruct recombination events.
@@
 the path traced by $A$'s ancestry switches at that point.
-In fact, $A$ is in blue because it is a \emph{recombinant} node --
-an ancestor that has a segment of genome inherited by a sample,
+In fact, $A$ is in blue because it is a \emph{recombinant}
+node -- an ancestor that has a segment of genome inherited by a sample,
 but that segment was inherited in two pieces from their parent.
@@
 events until each segment of ancestral material is only present in one lineage.
-The waiting time to next event is determined by these two competing processes
+The waiting time to the next event is determined by these two competing processes
 with exponentially distributed waiting times as outlined below.
@@
 of genomic length and time:
-so, if $T(t) = \sum_j |X_j(t)|$ is total length of ancestral material
+so, if $T(t) = \sum_j |X_j(t)|$ is the total length of ancestral material
 carried by lineages at time $t$,

### 2025-04-28 review-responses.tex
Bat back SINGER point

@@
 \reply{
+We think that the characterisation of SINGER being feasible for hundreds of
+samples is fair and reasonable. In the SINGER preprint, the maximum sample 
+size evaluated on simulations is 300, and 
+the flagship real data example 
+uses only 200 African samples from the 1000 Genomes dataset.
+How we define ``feasible''  is of course subjective, but 
+the scale of inferences done in the 
+method's paper seems like a reasonable approximation.
 }

### 2025-04-28 paper.tex
Hit a few comments

@@
 
-In Fig~\ref{fig:lik-surface} we begin with a realisation of the SMC
+In Fig.~\ref{fig:lik-surface} we begin with a realisation of the SMC
 containing complete information about every simulated event
@@
 
-Fig~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably \revpoint{2}{33}
+Fig.~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably \revpoint{2}{33}
 robust to the presences of polytomies in this example. For each of the three
@@
 removing 25\% of the internal
-nodes has negligible effect on the likelihood curve in Fig~\ref{fig:lik-surface},
+nodes has negligible effect on the likelihood curve in Fig.~\ref{fig:lik-surface},
 and even on the fully simplified
@@
 
-Fig~\ref{fig:lik-surface} also shows results for ARGs inferred from the  \revpoint{2}{35}
+Fig.~\ref{fig:lik-surface} also shows results for ARGs inferred from the  \revpoint{2}{35}
 simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
@@
 before dating. We can see that, as with the true data,
-removing all the unary nodes results in a significant bias \revpoint{2}{38}
+removing all the unary nodes results in a bias \revpoint{2}{38}
 away from the true value of the recombination rate parameter.

### 2025-04-28 paper.tex

@@
 $\exp(-r \mathcal{A}_e)$, where $r$ is the recombination rate
-and $\mathcal{A}_e$ is the total area (eligible links multiplied by length in time) \revpoint{2}{28}
+and $\mathcal{A}_e$ is the total ``area'' \revpoint{2}{28}
 of the edge.
-The length in time of the edge is just $t_p - t_c$,
-while the number of ``eligible links'' in the edge is the number of adjacent base pairs in the edge
+The area of an edge is its  length in time, $t_p - t_c$,
+multiplied by the number of ``eligible links'' along its
+span.
+The number of eligible links is the number of adjacent base pairs in the edge
 on which recombination did not occur, which is either $y-x$

### 2025-04-29 review-responses.tex
Hit a few comments
Tackle "eligible links" comment
Batch of responses

@@
 We are writing to submit a revised version of our manuscript,
-XXX
+GENETICS-2025-307907 
 titled
-``XXX''.
+``Likelihoods for a general class of ARGs under the SMC''.
 We appreciate the thoughtful comments by the reviewers.
 We've done our best to address all the comments,
-as detailed in our Response to Reviewers,
-XXX
+as detailed in our Response to Reviewers.
 
@@
 \reply{
+Reworded the definition of edge area to clarify.
 }
@@
 \reply{
+We would prefer to keep this sentence as it is, as ``retaining" here could 
+be confusing --- the calculation doesn't keep or retain nodes and edges, that is a 
+property of the input ARG.
 }
@@
 \begin{point}{\revref{} and subsequent paragraph} % 306-316:
- Is this number of polytomies too small and is then the impact on the likelihood small?
-It would be good to do one more extreme example for supplement to ensure this.
-This is the point where you can strengthen your paper the most.
+ Is this number of polytomies too small and is then the impact on the
+likelihood small? It would be good to do one more extreme example for
+supplement to ensure this. This is the point where you can strengthen your
+paper the most.
 \end{point}

### 2025-04-29 paper.tex
Batch of responses

@@
 \subsection*{Robustness to partial and incorrect information}
-
-Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any} \revpoint{2}{29}
-collection of nodes (ancestral haplotypes that lived at some time
-$t$ in the past) and edges (a record that child node $c$
+The ARG likelihood formulation in 
+Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for any \revpoint{2}{29}
+collection of nodes (haplotypes that existed at some time $t$) and 
+edges (a record that child node $c$
 inherited the genomic interval $[x, y)$ from parent node $p$)
@@
 
-In Fig.~\ref{fig:lik-surface} we begin with a realisation of the SMC
+In Fig.~\ref{fig:lik-surface} we begin with 
+with a sample from the SMC model 
 containing complete information about every simulated event
@@
 We next compute the likelihood ratio for the ARG obtained when we
-simplify out the recombination nodes from the full ARG
+simplify~\citep{kelleher_efficient_2018,wong_general_2023}
+ out the recombination nodes from the full ARG
 (``RE nodes removed''; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime
@@
 contain all the information that we need about recombination events
-for this application.
+for this application. \revpoint{2}{31}
 When we fully simplify the simulated ARG
 to remove all locally unary portions of nodes from trees
-(``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees)  \revpoint{2}{31}
+(``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees)  
 we can see there is a significant loss of information, \revpoint{2}{32}

### 2025-04-29 figures/generate-figs/plot_likelihood_surface.py
Update likelihood fig drawing

@@ -50,6 +50,10 @@ def plot(df_rat, path):
             if arg == "full_arg_hudson":
                 continue
             poly = "0"
+        marker = None
+        if arg == "full_arg_smc":
+            marker="o"
+
 
         y = df_rat[col]  # / df_rat[arg]
         line = ax.semilogx(
@@ -57,6 +61,8 @@ def plot(df_rat, path):
             y,
             color=colours[arg],
             ls=line_styles[poly],
+            marker=marker,
+            markersize=2,
             label="_nolegend_" if poly != "0" else titles[arg],
         )
 

### 2025-04-29 review-responses.tex
Update likelihood fig drawing
Add new supplemental figure.

@@
 \reply{
+We have added a supplemental figure which explores the effects of deleting
+up to 75\% of the internal nodes, and the overall picture remains the 
+same. Remarkably, this seems to improve
+performance in the fully-simplified case. Hopefully this will spark 
+some interest in follow-up studies.
 }
@@
 \reply{
+We have added small red dots on the red line to make it legible.
 }

### 2025-04-29 paper.tex
Add new supplemental figure.

@@
 ARG has only a very minor effect on location of the minimum.
+Supplemental Fig.~S1 explores more extreme examples in which we remove
+25\%, 50\% and 75\% of the internal nodes in these ARGs, and demonstrates 
+a remarkable robustness to the loss of the majority of the internal
+nodes in these examples. Interestingly, the loss of 75\% of the 
+internal nodes  in the fully simplified ARG 
+(1841 of 2455, leading  to an average of 27 polytomies per tree)
+appears to improve performance, substantially 
+reducing the bias away from the true parameter value.
 
@@
 the last parent a child node was connected to to detect recombination events.
+
+\setcounter{table}{0}
+\setcounter{figure}{0}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\begin{figure}[!ht]
+\centering
+\includegraphics{figures/likelihood_surface_suppl}
+    \caption{Log likelihood-ratio curves for three different ARGs 
+    with increasing fractions of polytomies.
+    The base ARGs are as described in Fig.~3, but here we remove 
+    25\%, 50\% and 75\% of the internal nodes to create polytomies.
+    The true value of recombination parameter is indicated by the 
+    dashed line, as in Fig.~3.
+    \label{fig:likelihood_suppl}}
+\end{figure}
 

### 2025-04-29 figures/generate-figs/plot_likelihood_surface.py

@@ -4,10 +4,37 @@ import matplotlib.pyplot as plt
 import matplotlib.lines as mlines
 import pandas as pd
 
+REC = 1.25e-8
+
+
+def plot_single(ax, df, title, include_legend=False):
+    print(list(df))
+
+    x = df.index
+    for col in df:
+        if col.startswith("poly"):
+            splits = col.split("_")
+            arg = "_".join(splits[1:-1])
+            poly = splits[-1]
+        else:
+            arg = col
+            if arg == "full_arg_hudson":
+                continue
+            poly = "0"
+
+        y = df[col]
+        line = ax.semilogx(x, y, label=f"{float(poly):.0%}")
+
+    if include_legend:
+        ax.legend(title="Polytomies")
+
+    ax.set_title(title)
+
+    ax.set_ylabel("Log likelihood ratio")
+    ax.axvline(REC, linestyle="dashed", color="grey")
+
 
 def plot(df_rat, path):
-
-    REC = 1.25e-8
 
     fig, ax = plt.subplots(1, 1)
 
@@ -32,12 +59,11 @@ def plot(df_rat, path):
         "0": "solid",
         "0.05": "dashed",
         "0.25": "dotted",
+        "0.5": "dashdot",
+        "0.75": "-",
     }
 
-    ls_patches = [
-        mlines.Line2D([0], [0], color="black", lw=1, ls=ls, label=f"{float(label):.0%}")
-        for label, ls in line_styles.items()
-    ]
+    used_line_styles = set()
 
     x = df_rat.index
     for col in df_rat:
@@ -52,8 +78,9 @@ def plot(df_rat, path):
             poly = "0"
         marker = None
         if arg == "full_arg_smc":
-            marker="o"
+            marker = "o"
 
+        used_line_styles.add(poly)
 
         y = df_rat[col]  # / df_rat[arg]
         line = ax.semilogx(
@@ -66,6 +93,18 @@ def plot(df_rat, path):
             label="_nolegend_" if poly != "0" else titles[arg],
         )
 
+    ls_patches = [
+        mlines.Line2D(
+            [0],
+            [0],
+            color="black",
+            lw=1,
+            ls=line_styles[label],
+            label=f"{float(label):.0%}",
+        )
+        for label in sorted(used_line_styles)
[…]

### 2025-04-29 paper.tex
Updates

@@
     msprime simulation of the SMC (100 diploid samples, genome length 1Mb,
-    $r=\mu=2.5\times10^{-8}$, $N_e=1\times10^4$).
+    $r=\mu=1.25\times10^{-8}$, $N_e=1\times10^4$).
     The ``RE nodes removed'' ARG is the
@@
 and even on the fully simplified
-ARG has only a very minor effect on location of the minimum.
+ARG has a very minor effect on location of the minimum.
 Supplemental Fig.~S1 explores more extreme examples in which we remove
@@
 
-Fig.~\ref{fig:lik-surface} also shows results for ARGs inferred from the  \revpoint{2}{35}
-simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
-using default parameters for tsinfer, and the true values of mutation
+Fig.~\ref{fig:lik-surface} also shows results for ARGs inferred from 
+the  \revpoint{2}{35}
+simulated data.
+To do this we first simulated mutations on the full ARG 
+at rate $1.25\times10^{-8}$ using msprime, resulting in an 
+ARG with 2,989 mutations.
+We then used tsinfer and tsdate~\citep{wohns_unified_2022}
+to infer a dated ARG,
+using default parameters for tsinfer and the true values of mutation
 rate and $N_e$ for tsdate.
@@
     Remaining parameters: genome length is 10kb, 20 diploid samples, and
-    $r=2.5\times10^{-8}$.
+    $r=1.25\times10^{-8}$.
     There are 148 edges in the resulting ARG.

### 2025-04-29 paper.tex
Cover tricky tsinfer points

@@
 rate and $N_e$ for tsdate.
-In the first, we use the default output of tsinfer
+In the first example, we use the default output of tsinfer
 (2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations). \revpoint{2}{36}
@@
 attempt to reconstruct the exact subtree-prune-and-regraft
-moves~\citep[e.g.,][]{rasmussen2023espalier}. Nonetheless, in
-this toy example at least, our likelihood function performs well \revpoint{2}{37}
-on the output of tsinfer, with the minimum close to the true
+moves~\citep[e.g.,][]{rasmussen2023espalier}. It is encouraging,
+therefore, that in this toy example our likelihood function is robust
+to these omitted and incorrect details in \revpoint{2}{37}
+the output of tsinfer, with the minimum close to the true
 parameter value. We also show the likelihood curve obtained when we

### 2025-04-29 paper.tex
Fixes

@@
 
-In Fig.~\ref{fig:lik-surface} we begin with 
+In Fig.~\ref{fig:lik-surface} we begin 
 with a sample from the SMC model 
@@
 and even on the fully simplified
-ARG has a very minor effect on location of the minimum.
+ARG has a very minor effect on the location of the minimum.
 Supplemental Fig.~S1 explores more extreme examples in which we remove
@@
     25\%, 50\% and 75\% of the internal nodes to create polytomies.
-    The true value of recombination parameter is indicated by the 
+    The true value of the recombination parameter is indicated by the 
     dashed line, as in Fig.~3.

### 2025-04-29 submitted_version.tex

--- 2025-04-29 paper.tex
@@
 \newcommand{\comment}[1]{{\it \color{orange} (#1)}}
-
-% Responses to reviews
-\input{review-response-commands}
-% set this to show line numbers and include responses to reviews or not
-\newif\ifreviewresponses
-\reviewresponsestrue  % include them
-% \reviewresponsesfalse  % don't include them
-\newcommand{\responsefile}{review-responses}  % name of the review responses file
 
@@
 In this paper we present a 
-backwards-time formulation of the SMC (conventionally regarded as
-an along-the-genome process)
+backwards-time formulation of the SMC 
 and derive a straightforward definition 
@@
 to be estimated, and is robust to the presence of polytomies. 
-We discuss the possibilities
-for ARG inference that this new formulation opens. \revpoint{2}{3}
+We discuss the possibilities for inference that this opens.
 \end{abstract}
@@
 there is now strong interest in applying Ancestral Recombination Graphs (ARGs)
-to a variety of questions in genetics~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}. \revpoint{2}{4}
+to a variety of questions in evolutionary
+biology~\citep{lewanski_era_2024,brandt_promise_2024,nielsen_inference_2025}.
 ARGs describe the intricately linked paths
@@
 recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
-and contain rich detail about ancestral processes. \revpoint{2}{5}
+and contain rich detail about ancestral processes.
 % TODO rephrase this a bit, would be a lot of duplication of references
@@
 % more natrual to talk about the advantages of model-driven stuff next.
-The first class of inference methods are probabilistic approaches \revpoint{2}{6}
+The first class of inference methods are Bayesian approaches
 that sample ARGs under a population genetics model such as the
@@
 heuristic and approximation driven, favouring computational
-scalability over a basis in a probabilistic model. \revpoint{2}{7}
+scalability over a basis in a statistical model.
 Secondly, they all produce a single
@@
 Similarly, there is often a fundamental lack of information about the 
-details of recombination events, and heuristic methods
-omit these details in different ways. \revpoint{2}{8}
-The significant progress made in
-scalability by these heuristic methods is in part because they don't
+details of recombination events, and heuristic methods (in different ways)
+omit these details. In many ways, the significant progress made in
+scalability by these heuristic methods is \emph{because} they don't
 attempt to precisely reconstruct recombination events.
@@
 posterior distribution provides a means of quantifying the uncertainty 
-around point estimates. \revpoint{2}{9}
+around estimates.
 On the other hand, heuristic methods have the potential
@@
 in terms of scalability over ARGweaver and other methods, inference is still 
-only feasible for hundreds of samples.  \revpoint{2}{1}
+only feasible for hundreds of samples. 
 Meanwhile, tsinfer, ARG-Needle and Threads have 
@@
 
-A key problem facing the field is that the ARGs estimated by heuristic \revpoint{2}{10}
+A key problem facing the field is that the ARGs estimated by heuristic large-scale 
 methods lack a concrete connection to population genetic theory because
-we cannot compute their likelihood under the SMC or similar models. \revpoint{2}{11}
+we cannot compute their likelihood under the SMC or similar models. 
 Computing the likelihood of an ARG under a model is a fundamental element of
@@
 program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
-which does not contain sufficient detail to reconstruct
-the ``full ARG'' required for likelihood
-computation~\citep{wong_general_2023}. \revpoint{2}{12}
+which does not contain sufficient detail for likelihood computation.
 The msprime
@@
 has much more lenient input requirements than KYF, and gracefully handles
-incomplete and imprecise ARGs. \revpoint{2}{13}
+incomplete and imprecise ARGs.
 % In this paper we provide a necessary first step towards this goal
@@
 By ``describes the history'',
-we mean that each node in the ARG represents a specific genome, \revpoint{2}{14}
-and that the ARG specifies the genealogical tree describing how the
-sampled genomes are related
-at each base pair. \revpoint{2}{15}
+we mean that each node in the ARG represents a specific (ancestral) genome,
+and that the ARG specifies the genealogical tree describing how the samples are related
+at each point on the genome.
 
@@
 the floating lineage coalesces randomly
-    with the remaining portion of the local tree following recombination.
+with a the remaining portion of the local tree following recombination.
 % By removing recombinant nodes, we limit the information on the time to each
@@
 An ARG can therefore be thought of either as a collection of inheritance
-relationships between
-sampled and ancestral haplotypes (haploid genomes), \revpoint{2}{16}
+relationships between ancestral haplotypes,
 or as a sequence of genealogical trees that have common node labels.
@@
 the path traced by $A$'s ancestry switches at that point.
-In fact, $A$ is in blue because it is a \emph{recombinant}
-node -- an ancestor that has a segment of genome inherited by a sample,
+In fact, $A$ is in blue because it is a \emph{recombinant} node --
+an ancestor that has a segment of genome inherited by a sample,
 but that segment was inherited in two pieces from their parent.
-Other nodes (labelled by black numbers) are either sample nodes
-or \emph{common ancestry} nodes -- \revpoint{2}{20}
+Other nodes (labelled by black numbers) are either samples
+or \emph{common ancestry} nodes --
 ancestors that are the most recent common ancestor for two 
@@
 while in the bottom plot, this is visible because node $f$ is parent to $d$
-in the second tree, but not to node $c$, and so is unary
-in the second tree. \revpoint{2}{21}
+in the second tree, but not to node $c$, and so is unary.
 
@@
 These locally unary nodes enable
-us to uniquely identify all lineages that were hit by a recombination event, \revpoint{2}{22}
+us to uniquely identify all lineages that were hit by a recombination event,
 even if we simplify 
@@
 but we do still have constraints on it;
-for instance, we know that $C$ happened sometime between the
-times of node $f$ and node $h$. \revpoint{2}{23}
-More generally, a recombinant node must happen in the time interval between the child node
+for instance, we know that $C$ happened somewhere between node $f$ and node $h$.
+More generally, a recombinant node must happen somewhere between the child node
 and the most recent of the parent nodes.
@@
 to simulate the coalescent with
-recombination was formulated backwards-in-time \revpoint{2}{24}
+recombination was formulated backwards in time
 by Hudson \citeyearpar{hudson_properties_1983}. 
@@
 obtain a process that is Markovian both backwards-in-time, 
-as well as along-the-genome. \revpoint{2}{25}
+as well as along-the-genome.
 In its backwards-in-time formulation, the SMC only requires a
@@
 genomic intervals may coalesce. The SMC' is a refinement 
-of the SMC which also allows lineages with adjoining genomic intervals \revpoint{2}{26}
-to coalesce~\citep{marjoram_fast_2006}.
+of the SMC which also allows lineages with abutting genomic intervals to 
+coalesce~\citep{marjoram_fast_2006}.
 Although \citet{mcvean_approximating_2005} verbally described the SMC
@@
 
-Linking the resulting tree-valued process \revpoint{2}{27}
-along the genome back to the relationship between the ``graph'' and
+Linking the resulting tree-valued process along the genome back to
+the relationship between the ``graph'' and
 ``local trees'' representation of ARGs, 
@@
 greatly simplifies the ability to keep track of coalescable pairs going
-backwards-in-time.
+backwards in time.
 At any point in time, the state of the process is 
@@
 % for some interval endpoints $x_{j} < y_{j}$.
-Because we look backwards-in-time, $t=0$ is today, and
+Because we look backwards in time, $t=0$ is today, and 
 $L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
@@
 events until each segment of ancestral material is only present in one lineage.
-The waiting time to the next event is determined by these two competing processes
+The waiting time to next event is determined by these two competing processes
 with exponentially distributed waiting times as outlined below.
@@
 of genomic length and time:
-so, if $T(t) = \sum_j |X_j(t)|$ is the total length of ancestral material
+so, if $T(t) = \sum_j |X_j(t)|$ is total length of ancestral material
 carried by lineages at time $t$,
@@
 thus, $|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
-and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$.  \revpoint{1}{2}
+and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$. 
 Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
@@
 In general, the likelihood for a continuous-time Markov process specified in this way
-is $\exp(-\Lambda) \prod_i \lambda_i$, \revpoint{1}{3}
+is $\exp(-\Lambda) \prod_i \lambda_i$,
 where $\lambda_i$ is the rate of the $i^\text{th}$ realised event,
@@
 $\exp(-r \mathcal{A}_e)$, where $r$ is the recombination rate
-and $\mathcal{A}_e$ is the total ``area'' \revpoint{2}{28}
-of the edge.
-The area of an edge is its  length in time, $t_p - t_c$,
-multiplied by the number of ``eligible links'' along its
-span.
-The number of eligible links is the number of adjacent base pairs in the edge
+and $\mathcal{A}_e$ is the total area (eligible links multiplied by length in time) of the edge.
+The length in time of the edge is just $t_p - t_c$,
+while the number of ``eligible links'' in the edge is the number of adjacent base pairs in the edge
 on which recombination did not occur, which is either $y-x$
@@
 Then, the cumulative hazard for common ancestry events of edge $e$
-between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(z) dz$. \revpoint{1}{4}
+between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(z) dz$.
 If the edge $e = (p,c,x,y)$
@@
 \subsection*{Robustness to partial and incorrect information}
-The ARG likelihood formulation in 
-Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for any \revpoint{2}{29}
-collection of nodes (haplotypes that existed at some time $t$) and 
-edges (a record that child node $c$
+Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
+collection of nodes (ancestral haplotypes that lived at some time 
+$t$ in the past) and edges (a record that child node $c$ 
 inherited the genomic interval $[x, y)$ from parent node $p$)
@@
     msprime simulation of the SMC (100 diploid samples, genome length 1Mb, 
-    $r=\mu=1.25\times10^{-8}$, $N_e=1\times10^4$).
+    $r=\mu=2.5\times10^{-8}$, $N_e=1\times10^4$).
     The ``RE nodes removed'' ARG is the 
@@
 
-In Fig.~\ref{fig:lik-surface} we begin with 
-with a sample from the SMC model 
+In Fig~\ref{fig:lik-surface} we begin with a realisation of the SMC 
 containing complete information about every simulated event
-(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). \revpoint{2}{30}
+(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). 
 We then plot the likelihood computed by Eq.~\eqref{eq:full-lik} 
@@
 We next compute the likelihood ratio for the ARG obtained when we 
-simplify~\citep{kelleher_efficient_2018,wong_general_2023}
- out the recombination nodes from the full ARG
+simplify out the recombination nodes from the full ARG
 (``RE nodes removed''; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime 
@@
 contain all the information that we need about recombination events
-for this application. \revpoint{2}{31}
+for this application.
 When we fully simplify the simulated ARG
@@
 (``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees) 
-we can see there is a significant loss of information, \revpoint{2}{32}
+we can see there is a significant loss of information,
 and the minimum of the 
@@
 
-Fig.~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably \revpoint{2}{33}
+Fig~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably
 robust to the presences of polytomies in this example. For each of the three
@@
 removing 25\% of the internal 
-nodes has negligible effect on the likelihood curve in Fig.~\ref{fig:lik-surface},
+nodes has negligible effect on the likelihood curve in Fig~\ref{fig:lik-surface},
 and even on the fully simplified 
-ARG has a very minor effect on location of the minimum.
-Supplemental Fig.~S1 explores more extreme examples in which we remove
-25\%, 50\% and 75\% of the internal nodes in these ARGs, and demonstrates 
-a remarkable robustness to the loss of the majority of the internal
-nodes in these examples. Interestingly, the loss of 75\% of the 
-internal nodes  in the fully simplified ARG 
-(1841 of 2455, leading  to an average of 27 polytomies per tree)
-appears to improve performance, substantially 
-reducing the bias away from the true parameter value.
-
-Fig.~\ref{fig:lik-surface} also shows results for ARGs inferred from 
-the  \revpoint{2}{35}
-simulated data.
-To do this we first simulated mutations on the full ARG 
-at rate $1.25\times10^{-8}$ using msprime, resulting in an 
-ARG with 2,989 mutations.
-We then used tsinfer and tsdate~\citep{wohns_unified_2022}
-to infer a dated ARG,
-using default parameters for tsinfer and the true values of mutation
+ARG has only a very minor effect on location of the minimum.
+
+Fig~\ref{fig:lik-surface} also shows results for ARGs inferred from the 
+simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
+using default parameters for tsinfer, and the true values of mutation
 rate and $N_e$ for tsdate.
 In the first, we use the default output of tsinfer 
-(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations). \revpoint{2}{36}
+(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations).
 This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
@@
 moves~\citep[e.g.,][]{rasmussen2023espalier}. Nonetheless, in
-this toy example at least, our likelihood function performs well \revpoint{2}{37}
+this toy example at least, our likelihood function performs well
 on the output of tsinfer, with the minimum close to the true
@@
 before dating. We can see that, as with the true data,
-removing all the unary nodes results in a bias \revpoint{2}{38}
+removing all the unary nodes results in a significant bias
 away from the true value of the recombination rate parameter.
@@
     Remaining parameters: genome length is 10kb, 20 diploid samples, and
-    $r=1.25\times10^{-8}$.
+    $r=2.5\times10^{-8}$. 
     There are 148 edges in the resulting ARG.
@@
 and imprecise information
-provided by some modern ARG inference tools, \revpoint{2}{39}
+provided by some modern ARG inference tools,
 using a novel backwards-time derivation of the SMC.
@@
 a significant increase in flexibility over the 
-strict input requirements for the classical KYF  \revpoint{2}{40}
+strict input requirements for the classical Kuhner-Yamato-Felsenstein 
 formulation. 
@@
 the details of the recent past in huge datasets such as 
-UK Biobank, passing over to a rigorous probabilistic approach \revpoint{2}{41}
-for the ancient past.
+UK Biobank, passing over to a rigorous model-based
+approach for the ancient past.
 This is similar in spirit to the widely-used approach 
@@
 and can be combined to great effect,
-large-scale heuristic and probabilistic inference  \revpoint{2}{42}
-methods have their applications, and there may be synergy in their combination.
+large-scale heuristic and Bayesian sampling inference methods have their 
+applications, and there may be synergy in their combination. 
 % This is pretty weak, tbh.
@@
 \url{https://github.com/gertjanbisschop/runsmc}.
-\revpoint{2}{43}
 
@@
 along with additional information that
-allows us to sequentially recover local trees (and reason about their differences) \revpoint{1}{5}
+allows us to sequentially recover (and reason about 
+their differences) 
 efficiently~\citep{kelleher_efficient_2016, ralph_efficiently_2020}.
@@
 the last parent a child node was connected to to detect recombination events.
-
-\setcounter{table}{0}
-\setcounter{figure}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\renewcommand{\thefigure}{S\arabic{figure}}
-
-\begin{figure}[!ht]
-\centering
-\includegraphics{figures/likelihood_surface_suppl}
-    \caption{Log likelihood-ratio curves for three different ARGs 
-    with increasing fractions of polytomies.
-    The base ARGs are as described in Fig.~3, but here we remove 
-    25\%, 50\% and 75\% of the internal nodes to create polytomies.
-    The true value of recombination parameter is indicated by the 
-    dashed line, as in Fig.~3.
-    \label{fig:likelihood_suppl}}
-\end{figure}
 
@@
 
-\includereviews
-
 \end{document}

### 2025-05-05 review-responses.tex
Updates
Cover tricky tsinfer points
peter edits

@@
 
+Thanks -- we're glad you appreciated the paper.
+
 \begin{point}{}
@@
 \reply{
+    Good catch -- fixed. \revref
 }
@@
 \reply{
+    Done -- hopefully this helps. \revref
 }
@@
 \reply{
+    Again, good catch! Fixed. \revref
 }
@@
 \end{quote}
+
+We're glad the paper found a warm welcome.
+And, thanks very much for your attention to detail in the comments.
 
@@
 \reply{
+    Nice idea -- we've done this, and hope it doesn't add too much to the complexity of the figures.
 }
@@
 \reply{
+We feel that ``tree-valued process'' is a fairly well established term in
+more mathematical literature and would prefer to keep it as-is.
 }
@@
 \reply{
+Rephrased.
 }
@@
 \reply{
+    Good point, we shouldn't let a nice acronym like this go unexplained.
+    We think the algorithm is at least relatively unaware of the times,
+    and so we're okay with the current acronym.
 }

### 2025-05-05 paper.tex

@@
 (locally) unary.
-Nodes A, B and C are recombinants, i.e, have two parents in the graph
+Nodes A, B and C are recombinants, i.e., have two parents in the graph
 and are unary in the local trees. The blue shaded regions indicate
@@
 multiplied by the number of overlapping pairs,
-i.e., $\lambda \sum_{i \neq j} I_{jk}$,
+i.e., $\lambda \sum_{j \neq k} I_{jk}$,
 where
@@
 % Cardin, and not important for our purposes.]
-For later use, also define $C_j(t)$ to be the set of
-lineages that can coalesce into lineage $j$, i.e.,
-$C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$;
-thus, $|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
-and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$.  \revpoint{1}{2}
+For later use, also define $I_j(t)$ to be the total rate at which lineages
+are coalescing into lineage $j$,
+and $C_j(t)$ to be the set of lineages that can coalesce into lineage $j$.
+In other words,
+$C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$
+and $|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
+and the total rate of coalescence at time $t$ is $\lambda \sum_{j} |C_j(t)|$.  \revpoint{1}{2}
 Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
@@
 (also called the ``total hazard'').
-As discussed in the Introduction, in many cases these events are not what is
+Each $\lambda_i$ would be either equal to $\lambda$ (if the event is a coalescence)
+or $r|X_j(t)|$ for some $j$ and $t$ (if it is a recombination),
+and since the total rates at time $t$ of recombination events is $rT(t)$
+and of coalescent events is $\lambda \sum_j |C_j(t)|$,
+the total hazard up to time $T$ is $\Lambda = \int_0^T r T(t) + \lambda \sum_j |C_j(t)| dt$.
+
+However,
+as discussed in the Introduction, in many cases these events are not what is
 estimated by inference methods, which produce, more fundamentally, a collection
@@
 Then, the cumulative hazard for common ancestry events of edge $e$
-between time $s$ and $t$ is given by $F(e, s, t) = \int_{s}^{t} I_{e}(z) dz$. \revpoint{1}{4}
+between time $s$ and $t$ is given by $\lambda$ multiplied by
+$F(e, s, t) = \int_{s}^{t} I_{e}(z) dz$. \revpoint{1}{4}
 If the edge $e = (p,c,x,y)$
@@
 then no recombination occurred along this edge and
-this edge simply contributes $\exp(-F(e,t_c,t_p))$ to the likelihood.
+this edge simply contributes $\exp(-\lambda F(e,t_c,t_p))$ to the likelihood.
 However, if the edge is \emph{not} the leftmost segment
@@
 \url{https://github.com/gertjanbisschop/smc-bit-paper},
-and an implementation of the algorithm is available at
+and an implementation of the algorithm,
+named ``Recombination (time) UNaware SMC'' (runsmc),
+is available at
 \url{https://github.com/gertjanbisschop/runsmc}.

### 2025-05-11 paper.tex
add pseudocode

@@
 \usepackage[hidelinks]{hyperref}
+
+% For algorithms in supplementary materials
+\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
 
@@
 the last parent a child node was connected to to detect recombination events.
+The pseudocode for the algorithm is included in Alg.~\ref{alg:full}. Note that Alg.~\ref{alg:depth} 
+details how to compute the integrals contained in Eq.~\eqref{eq:depth}.
 
@@
 \end{figure}
+
+% algorithm in pseudocode
+%% Algorithm to compute contribution of time component of each edge
+
+\begin{algorithm}[H]
+\caption{\textsc{LogDepth}: pseudocode for equation Eq.~\eqref{eq:depth}}
+\label{alg:depth}
+\KwIn{
+    Count vector $C$; interval boundaries $T$;\\
+    Parent time $t_p$; child time $t_c$;\\
+    Initial index $i_0$; recombination rate $r$;\\
+    Coalescence rate $c$; recombination flag $R$
+}
+\KwOut{log likelihood contribution in time dimension}
+\BlankLine
+
+Initialize cumulative area $A \leftarrow 0$\;
+Initialize result $S \leftarrow 0$\;
+Set index $i \leftarrow i_0$\;
+
+Define helper function $f(f_0, f_1)$:
+\begin{itemize}
+    \item If $f_0 = f_1$, return $0$
+    \item Else return $\displaystyle \frac{c}{r - c f_1} \cdot \frac{c}{r - c f_0} \cdot (f_1 - f_0) \cdot \frac{r}{c}$
+\end{itemize}
+
+Set stopping time:
+\[
+t_{\text{stop}} \leftarrow
+\begin{cases}
+t_c & \text{if } R = \text{False} \\
+t_p & \text{otherwise}
+\end{cases}
+\]
+
+\While{$T_i > t_{\text{stop}}$}{
+    $A \leftarrow A + (T_i - T_{i-1}) \cdot C_{i-1}$\;
+    $i \leftarrow i - 1$\;
+    \If{$i < 1$}{break}
+}
+
+\If{$R = \text{False}$}{
+    \Return $-c \cdot A$
+}
+
+\[
+B \leftarrow - \frac{c}{r - c C_{i-1}} \cdot \frac{r}{c} \cdot \exp(-r T_i - c A)
+\]
+
+\While{$i > 0$}{
+    $A \leftarrow A + (T_i - T_{i-1}) \cdot C_{i-1}$\;
+    $i \leftarrow i - 1$\;
+    \If{$T_i = t_c$}{break}
+    \[
+    S \leftarrow S + f(C_{i-1}, C_i) \cdot \exp(-r T_i - c A)
+    \]
+}
+
+\[
+S \leftarrow S + \frac{c}{r - c C_i} \cdot \frac{r}{c} \cdot \exp(-r T_i - c A)
+\]
+
+\[
+S \leftarrow S + B
+\]
+
+\Return $\log(S)$
+
+\end{algorithm}
+
+%% Main algorithm
+
+\begin{algorithm}[H]
+\caption{\textsc{LogLikelihood}: pseudocode for Eq.~\eqref{eq:full-lik}}
+\label{alg:full}
+\KwIn{
+    Time vector $I$ for all nodes;\\
+    Edge arrays containing parent, child, left and right coordinate of each edge: \texttt{edges\_parent}, \texttt{edges\_child}, \texttt{edges\_left}, \texttt{edges\_right};\\
+    Recombination rate $r$, coalescence rate $c$;\\
+    \texttt{treesequence};\\
+}
+\KwOut{Total log-likelihood value}
+\BlankLine
+
+Initialize log-likelihood: $\text{LL} \leftarrow 0$ \;
+Initialize lineage count vector $C \leftarrow 0$ \;
+Initialize arrays to track last parent and visited nodes \;
+
+\While{next tree exists in treesequence}{
+
+    \tcp{Process edge removals}
+    \ForEach{edge $e$ being removed}{
+        Identify parent $p$ and child $c$ from $e$ \;
+        Get node times $t_p$ for $p$, and $t_c$ for $c$ \;
+        Decrease lineage count $C$ between $t_c$ and $t_p$ \;
+        Record $p$ as last parent of $c$ \;
+    }
+
+    \tcp{Process edge insertions}
+    \ForEach{edge $e$ being inserted}{
+        Identify parent $p$ and child $c$ from $e$ \;
+        Get node times $t_p$ for $p$, and $t_c$ for $c$ \;
+
+        Set \texttt{rec\_event} $\leftarrow$ False \;
+        Set \texttt{left\_parent\_time} $\leftarrow \infty$ \;
+
+        \If{child $c$ had a previous parent $p^{\prime}$}{
+            Retrieve previous parent time $t_{p^{\prime}}$\;
+            \If{$p^{\prime}$ $\neq p$ and $p$ not visited}{
+                Set \texttt{rec\_event} $\leftarrow$ True \;
+            }
+        }
+
+        Set $t_{p_{\min}} \leftarrow \min(t_p, \texttt{left\_parent\_time})$ as effective parent time \;
+
+        \tcp{Add depth contribution}
+        $\text{LL} \leftarrow \text{LL} + \textsc{LogDepth}(C, I, t_{p_{\min}}, t_c, p, r, c, \texttt{rec\_event})$
+
+        \tcp{Add span contribution}
+        $\text{LL} \leftarrow \text{LL} -r \cdot (t_p - t_c) \cdot (\texttt{edges\_right}[e] - \texttt{edges\_left}[e])$
+
+        \tcp{Update lineage count}
+        Increase $C$ between $t_c$ and $t_p$ \;
+    }
+}
+
+Number of nodes with out-degree 0: $n_o$ \;
+Contribution coalescent events: $ll_c = \text{num\_edges} -\text{num\_nodes} + n_o$ \; 
+Number of nodes with in-degree 0: $n_r$ \;
+Contribution recombination events: $ll_r = \text{num\_edges} - \text{num\_nodes} + n_r$ \;
+Update likelihood: $\text{LL} \leftarrow \text{LL} +ll_c \cdot \log(c) + ll_r \cdot \log(r)$ \;
+
+\Return $\text{LL}$
+
+\end{algorithm}
 
